### 2012-02-03 papers/itsc_2012/draft.tex
fixed exceeding speed limit: agents were trying TOO hard to follow others

\documentclass[12pt]{article}

\title{Paper}
\author{Dustin Carlino and Mike Depinet}
\date{...}

\begin{document}
\maketitle

% what are our goals with this paper?
% * demonstrate use of OSM and motivate more data collection
% * present a simulation framework
% * present a pretty naive static planning
% * motivate and demonstrate wards for hierarchial planning (but also reasoning
%    about congestion)

% contribution: autonomous car / human car sim on EXISTING INFRASTRUCTURE

\section{Map Construction from OSM}

each step

limitations

proposals to change OSM to fix these issues

integrating with what other sources for spawning traffic, placing POI?

\section{Simulation}

limitations of timesteps and small edges. in reality, more computational power
and more reactivity

justify arbitrary constants, like deaccel for cars

\end{document}

### 2012-02-03 utexas/sim/Agent.scala

@@ -27,11 +27,6 @@ class Agent(val id: Int, val graph: Graph, start: Edge) {
   def step(dt_s: Double): Boolean = {
     assert(dt_s == cfg.dt_s)
 
-    // TODO cfg
-    if (speed > Util.mph_to_si(30)) {
-      Util.log(this + " at " + at + " is over the speed: " + speed + " > " + Util.mph_to_si(30))
-    }
-
     // To confirm determinism, enable one of these (more precision in doubles is
     // more likely to differ) and diff the logs.
     //Util.log(this + " at " + Agent.sim.tick + " is at " + at)
@@ -43,6 +38,9 @@ class Agent(val id: Int, val graph: Graph, start: Edge) {
     // Do physics to update current speed and figure out how far we've traveled in
     // this timestep.
     val new_dist = update_kinematics(dt_s)
+
+    // TODO cfg
+    assert(speed <= Util.mph_to_si(30))
 
     // Apply this distance. 
     var current_on = start_on

### 2012-02-04 papers/itsc_2012/draft.tex
re-enabled edge trimming with more restraints.. it's at least a bit better

@@
 
+interface... what invariants do agents have to follow, and how does the default
+behavior provide them?
+
+\section{notes}
+
+possible background/links
+http://www.caliper.com/transmodeler/default.htm
+http://ieeexplore.ieee.org/stamp/stamp.jsp?tp=&arnumber=6082793
+ https://www.google.com/search?sourceid=chrome&ie=UTF-8&q=Agent-Based+Traffic+Simulation+and+Traffic+Signal+Timing+Optimization+with+GPU
+
 \end{document}

### 2012-02-04 utexas/map/make/Builder.scala

@@ -27,7 +27,7 @@ object Builder {
         case "--show-dead" => { show_dead = value == "1" }
         case "--input"     => { ifn = value }
         case "--output"    => { ofn = value }
-        case _             => { println("Weird option " + key) }
+        case _             => { Util.log("Weird option " + key); sys.exit }
       }
     }
 

### 2012-02-04 utexas/map/make/Pass3.scala

@@ -361,8 +361,7 @@ class Pass3(old_graph: PreGraph2) {
     // Not adjusting the other line if we're not the rightmost lane... TODO does
     // this help?
     e.rightmost_lane.next_counterclockwise_to match {
-      // TODO re-enable when it stops destroying some lines entirely.
-      case Some(ccw) => //adjust_lines(e.lines.last, ccw.lines.head, e.is_rightmost)
+      case Some(ccw) => adjust_lines(e.lines.last, ccw.lines.head, e.is_rightmost)
       case _ => {}
     }
   }
@@ -374,12 +373,19 @@ class Pass3(old_graph: PreGraph2) {
         // With multiple lanes, this could happen multiple times. Either force
         // far-to-near order so lines only ever decrease, or just only permit
         // changes when it makes lines shorter...
-        if ((new Line(l1.x1, l1.y1, pt.x, pt.y)).length < l1.length) {
+
+        // But some changes, for some reason, make some lines MUCH shorter, so
+        // detect and avoid trimming those.
+        val max_trim = 1.0  // TODO cfg and make units make sense.
+
+        val possible1 = new Line(l1.x1, l1.y1, pt.x, pt.y)
+        if (possible1.length < l1.length && l1.length - possible1.length <= max_trim) {
           l1.x2 = pt.x
           l1.y2 = pt.y
         }
         if (both) {
-          if ((new Line(pt.x, pt.y, l2.x2, l2.y2)).length < l2.length) {
+          val possible2 = new Line(pt.x, pt.y, l2.x2, l2.y2)
+          if (possible2.length < l2.length && l2.length - possible2.length <= max_trim) {
             l2.x1 = pt.x
             l2.y1 = pt.y
           }

### 2012-02-06 papers/itsc_2012/draft.tex
paper draft: map construction

@@
 
+% TODO grab the format
+
 \title{Paper}
-\author{Dustin Carlino and Mike Depinet}
+\author{Dustin Carlino, Mike Depinet, Piyush Khandelwal, Peter Stone}
 \date{...}
@@
 % what are our goals with this paper?
+% theme: easy simulation in your city, in 2 minutes!
 % * demonstrate use of OSM and motivate more data collection
@@
 
+\section{Architecture}
+
+scala
+
+maps separate from simulation separate from UI (mostly)
+
+thread pools?
+
+programs: build map, UI (describe what it can do), headless
+
+api / extending it
+
 \section{Map Construction from OSM}
 
-each step
+% briefly introduce OSM, advantages, etc
 
-limitations
+To run a simulation on a map from OSM, there is a one-time conversion process to
+transform the map into AORTA's format, which is also encoded with XML. It is
+split into 3 passes, each feeding the next.
+
+\subsection{Map Model}
+
+roads, verts, edges, wards, traversables, turns...
+
+edges are made up of multiple lines.. this could be control pts of a curve
+(hereon, line means line segment)
+
+\subsection{Pass 1}
+
+The relevant parts of OSM's map model are nodes and ways. A node corresponds to
+a GPS location, and ways are links between many nodes, corresponding to roads
+and other paths. The first pass reads OSM's XML format, gathering ways as a
+series of coordinates. Ways intersect when a node is used by multiple ways.
+
+Since OSM also encodes pedestrian paths, bike routes, and even lakes with ways,
+pass 1 filters out anything that does not seem to be driveable. OSM does not
+seem to provide a unified list of valid roads, so these heuristics are
+imperfect.
+
+Finally, coordinates are normalized from GPS (longitude, latitude) to a planar
+coordinate system with no negative coordinates and the Y coordinate increasing
+upwards. This is done for convenince of manipulation in the UI, but the scale
+and offset used are retained, so a coordinate in the final form of the map can
+still be mapped back to its original GPS form.
+
+\subsection{Pass 2}
+
+OSM ways only implicitly intersect at common coordinates; AORTA's structure
+demands a traditional definition of edge linking exactly two vertices. Pass 2
+walks along each way's coordinates, dividing it into multiple undirected edges.
+
+\subsection{Pass 3}
+
+Finally, pass 3 performs successive transformations to produce the final graph,
+which is saved in XML.
+
+\subsubsection{Making Lanes}
+
+Undirected edges are first multiplied into directed lanes. The coordinates from
+a slice of an OSM way are interpreted to be the center line of the road, and
+lanes are placed by projecting the width of a line perpendicularly from the
+center line in either direction. OSM encodes one-way streets, so in this case,
+if there are an odd number of one-way lanes, the middle lane is placed where the
+center line would be, and the other lanes are shifted from there. If there are
+an even number, then the middle two are both equidistant away from the invisible
+center line. The number of lanes is guessed from OSM's annotation of the way's
+``type.'' OSM's specification includes a tag for number of lanes, but in
+practice as of the time of writing, it is rarely used.
+
+\subsubsection{Making Turns}
+
+Next, turns are constructed at each vertex between the incoming and outgoing
+edges. This is completely heuristic based: we classify a potential turn as a
+left, right, or straight turn based solely on angles between edges and if the
+edge's original OSM ID matches. U-turns are constructed at dead-ends. If several
+lanes all cross into fewer lanes, merging is forced for the rightmost lanes.
+These patterns are attempts at generalizations; this process results in many
+intersections having the incorrect turns. However, without a more detailed
+source, this is the best compromise. Finally, the geometry of a turn is
+extremely simplified as a single line between the ends of the edges it connects.
+In the future, a more realistic curve could be drawn.
+
+\subsubsection{Ensuring connectivity}
+
+Because turns are constructed imperfectly, most large graphs have edges that are
+not reachable from some portions of the graph. Simulation and pathfinding would
+suffer greatly if this was the case, so pass 3 uses Tarjan's linear-time
+algorithm for locating strongly connected components to remove the components of
+the directed graph with the least members. (There is also a debugging switch to
+leave these components in and highlight them in the UI.) The algorithm operates
+at the granularity of lanes. If several lanes of many are disconnected within
+one road, then the bad lanes are spliced out. This results in visually
+unappealing geometric gaps, but if the entire road were removed, then Tarjan's
+algorithm would have to be repeated to determine what edges the good lanes would
+then disconnect. The end result is a connected digraph, ideal for corner
+case-free pathfinding and simulation.
+
+\subsubsection{Cleaning up geometry}
+
+A road's center line correctly meets adjoining roads' center lines in the middle
+of an intersection. Since the lines for a lane are parallel to the center line
+and have the same length, they also meet in the intersection. This is
+geometrically incorrect; lanes stop along the border of the intersection. Thus,
+pass 3 includes heuristics to trim a lane's lines to the spot where they
+intersect with another lane's lines. This results in a more accurate model, but
+the process is imperfect, so sometimes lines extend too far or not enough.
+
+\subsubsection{Hierarchial divisions of the graph}
+
+% TODO i don't think this'll have much coverage in the rest of the paper; should
+% we describe wards?
+
+\subsection{Limitations}
 
 proposals to change OSM to fix these issues
+
+projecting lines sucks... but all it really does is clutter UI
+filtering out invalid ways in pass 1 is really not a clear process
+number of lanes... road types are wacky and wrong
+turns are brittle.. detecting right-turn only lane is rare. shared center lane?
+  not part of our model, or OSM's.
+removing sccs is not ideal
+
+\subsection{Integrating other data sources}
+
+how easy would it be?
+
+match on gps, or on road metadata like names
 
@@
 
-\section{notes}
+\section{Notes}
 

### 2012-02-06 utexas/sim/Headless.scala

@@ -26,7 +26,7 @@ object Headless {
 
   def main(args: Array[String]) = {
     val sim = process_args(args)
-    sim.spawn_army(1, 100)
+    sim.spawn_army(1, 1000)
 
     val timer = Util.timer("running the sim")
     Util.log("Starting simulation with time-steps of " + cfg.dt_s + "s")

### 2012-02-08 papers/itsc_2012/draft.tex
some work on the paper and build scripts without ant dependency

@@
 
-% TODO grab the format
+% TODO grab the ieee format
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \section{Architecture}
 
-scala
+AORTA's implementation uses Scala, a language implemented on the Java Virtual
+Machine. Scala provides the advantage of functional programming constructs
+without imposing them; parts of AORTA or extensions or clients can just as
+easily be written in the more familiar Java.
 
-maps separate from simulation separate from UI (mostly)
+The architecture of AORTA is described below.
 
@@
 api / extending it
+maps separate from simulation separate from UI (mostly)
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 (hereon, line means line segment)
+(turns are also traversable)
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \section{Simulation}
 
-limitations of timesteps and small edges. in reality, more computational power
-and more reactivity
+AORTA uses microscopic simulation, modeling individual drivers as agents.
+Discrete-time simulation is used, meaning an agent accelerates at a fixed rate
+for the duration of a ``tick'' to achieve a new position and velocity. The tick
+duration $dt$ is configurable and fixed for the simulation. Agents know this
+value and use it to reason about collision avoidance and satisfying other
+constraints, so setting this value to the standard ``reaction time'' of a human
+driver may be appropriate. Each tick, the simulation performs the following
+steps:
+\begin{itemize}
+  \item Introducing new agents into the map
+  \item Updating the position and velocity of agents
+  \item Checking for collisions
+  \item Allowing each agent to react
+\end{itemize}
+
+There are two structures to detect collisions. All traversables (edges and the
+straight line-approximation of turns) maintain a queue of occupying agents,
+ordering them so that the agent farthest along is at the head of the queue.
+A naive method of collision detection would detect an agent passing an agent on
+the same queue, or entering a new queue not at the end, immediately after an
+agent moves. However, agents take their steps sequentially and in an effectively
+random order. Ideally they could be reordered so that agents in the front of a
+queue always move first. Since the first agent may cross into the next queue,
+that introduces a depdendency between traversables. A topological sort of a
+directed graph containing circuits is not possible, though, and since we
+guarantee connectedness, there are obviously circuits. Instead, collisions are
+checked after every agent has stepped forward in time. Queues remember the
+original order of agents before a step, and afterwards can verify the order is
+still legal, with new agents on the end and existing agents in the same order.
+Likewise, intersections check for collisions after everybody has moved by
+examining the turn each agent is performing and verifying that no other agent is
+simultaneously performing a conflicting turn.
+
+invariants for agents. interface of behavior and intersection policies.
+
+lookahead... worst-case stop dist argument is everywhere. safe spawning, too.
+
+intersection policies... stop sign, traffic light.
 
@@
 
-interface... what invariants do agents have to follow, and how does the default
-behavior provide them?
+scale
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{Experimental Results}
+
+% upgrading stop signs to signals, optimizing timing
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 possible background/links
-http://www.caliper.com/transmodeler/default.htm
-http://ieeexplore.ieee.org/stamp/stamp.jsp?tp=&arnumber=6082793
- https://www.google.com/search?sourceid=chrome&ie=UTF-8&q=Agent-Based+Traffic+Simulation+and+Traffic+Signal+Timing+Optimization+with+GPU
+%http://www.caliper.com/transmodeler/default.htm
+%http://ieeexplore.ieee.org/stamp/stamp.jsp?tp=&arnumber=6082793
+%https://www.google.com/search?sourceid=chrome&ie=UTF-8&q=Agent-Based+Traffic+Simulation+and+Traffic+Signal+Timing+Optimization+with+GPU
 

### 2012-02-08 papers/itsc_2012/generate.sh

@@ -1,0 +1,2 @@
+#!/bin/sh
+pdflatex draft.tex

### 2012-02-08 utexas/sim/Agent.scala

@@ -167,7 +167,10 @@ object Agent {
 
 case class Position(val on: Traversable, val dist: Double) {
   assert(dist >= 0)
-  assert(dist <= on.length)
+  //assert(dist <= on.length)
+  if (dist > on.length) {
+    Util.log("safe_spawn_dist must be broken... " + dist + " > " + on.length + " on " + on)
+  }
 
   def location = on.location(dist)
   def dist_left = on.length - dist

### 2012-02-08 utexas/ui/MapCanvas.scala

@@ -536,57 +536,61 @@ class MapCanvas(sim: Simulation) extends ScrollingCanvas {
         // Let's find all vertices inside the polygon.
         val rds = sim.vertices.filter(v => polygon.contains(v.location.x, v.location.y)).flatMap(v => v.roads).toSet
         Util.log("Matched " + rds.size + " roads")
-        if (polygon_roads1.isEmpty) {
-          polygon_roads1 = rds
-          Util.log("Now select a second set of roads")
+        if (rds.isEmpty) {
+          Util.log("Try that again.")
         } else {
-          polygon_roads2 = rds
-          Util.log("Creating a new generator...")
+          if (polygon_roads1.isEmpty) {
+            polygon_roads1 = rds
+            Util.log("Now select a second set of roads")
+          } else {
+            polygon_roads2 = rds
+            Util.log("Creating a new generator...")
 
-          // Ask: fixed (how many) or continuous (how many per what time)
-          // TODO improve the UI.
-          Dialog.showOptions(
-            message = "Want a fixed, one-time burst or a continuous generator?",
-            optionType = Dialog.Options.YesNoCancel, initial = 0,
-            entries = Seq("Constant", "Continuous")
-          ) match {
-            case Dialog.Result.Yes => {
-              // Fixed
-              prompt_int("How many agents?") match {
-                case Some(num) => {
-                  sim.generators += new FixedSizeGenerator(
-                    sim,
-                    polygon_roads1.toList.flatMap(r => r.all_lanes),
-                    polygon_roads2.toList.flatMap(r => r.all_lanes),
-                    num
-                  )
-                }
-                case _ =>
-              }
-            }
-            case Dialog.Result.No => {
-              // Continuous
-              prompt_double("How often (in simulation-time seconds) do you want more agents?") match {
-                case Some(rate) => {
-                  prompt_int("How many agents?") match {
-                    case Some(num) => {
-                      sim.generators += new ContinuousGenerator(
-                        sim,
-                        polygon_roads1.toList.flatMap(r => r.all_lanes),
-                        polygon_roads2.toList.flatMap(r => r.all_lanes),
-                        rate, num
-                      )
-                    }
-                    case _ =>
+            // Ask: fixed (how many) or continuous (how many per what time)
+            // TODO improve the UI.
+            Dialog.showOptions(
+              message = "Want a fixed, one-time burst or a continuous generator?",
+              optionType = Dialog.Options.YesNoCancel, initial = 0,
+              entries = Seq("Constant", "Continuous")
+            ) match {
[…]

### 2012-02-09 papers/itsc_2012/root.tex
converted paper draft to ieee format

--- 2012-02-08 papers/itsc_2012/draft.tex
@@
-\documentclass[12pt]{article}
+\documentclass[letterpaper, 10 pt, conference]{ieeeconf}  % Comment this line out
+                                                          % if you need a4paper
+%\documentclass[a4paper, 10pt, conference]{ieeeconf}      % Use this line for a4
+                                                          % paper
 
-% TODO grab the ieee format
+\IEEEoverridecommandlockouts                              % This command is only
+                                                          % needed if you want to
+                                                          % use the \thanks command
+\overrideIEEEmargins
+% See the \addtolength command later in the file to balance the column lengths
+% on the last page of the document
 
-\title{Paper}
-\author{Dustin Carlino, Mike Depinet, Piyush Khandelwal, Peter Stone}
-\date{...}
+\title{\LARGE \bf
+TODO Name this thing
+}
+
+% TODO which style do we use for authors? and do we list all 4 of us here?
+
+%\author{ \parbox{3 in}{\centering Huibert Kwakernaak*
+%         \thanks{*Use the $\backslash$thanks command to put information here}\\
+%         Faculty of Electrical Engineering, Mathematics and Computer Science\\
+%         University of Twente\\
+%         7500 AE Enschede, The Netherlands\\
+%         {\tt\small h.kwakernaak@autsubmit.com}}
+%         \hspace*{ 0.5 in}
+%         \parbox{3 in}{ \centering Pradeep Misra**
+%         \thanks{**The footnote marks may be inserted manually}\\
+%        Department of Electrical Engineering \\
+%         Wright State University\\
+%         Dayton, OH 45435, USA\\
+%         {\tt\small pmisra@cs.wright.edu}}
+%}
+
+\author{Huibert Kwakernaak$^{1}$ and Pradeep Misra$^{2}$% <-this % stops a space
+\thanks{$^{1}$H. Kwakernaak is with Faculty of Electrical Engineering, Mathematics and Computer Science,
+        University of Twente, 7500 AE Enschede, The Netherlands
+        {\tt\small h.kwakernaak at papercept.net}}%
+\thanks{$^{2}$P. Misra is with the Department of Electrical Engineering, Wright State University,
+        Dayton, OH 45435, USA
+        {\tt\small p.misra at ieee.org}}%
+}
 
 \begin{document}
+
 \maketitle
+\thispagestyle{empty}
+\pagestyle{empty}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{abstract}
+Write this.
+\end{abstract}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{INTRODUCTION}
 
@@
 
+Write an intro.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\section{Architecture}
+\section{ARCHITECTURE}
 
@@
 
-\section{Map Construction from OSM}
+\section{MAP CONSTRUCTION FROM OSM}
 
-% briefly introduce OSM, advantages, etc
+% TODO briefly introduce OSM, advantages, etc
 
@@
 
-\section{Simulation}
+\section{SIMULATION}
 
@@
 
-\section{Experimental Results}
+\section{EXPERIMENTAL RESULTS}
 
@@
 
-\section{Notes}
+\section{CONCLUSIONS}
 
-possible background/links
+Write me.
+
+\addtolength{\textheight}{-12cm}  % This command serves to balance the column lengths
+                                  % on the last page of the document manually. It shortens
+                                  % the textheight of the last page by a suitable amount.
+                                  % This command does not take effect until the next page
+                                  % so it should come on the page before the last. Make
+                                  % sure that you do not shorten the textheight too much.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section*{APPENDIX}
+
+Probably stuff about getting and using the code.
+
+\section*{ACKNOWLEDGMENT}
+
+Dr. Quinlan has certainly been helpful.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+% Notes.
+
+%possible background/links
+
 %http://www.caliper.com/transmodeler/default.htm
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{thebibliography}{99}
+
+\bibitem{c1} G. O. Young, ÒSynthetic structure of industrial plastics (Book style with paper title and editor),Ó 	in Plastics, 2nd ed. vol. 3, J. Peters, Ed.  New York: McGraw-Hill, 1964, pp. 15Ð64.
+\end{thebibliography}
+
 \end{document}

### 2012-02-09 papers/itsc_2012/generate.sh

@@ -1,2 +1,2 @@
 #!/bin/sh
-pdflatex draft.tex
+pdflatex root.tex

### 2012-02-14 utexas/map/make/Checker.scala
fixed all stagnation issues except for a theoretical agent-following gridlock by optimistic lock yielding and squeezing onto tiny edges in reasonable ways.

@@ -9,7 +9,7 @@ object Checker {
 
     check_edges(g)
 
-    check_turn_length(g)
+    //check_turn_length(g)
 
     // TODO check connectivity more stringently? really not needed; I trust
     // Tarjan's.
@@ -34,7 +34,7 @@ object Checker {
 
   // This is a weak, but fast, way to measure connectivity
   def check_edges(g: Graph) = {
-    Util.log("Checking edges for connectivity")
+    Util.log("Checking edges for connectivity and length")
     Util.log_push
 
     for (e <- g.edges) {
@@ -43,6 +43,12 @@ object Checker {
       }
       if (e.prev_turns.size == 0) {
         Util.log("Nothing leads to " + e)
+      }
+
+      // Haha, we end up with some REALLY short edges sometimes...
+      if (e.length == 0.0) {
+        Util.log(e + " has ZERO length")
+        Util.log("  Points: " + e.road.points)
       }
     }
 

### 2012-02-14 utexas/map/make/Reader.scala

@@ -154,6 +154,12 @@ class Reader(fn: String) {
           edges(e_id) = e
           e.lane_num = e_lane
           e.set_lines(e_lines.toList)
+          // Sucks, but we have some REALLY small edges.
+          if (e.length <= 0.0) {
+            // Cheat. Agents NEED the ability to exist somewhere on this edge.
+            //Util.log("WARNING " + e + " is tiny")
+            e.length = 0.1
+          }
 
           // tell the road about the edge, too
           // TODO edges in the xml will be ordered by their id, which we created

### 2012-02-14 utexas/sim/Behaviors.scala

@@ -180,26 +180,49 @@ class AutonomousBehavior(a: Agent) extends Behavior(a) {
       return Act_Done_With_Route()
     } else {
       // So, three possible constraints.
-      val v1 = stop_how_far_away match {
-        case Some(dist) => accel_to_end(dist)
+      val a1 = stop_how_far_away match {
+        case Some(dist) => {
+          // in most normal circumstances, we take away end_threshold, since
+          // stopping right at the end of the edge makes us technically enter
+          // the intersection
+          // but when we're in the middle of a turn and that would make the
+          // distance be <= 0, we don't want to stop; that causes deadlock. it
+          // means the next edge is tiny. so just stop along it where we can.
+          val go_this_dist = a.at.on match {
+            // just go halfway whatever's available.
+            case t: Turn if dist <= cfg.end_threshold => accel_to_end(dist / 2)
+            case _ => dist - cfg.end_threshold
+          }
+          accel_to_end(go_this_dist)
+        }
         case _          => Double.MaxValue
       }
-      val v2 = follow_agent match {
+      val a2 = follow_agent match {
         case Some(f) => accel_to_follow(f, follow_agent_how_far_away)
         case _       => Double.MaxValue
       }
-      val v3 = accel_to_achieve(min_speed_limit)
+      val a3 = accel_to_achieve(min_speed_limit)
 
-      //if (a.id == 15) {
-      /*Util.log("%s's choices: %s, %s, speed lim %f".format(
-        a, stop_how_far_away, follow_agent, min_speed_limit
-      ))*/
-      /*Util.log("%s's choices: %.3f, %.3f, %.3f".format(
-        a, v1, v2, v3
-      ))*/
-      //}
+      val conservative_accel = math.min(a1, math.min(a2, a3))
 
-      val conservative_accel = math.min(v1, math.min(v2, v3))
+      // let the other intersections that are letting us go know that we
+      // won't be going just yet,
+      // IF we are not going to make a move this tick.
+
+      // This means we're holding a valuable resource...
+      if (a.speed == 0.0 && conservative_accel == 0.0) {
+        /*a.at.on match {
+          case t: Turn if a2 == 0.0 => {
+            // And the cause is an agent in front of us. If this situation persists,
+            // we could enter gridlock.
[…]

### 2012-02-14 utexas/sim/Intersection.scala

@@ -19,13 +19,14 @@ class Intersection(val v: Vertex) {
 
   // Just delegate.
   def should_stop(a: Agent, turn: Turn, far_away: Double) = policy.should_stop(a, turn, far_away)
+  def yield_lock(a: Agent) = policy.yield_lock(a)
   def unregister(a: Agent) = policy.unregister(a)
 
   // Multiple agents can be on the same turn; the corresponding queue will
   // handle collisions. So in fact, we want to track which turns are active...
   // but we have to know how many are on the turn to know when nobody's
   // attempting it.
-  var turns = MutableMap[Turn, Int]()
+  val turns = MutableMap[Turn, Int]()
 
   // Check for collisions by detecting agents abnormal changes in ordering.
   def end_step(): Unit = {
@@ -72,6 +73,7 @@ abstract class Policy(intersection: Intersection) {
   def should_stop(a: Agent, turn: Turn, far_away: Double): Boolean
   def validate_entry(a: Agent, turn: Turn): Boolean
   def handle_exit(a: Agent, turn: Turn)
+  def yield_lock(a: Agent)
   def unregister(a: Agent)
 
   // TODO common stuff for figuring out first-req
@@ -86,6 +88,7 @@ class NeverGoPolicy(intersection: Intersection) extends Policy(intersection) {
   def should_stop(a: Agent, turn: Turn, far_away: Double) = true
   def validate_entry(a: Agent, turn: Turn) = false
   def handle_exit(a: Agent, turn: Turn) = {}
+  def yield_lock(a: Agent) = {}
   def unregister(a: Agent) = {}
 }
 
@@ -95,8 +98,10 @@ class StopSignPolicy(intersection: Intersection) extends Policy(intersection) {
   // case, the first person has to pause a bit longer before continuing.
   var current_owner: Option[Agent] = None
   var queue = List[Agent]()
+  val yielders = new MutableSet[Agent]()
 
   def should_stop(a: Agent, turn: Turn, far_away: Double): Boolean = {
+    // TODO remove once stagnation fixed.
     Agent.sim.debug_agent match {
       case Some(agent) if agent == a => {
         Util.log(a + " wants to go, but locked from " + intersection.v + " by " + current_owner)
@@ -119,6 +124,14 @@ class StopSignPolicy(intersection: Intersection) extends Policy(intersection) {
     current_owner match {
       case Some(owner) if a == owner => return false
       case _       =>
+    }
+
+    // Did they have the lock, but temporarily yielded it to break deadlock? If
[…]

### 2012-02-16 papers/itsc_2012/root.tex
fixed all stagnation issues except for a theoretical agent-following gridlock by optimistic lock yielding and squeezing onto tiny edges in reasonable ways.
some initial stuff for the introdcution of the paper - still needs a ton of more work to be readable

@@
 \author{Huibert Kwakernaak$^{1}$ and Pradeep Misra$^{2}$% <-this % stops a space
-\thanks{$^{1}$H. Kwakernaak is with Faculty of Electrical Engineering, Mathematics and Computer Science,
-        University of Twente, 7500 AE Enschede, The Netherlands
-        {\tt\small h.kwakernaak at papercept.net}}%
-\thanks{$^{2}$P. Misra is with the Department of Electrical Engineering, Wright State University,
-        Dayton, OH 45435, USA
-        {\tt\small p.misra at ieee.org}}%
+% \thanks{$^{1}$H. Kwakernaak is with Faculty of Electrical Engineering, Mathematics and Computer Science,
+%         University of Twente, 7500 AE Enschede, The Netherlands
+%         {\tt\small h.kwakernaak at papercept.net}}%
+% \thanks{$^{2}$P. Misra is with the Department of Electrical Engineering, Wright State University,
+%         Dayton, OH 45435, USA
+%         {\tt\small p.misra at ieee.org}}%
 }
+
+\long\def\commentp#1{{\bf **Peter: #1**}}
+\long\def\commentpk#1{{\bf **Piyush: #1**}}
+\long\def\commentm#1{{\bf **Mike: #1**}}
+\long\def\commentd#1{{\bf **Dustin: #1**}}
 
@@
 
+\commentpk{This section is still under construction. Do not review.}
+
 % what are our goals with this paper?
@@
 
-Write an intro.
+Autonomous vehicles have seen tremendous progress in recent years.
+In 2004, no autonomous vehicle was able to complete the course set out in the first DARPA Challenge.
+By 2007, 5 vehicles finished the course of 130 miles in urban environments.
+By 2011, autonomous vehicles projects have now touched 160,000 miles and an autonomous car drove for 13,000 miles in Europe.
+Autonomous cars are here to stay, and this changes how traffic is dealt with in the forseeable future.
+The infrastructure will change to support autonomous vehicles, but it is hard to say more about our traffic demands.
+Will we be able to gain more roads in the future.
+Do we have the space to expand and will autonomous vehicles be able to reach your doorstep.
+
+This paper presents a new methodology for micro-simulating autonomous traffic for such situations where we will continue to use existing roads for autonomous vehicles.
+We believe that the autonomous traffic will be controlled via an overseer.
+The overseer can define behaviors for indivudal agents and individual intersections.
+The overseer itself can be a centralized and decentralized system that is capable for making such decisions in real-time.
+
+To implement such a methodology, we have come up with our own traffic simulator called AORTA.
+In AORTA we focus on the behavior of agents and intersections.
+Our simulator is generic enough to take advantage of implementing behaviors for cars which can represent human driving, along with standardized traffic light signals.
+We believe that that traffic light signals will be able to do better once we start using Autonomous Intersection Managers such as AIM.
+
+To use road data, we can use OpenStreetMap.
+xml files exported using OSM are directly read by a tool and parsed to create an understandable map.
+Our simulator then uses this map to create the simulation and can introduce an army of agents to simulate traffic inside the simulator.
+
+\commentpk{Needs introduction to the rest of the paper here}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{RELATED WORKS}
+
+\commentpk{This section is still under construction. Do not review.}
+
+A number of good micro-simulators have been created in the past decade. In this section we compare some of the similarities and differences with other such simulators.
+
+One such
 
@@
 
+systems ideas.. trylocks, atomic sim update, deadlock issues
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2012-02-16 papers/itsc_2012/root.tex
WIP fleshing out agent section of paper

@@
 %    about congestion)
+% be sure to emphasize open source!
 
@@
 steps:
-\begin{itemize}
+\begin{enumerate}
   \item Introducing new agents into the map
@@
   \item Allowing each agent to react
-\end{itemize}
+\end{enumerate}
+
+Agents have few enforced constraints -- they cannot exceed their physical
+acceleration capabilities, travel in reverse, or travel past the end of a
+traversable without appearing on the next. Intersection policies and avoiding
+collisions on the same edge are the source of the other constraint in the
+simulation, but it is the job of a behavior to obey. There are three main
+components of the simulation, each configurable by extending default classes:
+agent behaviors, routing strategies, and intersection policies. Each will be
+described below.
+
+justify arbitrary constants, like deaccel for cars
+
+scale
+
+systems ideas, or limitations.. trylocks, atomic sim update, deadlock issues
+
+limit: no lane-changing
+
+\subsection{Spawning}
+
+in realtime!
+
+\subsection{Collision checking}
 
@@
 guarantee connectedness, there are obviously circuits. Instead, collisions are
-checked after every agent has stepped forward in time. Queues remember the
-original order of agents before a step, and afterwards can verify the order is
-still legal, with new agents on the end and existing agents in the same order.
-Likewise, intersections check for collisions after everybody has moved by
-examining the turn each agent is performing and verifying that no other agent is
-simultaneously performing a conflicting turn.
+checked after every agent has stepped forward in time, guaranteeing that a
+time-step is atomic. Queues remember the original order of agents before a step,
+and afterwards can verify the order is still legal, with new agents on the end
+and existing agents in the same order.  Likewise, intersections check for
+collisions after everybody has moved by examining the turn each agent is
+performing and verifying that no other agent is simultaneously performing a
+conflicting turn.
 
-invariants for agents. interface of behavior and intersection policies.
+\section*{Agent Reactions}
 
-lookahead... worst-case stop dist argument is everywhere. safe spawning, too.
+Each agent has a ``behavior'' governing it. When the agent travels past the end
+of an edge during its step, the behavior picks the turn to pursue. Each tick,
+the behavior can make the agent react by performing one of two actions:
+disappearing from the map (only when the agent is at rest and is done with its
+route) or setting an acceleration for the next step. In the future, another
+possible action will be initiating a lane-change or merge.
 
-intersection policies... stop sign, traffic light.
+Currently, the primary behavior used by all agents is the ``route-following
+behavior.'' It is a generalized, baseline behavior that guarantees not to
+collide with another agent or enter an intersection at the wrong time. It avoids
+some forms of deadlock (in which no agent can move due to interdependencies),
+but there is currently one known flaw. This baseline behavior picks the fasrest
+conservative (safe) choice of acceleration at each step. It could be easily
+extended to mimic human drivers by introducing a higher reaction time delay, for
+instance.
 
-justify arbitrary constants, like deaccel for cars
+% TODO diagram for lookahead
 
-scale
+lookahead... worst-case stop dist argument is everywhere.
 
-systems ideas.. trylocks, atomic sim update, deadlock issues
+\section*{Routing Strategies}
+
+\section*{Intersection Policies}
+
+and future plans
 

### 2012-02-17 papers/itsc_2012/root.tex
a "bit" more on sim

@@
 
+Section 2 will review other traffic simulators. Section 3 will introduce the
+overall architecture of AORTA. Section 4 describes how OSM maps are used.
+Section 5 explains the simulation engine and the properties AORTA posses that
+are useful for general traffic experiments. Section 6 gives some initial
+results from such an experiment, and Section 7 concludes.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-programs: build map, UI (describe what it can do), headless
+programs: build map, UI (describe what it can do -- particularly demand
+generation), headless
 
@@
 
-% TODO briefly introduce OSM, advantages, etc
+\commentd{briefly introduce OSM, advantages, etc}
 
@@
 
-% TODO i don't think this'll have much coverage in the rest of the paper; should
-% we describe wards?
+\commentd{i don't think this'll have much coverage in the rest of the paper; should
+ we describe wards?}
 
@@
 Discrete-time simulation is used, meaning an agent accelerates at a fixed rate
-for the duration of a ``tick'' to achieve a new position and velocity. The tick
-duration $dt$ is configurable and fixed for the simulation. Agents know this
-value and use it to reason about collision avoidance and satisfying other
-constraints, so setting this value to the standard ``reaction time'' of a human
-driver may be appropriate. Each tick, the simulation performs the following
-steps:
+for the duration of a ``tick'' (also referred to as ``step'') to achieve a new
+position and velocity. The tick duration $dt$ is configurable and fixed for the
+simulation. Agents know this value and use it to reason about collision
+avoidance and satisfying other constraints, so setting this value to the
+standard ``reaction time'' of a human driver may be appropriate. Each tick, the
+simulation performs the following steps:
+
 \begin{enumerate}
@@
 
-in realtime!
+in realtime, safely!
+
+and architecture review (generators and thread pool)
 
@@
 
-\section*{Agent Reactions}
+\section{Agent Reactions}
 
@@
 extended to mimic human drivers by introducing a higher reaction time delay, for
-instance.
+instance, or tune rates of acceleration for optimizing fuel efficiency.
 
-% TODO diagram for lookahead
+Since agents know the fixed $dt$ duration, they may reason about what could
+happen during the next tick and avoid collision and illegal intersection entry.
+The conservative analysis is to determine worst case -- how far could an agent
+travel by the end of the next tick? If they accelerate as fast as possible for
+$dt$, this gives them the longest possible traveling distance and highest
+possible speed. Adding the stopping distance for this speed to the traveling
+distance gives the worst-case distance.
+\commentd{how much formula? pretty simple so far, but can certainly include them.
+ should this part be worded like a proof? (because it can be)}
 
-lookahead... worst-case stop dist argument is everywhere.
+The worst-case distance may exceed the distance remaining on the agent's current
+traversable. That means they could start or finish a turn during the next tick.
+Because edges are often extremely short (less than $???$ meters), they may
+travel through several traversables during the next move. Thus, any conservative
+behavior must look ahead until the worst-case distance is exhausted, asking the
+route strategy which turn an agent will choose in the future. At each
+traversable considered, there are 3 constraints to satisfy: obeying the
+intersection, avoiding collision with another agent, and obeying speed limits.
+Each constraint contributes the max acceleration that will obey the limit, so
+ultimately the result of lookahead will be the minimum of these candidates.
 
-\section*{Routing Strategies}
+A behavior's interaction with an intersection policy amounts to polling it with
+the turn the agent wants to perform, the distance along the incoming edge the
+agent will be, and how long the agent has been waiting (so policies can enforce
+a required delay before entering the intersection). The intersection policy
+simply orders the agent to ``stop'' or ``continue.'' The lookahead considers no
+intersections beyond the first that issues ``stop'', since the behavior cannot
+legally proceed past that one anyway.  The acceleration to avoid entering an
+intersection attempts to end a configurable threshold back from the end of a
+traversable to account for crosswalks and as an epsilon value, since achieveing
+a distance along an edge equal to its length means spilling over into the next
+traversable, which the intersection has banned. Sometimes stopping behind this
+threshold is not possible -- namely, when the edge length is shorter -- and in
+that case, anywhere before the end is acceptable.
+\commentd{include some form of piyush's proof about accel to end?}
 
-\section*{Intersection Policies}
+When the lookahead analysis is considering some future traversable, it only has
+to consider the agent closest to the start in the present. An edge may only be
+entered from a turn, so the intersection policy has the burden of ensuring no
+agent performs a turn that leads to the same edge simultaneously. Following an
+agent safely means maintaining a configurable following distance behind the
+agent. Worst-case analysis is again used: the most dangerous move the agent in
+front could perform is decelerating as fast as possible the next tick, and the
+worst-case traveling distance for the agent behind is already known from
+lookahead. Maintaining following distance from agents actually enables a form of
+deadlock in the simulation. If an agent is forced to stop in an intersection
+mid-turn due to its destination edge filling up with other agents, then nobody
+may cross the blocked intersection. With many agents in a small area, this
+situation may occur several times in an interdependent way. The solution
+currently being investigated will require behaviors to ensure a turn may be
+fully completed before starting it. This will prevent intersections from ever
+being blocked.
+\commentd{diagram showing deadlock}
+\commentd{i really hope to fix this issue soon, but it's not an immediately easy
+fix}
+
+Finally, obeying a road's speed limit is a simple constraint. Like many
+parameters of the map, speed limits are inferred from the road type claimed by
+the OSM source.
+
+\commentd{diagram for example of lookahead?}
+
+\commentd{mention epsilon issues?}
+
+\section{Routing Strategies}
+
+Picking accelerations that optimize some objective functions is a separate task
+from picking a path to travel. Thus, the behavior described above can consult
+any route strategy. Currently two simple implementations are available. One
+statically plans a route to some goal using standard the A* search; the other
+performs a drunken walk, making random choices at each intersection. The latter
+lazily plans several steps ahead, only when the behavior's lookahead engine
+needs to know. Future experiments in dynamic replanning or hierarchial planning
+can be performed without modifying any code other than a creating new route
+implementation.
+
+\section{Intersection Policies}
+
+when is an agent waiting?
 
 and future plans
+
+\section{Determinism and other simulation properties}
 

### 2012-02-17 papers/itsc_2012/root.tex
paper: intersection policies

@@
 
+illustration of turn conflicts
+
 \subsection{Pass 1}
@@
 
-in realtime, safely!
+\commentd{architecture review (generators and thread pool)}
 
-and architecture review (generators and thread pool)
+Once an agent's route is ready, the agent waits alongside its starting edge. At
+the beginning of each tick, they enter with zero initial speed, if it is safe to
+do so. The queue for the edge determines safety. Because behaviors must enforce
+speed limit along an edge by the time an agent enters the road, then if there is
+no agent within the worst-case distance (travel distance at the speed limit,
+plus stopping distance) before the spawn point, then the new agent may enter,
+whether the simulation is running or paused. It is unnecessary to predict which
+agents could enter the edge soon. As a result, there is a buffer space at the
+beginning of the edge where a new agent may never spawn. This disqualifies some
+short edges from being spawn points, but this is not an issue.
 
@@
 
-when is an agent waiting?
+Agents and behaviors (on behalf of an agent) poll an intersection as they
+approach it, asking if they should stop or not. A policy is free to prevent
+agents from performing conflicting turns concurrently by any means. Two main
+policies exist already, but implementing the interface for extensions is simple.
 
-and future plans
+The stop sign policy works by maintaining a current owner and a queue of waiting
+agents. An agent can enter the queue only when they are within some threshold of
+the start of their turn (a constraint satisfied by the default lookahead
+behavior), even if that is several (rather short) edges and turns away. After
+the head of the queue has waited for a configurable pause, they obtain the lock
+and have exclusive access to the intersection.
+
+A form of deadlock is possible in a sequence of short edges and turns. Agent
+$A_1$ acquires the lock to intersection $V_1$, but needs the lock to $V_2$ to
+proceed. Meanwhile, $A_2$ has the lock for $V_2$ and needs the lock for $V_1$.
+To deal with this circumstance, an agent's behavior can yield any held locks (a
+notion generalized to any intersection policy) when its speed is $0$ and its
+conservative acceleration is to remain at rest. The stop sign policy only
+removes an agent's lock if it has not started its turn (because otherwise, the
+agent is stalling in the intersection). The next time that agent polls, the
+policy restores the agent to the front of the queue.
+\commentd{hopefully we'll have a neater solution than this soon}
+
+\commentd{describe traffic signal policy, once it's working}
+
+Intersection policies are one of the most intriguing piece of the simulation to
+tweak. Future policies will include stop lights that adaptively learn better
+timings and an AIM-esque reservation system.
 
 \section{Determinism and other simulation properties}
+
+TODO
 
@@
 Write me.
+
+big future ideas:
+  contraflow
+  microtolling
+  dynamic replanning
+    autonomous intersections
 

### 2012-02-17 papers/itsc_2012/root.tex
some minor changes to abstract

--- 2012-02-17 papers/itsc_2012/root.tex
@@
 \begin{abstract}
-Write this.
+Autonomous vehicle technology have seen great advancements in recent years and are closer to being commercially viable.
+Driverless cars provide opportunities for optimizing traffic across cities in ways not possible before.
+This paper explores such scenarios and introduces a new traffic micro-simulator named AORTA.
+AORTA emphasizes research using scale simulations generated from OpenStreetMap data.
+This paper presents parsing map through the simulator and using the AORTA simulator.
+It concludes with some preliminary results that demonstrate the power of this simulator.
 \end{abstract}
@@
 Autonomous vehicles have seen tremendous progress in recent years.
-In 2004, no autonomous vehicle was able to complete the course set out in the first DARPA Challenge.
+The DARPA Challenge in 2004 saw the farthese distance being traversed autonomously as In 2004, no autonomous vehicle was able to complete the course set out in the first DARPA Challenge.
 By 2007, 5 vehicles finished the course of 130 miles in urban environments.

### 2012-02-17 papers/itsc_2012/root.tex
Added a bit to the paper.  Now back to the code.

--- 2012-02-17 papers/itsc_2012/root.tex
@@
 The infrastructure will change to support autonomous vehicles, but it is hard to say more about our traffic demands.
-Will we be able to gain more roads in the future.
-Do we have the space to expand and will autonomous vehicles be able to reach your doorstep.
+Will we be able to gain more roads in the future?
+Do we have the space to expand and will autonomous vehicles be able to reach your doorstep?
 
@@
 The overseer can define behaviors for indivudal agents and individual intersections.
-The overseer itself can be a centralized and decentralized system that is capable for making such decisions in real-time.
+The overseer itself can be a centralized or decentralized system that is capable of making such decisions in real-time.
 
@@
 coordinate system with no negative coordinates and the Y coordinate increasing
-upwards. This is done for convenince of manipulation in the UI, but the scale
+upwards. This is done for convenience of manipulation in the UI, but the scale
 and offset used are retained, so a coordinate in the final form of the map can
@@
 a slice of an OSM way are interpreted to be the center line of the road, and
-lanes are placed by projecting the width of a line perpendicularly from the
+lanes are placed by projecting the width of a lane perpendicularly from the
 center line in either direction. OSM encodes one-way streets, so in this case,
@@
 source, this is the best compromise. Finally, the geometry of a turn is
-extremely simplified as a single line between the ends of the edges it connects.
+simplified to a single line between the ends of the edges it connects.
 In the future, a more realistic curve could be drawn.
@@
 algorithm for locating strongly connected components to remove the components of
-the directed graph with the least members. (There is also a debugging switch to
+the directed graph with the fewest members. (There is also a debugging switch to
 leave these components in and highlight them in the UI.) The algorithm operates
@@
 intersect with another lane's lines. This results in a more accurate model, but
-the process is imperfect, so sometimes lines extend too far or not enough.
+the process is imperfect, so sometimes lines extend too far or not far enough.
 
@@
  we describe wards?}
+\commentm{We aren't actually using wards at all yet.  I think we should save this
+for a later paper.}
 
@@
 filtering out invalid ways in pass 1 is really not a clear process
-number of lanes... road types are wacky and wrong
+number of lanes and speed limits... road types are wacky and wrong
 turns are brittle.. detecting right-turn only lane is rare. shared center lane?
@@
 
-how easy would it be?
+In the future, we intend to support the ability to augment an AORTA map with
+data from non-OSM sources, such as information about the presence of extra lanes
+for left turns.  Using data such as GPS coordinates and road names, we could fuse
+this extra information into an osm-based AORTA map to make simulations on that map
+that much more realistic.
 
-match on gps, or on road metadata like names
-
-integrating with what other sources for spawning traffic, placing POI?
+We are also experimenting with explicitly stating where traffic should spawn and where
+it should go.  This allows more accurate (and consistent) simulations of situations such
+as rush hour.  In the future, we hope to be able to generate these demand descriptions
+from government-collected data about a municipality's road usage.
 
@@
 
-scale
+scale \commentm{I'm going to change this back to a 1:1 correlation with the real world.  (Hopefully this weekend.)}
 
@@
 that introduces a depdendency between traversables. A topological sort of a
-directed graph containing circuits is not possible, though, and since we
-guarantee connectedness, there are obviously circuits. Instead, collisions are
-checked after every agent has stepped forward in time, guaranteeing that a
-time-step is atomic. Queues remember the original order of agents before a step,
-and afterwards can verify the order is still legal, with new agents on the end
+directed graph containing circuits is not possible though, and since we
+guarantee connectedness, there are necessarily circuits in most maps. Instead, collisions are
+checked after every agent has stepped forward in time, treating each time-step
+as atomic. Queues remember the original order of agents before a step,
+and afterwards can verify that the order is still legal, with new agents on the end
 and existing agents in the same order.  Likewise, intersections check for
@@
 some forms of deadlock (in which no agent can move due to interdependencies),
-but there is currently one known flaw. This baseline behavior picks the fasrest
+but there is currently one known flaw. This baseline behavior picks the fastest
 conservative (safe) choice of acceleration at each step. It could be easily
@@
 intersection, avoiding collision with another agent, and obeying speed limits.
-Each constraint contributes the max acceleration that will obey the limit, so
+Each constraint contributes to the max acceleration that will obey the limit, so
 ultimately the result of lookahead will be the minimum of these candidates.

### 2012-02-17 papers/itsc_2012/root.tex
finished abstract and first 2 paragraphs of introduction

--- 2012-02-17 papers/itsc_2012/root.tex
@@
 \begin{abstract}
-Autonomous vehicle technology have seen great advancements in recent years and are closer to being commercially viable.
-Driverless cars provide opportunities for optimizing traffic across cities in ways not possible before.
-This paper explores such scenarios and introduces a new traffic micro-simulator named AORTA.
-AORTA emphasizes research using scale simulations generated from OpenStreetMap data.
-This paper presents parsing map through the simulator and using the AORTA simulator.
-It concludes with some preliminary results that demonstrate the power of this simulator.
+Autonomous vehicles have seen great advancements in recent years and such vehicles are now ever closer to being commercially available.
+The advent of driverless cars provides opportunities for optimizing traffic in ways not possible before.
+This paper introduces an open source multi-agent traffic simulator called AORTA, which stands for \textit{Approximately Orchestrated Road Network Automata}, and demonstrates its usage for optimizing autonomous traffic at a city-wide scale.
+AORTA creates scale simulations by generating maps using publicly available road data from \textit{OpenStreetMaps} (OSM).
+Simulations can be set up in the few minutes it takes to export data for the region of interest from OSM.
+AORTA focuses on interactions between agents by defining \textit{behaviors} for individual vehicles and intersections.
+These behaviors allow agents to interact with each other as well as with an external \textit{overseer} that is responsible for optimizing traffic flow.
+Overseers can interact with these agents to allow for dynamic route re-planning to reduce traffic load, optimized intersection navigation, and improved car following.
+\commentpk{rest devoted to application}
 \end{abstract}
@@
 \section{INTRODUCTION}
-
-\commentpk{This section is still under construction. Do not review.}
 
@@
 
-Autonomous vehicles have seen tremendous progress in recent years.
-The DARPA Challenge in 2004 saw the farthese distance being traversed autonomously as In 2004, no autonomous vehicle was able to complete the course set out in the first DARPA Challenge.
-By 2007, 5 vehicles finished the course of 130 miles in urban environments.
-By 2011, autonomous vehicles projects have now touched 160,000 miles and an autonomous car drove for 13,000 miles in Europe.
-Autonomous cars are here to stay, and this changes how traffic is dealt with in the forseeable future.
+Autonomous vehicle technology has made tremendous progress in the last decade.
+In 2004, the farthest distance travelled by a vehicle autonomously in the the DARPA Grand Challenge of 2004 was $11.9km$ \cite{cnnGrandChallenge2004}.
+By 2007, six of the competing teams completed the $96km$ course set for the DARPA Urban Challenge \cite{spectrumUrbanChallenge2007}.
+They did so while following the same traffic laws followed by human drivers, navigating around moving vehicles and following correct intersection precedence order.
+Since then, Google's driverless cars have clocked more than $250,000km$ on public roads in urban California, USA \cite{tedThrun2011}.
+In 2010, researchers from the University of Parma successfully completed an autonomous \textit{intercontinental} run from Parma, Italy to Shanghai, China \cite{cnnVislab2010}.
+The successful completion of all these milestones indicates that autonomous cars are here to stay, and are ever closer to becoming commercially available.
+
+The arrival of driverless cars presents a number of challenges that need to addressed in the near term.
+Road infrastructure needs to be updated to better support autonomous vehicles.
+A number of laws and safety concerns need to be addressed in the context of driverless cars.
+Human drivers need new methodologies to interact with autonomous drivers on public roads.
+At the same time, driverless cars provide opportunities of solving traffic problems in ways that were never before possible.
+For instance, autonomous cars can be quickly diverted to make way for emergency vehicles responding to an accident.
+Even day-to-day traffic issues such as congestion could potentially be greatly reduced by effective management of traffic flow.
+This paper attempts to address some of these problems.
+
 The infrastructure will change to support autonomous vehicles, but it is hard to say more about our traffic demands.
@@
 Our simulator then uses this map to create the simulation and can introduce an army of agents to simulate traffic inside the simulator.
+Real road data is necessary for such simulations due to ever limited space inside cities, which makes it so that in the near future all the infrastructure will become friendly towards autonomous vehicles.
+But we will still continue to use the same roads.
 
@@
 
-\begin{thebibliography}{99}
-
-\bibitem{c1} G. O. Young, ÒSynthetic structure of industrial plastics (Book style with paper title and editor),Ó 	in Plastics, 2nd ed. vol. 3, J. Peters, Ed.  New York: McGraw-Hill, 1964, pp. 15Ð64.
-\end{thebibliography}
+\bibliographystyle{IEEEtran}
+\bibliography{root}
 

### 2012-02-17 papers/itsc_2012/generate.sh

@@ -1,2 +1,3 @@
 #!/bin/sh
 pdflatex root.tex
+bibtex root

### 2012-02-17 papers/itsc_2012/root.tex
paper: architecture, map model, determinism properties

--- 2012-02-17 papers/itsc_2012/root.tex
@@
 
+\commentd{nevada even made autonomous cars legal:
+http://www.pcmag.com/article2/0,2817,2400400,00.asp}
 This paper presents a new methodology for micro-simulating autonomous traffic for such situations where we will continue to use existing roads for autonomous vehicles.
@@
 
+\commentd{UML diagram?}
+
+The architecture and implementation notes of AORTA follow. AORTA is divided into
+roughly three components, each somewhat independent of the others. The map model
+handles everything from building graphs from OSM source to pathfinding. This
+component is cleanly separated from the simulation engine, which adds the notion
+of agents, structures on top of map components that detect collisions, and the
+machinery to implement AORTA's model of car dynamics. If a future experiment
+demands macrosimulation or cellular automata method, no changes to the map
+infrastructure are needed. Finally, there is a UI using Java Swing, which is
+(out of necessity) tied to both the map and simulation components. The software
+has a documented API, is open-source, and contains all of the code for the
+experiments described in Section 6: it is designed for extensibility.
+
 AORTA's implementation uses Scala, a language implemented on the Java Virtual
-Machine. Scala provides the advantage of functional programming constructs
-without imposing them; parts of AORTA or extensions or clients can just as
-easily be written in the more familiar Java.
+Machine. Scala provides the advantage of functional programming constructs while
+still permitting imperative style. Parts of AORTA or extensions or clients can
+just as easily be written in Java.
 
-The architecture of AORTA is described below.
-
-thread pools?
-
-programs: build map, UI (describe what it can do -- particularly demand
-generation), headless
-
-api / extending it
-maps separate from simulation separate from UI (mostly)
+There are three standalone applications in AORTA. The map builder takes an OSM
+file as input and produces an AORTA graph as output. There is a UI, using Swing,
+to interactively explore the map and run simulations. Specific traffic patterns
+can quickly be simulated by selecting a source and destination group of roads in
+the UI. Finally, there is a headless mode that runs some specific experiment
+without the overhead of visualization. The UI and headless mode both take a
+graph as input, meaning the map construction step must be run once per OSM
+input, but the same graph can be reused indefinitely.
 
@@
 
-\commentd{briefly introduce OSM, advantages, etc}
+Our goal is to simulate traffic on existing infrastructure, not contrived,
+randomly generated maps. We do this by utilizing OpenStreetMaps, a community
+project offering user-editable, free maps.
 
@@
 
-roads, verts, edges, wards, traversables, turns...
+AORTA's map format reflects our map model, which will now be described.
 
-edges are made up of multiple lines.. this could be control pts of a curve
-(hereon, line means line segment)
-(turns are also traversable)
+\commentd{terminology sucks below. suggestions?}
 
-illustration of turn conflicts
+\emph{Roads} are undirected edges between \emph{vertices}, which are
+intersections between multiple roads. A road contains one or two sets of
+directed \emph{edges}, each of which represents a single lane of travel. One-way
+roads are supported. An edge leads from one vertex to another; the second vertex
+then contains a mapping from each incoming edge to some outgoign edge; each pair
+is a \emph{turn}. Strictly speaking, the multigraph consists of \emph{lanes} as
+vertices (this is confusing, because lanes are \emph{edges} in our model) and
+\emph{turns} as edges. \emph{Roads} are a super-structure with meta-data,
+bundling several \emph{lanes} together, and \emph{vertices} serve to bundle
+together \emph{turns}.
+
+Geometrically speaking, a road has a sequence of points defining the line
+segments constituting the center yellow line. These points could also be
+interpreted as control points for a curved road, but AORTA approximates using
+straight lines. Edges mimic this geometry, projecting the center line a lane's
+width away. Turns are modeled with a single straight line connecting the end of
+one edge to the start of another. Since both edges and turns can support
+traffic, they are grouped together as \emph{traversables}, meaning they have
+some sequence of line segments and a total length.
 
@@
 
+\commentd{illustration of turn conflicts}
+
+AORTA has a low-granularity model of turn conflict. If two vehicles at any point
+along two different turns could ever come into contact, then the turns are
+always in conflict, no matter where the agent specifically is along that turn.
+The alternative of tiling intersections would provide more granularity, but it
+adds computational complexity and would be impractical for many of the
+intersection geometries inferred from OSM.
+
 \subsubsection{Ensuring connectivity}
@@
 
-We are also experimenting with explicitly stating where traffic should spawn and where
-it should go.  This allows more accurate (and consistent) simulations of situations such
-as rush hour.  In the future, we hope to be able to generate these demand descriptions
-from government-collected data about a municipality's road usage.
+Section 5.1 details a way to specify where traffic should originate and move
+towards using the UI. This allows a more accurate simulation of situations such
+as rush hour. In the future, we hope to be able to generate these demand
+descriptions from government-collected data about a municipality's road usage.
 
@@
 
-justify arbitrary constants, like deaccel for cars
-
-scale \commentm{I'm going to change this back to a 1:1 correlation with the real world.  (Hopefully this weekend.)}
-
-systems ideas, or limitations.. trylocks, atomic sim update, deadlock issues
-
-limit: no lane-changing
+\commentd{justify arbitrary constants, like deaccel for cars;
+          systems ideas, or limitations.. trylocks, atomic sim update, deadlock
+          issues;
+          limit: no lane-changing}
 
@@
 
-\commentd{architecture review (generators and thread pool)}
+\commentd{(generators and thread pool)}
 
@@
 
-TODO
+AORTA is designed for controlled experiments. Consequently, simulations are
+deterministically reprodcible. A single seed for the pseudo-random number
+generator, the input graph, and the generators completely determine the single
+outcome of any simulation. This determinism breaks down when generators use
+worker threads, since routes may be computed in different orders, causing agents
+to be ready to enter the road at different times.
+
+\commentd{and how will we solve this issue with threads and determinism?}
+
+\commentd{give some informal or formal (with numbers) idea of scale / how many
+agents we can simulate in different map sizes}
 
@@
 
+What credits do we need to give to OSM?
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2012-02-17 utexas/map/Road.scala

@@ -50,29 +50,31 @@ class Road(var id: Int, val points: List[Coordinate], val name: String,
   def pairs_of_points = points zip points.tail
 
   def speed_limit = Util.mph_to_si(road_type match {
-    case "residential"		=> 30
-    case "motorway"			=> 80
-    case "motorway_link"	=> 70 //Actually these don't have a speed limit legally...  35 is suggested, but NOBODY does that
-    case "trunk" 			=> 70
-    case "trunk_link"		=> 60
-    case "primary"			=> 65
-    case "primary_link"		=> 55
-    case "secondary"		=> 55
-    case "secondary_link"	=> 45
-    case "tertiary"			=> 45
-    case "tertiary_link"	=> 35
+    case "residential"    => 30
+    case "motorway"       => 80
+    // Actually these don't have a speed limit legally...  35 is suggested, but NOBODY does that
+    case "motorway_link"  => 70
+    case "trunk"          => 70
+    case "trunk_link"     => 60
+    case "primary"        => 65
+    case "primary_link"   => 55
+    case "secondary"      => 55
+    case "secondary_link" => 45
+    case "tertiary"       => 45
+    case "tertiary_link"  => 35
+    //case "unclassified"   => 40
+    //case "road"           => 40
+    case "living_street"  => 20
+    // TODO some of these we filter out in Pass 1... cross-ref with that list
+    case "service"        => 10 // This is apparently parking-lots basically, not feeder roads
+    case "services"       => 10
+    //case "track"          => 35
+    // I feel the need.  The need for speed.  Where can we find one of these?
+    case "raceway"        => 300
+    //case "null"           => 30
+    //case "proposed"       => 35
+    //case "construction"     => 20
     
-//    case "unclassified"		=> 40
-//    case "road"				=> 40
-    case "living_street"	=> 20
-    case "service"			=> 10 //This is apparently parking-lots basically, not feeder roads
-    case "services"			=> 10
-//    case "track"			=> 35
-    case "raceway"			=> 300 //I feel the need.  The need for speed.  Where can we find one of these?
-//    case "null"				=> 30
-//    case "proposed"			=> 35
-//    case "construction"		=> 20
-    
-    case _					=> 35 //Generally a safe speed, right?
+    case _                => 35 // Generally a safe speed, right?
   })
 }

### 2012-02-17 utexas/map/make/Checker.scala

@@ -46,8 +46,8 @@ object Checker {
       }
 
       // Haha, we end up with some REALLY short edges sometimes...
-      if (e.length == 0.0) {
-        Util.log(e + " has ZERO length")
+      if (e.length == 0.0 || e.length.isNaN) {
+        Util.log(e + " has length " + e.length)
         Util.log("  Points: " + e.road.points)
       }
     }

### 2012-02-17 utexas/map/make/Reader.scala

@@ -13,7 +13,7 @@ import utexas.map.{Coordinate, Road, Vertex, Edge, Direction, Line, TurnType,
                    Turn, Graph, Ward}
 import utexas.sim.Simulation
 
-import utexas.Util
+import utexas.{Util, cfg}
 
 class Reader(fn: String) {
   // per road
@@ -155,7 +155,7 @@ class Reader(fn: String) {
           e.lane_num = e_lane
           e.set_lines(e_lines.toList)
           // Sucks, but we have some REALLY small edges.
-          if (e.length <= 0.0) {
+          if (e.length <= cfg.min_lane_length) {
             // Cheat. Agents NEED the ability to exist somewhere on this edge.
             //Util.log("WARNING " + e + " is tiny")
             e.length = 0.1

### 2012-02-17 papers/itsc_2012/root.tex
finished first draft of abstract and introduction

--- 2012-02-17 papers/itsc_2012/root.tex
@@
 Autonomous vehicles have seen great advancements in recent years and such vehicles are now ever closer to being commercially available.
-The advent of driverless cars provides opportunities for optimizing traffic in ways not possible before.
+The advent of driver-less cars provides opportunities for optimizing traffic in ways not possible before.
 This paper introduces an open source multi-agent traffic simulator called AORTA, which stands for \textit{Approximately Orchestrated Road Network Automata}, and demonstrates its usage for optimizing autonomous traffic at a city-wide scale.
-AORTA creates scale simulations by generating maps using publicly available road data from \textit{OpenStreetMaps} (OSM).
+AORTA creates scale simulations by generating maps using publicly available road data from \textit{OpenStreetMap} (OSM).
 Simulations can be set up in the few minutes it takes to export data for the region of interest from OSM.
@@
 Autonomous vehicle technology has made tremendous progress in the last decade.
-In 2004, the farthest distance travelled by a vehicle autonomously in the the DARPA Grand Challenge of 2004 was $11.9km$ \cite{cnnGrandChallenge2004}.
+In 2004, the farthest distance traveled by a vehicle autonomously in the DARPA Grand Challenge was $11.9km$ \cite{cnnGrandChallenge2004}.
 By 2007, six of the competing teams completed the $96km$ course set for the DARPA Urban Challenge \cite{spectrumUrbanChallenge2007}.
-They did so while following the same traffic laws followed by human drivers, navigating around moving vehicles and following correct intersection precedence order.
-Since then, Google's driverless cars have clocked more than $250,000km$ on public roads in urban California, USA \cite{tedThrun2011}.
+They did so while following the same traffic laws followed by human drivers, navigating along with other moving vehicles, and following correct intersection precedence order.
+Since then, Google's driver-less cars have clocked more than $250,000km$ on public roads in urban California, USA \cite{tedThrun2011}.
 In 2010, researchers from the University of Parma successfully completed an autonomous \textit{intercontinental} run from Parma, Italy to Shanghai, China \cite{cnnVislab2010}.
@@
 
-The arrival of driverless cars presents a number of challenges that need to addressed in the near term.
+The arrival of driver-less cars presents a number of challenges that need to addressed in the near term.
 Road infrastructure needs to be updated to better support autonomous vehicles.
-A number of laws and safety concerns need to be addressed in the context of driverless cars.
+A number of laws and safety concerns need to be addressed in the context of driver-less cars.
 Human drivers need new methodologies to interact with autonomous drivers on public roads.
-At the same time, driverless cars provide opportunities of solving traffic problems in ways that were never before possible.
+At the same time, driver-less cars provide opportunities of solving traffic problems in ways that were never before possible.
 For instance, autonomous cars can be quickly diverted to make way for emergency vehicles responding to an accident.
 Even day-to-day traffic issues such as congestion could potentially be greatly reduced by effective management of traffic flow.
-This paper attempts to address some of these problems.
+This work attempts to address some of these problems.
 
-The infrastructure will change to support autonomous vehicles, but it is hard to say more about our traffic demands.
-Will we be able to gain more roads in the future.
-Do we have the space to expand and will autonomous vehicles be able to reach your doorstep.
+This paper introduces an open source multi-agent traffic micro-simulator called AORTA, which stands for \textit{Approximately Orchestrated Road Traffic Automata}.
+The AORTA simulator focuses on the interactions between individual vehicles and intersections, both of which act as \textit{agents} in AORTA's framework.
+Vehicles interact with other vehicles to pass and change lanes, and interact with intersections to decide when to cross them.
+Researchers developing applications on top of the AORTA simulator define \textit{behaviors} that these agents follow to interact with one another.
+To optimize traffic flow, external \textit{overseers} can coordinate with these agents to provide recommendations such as route suggestions or intersection signaling policies.
+An advantage of such an approach is that by assigning human-like behaviors such as the "car following model" \cite{brackstone1999car} to vehicles and traffic signal like policies to intersection, AORTA can just as easily simulate human traffic.
 
-This paper presents a new methodology for micro-simulating autonomous traffic for such situations where we will continue to use existing roads for autonomous vehicles.
-We believe that the autonomous traffic will be controlled via an overseer.
-The overseer can define behaviors for indivudal agents and individual intersections.
-The overseer itself can be a centralized and decentralized system that is capable for making such decisions in real-time.
+Like any other traffic micro-simulator \commentpk{check this term}, AORTA needs maps to run simulations.
+We generate our maps using road data available from OpenStreetMap (OSM) \cite{osm}, a Wikipedia like interface for world maps.
+Researchers using AORTA can export the relevant map area from OSM that is then parsed by AORTA to create scale simulations of the real world.
+Using real world road data is advantageous for a number of reasons.
+First, researchers using AORTA for human traffic simulations would find it advantageous to also have the ability to use real-world roads.
+Second, even as infrastructural changes will be made to the road infrastructure to accommodate driver-less cars, it is reasonable to assume that the \textit{same} road network will still be in use.
+Third, using real-world data from OSM allows setting up simulations for an area of interest in a matter of minutes, thereby allowing easy transfer of research results to any desired region.
 
-To implement such a methodology, we have come up with our own traffic simulator called AORTA.
-In AORTA we focus on the behavior of agents and intersections.
-Our simulator is generic enough to take advantage of implementing behaviors for cars which can represent human driving, along with standardized traffic light signals.
-We believe that that traffic light signals will be able to do better once we start using Autonomous Intersection Managers such as AIM.
+\commentpk{Needs 1 more para based on final application, and image on the top right, and introduction to the rest of the paper here once we finalize the sections}
 
-To use road data, we can use OpenStreetMap.
-xml files exported using OSM are directly read by a tool and parsed to create an understandable map.
-Our simulator then uses this map to create the simulation and can introduce an army of agents to simulate traffic inside the simulator.
-Real road data is necessary for such simulations due to ever limited space inside cities, which makes it so that in the near future all the infrastructure will become friendly towards autonomous vehicles.
-But we will still continue to use the same roads.
-
-\commentpk{Needs introduction to the rest of the paper here}
-
-Section 2 will review other traffic simulators. Section 3 will introduce the
-overall architecture of AORTA. Section 4 describes how OSM maps are used.
-Section 5 explains the simulation engine and the properties AORTA posses that
-are useful for general traffic experiments. Section 6 gives some initial
-results from such an experiment, and Section 7 concludes.
+% Section 2 will review other traffic simulators. Section 3 will introduce the
+% overall architecture of AORTA. Section 4 describes how OSM maps are used.
+% Section 5 explains the simulation engine and the properties AORTA posses that
+% are useful for general traffic experiments. Section 6 gives some initial
+% results from such an experiment, and Section 7 concludes.
 

### 2012-02-17 papers/itsc_2012/root.tex

@@
 
+% \commentd{nevada even made autonomous cars legal:
+% http://www.pcmag.com/article2/0,2817,2400400,00.asp}
+% This paper presents a new methodology for micro-simulating autonomous traffic for such situations where we will continue to use existing roads for autonomous vehicles.
+We believe that the autonomous traffic will be controlled via an overseer.
+The overseer can define behaviors for indivudal agents and individual intersections.
+The overseer itself can be a centralized or decentralized system that is capable of making such decisions in real-time.
+
 This paper introduces an open source multi-agent traffic micro-simulator called AORTA, which stands for \textit{Approximately Orchestrated Road Traffic Automata}.

### 2012-02-18 papers/itsc_2012/root.tex
finished related work section

@@
 \section{INTRODUCTION}
+\label{sec:introduction}
 
@@
 Road infrastructure needs to be updated to better support autonomous vehicles.
-A number of laws and safety concerns need to be addressed in the context of driver-less cars.
+A number of legal and safety concerns need to be addressed in the context of driver-less cars.
 Human drivers need new methodologies to interact with autonomous drivers on public roads.
@@
 
-This paper introduces an open source multi-agent traffic micro-simulator called AORTA, which stands for \textit{Approximately Orchestrated Road Traffic Automata}.
+This paper introduces an open source multi-agent microscopic traffic simulator called AORTA, which stands for \textit{Approximately Orchestrated Road Traffic Automata}.
 The AORTA simulator focuses on the interactions between individual vehicles and intersections, both of which act as \textit{agents} in AORTA's framework.
@@
 
-Like any other traffic micro-simulator \commentpk{check this term}, AORTA needs maps to run simulations.
+Like any other microscopic traffic simulator, AORTA needs maps to run simulations.
 We generate our maps using road data available from OpenStreetMap (OSM) \cite{osm}, a Wikipedia like interface for world maps.
@@
 
-\section{RELATED WORKS}
+\section{RELATED WORK}
+\label{sec:related_work}
 
-\commentpk{This section is still under construction. Do not review.}
+Computational processing power has made excellent advancements in the last two decades.
+Parallel computing has proved significant for traffic simulation through the use of Geographic Information System (GIS) software \cite{pursula1999simulation}.
+Advanced computing techniques have enabled microscopic models of traffic simulation to generate results at a meaningful scale (city-wide or greater).
+As a result, a number of good micro-simulators have come up in the past decade.
+We review a number of such simulators in this section along with other relevant work.
 
-A number of good micro-simulators have been created in the past decade. In this section we compare some of the similarities and differences with other such simulators.
+A number of traffic simulators already make use of OSM data.
+MATSim is one such multi-agent simulator that focuses on performing large-scale simulations in a relatively small amount of time \cite{balmer2009matsim}.
+Traffic demand is supplied to MATSim in the form of \textit{plans} that an individual may intend to follow in a given day.
+MATSim aims to iteratively improve these plans though offline computation to maximize traffic throughput.
+In other words, it helps individuals understand the best way to go about their daily routine to minimize time spent in transit.
+This goal is somewhat different from that of AORTA, where the demand supplied by individuals is taken as is.
+Instead, AORTA aims to reduce traffic congestion through interactions with individuals in real-time based on the current demand.
 
-One such
+Another popular open source traffic simulator that can use OSM data is SUMO \cite{SUMO2011}, which shares the same goals as AORTA and is fairly similar at a high level.
+SUMO has been used to study the effect of automated transportation systems, route planning of individual vehicles, as well as dynamically adapting traffic signal policies to increase traffic efficiency.
+AORTA and SUMO differ in the manner in which intersections are handled.
+SUMO uses traditional traffic signals, while AORTA adopts a much more general \textit{reservation} system.
+A reservation tells a car when it is safe to take its desired path through the intersection.
+Reservations can be alotted to vehicles through more traditional means such as traffic signals or stop-sign based intersection based precedence, or through more efficient methods designed specificaly for autonomous intersection management \cite{JAIR08-dresner}.
+\commentpk{does this difference seem a bit pedantic?}
+
+On the other hand, there are a number of simulators that specifically target autonomous traffic systems.
+One such system is AIM \cite{JAIR08-dresner}, which aims to optimize traffic flow between autonomous vehicles at a given intersection.
+AIM studies problems at the scale of a single intersection, while AORTA has the potential to apply some of the same research at a city-wide scale to real-world road networks.
+A few simulators focus on autonomous vehicles with sensors and actuators \cite{figueiredo2009approach}.
+These simulators are used to ensure that autonomous vehicles are behaving correctly while interacting with other vehicles and infrastructure.
+On the other hand, AORTA assumes that autonomous vehicles are capable of reasonable driving, and focuses on interactions between agents to optimize traffic flow.
+
+Autonomous agents have also been used in the past to model human traffic as well.
+For instance, an approach models human merging and lane changing behavior through the use of autonomous agents interacting with one another \cite{hidas2002modelling}.
+Another approach attempted to improve traffic congestion for human drivers by using a network of autonomous intersection.
+These intersections have the ability to interact with one another and execute a dynamic traffic signal policy minimizing overall wait time \cite{manikonda2001autonomous}.
+
+\commentpk{I suspect there are still more papers that might be more relevant - we'll definitely have some when we write the application}
+
+% In the realm of applications
+% It is also necessary to analyze some other simulators that use GIS data for microscopic traffic simulations, this is because it may be possible to export OSM data so that it can be used by these simulators, though the authors are not aware of such a connection.
+Incident response has been studied by the following paper (though perhaps only a minor citation as it is an application and does not really compare against AORTA, but it is something we would like to study this as well.
+\cite{huang2007gis}
 

### 2012-02-18 papers/itsc_2012/root.tex
headless reports how many agents actually move during ticks, to detect deadlock. small updates to paper.

@@
 
+\commentd{sounds fine to me, only thing is that we haven't actually focused on
+  dynamic re-planning or any kind of overseer yet, although theres nothing
+  stopping us from trying that idea again}
+
 \begin{abstract}
@@
 
-% \commentd{nevada even made autonomous cars legal:
-% http://www.pcmag.com/article2/0,2817,2400400,00.asp}
-% This paper presents a new methodology for micro-simulating autonomous traffic for such situations where we will continue to use existing roads for autonomous vehicles.
-We believe that the autonomous traffic will be controlled via an overseer.
-The overseer can define behaviors for indivudal agents and individual intersections.
-The overseer itself can be a centralized or decentralized system that is capable of making such decisions in real-time.
+\commentd{nevada even made autonomous cars legal:
+http://www.pcmag.com/article2/0,2817,2400400,00.asp}
+
+\commentd{another example long-term usage: fuel. I went to a talk by
+  somebody on the engr side about powermiling and optimizing battery usage}
+
+\commentd{we don't support lane-changing yet, and overseers arent re-implemented
+yet either}
 
@@
 
+\commentd{and we can do the above too. we talk about how to make the current
+  behavior more flawed and humanlike in the agent reactions section.}
+
 \commentpk{I suspect there are still more papers that might be more relevant - we'll definitely have some when we write the application}
@@
 
-\commentd{UML diagram?}
+\commentd{? UML diagram: 3 components with their capabilities listed inside, and
+arrows of interactions between them?}
 
@@
 demands macrosimulation or cellular automata method, no changes to the map
-infrastructure are needed. Finally, there is a UI using Java Swing, which is
-(out of necessity) tied to both the map and simulation components. The software
-has a documented API, is open-source, and contains all of the code for the
-experiments described in Section 6: it is designed for extensibility.
+infrastructure are needed, due to the software engineering. Finally, there is a
+UI using Java Swing, which is (out of necessity) tied to both the map and
+simulation components. The software has a documented API, is open-source, and
+contains all of the code for the experiments described in Section 6 -- it is
+designed for extensibility.
 
@@
 randomly generated maps. We do this by utilizing OpenStreetMaps, a community
-project offering user-editable, free maps.
-
-To run a simulation on a map from OSM, there is a one-time conversion process to
-transform the map into AORTA's format, which is also encoded with XML. It is
-split into 3 passes, each feeding the next.
+project offering user-editable, free maps.  To run a simulation on a map from
+OSM, there is a one-time conversion process to transform the map into AORTA's
+format, which is also encoded with XML. It is split into 3 passes, each feeding
+the next.
 
@@
 
-AORTA's map format reflects our map model, which will now be described.
+\commentd{a simple diagram of a road with a few edges connected maybe to a
+one-way}
 
-\commentd{terminology sucks below. suggestions?}
+\commentd{terminology is awkward. if we bold-face our notions of edge and vertex
+every time we use them, would that better distinguish them from a graph's
+verts/edges?}
 
 \emph{Roads} are undirected edges between \emph{vertices}, which are
-intersections between multiple roads. A road contains one or two sets of
+intersections between multiple roads. A \emph{road} contains one or two sets of
 directed \emph{edges}, each of which represents a single lane of travel. One-way
-roads are supported. An edge leads from one vertex to another; the second vertex
-then contains a mapping from each incoming edge to some outgoign edge; each pair
-is a \emph{turn}. Strictly speaking, the multigraph consists of \emph{lanes} as
-vertices (this is confusing, because lanes are \emph{edges} in our model) and
-\emph{turns} as edges. \emph{Roads} are a super-structure with meta-data,
-bundling several \emph{lanes} together, and \emph{vertices} serve to bundle
-together \emph{turns}.
+\emph{roads} are supported. An \emph{edge} leads from one \emph{vertex} to
+another; the second \emph{vertex} then contains a mapping from each incoming
+\emph{edge to some outgoing \emph{edge}; each pair is a \emph{turn}. Strictly
+speaking and with confusing terminology, the multigraph consists of lanes
+(\emph{edges}) as vertices and \emph{turns} as edges. \emph{Roads} are a
+super-structure with meta-data, bundling several \emph{edges} together, and
+\emph{vertices} serve to bundle together \emph{turns}.
 
-Geometrically speaking, a road has a sequence of points defining the line
-segments constituting the center yellow line. These points could also be
-interpreted as control points for a curved road, but AORTA approximates using
-straight lines. Edges mimic this geometry, projecting the center line a lane's
-width away. Turns are modeled with a single straight line connecting the end of
-one edge to the start of another. Since both edges and turns can support
-traffic, they are grouped together as \emph{traversables}, meaning they have
-some sequence of line segments and a total length.
+Geometrically, a road has a sequence of points defining the line segments
+constituting the center yellow line. These points could also be interpreted as
+control points for a curved road, but AORTA approximates using straight lines.
+Edges mimic this geometry, projecting the center line some multiple of a lane's
+width away in the appropriate direction. Turns are modeled with a single
+straight line connecting the end of one edge to the start of another (so right
+turns often have $0$ length). Since both edges and turns can support traffic,
+they are grouped together as \emph{traversables}, meaning they have some
+sequence of line segments and a total length.
 
@@
 The relevant parts of OSM's map model are nodes and ways. A node corresponds to
-a GPS location, and ways are links between many nodes, corresponding to roads
-and other paths. The first pass reads OSM's XML format, gathering ways as a
-series of coordinates. Ways intersect when a node is used by multiple ways.
+a GPS location, and ways are links between many nodes, corresponding to
+driveable streets and other paths. The first pass reads OSM's XML format,
+gathering ways as a series of coordinates. Ways implicitly intersect when a node
+is used by multiple ways.
 
@@
 
-Finally, coordinates are normalized from GPS (longitude, latitude) to a planar
-coordinate system with no negative coordinates and the Y coordinate increasing
-upwards. This is done for convenience of manipulation in the UI, but the scale
-and offset used are retained, so a coordinate in the final form of the map can
-still be mapped back to its original GPS form.
+Finally, coordinates are normalized from GPS (longitude, latitude) to a
+Cartesian coordinate system with no negative coordinates and the Y coordinate
+increasing upwards. This is only done for convenience of manipulation in the UI.
+Any coordinate is translated back to its original GPS form when used for
+computing traversable length.
 
@@
 
-OSM ways only implicitly intersect at common coordinates; AORTA's structure
-demands a traditional definition of edge linking exactly two vertices. Pass 2
-walks along each way's coordinates, dividing it into multiple undirected edges.
+AORTA's structure demands a traditional definition of edge linking exactly two
+vertices, rather than OSM's ways going through many common nodes. Pass 2 walks
+along each way's coordinates, dividing it into multiple undirected edges.
 
@@
 \subsubsection{Making Lanes}
+
+\commentd{the geometry was already explained in Map Model; I think it belongs
+there if we have to cut one}
 
@@
 intersections having the incorrect turns. However, without a more detailed
-source, this is the best compromise. Finally, the geometry of a turn is
-simplified to a single line between the ends of the edges it connects.
-In the future, a more realistic curve could be drawn.
+source, this is the best compromise. At the time of writing, even sources like
+Google Maps lack this information. (TODO how do you cite a lack of something?)
+Finally, the geometry of a turn is simplified to a single line between the ends
+of the edges it connects.  In the future, a more realistic curve could be drawn.
 
@@
 always in conflict, no matter where the agent specifically is along that turn.
-The alternative of tiling intersections would provide more granularity, but it
-adds computational complexity and would be impractical for many of the
-intersection geometries inferred from OSM.
+The alternative of tiling intersections (like \cite{JAIR08-dresner} does) would
+provide more granularity, but it adds computational complexity and would be
+impractical for many of the complex intersection geometries inferred from OSM.
 
@@
 not reachable from some portions of the graph. Simulation and pathfinding would
-suffer greatly if this was the case, so pass 3 uses Tarjan's linear-time
+suffer greatly if this remained the case, so pass 3 uses Tarjan's linear-time
 algorithm for locating strongly connected components to remove the components of
@@
 
+\commentd{this maybe doesn't belong in this paper}
+
 proposals to change OSM to fix these issues
@@
 for left turns. Using data such as GPS coordinates and road names, we could fuse
-this extra information into an osm-based AORTA map to make simulations on that map
+this extra information into an OSM-based AORTA map to make simulations on that map
 that much more realistic.

### 2012-02-18 papers/itsc_2012/generate.sh

@@ -1,3 +1,7 @@
 #!/bin/sh
 pdflatex root.tex
 bibtex root
+# pretty sure we have to re-run this twice to include the newly generated bib
+# and references
+pdflatex root.tex
+pdflatex root.tex

### 2012-02-18 utexas/map/Coordinate.scala

@@ -50,7 +50,7 @@ object Coordinate {
     
     // a is the square of half the chord length between the points
     val a = math.pow(math.sin(dLat / 2), 2) +
-    	      math.pow(math.sin(dLon / 2), 2) * Math.cos(lat1) * Math.cos(lat2)
+    	      math.pow(math.sin(dLon / 2), 2) * math.cos(lat1) * math.cos(lat2)
     // c is the angular distance in radians
     val c = 2 * math.atan2(math.sqrt(a), math.sqrt(1 - a))
     return radius * c

### 2012-02-18 utexas/sim/Agent.scala

@@ -27,7 +27,8 @@ class Agent(val id: Int, val graph: Graph, val start: Edge, val start_dist: Doub
 
   override def toString = "Agent " + id
 
-  def step(dt_s: Double) = {
+  // Returns true if we move or do anything at all
+  def step(dt_s: Double): Boolean = {
     assert(dt_s == cfg.dt_s)
 
     // To confirm determinism, enable one of these (more precision in doubles is
@@ -110,6 +111,7 @@ class Agent(val id: Int, val graph: Graph, val start: Edge, val start_dist: Doub
     }
 
     // TODO deal with lane-changing
+    return new_dist > 0.0
   }
 
   def how_long_idle = if (idle_since == -1.0)

### 2012-02-18 utexas/sim/Headless.scala

@@ -44,18 +44,22 @@ object Headless {
     Util.log("Starting simulation with time-steps of " + cfg.dt_s + "s")
     var last_real_time = 0.0
     var last_virtual_time = 0.0
+    var max_movements = 0
     while (!sim.done) {
       // Print every 1 literal second
       val now = timer.so_far
       //val now = sim.tick    // or every 1 virtual second
       if (now - last_real_time >= 1.0) {
-        Util.log("At t=%.1f (%.1fs later): %s".format(
-          sim.tick, sim.tick - last_virtual_time, sim.describe_agents
+        Util.log("At t=%.1f (%.1fs later): %s [%d agents moved]".format(
+          sim.tick, sim.tick - last_virtual_time, sim.describe_agents,
+          max_movements
         ))
         last_real_time = now
         last_virtual_time = sim.tick
+        max_movements = 0
       }
-      sim.step(cfg.dt_s)
+      val moves = sim.step(cfg.dt_s)
+      max_movements = math.max(max_movements, moves)
     }
     Util.log("Simulation took " + sim.tick + " virtual seconds")
     timer.stop

### 2012-02-18 utexas/sim/Simulation.scala

@@ -100,13 +100,16 @@ class Simulation(roads: List[Road], edges: List[Edge], vertices: List[Vertex],
     ready_to_spawn = ready_to_spawn.filter(a => !try_spawn(a))
   }
 
-  def step(dt_s: Double) = {
+  // Returns the number of agents that moved
+  def step(dt_s: Double): Int = {
     pre_step
 
     // This value is dt in simulation time, not real time
     val this_time = dt_s * time_speed
     dt_accumulated += this_time
     tick += this_time
+
+    var moved_count = 0
 
     // Agents can't react properly in the presence of huge time-steps. So chop
     // up this time-step into exactly consistent/equal pieces, if needed.
@@ -124,7 +127,11 @@ class Simulation(roads: List[Road], edges: List[Edge], vertices: List[Vertex],
       //t2.stop
 
       //val t3 = Util.timer("agent step")
-      agents.foreach(a => a.step(cfg.dt_s))
+      agents.foreach(a => {
+        if (a.step(cfg.dt_s)) {
+          moved_count += 1
+        }
+      })
       //t3.stop
 
       //val t4 = Util.timer("queue stop")
@@ -147,6 +154,7 @@ class Simulation(roads: List[Road], edges: List[Edge], vertices: List[Vertex],
       //Util.log_pop
       //t0.stop
     }
+    return moved_count
   }
 
   // True if we've correctly promoted into real agents. Does the work of

### 2012-02-18 papers/itsc_2012/root.tex
cleaned rest of paper so far and working on diagrams

@@
 % on the last page of the document
+
+\usepackage[pdftex]{graphicx}
 
@@
 The advent of driver-less cars provides opportunities for optimizing traffic in ways not possible before.
-This paper introduces an open source multi-agent traffic simulator called AORTA, which stands for \textit{Approximately Orchestrated Road Network Automata}, and demonstrates its usage for optimizing autonomous traffic at a city-wide scale.
+This paper introduces an open source multi-agent traffic simulator called AORTA, which stands for \textit{Approximately Orchestrated Road Traffic Automata}, and demonstrates its usage for optimizing autonomous traffic at a city-wide scale.
 AORTA creates scale simulations by generating maps using publicly available road data from \textit{OpenStreetMap} (OSM).
@@
 
+\commentd{The first reason for real world data almost feels too obvious to state
+like that}
+
 \commentpk{Needs 1 more para based on final application, and image on the top right, and introduction to the rest of the paper here once we finalize the sections}
@@
 
-\commentd{? UML diagram: 3 components with their capabilities listed inside, and
-arrows of interactions between them?}
+{ 
+  \begin{center}
+    \setlength\fboxsep{0pt}
+    \setlength\fboxrule{0.5pt}
+    \fbox{\includegraphics[scale=0.3]{architecture.eps}}
+  \end{center}
+}
 
@@
 another; the second \emph{vertex} then contains a mapping from each incoming
-\emph{edge to some outgoing \emph{edge}; each pair is a \emph{turn}. Strictly
+\emph{edge} to some outgoing \emph{edge}; each pair is a \emph{turn}. Strictly
 speaking and with confusing terminology, the multigraph consists of lanes
-(\emph{edges}) as vertices and \emph{turns} as edges. \emph{Roads} are a
-super-structure with meta-data, bundling several \emph{edges} together, and
+(\emph{edges}) as vertices and \emph{turns} as edges. \emph{Roads} are
+super-structures with meta-data, bundling several \emph{edges} together, and
 \emph{vertices} serve to bundle together \emph{turns}.
@@
 avoidance and satisfying other constraints, so setting this value to the
-standard ``reaction time'' of a human driver may be appropriate. Each tick, the
+average ``reaction time'' of a human driver may be appropriate. Each tick, the
 simulation performs the following steps:
@@
 traversable without appearing on the next. Intersection policies and avoiding
-collisions on the same edge are the source of the other constraint in the
-simulation, but it is the job of a behavior to obey. There are three main
-components of the simulation, each configurable by extending default classes:
-agent behaviors, routing strategies, and intersection policies. Each will be
-described below.
+collisions on the same traversable are the source of the other constraint in the
+simulation, but it is the job of a behavior to satisfy those. There are three
+main components of the simulation, each configurable by extending default
+classes: agent behaviors, routing strategies, and intersection policies. Each
+will be described below.
 
@@
 
-\commentd{(generators and thread pool)}
+To introduce new agents into the simulation in real-time, users create
+\emph{generators} programatically or using the UI. A generator picks a random
+starting position and random goal from two sets of roads chosen by the user. A
+generator may cover the entire map, uniformly distributing traffic everywhere,
+or it can mimic rush-hour scenarios. Every tick, a generator creates some number
+of new agents.  If the agent's route policy needs work to be done (such as
+pathfinding), the generator either performs it immediately (incurring a delay in
+simulation if it is running) or sends it to a pool of worker threads to compute
+in the background. When a generator finishes the work for an agent, it promotes
+it to the ``ready'' state.
 
-Once an agent's route is ready, the agent waits alongside its starting edge. At
-the beginning of each tick, they enter with zero initial speed, if it is safe to
-do so. The queue for the edge determines safety. Because behaviors must enforce
-speed limit along an edge by the time an agent enters the road, then if there is
-no agent within the worst-case distance (travel distance at the speed limit,
-plus stopping distance) before the spawn point, then the new agent may enter,
-whether the simulation is running or paused. It is unnecessary to predict which
-agents could enter the edge soon. As a result, there is a buffer space at the
-beginning of the edge where a new agent may never spawn. This disqualifies some
-short edges from being spawn points, but this is not an issue.
+Once an agent's route is ready, the agent waits alongside its starting edge (as
+if they are in a driveway or parking lot). At the beginning of each tick, they
+enter with zero initial speed, if it is safe to do so. The list of agents
+currently on the edge determines safety. Because behaviors must enforce speed
+limit along an edge by the time an agent enters the road, then if there is no
+agent within the worst-case distance (travel distance at the speed limit, plus
+stopping distance from this speed) before the spawn point, then the new agent
+may enter. To render unnecessary the prediction of which agents could enter the
+edge in the near future, there is a buffer space at the beginning of the edge
+where a new agent may never spawn. This disqualifies some short edges from being
+spawn points, but this is not a troublesome issue in practice.
 
@@
 There are two structures to detect collisions. All traversables (edges and the
-straight line-approximation of turns) maintain a queue of occupying agents,
-ordering them so that the agent farthest along is at the head of the queue.
-A naive method of collision detection would detect an agent passing an agent on
-the same queue, or entering a new queue not at the end, immediately after an
-agent moves. However, agents take their steps sequentially and in an effectively
-random order. Ideally they could be reordered so that agents in the front of a
-queue always move first. Since the first agent may cross into the next queue,
-that introduces a depdendency between traversables. A topological sort of a
-directed graph containing circuits is not possible though, and since we
-guarantee connectedness, there are necessarily circuits in most maps. Instead, collisions are
-checked after every agent has stepped forward in time, treating each time-step
-as atomic. Queues remember the original order of agents before a step,
-and afterwards can verify that the order is still legal, with new agents on the end
-and existing agents in the same order. Likewise, intersections check for
-collisions after everybody has moved by examining the turn each agent is
-performing and verifying that no other agent is simultaneously performing a
-conflicting turn.
+straight line-approximation of turns) maintain a \emph{queue} of occupying
+agents, ordering them so that the agent farthest along is at the head of the
+queue. A naive method of collision detection would detect an agent passing an
+agent on the same queue, or entering a new queue not at the end, immediately
+after an agent moves. However, agents take their steps sequentially and in an
+effectively random order. Ideally they could be reordered so that agents in the
+front of a queue always move first. Since the first agent may cross into the
+next queue, that introduces a depdendency between traversables. A topological
+sort of a directed graph containing circuits is not possible though, and since
+we guarantee connectedness, there are necessarily circuits in maps.  Instead,
+collisions are checked after every agent has stepped forward in time, treating
+each time-step as atomic. Queues remember the original order of agents before a
+step, and afterwards can verify that the order is still legal, with new agents
+on the end and existing agents in the same order.
 
-\section{Agent Reactions}
+Likewise, intersections check for collisions after everybody has moved by
+examining the turn each agent is performing and verifying that no other agent is
+simultaneously performing a conflicting turn.
+
+\subsection{Agent Reactions}
 
@@
 behavior.'' It is a generalized, baseline behavior that guarantees not to
-collide with another agent or enter an intersection at the wrong time. It avoids
-some forms of deadlock (in which no agent can move due to interdependencies),
-but there is currently one known flaw. This baseline behavior picks the fastest
-conservative (safe) choice of acceleration at each step. It could be easily
-extended to mimic human drivers by introducing a higher reaction time delay, for
-instance, or tune rates of acceleration for optimizing fuel efficiency.
+collide with another agent or enter an intersection at the wrong time. Since it
+picks the fastest conservative (safe) choice of acceleration at each step, it
+could be easily extended to mimic human drivers by traveling at some random
+amount less than this or to optimize fuel efficiency by tuning acceleration.
+
+\subsubsection{Worst-case analysis and lookahead}
 
 Since agents know the fixed $dt$ duration, they may reason about what could
-happen during the next tick and avoid collision and illegal intersection entry.
+happen during the next tick to avoid collision and illegal intersection entry.
 The conservative analysis is to determine worst case -- how far could an agent
@@
 distance gives the worst-case distance.
+
 \commentd{how much formula? pretty simple so far, but can certainly include them.
@@
 traversable. That means they could start or finish a turn during the next tick.
-Because edges are often extremely short (less than $???$ meters), they may
-travel through several traversables during the next move. Thus, any conservative
-behavior must look ahead until the worst-case distance is exhausted, asking the
-route strategy which turn an agent will choose in the future. At each
-traversable considered, there are 3 constraints to satisfy: obeying the
-intersection, avoiding collision with another agent, and obeying speed limits.
-Each constraint contributes to the max acceleration that will obey the limit, so
-ultimately the result of lookahead will be the minimum of these candidates.
+Because edges are often extremely short due to mis-marked ways in OSM (less than
+$???$ meters), they may travel through several traversables during the next
+move. Thus, any conservative behavior must look ahead until the worst-case
+distance is exhausted, asking the route strategy which turn an agent will choose
+in the future. At each traversable considered, there are 3 constraints to
+satisfy: obeying the intersection, avoiding collision with another agent, and
+obeying speed limits. Each constraint contributes to the max acceleration that
+will obey the limit, so ultimately the result of lookahead will be the minimum
+of these candidates.
+
+Obeying a road's speed limit is a simple constraint. Like many parameters of the
+map, speed limits are inferred from the road type claimed by the OSM source.
+
+\subsubsection{Obeying intersections}
 
 A behavior's interaction with an intersection policy amounts to polling it with
-the turn the agent wants to perform, the distance along the incoming edge the
-agent will be, and how long the agent has been waiting (so policies can enforce
-a required delay before entering the intersection). The intersection policy
-simply orders the agent to ``stop'' or ``continue.'' The lookahead considers no
-intersections beyond the first that issues ``stop'', since the behavior cannot
-legally proceed past that one anyway. The acceleration to avoid entering an
-intersection attempts to end a configurable threshold back from the end of a
-traversable to account for crosswalks and as an epsilon value, since achieveing
-a distance along an edge equal to its length means spilling over into the next
-traversable, which the intersection has banned. Sometimes stopping behind this
-threshold is not possible -- namely, when the edge length is shorter -- and in
-that case, anywhere before the end is acceptable.
+the turn the agent wants to perform, the distance away the agent currently is,
+and how long the agent has been waiting (so policies can enforce a required
+delay before entering the intersection). The intersection policy simply orders
+the agent to ``stop'' or ``continue.'' The lookahead considers no intersections
+beyond the first that issues ``stop'', since the behavior cannot legally proceed
+past that one anyway. The acceleration to avoid entering an intersection
+attempts to end a configurable threshold back from the end of a traversable to
+account for crosswalks and as an epsilon value, since achieveing a distance
+along an edge equal to its length means spilling over into the next traversable,
+which the intersection has banned. Sometimes stopping behind this threshold is
+not possible -- namely, when the edge length is shorter -- and in that case,
+anywhere before the end is acceptable.
+
 \commentd{include some form of piyush's proof about accel to end?}
 
+\subsubsection{Optimal tailgating}
+
+\commentd{diagram with overlapping distances}
+
 When the lookahead analysis is considering some future traversable, it only has
-to consider the agent closest to the start in the present. An edge may only be
-entered from a turn, so the intersection policy has the burden of ensuring no
-agent performs a turn that leads to the same edge simultaneously. Following an
-agent safely means maintaining a configurable following distance behind the
-agent. Worst-case analysis is again used: the most dangerous move the agent in
-front could perform is decelerating as fast as possible the next tick, and the
-worst-case traveling distance for the agent behind is already known from
-lookahead. Maintaining following distance from agents actually enables a form of
-deadlock in the simulation. If an agent is forced to stop in an intersection
-mid-turn due to its destination edge filling up with other agents, then nobody
-may cross the blocked intersection. With many agents in a small area, this
-situation may occur several times in an interdependent way. The solution
-currently being investigated will require behaviors to ensure a turn may be
-fully completed before starting it. This will prevent intersections from ever
-being blocked.
-\commentd{diagram showing deadlock}
+to consider the agent closest to the start of it in the present. An edge may
+only be entered from a turn, so the intersection policy has the burden of
+ensuring no agent performs a turn that leads to the same edge simultaneously.
+Following an agent safely means maintaining a configurable following distance
+behind the agent. Worst-case analysis is again used: the most dangerous move the
+agent in front could perform is decelerating as fast as possible the next tick,
+and the worst-case traveling distance for the agent following is already known
+from lookahead (speeding up as much as possible).
+
+Maintaining following distance from agents actually enables a form of deadlock
+in the simulation. If an agent is forced to stop in an intersection mid-turn due
+to its destination edge filling up with other agents, then nobody may cross the
+blocked intersection. With many agents in a small area, this situation may
+occur several times in an interdependent way. The solution currently being
+investigated will require behaviors to ensure a turn may be fully completed
+before starting it. This will prevent intersections from ever being blocked.
+
+\commentd{diagram showing deadlock, or lookahead in gen eral}
 \commentd{i really hope to fix this issue soon, but it's not an immediately easy
@@
 
-Finally, obeying a road's speed limit is a simple constraint. Like many
-parameters of the map, speed limits are inferred from the road type claimed by
-the OSM source.
-
-\commentd{diagram for example of lookahead?}
-
-\commentd{mention epsilon issues?}
-
-\section{Routing Strategies}
+\subsection{Routing Strategies}
 
@@
 statically plans a route to some goal using standard the A* search; the other
-performs a drunken walk, making random choices at each intersection. The latter
-lazily plans several steps ahead, only when the behavior's lookahead engine
-needs to know. Future experiments in dynamic replanning or hierarchial planning
-can be performed without modifying any code other than a creating new route
-implementation.
+performs a drunken walk, lazily planning random choices at each intersection as
+the lookahead engine needs to know. Future experiments in dynamic replanning or
+hierarchial planning can be performed without modifying any code other than a
+creating new route implementation. Route policies could also interact with a
+singleton overseer to have some sort of global insight.
 
-\section{Intersection Policies}
+\subsection{Intersection Policies}
 
@@
 the head of the queue has waited for a configurable pause, they obtain the lock
-and have exclusive access to the intersection.
+and have exclusive access to the intersection, releasing it only when they
+complete their turn.
 
@@
 policy restores the agent to the front of the queue.
-\commentd{hopefully we'll have a neater solution than this soon}
+
+\commentd{hopefully we'll have a neater solution than the above soon}
 
@@
 tweak. Future policies will include stop lights that adaptively learn better
-timings and an AIM-esque reservation system.
+timings and an AIM-esque reservation system \cite{JAIR08-dresner}.
 
-\section{Determinism and other simulation properties}
+\subsection{Determinism and other simulation properties}
 
 AORTA is designed for controlled experiments. Consequently, simulations are
-deterministically reprodcible. A single seed for the pseudo-random number
-generator, the input graph, and the generators completely determine the single
-outcome of any simulation. This determinism breaks down when generators use
-worker threads, since routes may be computed in different orders, causing agents
-to be ready to enter the road at different times.
+deterministically reproducible. A single seed for the pseudo-random number
+generator, the input graph, and the generators' configuration completely
+determine the outcome of any simulation. This determinism does break down when
+generators use worker threads, since routes may be computed in different orders,
+causing agents to be ready to enter the road at different times.
+
+\commentd{saving a ``scenario'' can soon be a feature}
 
@@
 
-% upgrading stop signs to signals, optimizing timing
+\commentd{upgrading stop signs to signals, optimizing timing?}
 
@@
 
-Write me.
+Write me. big future ideas:
 
-big future ideas:
 contraflow
+
 microtolling
+
 dynamic replanning
+
 autonomous intersections

### 2012-02-18 papers/itsc_2012/root.tex
fixed authors and adding diagrams

@@
 \usepackage[pdftex]{graphicx}
-
-\title{\LARGE \bf
-TODO Name this thing
+% TODO remember we need to take out all the color
+\newcommand{\pix}[3]{
+  \begin{figure}[h]
+    \centering \includegraphics[#3]{#1}
+    \caption{#2}
+  \end{figure}
 }
 
-% TODO which style do we use for authors? and do we list all 4 of us here?
+% TODO better name!
+\title{\LARGE \bf
+Approximately Orchestrated Road Traffic Automata\\
+a General Traffic Simulation Framework
+}
 
-%\author{ \parbox{3 in}{\centering Huibert Kwakernaak*
-%         \thanks{*Use the $\backslash$thanks command to put information here}\\
-%         Faculty of Electrical Engineering, Mathematics and Computer Science\\
-%         University of Twente\\
-%         7500 AE Enschede, The Netherlands\\
-%         {\tt\small h.kwakernaak@autsubmit.com}}
-%         \hspace*{ 0.5 in}
-%         \parbox{3 in}{ \centering Pradeep Misra**
-%         \thanks{**The footnote marks may be inserted manually}\\
-%        Department of Electrical Engineering \\
-%         Wright State University\\
-%         Dayton, OH 45435, USA\\
-%         {\tt\small pmisra@cs.wright.edu}}
-%}
-
-\author{Huibert Kwakernaak$^{1}$ and Pradeep Misra$^{2}$% <-this % stops a space
-% \thanks{$^{1}$H. Kwakernaak is with Faculty of Electrical Engineering, Mathematics and Computer Science,
-%         University of Twente, 7500 AE Enschede, The Netherlands
-%         {\tt\small h.kwakernaak at papercept.net}}%
-% \thanks{$^{2}$P. Misra is with the Department of Electrical Engineering, Wright State University,
-%         Dayton, OH 45435, USA
-%         {\tt\small p.misra at ieee.org}}%
-}
+\author{Dustin Carlino, Mike Depinet, Piyush Khandelwal, and Peter Stone\\
+        Department of Computer Science\\
+        The University of Texas at Austin\\
+        Austin, TX 78712\\
+        % TODO mike, what's your utcs unixname?
+        {\tt \small\{dcarlino,msdepinet,piyushk,pstone\}@cs.utexas.edu}}
 
@@
 yet either}
+
+\pix{downtown_atx.png}{A simulation of downtown Austin, Texas.}{scale=0.25}
 
@@
 
-{ 
-  \begin{center}
-    \setlength\fboxsep{0pt}
-    \setlength\fboxrule{0.5pt}
-    \fbox{\includegraphics[scale=0.3]{architecture.eps}}
-  \end{center}
-}
+\pix{architecture.eps}{An illustration of AORTA's components.}{scale=0.3}
 
@@
 
-\commentd{a simple diagram of a road with a few edges connected maybe to a
-one-way}
+\pix{map_model.png}
+    {Roads contain several edges, and vertices contain several turns between edges.}
+    {scale=0.5}
 
@@
 
-\commentd{illustration of turn conflicts}
+\pix{turn_conflicts.png}
+    {An agent cannot turn left while agents on the perpendicular road cross}
+    {scale=0.5}
 

### 2012-02-18 papers/itsc_2012/root.tex
Minor revisions and my comments

@@
         Austin, TX 78712\\
-        % TODO mike, what's your utcs unixname?
-        {\tt \small\{dcarlino,msdepinet,piyushk,pstone\}@cs.utexas.edu}}
+        {\tt \small\{dcarlino,msd775,piyushk,pstone\}@cs.utexas.edu}}
 
@@
 
-The arrival of driver-less cars presents a number of challenges that need to addressed in the near term.
+The arrival of driver-less cars presents a number of challenges that need to be addressed in the near term.
 Road infrastructure needs to be updated to better support autonomous vehicles.
@@
 verts/edges?}
+\commentm{Why don't we limit ourselves to the words used in traffic for our
+structures (intersections, roads, lanes, etc.) and use graph-related terms
+(edges, vertices) to mean their strict graph theory definitions?  Emphasizing
+our structures the first time we use them may be helpful too.}
 

### 2012-02-19 papers/itsc_2012/root.tex
checking in my final comments on the first draft

@@
 
+% \long\def\commentp#1{}
+% \long\def\commentpk#1{}
+% \long\def\commentm#1{}
+% \long\def\commentd#1{}
+
 \begin{document}
@@
 \begin{abstract} 
-Autonomous vehicles have seen great advancements in recent years and such vehicles are now ever closer to being commercially available.
-The advent of driver-less cars provides opportunities for optimizing traffic in ways not possible before.
-This paper introduces an open source multi-agent traffic simulator called AORTA, which stands for \textit{Approximately Orchestrated Road Traffic Automata}, and demonstrates its usage for optimizing autonomous traffic at a city-wide scale.
-AORTA creates scale simulations by generating maps using publicly available road data from \textit{OpenStreetMap} (OSM).
-Simulations can be set up in the few minutes it takes to export data for the region of interest from OSM.
-AORTA focuses on interactions between agents by defining \textit{behaviors} for individual vehicles and intersections.
-These behaviors allow agents to interact with each other as well as with an external \textit{overseer} that is responsible for optimizing traffic flow.
-Overseers can interact with these agents to allow for dynamic route re-planning to reduce traffic load, optimized intersection navigation, and improved car following.
+Autonomous vehicles have seen great advancements in recent
+years and such vehicles are now ever closer to being commercially available.
+The advent of driver-less cars provides opportunities for optimizing traffic in
+ways not possible before. This paper introduces an open source multi-agent
+traffic simulator called AORTA, which stands for \textit{Approximately
+Orchestrated Road Traffic Automata}, and demonstrates its usage for optimizing
+autonomous traffic at a city-wide scale. AORTA creates scale simulations by
+generating maps using publicly available road data from \textit{OpenStreetMap}
+(OSM). Simulations can be set up in the few minutes it takes to export data for
+the region of interest from OSM. AORTA focuses on interactions between agents
+by defining \textit{behaviors} for individual vehicles and intersections. These
+behaviors allow agents to interact with each other as well as with an external
+\textit{overseer} that is responsible for optimizing traffic flow. Overseers
+can interact with these agents to allow for dynamic route re-planning to reduce
+traffic load, optimized intersection navigation, and improved car following.
 \commentpk{rest devoted to application} 
@@
 Autonomous vehicle technology has made tremendous progress in the last decade.
-In 2004, the farthest distance traveled by a vehicle autonomously in the DARPA Grand Challenge was $11.9km$ \cite{cnnGrandChallenge2004}.
-By 2007, six of the competing teams completed the $96km$ course set for the DARPA Urban Challenge \cite{spectrumUrbanChallenge2007}.
-They did so while following the same traffic laws followed by human drivers, navigating along with other moving vehicles, and following correct intersection precedence order.
-Since then, Google's driver-less cars have clocked more than $250,000km$ on public roads in urban California, USA \cite{tedThrun2011}.
-In 2010, researchers from the University of Parma successfully completed an autonomous \textit{intercontinental} run from Parma, Italy to Shanghai, China \cite{cnnVislab2010}.
-The successful completion of all these milestones indicates that autonomous cars are here to stay, and are ever closer to becoming commercially available.
+In 2004, the farthest distance traveled by a vehicle autonomously in the DARPA
+Grand Challenge was $11.9km$ \cite{cnnGrandChallenge2004}. By 2007, six of the
+competing teams completed the $96km$ course set for the DARPA Urban Challenge
+\cite{spectrumUrbanChallenge2007}. They did so while following the same traffic
+laws followed by human drivers, navigating along with other moving vehicles,
+and following correct intersection precedence order. Since then, Google's
+driver-less cars have clocked more than $250,000km$ on public roads in urban
+California, USA \cite{tedThrun2011}. In 2010, researchers from the University
+of Parma successfully completed an autonomous \textit{intercontinental} run
+from Parma, Italy to Shanghai, China \cite{cnnVislab2010}. The successful
+completion of all these milestones indicates that autonomous cars are here to
+stay, and are ever closer to becoming commercially available.
 
-The arrival of driver-less cars presents a number of challenges that need to be addressed in the near term.
-Road infrastructure needs to be updated to better support autonomous vehicles.
-A number of legal and safety concerns need to be addressed in the context of driver-less cars.
-Human drivers need new methodologies to interact with autonomous drivers on public roads.
-At the same time, driver-less cars provide opportunities of solving traffic problems in ways that were never before possible.
-For instance, autonomous cars can be quickly diverted to make way for emergency vehicles responding to an accident.
-Even day-to-day traffic issues such as congestion could potentially be greatly reduced by effective management of traffic flow.
-This work attempts to address some of these problems.
+The arrival of driver-less cars presents a number of challenges that need to be
+addressed in the near term. Road infrastructure needs to be updated to better
+support autonomous vehicles. A number of legal and safety concerns need to be
+addressed in the context of driver-less cars. Human drivers need new
+methodologies to interact with autonomous drivers on public roads. At the same
+time, driver-less cars provide opportunities of solving traffic problems in
+ways that were never before possible. For instance, autonomous cars can be
+quickly diverted to make way for emergency vehicles responding to an accident.
+Even day-to-day traffic issues such as congestion could potentially be greatly
+reduced by effective management of traffic flow. This work attempts to address
+some of these problems.
 
@@
 
-This paper introduces an open source multi-agent microscopic traffic simulator called AORTA, which stands for \textit{Approximately Orchestrated Road Traffic Automata}.
-The AORTA simulator focuses on the interactions between individual vehicles and intersections, both of which act as \textit{agents} in AORTA's framework.
-Vehicles interact with other vehicles to pass and change lanes, and interact with intersections to decide when to cross them.
-Researchers developing applications on top of the AORTA simulator define \textit{behaviors} that these agents follow to interact with one another.
-To optimize traffic flow, external \textit{overseers} can coordinate with these agents to provide recommendations such as route suggestions or intersection signaling policies.
-An advantage of such an approach is that by assigning human-like behaviors such as the "car following model" \cite{brackstone1999car} to vehicles and traffic signal like policies to intersection, AORTA can just as easily simulate human traffic.
+This paper introduces an open source multi-agent microscopic traffic simulator
+called AORTA, which stands for \textit{Approximately Orchestrated Road Traffic
+Automata}. The AORTA simulator focuses on the interactions between individual
+vehicles and intersections, both of which act as \textit{agents} in AORTA's
+framework. Vehicles interact with other vehicles to pass and change lanes, and
+interact with intersections to decide when to cross them. Researchers
+developing applications on top of the AORTA simulator define \textit{behaviors}
+that these agents follow to interact with one another. To optimize traffic
+flow, external \textit{overseers} can coordinate with these agents to provide
+recommendations such as route suggestions or intersection signaling policies.
+An advantage of such an approach is that by assigning human-like behaviors such
+as the "car following model" \cite{brackstone1999car} to vehicles and traffic
+signal like policies to intersection, AORTA can just as easily simulate human
+traffic.
 
-Like any other microscopic traffic simulator, AORTA needs maps to run simulations.
-We generate our maps using road data available from OpenStreetMap (OSM) \cite{osm}, a Wikipedia like interface for world maps.
-Researchers using AORTA can export the relevant map area from OSM that is then parsed by AORTA to create scale simulations of the real world.
-Using real world road data is advantageous for a number of reasons.
-First, researchers using AORTA for human traffic simulations would find it advantageous to also have the ability to use real-world roads.
-Second, even as infrastructural changes will be made to the road infrastructure to accommodate driver-less cars, it is reasonable to assume that the \textit{same} road network will still be in use.
-Third, using real-world data from OSM allows setting up simulations for an area of interest in a matter of minutes, thereby allowing easy transfer of research results to any desired region.
+Like any other microscopic traffic simulator, AORTA needs maps to run
+simulations. We generate our maps using road data available from OpenStreetMap
+(OSM) \cite{osm}, a Wikipedia like interface for world maps. Researchers using
+AORTA can export the relevant map area from OSM that is then parsed by AORTA to
+create scale simulations of the real world. Using real world road data is
+advantageous for a number of reasons. First, researchers using AORTA for human
+traffic simulations would find it advantageous to also have the ability to use
+real-world roads. Second, even as infrastructural changes will be made to the
+road infrastructure to accommodate driver-less cars, it is reasonable to assume
+that the \textit{same} road network will still be in use. Third, using
+real-world data from OSM allows setting up simulations for an area of interest
+in a matter of minutes, thereby allowing easy transfer of research results to
+any desired region.
 
@@
 
-\commentpk{Needs 1 more para based on final application, and image on the top right, and introduction to the rest of the paper here once we finalize the sections}
+\commentpk{Needs 1 more para based on final application}
 
@@
 
-Computational processing power has made excellent advancements in the last two decades.
-Parallel computing has proved significant for traffic simulation through the use of Geographic Information System (GIS) software \cite{pursula1999simulation}.
-Advanced computing techniques have enabled microscopic models of traffic simulation to generate results at a meaningful scale (city-wide or greater).
-As a result, a number of good micro-simulators have come up in the past decade.
-We review a number of such simulators in this section along with other relevant work.
+Computational processing power has made excellent advancements in the last two
+decades. Parallel computing has proved significant for traffic simulation
+through the use of Geographic Information System (GIS) software
+\cite{pursula1999simulation}. Advanced computing techniques have enabled
+microscopic models of traffic simulation to generate results at a meaningful
+scale (city-wide or greater). As a result, a number of good micro-simulators
+have come up in the past decade. We review a number of such simulators in this
+section along with other relevant work.
 
-A number of traffic simulators already make use of OSM data.
-MATSim is one such multi-agent simulator that focuses on performing large-scale simulations in a relatively small amount of time \cite{balmer2009matsim}.
-Traffic demand is supplied to MATSim in the form of \textit{plans} that an individual may intend to follow in a given day.
-MATSim aims to iteratively improve these plans though offline computation to maximize traffic throughput.
-In other words, it helps individuals understand the best way to go about their daily routine to minimize time spent in transit.
-This goal is somewhat different from that of AORTA, where the demand supplied by individuals is taken as is.
-Instead, AORTA aims to reduce traffic congestion through interactions with individuals in real-time based on the current demand.
+A number of traffic simulators already make use of OSM data. MATSim is one such
+multi-agent simulator that focuses on performing large-scale simulations in a
+relatively small amount of time \cite{balmer2009matsim}. Traffic demand is
+supplied to MATSim in the form of \textit{plans} that an individual may intend
+to follow in a given day. MATSim aims to iteratively improve these plans though
+offline computation to maximize traffic throughput. In other words, it helps
+individuals understand the best way to go about their daily routine to minimize
+time spent in transit. This goal is somewhat different from that of AORTA,
+where the demand supplied by individuals is taken as is. Instead, AORTA aims to
+reduce traffic congestion through interactions with individuals in real-time
+based on the current demand.
 
-Another popular open source traffic simulator that can use OSM data is SUMO \cite{SUMO2011}, which shares the same goals as AORTA and is fairly similar at a high level.
-SUMO has been used to study the effect of automated transportation systems, route planning of individual vehicles, as well as dynamically adapting traffic signal policies to increase traffic efficiency.
-AORTA and SUMO differ in the manner in which intersections are handled.
-SUMO uses traditional traffic signals, while AORTA adopts a much more general \textit{reservation} system.
-A reservation tells a car when it is safe to take its desired path through the intersection.
-Reservations can be alotted to vehicles through more traditional means such as traffic signals or stop-sign based intersection based precedence, or through more efficient methods designed specificaly for autonomous intersection management \cite{JAIR08-dresner}.
-\commentpk{does this difference seem a bit pedantic?}
+Another popular open source traffic simulator that can use OSM data is SUMO
+\cite{SUMO2011}, which shares the same goals as AORTA and is fairly similar at
+a high level. SUMO has been used to study the effect of automated
+transportation systems, route planning of individual vehicles, as well as
+dynamically adapting traffic signal policies to increase traffic efficiency.
+AORTA and SUMO differ in the manner in which intersections are handled. SUMO
+uses traditional traffic signals, while AORTA adopts a much more general
+\textit{reservation} system. A reservation tells a car when it is safe to take
+its desired path through the intersection. Reservations can be allotted to
+vehicles through more traditional means such as traffic signals or stop-sign
+based intersection based precedence, or through more efficient methods designed
+specifically for autonomous intersection management \cite{JAIR08-dresner}.
+\commentpk{does the above difference seem a bit pedantic?}
 
-On the other hand, there are a number of simulators that specifically target autonomous traffic systems.
-One such system is AIM \cite{JAIR08-dresner}, which aims to optimize traffic flow between autonomous vehicles at a given intersection.
-AIM studies problems at the scale of a single intersection, while AORTA has the potential to apply some of the same research at a city-wide scale to real-world road networks.
-A few simulators focus on autonomous vehicles with sensors and actuators \cite{figueiredo2009approach}.
-These simulators are used to ensure that autonomous vehicles are behaving correctly while interacting with other vehicles and infrastructure.
-On the other hand, AORTA assumes that autonomous vehicles are capable of reasonable driving, and focuses on interactions between agents to optimize traffic flow.
+On the other hand, there are a number of simulators that specifically target
+autonomous traffic systems. One such system is AIM \cite{JAIR08-dresner}, which
+aims to optimize traffic flow between autonomous vehicles at a given
+intersection. AIM studies problems at the scale of a single intersection, while
+AORTA has the potential to apply some of the same research at a city-wide scale
+to real-world road networks. A few simulators focus on autonomous vehicles with
+sensors and actuators \cite{figueiredo2009approach}. These simulators are used
+to ensure that autonomous vehicles are behaving correctly while interacting
+with other vehicles and infrastructure. On the other hand, AORTA assumes that
+autonomous vehicles are capable of reasonable driving, and focuses on
+interactions between agents to optimize traffic flow.
 
-Autonomous agents have also been used in the past to model human traffic as well.
-For instance, an approach models human merging and lane changing behavior through the use of autonomous agents interacting with one another \cite{hidas2002modelling}.
-Another approach attempted to improve traffic congestion for human drivers by using a network of autonomous intersection.
-These intersections have the ability to interact with one another and execute a dynamic traffic signal policy minimizing overall wait time \cite{manikonda2001autonomous}.
+Autonomous agents have also been used in the past to model human traffic as
+well. For instance, an approach models human merging and lane changing behavior
+through the use of autonomous agents interacting with one another
+\cite{hidas2002modelling}. Another approach attempted to improve traffic
+congestion for human drivers by using a network of autonomous intersection.
+These intersections have the ability to interact with one another and execute a
+dynamic traffic signal policy minimizing overall wait time
+\cite{manikonda2001autonomous}.
 
@@
   behavior more flawed and humanlike in the agent reactions section.}
-
-\commentpk{I suspect there are still more papers that might be more relevant - we'll definitely have some when we write the application}
-
-% In the realm of applications
-% It is also necessary to analyze some other simulators that use GIS data for microscopic traffic simulations, this is because it may be possible to export OSM data so that it can be used by these simulators, though the authors are not aware of such a connection.
-Incident response has been studied by the following paper (though perhaps only a minor citation as it is an application and does not really compare against AORTA, but it is something we would like to study this as well.
-\cite{huang2007gis}
 
@@
 
-\pix{architecture.eps}{An illustration of AORTA's components.}{scale=0.3}
+\pix{architecture.pdf}{An illustration of AORTA's components.}{scale=0.3}
 
-The architecture and implementation notes of AORTA follow. AORTA is divided into
-roughly three components, each somewhat independent of the others. The map model
-handles everything from building graphs from OSM source to pathfinding. This
-component is cleanly separated from the simulation engine, which adds the notion
-of agents, structures on top of map components that detect collisions, and the
-machinery to implement AORTA's model of car dynamics. If a future experiment
-demands macrosimulation or cellular automata method, no changes to the map
-infrastructure are needed, due to the software engineering. Finally, there is a
-UI using Java Swing, which is (out of necessity) tied to both the map and
+\commentpk {This section is needlessly verbose. I recommend reducing the text down to 1 small para.}
+
+The architecture and implementation notes of AORTA follow. AORTA is divided
+into roughly three components, each somewhat independent of the others. The map
+model handles everything from building graphs from OSM source to pathfinding.
+This component is cleanly separated from the simulation engine, which adds the
+notion of agents, structures on top of map components that detect collisions,
+and the machinery to implement AORTA's model of car dynamics. If a future
+experiment demands macro-simulation or cellular automata method, no changes to
+the map infrastructure are needed, due to the software engineering
+\commentpk{just saying software engineering seems a bit odd} . Finally, there
+is a UI using Java Swing, which is (out of necessity) tied to both the map and
 simulation components. The software has a documented API, is open-source, and
 contains all of the code for the experiments described in Section 6 -- it is
-designed for extensibility.
+designed for extensibility. \commentpk{Insert a footnote to the code here instead od using an Appendix}
 
@@
 
-There are three standalone applications in AORTA. The map builder takes an OSM
-file as input and produces an AORTA graph as output. There is a UI, using Swing,
-to interactively explore the map and run simulations. Specific traffic patterns
-can quickly be simulated by selecting a source and destination group of roads in
-the UI. Finally, there is a headless mode that runs some specific experiment
-without the overhead of visualization. The UI and headless mode both take a
-graph as input, meaning the map construction step must be run once per OSM
-input, but the same graph can be reused indefinitely.
+\commentpk{the next para could be merged with the first one} There are three
+standalone applications in AORTA. The map builder takes an OSM file as input
+and produces an AORTA graph as output. There is a UI, using Swing, to
+interactively explore the map and run simulations. Specific traffic patterns
+can quickly be simulated by selecting a source and destination group of roads
+in the UI \commentpk{I would emphasize the demand generation a bit more}.
+Finally, there is a headless mode that runs some specific experiment without
+the overhead of visualization. The UI and headless mode both take a graph as
+input, meaning the map construction step must be run once per OSM input, but
+the same graph can be reused indefinitely.
 
@@
 \section{MAP CONSTRUCTION FROM OSM}
+
+\commentpk{Since OSM data has been used in the past in a couple of approaches,
+I would recommend re-writing this entire section down to a 2-3 paragraph
+sub-section. This is sort of the infrastructural hard work that gets swept
+under the rug, but you can write about it in future papers (journal articles
+etc). Remember to explicitly save a copy of the text before editing. I have
+written a few comments that can potentially be used in the future}
 
@@
 Any coordinate is translated back to its original GPS form when used for
-computing traversable length.
+computing traversable length.\commentpk{Is your cartesian coordinate system arbitrary?
+If so, I would highly recommend using the UTM system (google wiki UTM).
+We already have the code to convert to it from lat/long coordinates.}
 
@@
 AORTA's structure demands a traditional definition of edge linking exactly two
-vertices, rather than OSM's ways going through many common nodes. Pass 2 walks
+vertices, rather than OSM's ways going through many common nodes \commentpk{I am not sure what you mean by the first statement}. Pass 2 walks
 along each way's coordinates, dividing it into multiple undirected edges.
@@
 \subsubsection{Making Lanes}
+
+\commentpk{see if you can remove the passive voice. Maybe name the utility \textit{map\_builder} and use that instead}
 
@@
 Next, turns are constructed at each vertex between the incoming and outgoing
-edges. This is completely heuristic based: we classify a potential turn as a
+edges. This is completely heuristic based: we \commentpk{don't use we} classify a potential turn as a
 left, right, or straight turn based solely on angles between edges and if the
@@
 
-AORTA has a low-granularity model of turn conflict. If two vehicles at any point
+\commentpk{The following paragraph should not be here. It should be in intersection behaviors}, AORTA has a low-granularity model of turn conflict. If two vehicles at any point
 along two different turns could ever come into contact, then the turns are
 always in conflict, no matter where the agent specifically is along that turn.
-The alternative of tiling intersections (like \cite{JAIR08-dresner} does) would
+The alternative of tiling intersections (like \cite{JAIR08-dresner} \commentpk{I made this same mistake a month ago, dont use citations as a noun} does) would
 provide more granularity, but it adds computational complexity and would be
@@
 algorithm for locating strongly connected components to remove the components of
-the directed graph with the fewest members. (There is also a debugging switch to
+the directed graph with the fewest members \commentpk{missing citation?}. (There is also a debugging switch to
 leave these components in and highlight them in the UI.) The algorithm operates
@@
 
-\commentd{this maybe doesn't belong in this paper}
+\commentd{this maybe doesn't belong in this paper} \commentpk{yes. You can state that OSM data is not perfect, but I would definitely remove this for now}
 
@@
 \subsection{Integrating other data sources}
+
+\commentpk{This needs to be removed as well, it should be in future work}
 
@@
 for the duration of a ``tick'' (also referred to as ``step'') to achieve a new
-position and velocity. The tick duration $dt$ is configurable and fixed for the
-simulation. Agents know this value and use it to reason about collision
-avoidance and satisfying other constraints, so setting this value to the
-average ``reaction time'' of a human driver may be appropriate. Each tick, the
-simulation performs the following steps:
+position and velocity. The tick duration $dt$ is configurable and fixed
+\commentpk{configurable \textit{and} fixed is confusing} for the simulation.
+Agents know this value and use it to reason about collision avoidance and
+satisfying other constraints, so setting this value to the average ``reaction
+time'' of a human driver may be appropriate. Each tick, the simulation performs
+the following steps:
 
@@
 \subsection{Spawning}
+
+\commentpk{I wonder if demand data generation in the ui should go here.}
 
@@
 front of a queue always move first. Since the first agent may cross into the
-next queue, that introduces a depdendency between traversables. A topological
+next queue, that introduces a dependency between traversables. A topological
 sort of a directed graph containing circuits is not possible though, and since
@@
 
-\commentd{include some form of piyush's proof about accel to end?}
+\commentd{include some form of piyush's proof about accel to end?} \commentpk{I would say it is unnecessary}
 
@@
 Following an agent safely means maintaining a configurable following distance
-behind the agent. Worst-case analysis is again used: the most dangerous move the
+behind the agent \commentpk{not sure what you mean by configurable here}. Worst-case analysis is again used: the most dangerous move the
 agent in front could perform is decelerating as fast as possible the next tick,
@@
 \commentd{give some informal or formal (with numbers) idea of scale / how many
-agents we can simulate in different map sizes}
+agents we can simulate in different map sizes} \commentpk{I would say that the the last point is fairly important}
 

### 2012-02-19 papers/itsc_2012/generate.sh

@@ -3,6 +3,7 @@
 # generate graphviz diagrams
 for f in *.dot; do
   dot -Tps2 -o ${f%dot}eps $f
+  epstopdf ${f%dot}eps
 done
 
 pdflatex root.tex

### 2012-02-19 papers/itsc_2012/peter.tex
added peter's tex file to get latex-diff

--- 2012-02-19 papers/itsc_2012/root.tex
@@
   stopping us from trying that idea again}
+\commentp{If it's not implemented, we shouldn't talk about it.}
 
 \begin{abstract} 
-Autonomous vehicles have seen great advancements in recent
-years and such vehicles are now ever closer to being commercially available.
-The advent of driver-less cars provides opportunities for optimizing traffic in
-ways not possible before. This paper introduces an open source multi-agent
-traffic simulator called AORTA, which stands for \textit{Approximately
-Orchestrated Road Traffic Automata}, and demonstrates its usage for optimizing
-autonomous traffic at a city-wide scale. AORTA creates scale simulations by
-generating maps using publicly available road data from \textit{OpenStreetMap}
-(OSM). Simulations can be set up in the few minutes it takes to export data for
-the region of interest from OSM. AORTA focuses on interactions between agents
-by defining \textit{behaviors} for individual vehicles and intersections. These
-behaviors allow agents to interact with each other as well as with an external
-\textit{overseer} that is responsible for optimizing traffic flow. Overseers
-can interact with these agents to allow for dynamic route re-planning to reduce
-traffic load, optimized intersection navigation, and improved car following.
+Autonomous vehicles have seen great advancements in recent years and
+such vehicles are now ever closer to being commercially available.
+The advent of driver-less cars provides opportunities for optimizing
+traffic in ways not possible before. This paper introduces an open
+source multi-agent traffic simulator called AORTA, which stands for
+\textit{Approximately Orchestrated Road Traffic Automata}, and
+demonstrates its usage for optimizing autonomous traffic at a
+city-wide scale. AORTA creates scale simulations by generating maps
+using publicly available road data from \textit{OpenStreetMap}
+(OSM). Simulations of specific regions throughout the world can be set
+up in the few minutes it takes to export data for the region of
+interest from OSM. AORTA focuses on interactions between agents by
+defining \textit{behaviors} for individual vehicles and
+intersections. These behaviors allow agents to interact with each
+other as well as with an external \textit{overseer} that is
+responsible for optimizing traffic flow. Overseers can interact with
+these agents to allow for dynamic route re-planning to reduce traffic
+load, optimized intersection navigation, and improved car following.
 \commentpk{rest devoted to application}
@@
 from Parma, Italy to Shanghai, China \cite{cnnVislab2010}. The successful
-completion of all these milestones indicates that autonomous cars are here to
+completion of all these milestones suggests that autonomous cars are here to
 stay, and are ever closer to becoming commercially available.
 
-The arrival of driver-less cars presents a number of challenges that need to be
+The arrival of driver-less cars presents several challenges that need to be
 addressed in the near term. Road infrastructure needs to be updated to better
-support autonomous vehicles. A number of legal and safety concerns need to be
+support autonomous vehicles.\commentp{Is this necessarily the case?
+  It may be useful, but is it a need?} A number of legal and safety concerns need to be
 addressed in the context of driver-less cars. Human drivers need new
@@
 Even day-to-day traffic issues such as congestion could potentially be greatly
-reduced by effective management of traffic flow. This work attempts to address
-some of these problems.
+reduced by effective management of traffic flow. This work addresses
+some of these problems.\commentp{Be more specific in this paragraph.
+  Which of these challenges are eing addressed elsewhere (cite), and
+  which are addressed in this paper?}
+\commentp{Overall, this paragraph is a bit scattered.  I suggest
+  something more like the following (after the first sentence).
+  Several of the challenges are already being addressed.  For example
+  .....
+  This paper focusses on the specific challenge(s) of....}
 
@@
 
+\commentp{legal issues are addressed here:  http://www.path.berkeley.edu/PATH/Publications/PDF/PRR/2009/PRR-2009-28.pdf}
+
 \pix{downtown_atx.png}{A simulation of downtown Austin, Texas.}{scale=0.25}
 
-This paper introduces an open source multi-agent microscopic traffic simulator
-called AORTA, which stands for \textit{Approximately Orchestrated Road Traffic
-Automata}. The AORTA simulator focuses on the interactions between individual
-vehicles and intersections, both of which act as \textit{agents} in AORTA's
-framework. Vehicles interact with other vehicles to pass and change lanes, and
-interact with intersections to decide when to cross them. Researchers
-developing applications on top of the AORTA simulator define \textit{behaviors}
-that these agents follow to interact with one another. To optimize traffic
-flow, external \textit{overseers} can coordinate with these agents to provide
-recommendations such as route suggestions or intersection signaling policies.
-An advantage of such an approach is that by assigning human-like behaviors such
-as the "car following model" \cite{brackstone1999car} to vehicles and traffic
-signal like policies to intersection, AORTA can just as easily simulate human
-traffic.
+This paper introduces an open source multi-agent microscopic traffic
+simulator called AORTA, which stands for \textit{Approximately
+Orchestrated Road Traffic Automata}. The AORTA simulator focuses on
+the interactions between individual vehicles and intersections, both
+of which act as \textit{agents} in AORTA's framework. Vehicles
+interact with other vehicles to pass and change lanes, and interact
+with intersections to decide when to cross them. Researchers
+developing applications on top of the AORTA simulator define
+\textit{behaviors} that these agents follow to interact with one
+another. To optimize traffic flow, external \textit{overseers} can
+coordinate with these agents to provide recommendations such as route
+suggestions or intersection signaling policies.  An advantage of such
+an approach is that by assigning human-like behaviors such as the
+``car following model'' \cite{brackstone1999car} to vehicles, and
+traffic-signal-like policies to intersection, AORTA can just as easily
+simulate human traffic.
 
 Like any other microscopic traffic simulator, AORTA needs maps to run
-simulations. We generate our maps using road data available from OpenStreetMap
-(OSM) \cite{osm}, a Wikipedia like interface for world maps. Researchers using
+simulations. One of AORTA's key features is that it generates maps using road data available from OpenStreetMap
+(OSM) \cite{osm}, a Wikipedia-like interface for world maps. Researchers using
 AORTA can export the relevant map area from OSM that is then parsed by AORTA to
@@
 road infrastructure to accommodate driver-less cars, it is reasonable to assume
-that the \textit{same} road network will still be in use. Third, using
+that the \textit{same} road network will still be in use.\commentp{I
+  don't understand this second point} Third, using
 real-world data from OSM allows setting up simulations for an area of interest
@@
 like that}
+\commentp{I think all 3 reasons are similar.  I'd just say simplty
+  that AORTA's key features are being open source, easy to install and
+  get running, and flexibility to quickly and easily incorporate
+  models from real road networks around the world.}
 
@@
 
-Computational processing power has made excellent advancements in the last two
-decades. Parallel computing has proved significant for traffic simulation
-through the use of Geographic Information System (GIS) software
-\cite{pursula1999simulation}. Advanced computing techniques have enabled
-microscopic models of traffic simulation to generate results at a meaningful
-scale (city-wide or greater). As a result, a number of good micro-simulators
-have come up in the past decade. We review a number of such simulators in this
-section along with other relevant work.
+Computational processing power has made excellent advancements in the
+last two decades. Parallel computing has proved significant for
+traffic simulation through the use of Geographic Information System
+(GIS) software \cite{pursula1999simulation}. Advanced computing
+techniques have enabled microscopic models of traffic simulation to
+generate results at a meaningful scale (city-wide or greater). As a
+result, a number of good micro-simulators have been introduced in the
+past decade. We review a number of such simulators in this section
+along with other relevant work.
 
 A number of traffic simulators already make use of OSM data. MATSim is one such
-multi-agent simulator that focuses on performing large-scale simulations in a
+multiagent simulator that focuses on performing large-scale simulations in a
 relatively small amount of time \cite{balmer2009matsim}. Traffic demand is
@@
 vehicles through more traditional means such as traffic signals or stop-sign
-based intersection based precedence, or through more efficient methods designed
-specifically for autonomous intersection management \cite{JAIR08-dresner}.
-\commentpk{does the above difference seem a bit pedantic?}
+based intersection precedence, or through more efficient methods designed
+specifically for autonomous vehicles \cite{JAIR08-dresner}.
+%\commentpk{does the above difference seem a bit pedantic?}
 
-On the other hand, there are a number of simulators that specifically target
-autonomous traffic systems. One such system is AIM \cite{JAIR08-dresner}, which
-aims to optimize traffic flow between autonomous vehicles at a given
-intersection. AIM studies problems at the scale of a single intersection, while
-AORTA has the potential to apply some of the same research at a city-wide scale
-to real-world road networks. A few simulators focus on autonomous vehicles with
-sensors and actuators \cite{figueiredo2009approach}. These simulators are used
-to ensure that autonomous vehicles are behaving correctly while interacting
-with other vehicles and infrastructure. On the other hand, AORTA assumes that
-autonomous vehicles are capable of reasonable driving, and focuses on
-interactions between agents to optimize traffic flow.
+On the other hand, there are a number of simulators that specifically
+target autonomous traffic systems. One such system is AIM
+\cite{JAIR08-dresner}, which aims to optimize traffic flow between
+autonomous vehicles at a given intersection. AIM studies problems at
+the scale of roughly a dozen intersections arranged in a regular grid,
+while AORTA has the potential to apply some of the same research at a
+city-wide scale to real-world road networks. A few simulators focus on
+autonomous vehicles with sensors and actuators
+\cite{figueiredo2009approach}. These simulators are used to ensure
+that autonomous vehicles are behaving correctly while interacting with
+other vehicles and infrastructure. On the other hand, AORTA assumes
+that autonomous vehicles are capable of reasonable driving, and
+focuses on interactions between agents to optimize traffic flow.
 
-Autonomous agents have also been used in the past to model human traffic as
-well. For instance, an approach models human merging and lane changing behavior
-through the use of autonomous agents interacting with one another
-\cite{hidas2002modelling}. Another approach attempted to improve traffic
-congestion for human drivers by using a network of autonomous intersection.
-These intersections have the ability to interact with one another and execute a
-dynamic traffic signal policy minimizing overall wait time
-\cite{manikonda2001autonomous}.
+Autonomous agents have also been used in the past to model human
+traffic. For instance, one approach models human merging and lane
+changing behavior through the use of autonomous agents interacting
+with one another \cite{hidas2002modelling}. Another approach attempted
+to improve traffic congestion for human drivers by using a network of
+autonomous intersections.  These intersections have the ability to
+interact with one another and execute a dynamic traffic signal policy
+minimizing overall wait time \cite{manikonda2001autonomous}.
 
@@
 The architecture and implementation notes of AORTA follow. AORTA is divided
-into roughly three components, each somewhat independent of the others. The map
+into three modular components. First, the map
 model handles everything from building graphs from OSM source to pathfinding.
-This component is cleanly separated from the simulation engine, which adds the
+Second, this component is cleanly separated from the simulation engine, which adds the
 notion of agents, structures on top of map components that detect collisions,
@@
 the map infrastructure are needed, due to the software engineering
-\commentpk{just saying software engineering seems a bit odd} . Finally, there
+\commentpk{just saying software engineering seems a bit
+  odd}\commentp{I'd delete that whole senetence.}. Third, there
 is a UI using Java Swing, which is (out of necessity) tied to both the map and
@@
 Machine. Scala provides the advantage of functional programming constructs while
-still permitting imperative style. Parts of AORTA or extensions or clients can
+still permitting imperative style. Parts of AORTA, including extensions and clients, can
 just as easily be written in Java.
@@
 in the UI \commentpk{I would emphasize the demand generation a bit more}.
-Finally, there is a headless mode that runs some specific experiment without
+Finally, there is a headless mode that enables running experiments without
 the overhead of visualization. The UI and headless mode both take a graph as
@@
 our structures the first time we use them may be helpful too.}
+\commentp{I agree with Mike - keep roads, intersection, and turns
+  separate from vertices and edges.  Just say something like ``A map is
+  represented as a multigraph, with edges for lanes and vertices for
+  intersections and changes of direction.  Thus a road consisting of multiple lanes is
+  represented by multiple edges, and a lane that curves is
+  approximated by several line segments.  Intersections consist
+  of the set of possible turns from incoming lanes to
+  outgoing lanes, each represented as a line segment (right turns may
+  have 0 length).''  I wouldn't emphasize all
+  the terms as you do below.}
 
@@
 constituting the center yellow line. These points could also be interpreted as
-control points for a curved road, but AORTA approximates using straight lines.
+control points for a curved road, but AORTA uses straight lines as an approximation.
 Edges mimic this geometry, projecting the center line some multiple of a lane's
@@
 \subsection{Pass 1}
+\commentp{Use a more descriptive title, such as Map Construction: Pass
+  1}
 
-The relevant parts of OSM's map model are nodes and ways. A node corresponds to
-a GPS location, and ways are links between many nodes, corresponding to
-driveable streets and other paths. The first pass reads OSM's XML format,
-gathering ways as a series of coordinates. Ways implicitly intersect when a node
-is used by multiple ways.
+For converting OSM data to AORTA's map model, the relevant OSM fields
+are nodes and ways. Nodes correspond to  GPS locations, and
+ways are links between many nodes, corresponding to driveable streets
+and other paths. The first pass reads OSM's XML format, gathering ways
+as a series of coordinates. Ways implicitly intersect when a node is
+used by multiple ways.
 
@@
 pass 1 filters out anything that does not seem to be driveable. OSM does not
-seem to provide a unified list of valid roads, so these heuristics are
+provide a unified list of valid roads, so these heuristics are
 imperfect.
@@
 Any coordinate is translated back to its original GPS form when used for
-computing traversable length.\commentpk{Is your cartesian coordinate system arbitrary?
-If so, I would highly recommend using the UTM system (google wiki UTM).
-We already have the code to convert to it from lat/long coordinates.}
+computing traversable length.\commentpk{Is your cartesian coordinate
+  system arbitrary? If so, I would highly recommend using the UTM
+  system (google wiki UTM). We already have the code to convert to it
+  from lat/long coordinates.}
+\commentp{Either way, this is too much of an implementation detail for
+  inclusion in the paper.  There's no real technical significance to
+  the translation.}
 
@@
 center line would be, and the other lanes are shifted from there. If there are
-an even number, then the middle two are both equidistant away from the invisible
+an even number, then the middle two are both equidistant from the invisible
 center line. The number of lanes is guessed from OSM's annotation of the way's
@@
 These patterns are attempts at generalizations; this process results in many
-intersections having the incorrect turns. However, without a more detailed
-source, this is the best compromise. At the time of writing, even sources like
-Google Maps lack this information. (TODO how do you cite a lack of something?)
-Finally, the geometry of a turn is simplified to a single line between the ends
+intersections having incorrect turns. However, without a more detailed
+source, this compromise is the best available. At the time of writing, even sources like
+Google Maps lack this information. (TODO how do you cite a lack of
+something?)
+\commentp{Just cite google maps - prior to the words ``lack this...''}
+Finally, the geometry of a turn is simplified to a single line segment
+between the ends
 of the edges it connects.  In the future, a more realistic curve could be drawn.
@@
 
-\commentpk{The following paragraph should not be here. It should be in intersection behaviors}, AORTA has a low-granularity model of turn conflict. If two vehicles at any point
+\commentpk{The following paragraph should not be here. It should be in intersection behaviors} AORTA has a low-granularity model of turn conflict. If two vehicles at any point
 along two different turns could ever come into contact, then the turns are
@@
 at the granularity of lanes. If several lanes of many are disconnected within
-one road, then the bad lanes are spliced out. This results in visually
+one road, then the disconnected lanes are spliced out. This process
+can result in visually
 unappealing geometric gaps, but if the entire road were removed, then Tarjan's
-algorithm would have to be repeated to determine what edges the good lanes would
-then disconnect. The end result is a connected digraph, ideal for corner
-case-free pathfinding and simulation.
+algorithm would have to be repeated to determine what edges the remaining lanes would
+then disconnect. The end result is a connected digraph, ideal for corner-case-free pathfinding and simulation.
 
@@
 A road's center line correctly meets adjoining roads' center lines in the middle
-of an intersection. Since the lines for a lane are parallel to the center line
-and have the same length, they also meet in the intersection. This is
+of an intersection. Since the line segments for a lane are parallel to the center line
+and have the same length, they also meet in the
+intersection.\commentp{How does the latter follow from the former?
+  Maybe a diagram is needed.} This is
 geometrically incorrect; lanes stop along the border of the intersection. Thus,
 pass 3 includes heuristics to trim a lane's lines to the spot where they
-intersect with another lane's lines. This results in a more accurate model, but
-the process is imperfect, so sometimes lines extend too far or not far enough.
+intersect with another lane's lines. The result is a more accurate model, but
+the process is imperfect, so sometimes lines extend too far or not far
+enough.
+\commentp{why?}
 
@@
 
-\commentd{this maybe doesn't belong in this paper} \commentpk{yes. You can state that OSM data is not perfect, but I would definitely remove this for now}
+\commentd{this maybe doesn't belong in this paper} \commentpk{yes. You
+  can state that OSM data is not perfect, but I would definitely
+  remove this for now}
+\commentp{Proposed improvements to OSM data would be great for a
+  discussion or conclusion section at the end of the paper.}
 
@@
 \commentpk{This needs to be removed as well, it should be in future work}
+\commentp{I agree.}
 
@@
 
-Agents have few enforced constraints -- they cannot exceed their physical
+Agents have a few enforced constraints -- they cannot exceed their physical
 acceleration capabilities, travel in reverse, or travel past the end of a
@@
 classes: agent behaviors, routing strategies, and intersection policies. Each
-will be described below.
+is described below.
 
@@
 where a new agent may never spawn. This disqualifies some short edges from being
-spawn points, but this is not a troublesome issue in practice.
+spawn points, but this is not a big limitation in practice.
 
@@
 next queue, that introduces a dependency between traversables. A topological
-sort of a directed graph containing circuits is not possible though, and since
+sort of a directed graph containing circuits\commentp{do you mean cycles?} is not possible though, and since
 we guarantee connectedness, there are necessarily circuits in maps.  Instead,
@@
 step, and afterwards can verify that the order is still legal, with new agents
-on the end and existing agents in the same order.
+on the end and existing agents in the same order.\commentp{I don't
+  follow the point of the last sentence.}
 
@@
 simultaneously performing a conflicting turn.
+
+\commentp{What happens when a collision is detected? }
 
@@
 route) or setting an acceleration for the next step. In the future, another
-possible action will be initiating a lane-change or merge.
+possible action will be initiating a lane-change or
+merge.\commentp{Say ``Were lane changing or merging modeled, that
+  would be another possible action.''}
 
@@
 could be easily extended to mimic human drivers by traveling at some random
-amount less than this or to optimize fuel efficiency by tuning acceleration.
+speed slower than this or to optimize fuel efficiency by tuning acceleration.
 
@@
  should this part be worded like a proof? (because it can be)}
+\commentp{Use a formula if you can use it to mkae the description more
+  concise (and precise).}
 
@@
 Obeying a road's speed limit is a simple constraint. Like many parameters of the
-map, speed limits are inferred from the road type claimed by the OSM source.
+map, speed limits are inferred from the road type claimed by the OSM
+source.
+
+\commentp{This section seems belabored.  Can't you just say that the
+  agent sets its acceleration to be the maximum possible that satifies
+  3 constraints: staying at or below the speed limit, avoiding
+  collisions, and not entering an intersection illegally.}
 
@@
 behind the agent \commentpk{not sure what you mean by configurable here}. Worst-case analysis is again used: the most dangerous move the
-agent in front could perform is decelerating as fast as possible the next tick,
+agent in front could perform is decelerating as fast as possible
+during the next tick,
 and the worst-case traveling distance for the agent following is already known
@@
 blocked intersection. With many agents in a small area, this situation may
-occur several times in an interdependent way. The solution currently being
+occur several times in interdependent ways. The solution currently being
 investigated will require behaviors to ensure a turn may be fully completed
@@
 fix}
+\commentp{Cite (and ideally use ideas from) paper commented out in latex file
+    % @inproceedings{AAAI11-au,
+    % Author = {Tsz-Chiu Au and Neda Shahidi and Peter Stone},
+    % Booktitle= {Proceedings of the Twenty-Fifth Conference on Artificial Intelligence},
+    % Title = {Enforcing Liveness in Autonomous Traffic Management},
+    % Year = {2011},
+    % Month= {August},
+}
 
@@
 any route strategy. Currently two simple implementations are available. One
-statically plans a route to some goal using standard the A* search; the other
+statically plans a route to some goal using standard A* search; the other
 performs a drunken walk, lazily planning random choices at each intersection as
@@
 The stop sign policy works by maintaining a current owner and a queue of waiting
-agents. An agent can enter the queue only when they are within some threshold of
-the start of their turn (a constraint satisfied by the default lookahead
+agents. An agent can enter the queue only when it is within some threshold of
+the start of its turn (a constraint satisfied by the default lookahead
 behavior), even if that is several (rather short) edges and turns away. After
-the head of the queue has waited for a configurable pause, they obtain the lock
-and have exclusive access to the intersection, releasing it only when they
-complete their turn.
+the agent at the head of the queue has waited for a configurable pause, it obtains the lock
+and has exclusive access to the intersection, releasing it only when it
+completes its turn.
 
@@
 
-Intersection policies are one of the most intriguing piece of the simulation to
-tweak. Future policies will include stop lights that adaptively learn better
-timings and an AIM-esque reservation system \cite{JAIR08-dresner}.
+Intersection policies are one of the most intriguing aspects of the simulation to
+tweak, especially in light of autonomous vehicles. Future policies could include stop lights that adaptively learn better
+timings and an AIM-based reservation system \cite{JAIR08-dresner}.
 
@@
 What credits do we need to give to OSM?
+\commentp{Just citing it is good enough.}
 

### 2012-02-19 papers/itsc_2012/generate-diff.sh

@@ -1,0 +1,14 @@
+#!/bin/sh
+
+# generate graphviz diagrams
+for f in *.dot; do
+  dot -Tps2 -o ${f%dot}eps $f
+  epstopdf ${f%dot}eps
+done
+
+pdflatex diff.tex
+bibtex diff
+# pretty sure we have to re-run this twice to include the newly generated bib
+# and references
+pdflatex diff.tex
+pdflatex diff.tex

### 2012-02-20 papers/itsc_2012/draft2.tex
modifying draft based on first round of comments: arch, map model

--- 2012-02-19 papers/itsc_2012/peter.tex
@@
 
-% what are our goals with this paper?
 % theme: easy simulation in your city, in 2 minutes!
-% * demonstrate use of OSM and motivate more data collection
-% * present a simulation framework
-% * present a pretty naive static planning
-% * motivate and demonstrate wards for hierarchial planning (but also reasoning
-%    about congestion)
 % be sure to emphasize open source!
-
 % contribution: autonomous car / human car sim on EXISTING INFRASTRUCTURE
@@
 
+\commentd{consider re-grouping sections... maybe this is all 'methodology'?}
+
 \pix{architecture.pdf}{An illustration of AORTA's components.}{scale=0.3}
 
-\commentpk {This section is needlessly verbose. I recommend reducing the text down to 1 small para.}
-
-The architecture and implementation notes of AORTA follow. AORTA is divided
-into three modular components. First, the map
-model handles everything from building graphs from OSM source to pathfinding.
-Second, this component is cleanly separated from the simulation engine, which adds the
-notion of agents, structures on top of map components that detect collisions,
-and the machinery to implement AORTA's model of car dynamics. If a future
-experiment demands macro-simulation or cellular automata method, no changes to
-the map infrastructure are needed, due to the software engineering
-\commentpk{just saying software engineering seems a bit
-  odd}\commentp{I'd delete that whole senetence.}. Third, there
-is a UI using Java Swing, which is (out of necessity) tied to both the map and
-simulation components. The software has a documented API, is open-source, and
-contains all of the code for the experiments described in Section 6 -- it is
-designed for extensibility. \commentpk{Insert a footnote to the code here instead od using an Appendix}
+AORTA is divided into three modular components: the map model, a
+micro-simulation engine, and a UI. The map model transforms OSM maps into AORTA
+graphs, then answers pathfinding and geometry queries. The simulation engine
+adds a notion of agents, vehicle dynamics, and collisions. A headless mode
+enables running experiments without the overhead of visualization. Finally, the
+UI, which uses Swing, can interactively render the map and agents. For
+convenience, specific traffic patterns can be constructed between two regions of
+the map by simply drawing a polygon around each set of roads.
 
@@
 still permitting imperative style. Parts of AORTA, including extensions and clients, can
-just as easily be written in Java.
-
-\commentpk{the next para could be merged with the first one} There are three
-standalone applications in AORTA. The map builder takes an OSM file as input
-and produces an AORTA graph as output. There is a UI, using Swing, to
-interactively explore the map and run simulations. Specific traffic patterns
-can quickly be simulated by selecting a source and destination group of roads
-in the UI \commentpk{I would emphasize the demand generation a bit more}.
-Finally, there is a headless mode that enables running experiments without
-the overhead of visualization. The UI and headless mode both take a graph as
-input, meaning the map construction step must be run once per OSM input, but
-the same graph can be reused indefinitely.
+just as easily be written in Java. The software has a documented API, is open-source, and
+contains all of the code for the experiments described in Section 6 -- it is
+designed for extensibility\footnote{http://code.google.com/p/road-rage}.
 
@@
 
-\commentpk{Since OSM data has been used in the past in a couple of approaches,
-I would recommend re-writing this entire section down to a 2-3 paragraph
-sub-section. This is sort of the infrastructural hard work that gets swept
-under the rug, but you can write about it in future papers (journal articles
-etc). Remember to explicitly save a copy of the text before editing. I have
-written a few comments that can potentially be used in the future}
-
 Our goal is to simulate traffic on existing infrastructure, not contrived,
@@
 OSM, there is a one-time conversion process to transform the map into AORTA's
-format, which is also encoded with XML. It is split into 3 passes, each feeding
-the next.
+format, which is also encoded with XML.
 
@@
 \pix{map_model.png}
-    {Roads contain several edges, and vertices contain several turns between edges.}
+    {Roads contain several lanes, and intersections contain several turns
+     between lanes.}
     {scale=0.5}
 
-\commentd{terminology is awkward. if we bold-face our notions of edge and vertex
-every time we use them, would that better distinguish them from a graph's
-verts/edges?}
-\commentm{Why don't we limit ourselves to the words used in traffic for our
-structures (intersections, roads, lanes, etc.) and use graph-related terms
-(edges, vertices) to mean their strict graph theory definitions?  Emphasizing
-our structures the first time we use them may be helpful too.}
-\commentp{I agree with Mike - keep roads, intersection, and turns
-  separate from vertices and edges.  Just say something like ``A map is
-  represented as a multigraph, with edges for lanes and vertices for
-  intersections and changes of direction.  Thus a road consisting of multiple lanes is
-  represented by multiple edges, and a lane that curves is
-  approximated by several line segments.  Intersections consist
-  of the set of possible turns from incoming lanes to
-  outgoing lanes, each represented as a line segment (right turns may
-  have 0 length).''  I wouldn't emphasize all
-  the terms as you do below.}
+\commentd{is this clear now? it's not a multigraph and there are no loops, since
+  edges are lanes and turns. I have to think a bit about intersections as
+  vertices strictly, though...}
 
-\emph{Roads} are undirected edges between \emph{vertices}, which are
-intersections between multiple roads. A \emph{road} contains one or two sets of
-directed \emph{edges}, each of which represents a single lane of travel. One-way
-\emph{roads} are supported. An \emph{edge} leads from one \emph{vertex} to
-another; the second \emph{vertex} then contains a mapping from each incoming
-\emph{edge} to some outgoing \emph{edge}; each pair is a \emph{turn}. Strictly
-speaking and with confusing terminology, the multigraph consists of lanes
-(\emph{edges}) as vertices and \emph{turns} as edges. \emph{Roads} are
-super-structures with meta-data, bundling several \emph{edges} together, and
-\emph{vertices} serve to bundle together \emph{turns}.
+A map is represented as a directed graph with \emph{lanes} and \emph{turns} as edges and
+\emph{intersections} as vertices. A \emph{road} groups one or two groups of
+lanes together, each group being a different direction of travel. Intersections
+map the possible turns from incoming to outgoing lanes. Since both lanes and
+turns support traffic, they are grouped together as the \emph{traversables} or
+edges, meaning they have some total length.
 
-Geometrically, a road has a sequence of points defining the line segments
-constituting the center yellow line. These points could also be interpreted as
-control points for a curved road, but AORTA uses straight lines as an approximation.
-Edges mimic this geometry, projecting the center line some multiple of a lane's
-width away in the appropriate direction. Turns are modeled with a single
-straight line connecting the end of one edge to the start of another (so right
-turns often have $0$ length). Since both edges and turns can support traffic,
-they are grouped together as \emph{traversables}, meaning they have some
-sequence of line segments and a total length.
+The map also has a geometry interpretation, used for visualization and physical
+distances. A road has a sequence of points giving the line segments of the
+center yellow line. These points could also be interpreted as control points for
+a curved road, but AORTA uses straight lines as an approximation. Lanes are
+physically placed a lane's width away from and parallel to the center line. (In
+the case of one-ways, there is no center line; instead, the lanes simply cluster
+around the invisible center lane, all pointing the same way.) Turns are modeled
+with a single straight line connecting the end of one edge to the start of
+another (so right turns often have $0$ length).
 
-\subsection{Pass 1}
-\commentp{Use a more descriptive title, such as Map Construction: Pass
-  1}
+\subsection{Map Construction Passes}
 
-For converting OSM data to AORTA's map model, the relevant OSM fields
-are nodes and ways. Nodes correspond to  GPS locations, and
-ways are links between many nodes, corresponding to driveable streets
-and other paths. The first pass reads OSM's XML format, gathering ways
-as a series of coordinates. Ways implicitly intersect when a node is
-used by multiple ways.
+\commentd{gotta work on passive voice}
 
-Since OSM also encodes pedestrian paths, bike routes, and even lakes with ways,
-pass 1 filters out anything that does not seem to be driveable. OSM does not
-provide a unified list of valid roads, so these heuristics are
-imperfect.
-
-Finally, coordinates are normalized from GPS (longitude, latitude) to a
-Cartesian coordinate system with no negative coordinates and the Y coordinate
-increasing upwards. This is only done for convenience of manipulation in the UI.
-Any coordinate is translated back to its original GPS form when used for
-computing traversable length.\commentpk{Is your cartesian coordinate
-  system arbitrary? If so, I would highly recommend using the UTM
-  system (google wiki UTM). We already have the code to convert to it
-  from lat/long coordinates.}
-\commentp{Either way, this is too much of an implementation detail for
-  inclusion in the paper.  There's no real technical significance to
-  the translation.}
-
-\subsection{Pass 2}
-
-AORTA's structure demands a traditional definition of edge linking exactly two
-vertices, rather than OSM's ways going through many common nodes \commentpk{I am not sure what you mean by the first statement}. Pass 2 walks
-along each way's coordinates, dividing it into multiple undirected edges.
-
-\subsection{Pass 3}
-
-Finally, pass 3 performs successive transformations to produce the final graph,
-which is saved in XML.
-
-\subsubsection{Making Lanes}
-
-\commentpk{see if you can remove the passive voice. Maybe name the utility \textit{map\_builder} and use that instead}
-
-\commentd{the geometry was already explained in Map Model; I think it belongs
-there if we have to cut one}
-
-Undirected edges are first multiplied into directed lanes. The coordinates from
-a slice of an OSM way are interpreted to be the center line of the road, and
-lanes are placed by projecting the width of a lane perpendicularly from the
-center line in either direction. OSM encodes one-way streets, so in this case,
-if there are an odd number of one-way lanes, the middle lane is placed where the
-center line would be, and the other lanes are shifted from there. If there are
-an even number, then the middle two are both equidistant from the invisible
-center line. The number of lanes is guessed from OSM's annotation of the way's
-``type.'' OSM's specification includes a tag for number of lanes, but in
-practice as of the time of writing, it is rarely used.
-
-\subsubsection{Making Turns}
-
-Next, turns are constructed at each vertex between the incoming and outgoing
-edges. This is completely heuristic based: we \commentpk{don't use we} classify a potential turn as a
-left, right, or straight turn based solely on angles between edges and if the
-edge's original OSM ID matches. U-turns are constructed at dead-ends. If several
-lanes all cross into fewer lanes, merging is forced for the rightmost lanes.
-These patterns are attempts at generalizations; this process results in many
-intersections having incorrect turns. However, without a more detailed
-source, this compromise is the best available. At the time of writing, even sources like
-Google Maps lack this information. (TODO how do you cite a lack of
-something?)
-\commentp{Just cite google maps - prior to the words ``lack this...''}
-Finally, the geometry of a turn is simplified to a single line segment
-between the ends
-of the edges it connects.  In the future, a more realistic curve could be drawn.
-
-\pix{turn_conflicts.png}
-    {An agent cannot turn left while agents on the perpendicular road cross}
-    {scale=0.5}
-
-\commentpk{The following paragraph should not be here. It should be in intersection behaviors} AORTA has a low-granularity model of turn conflict. If two vehicles at any point
-along two different turns could ever come into contact, then the turns are
-always in conflict, no matter where the agent specifically is along that turn.
-The alternative of tiling intersections (like \cite{JAIR08-dresner} \commentpk{I made this same mistake a month ago, dont use citations as a noun} does) would
-provide more granularity, but it adds computational complexity and would be
-impractical for many of the complex intersection geometries inferred from OSM.
-
-\subsubsection{Ensuring connectivity}
-
-Because turns are constructed imperfectly, most large graphs have edges that are
-not reachable from some portions of the graph. Simulation and pathfinding would
-suffer greatly if this remained the case, so pass 3 uses Tarjan's linear-time
-algorithm for locating strongly connected components to remove the components of
-the directed graph with the fewest members \commentpk{missing citation?}. (There is also a debugging switch to
-leave these components in and highlight them in the UI.) The algorithm operates
-at the granularity of lanes. If several lanes of many are disconnected within
-one road, then the disconnected lanes are spliced out. This process
-can result in visually
-unappealing geometric gaps, but if the entire road were removed, then Tarjan's
-algorithm would have to be repeated to determine what edges the remaining lanes would
-then disconnect. The end result is a connected digraph, ideal for corner-case-free pathfinding and simulation.
-
-\subsubsection{Cleaning up geometry}
-
-A road's center line correctly meets adjoining roads' center lines in the middle
-of an intersection. Since the line segments for a lane are parallel to the center line
-and have the same length, they also meet in the
-intersection.\commentp{How does the latter follow from the former?
-  Maybe a diagram is needed.} This is
-geometrically incorrect; lanes stop along the border of the intersection. Thus,
-pass 3 includes heuristics to trim a lane's lines to the spot where they
-intersect with another lane's lines. The result is a more accurate model, but
-the process is imperfect, so sometimes lines extend too far or not far
-enough.
-\commentp{why?}
-
-\subsubsection{Hierarchial divisions of the graph}
-
-\commentd{i don't think this'll have much coverage in the rest of the paper; should
- we describe wards?}
-\commentm{We aren't actually using wards at all yet.  I think we should save this
-for a later paper.}
-
-\subsection{Limitations}
-
-\commentd{this maybe doesn't belong in this paper} \commentpk{yes. You
-  can state that OSM data is not perfect, but I would definitely
-  remove this for now}
-\commentp{Proposed improvements to OSM data would be great for a
-  discussion or conclusion section at the end of the paper.}
-
-proposals to change OSM to fix these issues
-
-projecting lines sucks... but all it really does is clutter UI
-filtering out invalid ways in pass 1 is really not a clear process
-number of lanes and speed limits... road types are wacky and wrong
-turns are brittle.. detecting right-turn only lane is rare. shared center lane?
-  not part of our model, or OSM's.
-removing sccs is not ideal
-
-\subsection{Integrating other data sources}
-
-\commentpk{This needs to be removed as well, it should be in future work}
-\commentp{I agree.}
-
-In the future, we intend to support the ability to augment an AORTA map with
-data from non-OSM sources, such as information about the presence of extra lanes
-for left turns. Using data such as GPS coordinates and road names, we could fuse
-this extra information into an OSM-based AORTA map to make simulations on that map
-that much more realistic.
-
-Section 5.1 details a way to specify where traffic should originate and move
-towards using the UI. This allows a more accurate simulation of situations such
-as rush hour. In the future, we hope to be able to generate these demand
-descriptions from government-collected data about a municipality's road usage.
+Map construction is split into several passes, each feeding the next.
+\begin{enumerate}
+  \item OSM encodes many paths besides driveable roads, so these are first
+        filtered out. GPS locations common to multiple roads are marked, since
+        they implicitly indicate intersections.
+  \item OSM \emph{ways} (sequences of GPS locations) include many
+        intersections, so the builder next splits ways into undirected
+        segments of roads between exactly two intersections.
+  \item Undirected roads are multiplied into directed lanes in each direction
+        (unless the road is marked one-way). The number of lanes is guessed from
+        OSM's ``road type'' tag or an explicit number of lanes, when given.
+  \item Turns are next constructed between incoming and outgoing edges at each
+        vertex. Left, right, straight, and U-turns are heuristically chosen
+        based on the angles between edges. When several lanes all cross into
+        fewer lanes, merging is arbitrarily forced on the rightmost lanes.
+        Although inaccurate, these patterns are the best compromise. At the
+        time of writing, even sources like Google Maps (TODO cite) lack this
+        information.
+  \item The strongly connected components of the digraph with the fewest
+        members are removed. Tarjan's algorithm locates these in linear time
+        (TODO cite). Disconnected lanes are spliced out of larger roads, leaving
+        unappealing geometric gaps, but correctly ensuring connectivity for
+        corner-case free pathfinding.
+  \item The builder serializes the graph in a simple XML format.
+\end{enumerate}
 
@@
 simultaneously performing a conflicting turn.
+
+\commentpk{The following paragraph should not be here. It should be in intersection behaviors} AORTA has a low-granularity model of turn conflict. If two vehicles at any point
+along two different turns could ever come into contact, then the turns are
+always in conflict, no matter where the agent specifically is along that turn.
+The alternative of tiling intersections (like \cite{JAIR08-dresner} \commentpk{I made this same mistake a month ago, dont use citations as a noun} does) would
+provide more granularity, but it adds computational complexity and would be
+impractical for many of the complex intersection geometries inferred from OSM.
+
+\pix{turn_conflicts.png}
+    {An agent cannot turn left while agents on the perpendicular road cross}
+    {scale=0.5}
 
@@
 
+% TODO and mike's variant that wanders in the right direction
+
 \subsection{Intersection Policies}
@@
 
+\commentp{Proposed improvements to OSM data would be great for a
+  discussion or conclusion section at the end of the paper.}
+proposals to change OSM to fix these issues
+  projecting lines sucks... but all it really does is clutter UI
+  filtering out invalid ways in pass 1 is really not a clear process
+  number of lanes and speed limits... road types are wacky and wrong
+  turns are brittle.. detecting right-turn only lane is rare. shared center lane?
+    not part of our model, or OSM's.
+  removing sccs is not ideal
+
 contraflow
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section*{APPENDIX}
-
-Probably stuff about getting and using the code.
 

### 2012-02-20 papers/itsc_2012/clean.sh

@@ -1,0 +1,3 @@
+#!/bin/sh
+
+rm -f *.aux *.log *.pdf *.bbl *.blg *.eps

### 2012-02-20 papers/itsc_2012/generate-diff.sh

@@ -1,14 +1,0 @@
-#!/bin/sh
-
-# generate graphviz diagrams
-for f in *.dot; do
-  dot -Tps2 -o ${f%dot}eps $f
-  epstopdf ${f%dot}eps
-done
-
-pdflatex diff.tex
-bibtex diff
-# pretty sure we have to re-run this twice to include the newly generated bib
-# and references
-pdflatex diff.tex
-pdflatex diff.tex

### 2012-02-20 papers/itsc_2012/make-diff.sh

@@ -1,0 +1,8 @@
+#!/bin/sh
+
+V1=draft1.tex
+V2=peter.tex
+
+latexdiff $V1 $V2 > diff.tex
+
+. generate.sh diff.tex

### 2012-02-21 utexas/map/Turn.scala
worker threads dont do dumb work, deterministic cycled light policy with bad math

@@ -14,8 +14,10 @@ object TurnType extends Enumeration {
 }
 
 class Turn(val id: Int, val from: Edge, val turn_type: TurnType.TurnType, val to: Edge)
-  extends Traversable
+  extends Traversable with Ordered[Turn]
 {
+  override def compare(other: Turn) = id.compare(other.id)
+
   // id is just for comparisons, not indexing
   def to_xml(out: FileWriter) = {
     out.write(

### 2012-02-21 utexas/map/Vertex.scala

@@ -6,7 +6,8 @@ import scala.collection.mutable.MutableList
 
 // TODO var id due to tarjan
 class Vertex(val location: Coordinate, var id: Int) {
-  // TODO we could keep a map for faster lookup, sure
+  // TODO we could keep a map for faster lookup, sure, but determinism's cool
+  // too.
   var turns = new MutableList[Turn]
 
   // TODO construction sucks

### 2012-02-21 utexas/sim/Agent.scala

@@ -60,8 +60,6 @@ class Agent(val id: Int, val graph: Graph, val start: Edge, val start_dist: Doub
     var current_on = start_on
     var current_dist = old_dist + new_dist
 
-    // TODO this is cheating, it's temporary.
-    try {
     // TODO
     while (current_dist >= current_on.length) {
       current_dist -= current_on.length
@@ -93,14 +91,6 @@ class Agent(val id: Int, val graph: Graph, val start: Edge, val start_dist: Doub
       behavior.transition(current_on, next)
       current_on = next
     }
-    // TODO
-    } catch {
-      case _ => {
-        Util.log("bad time to choose a turn, apparently.")
-        dump_info
-      }
-    }
-    // TODO
 
     // so we finally end up somewhere...
     if (start_on == current_on) {

### 2012-02-21 utexas/sim/Behaviors.scala

@@ -107,7 +107,7 @@ class RouteFollowingBehavior(a: Agent, route: Route) extends Behavior(a) {
           case e: Edge => {
             val i = Agent.sim.intersections(e.to)
             a.upcoming_intersections += i   // remember we've registered here
-            i.should_stop(a, next_turn, how_far_away)
+            !i.can_go(a, next_turn, how_far_away)
           }
         })
         if (stop_at_end) {

### 2012-02-21 utexas/sim/Generator.scala

@@ -1,6 +1,6 @@
 package utexas.sim
 
-import java.util.concurrent.{Executors, FutureTask}
+import java.util.concurrent.{Executors, FutureTask, Callable}
 
 import utexas.map.{Edge, Traversable}
 
@@ -18,7 +18,8 @@ abstract class Generator(sim: Simulation, desired_starts: List[Edge], val end_ca
   // Prune the desired_starts for road type and length
   val start_candidates = desired_starts.filter(e => sim.queues(e).ok_to_spawn)
 
-  protected var pending: List[(Agent, FutureTask[List[Traversable]])] = Nil
+  // there may be no task scheduled
+  protected var pending: List[(Agent, Option[FutureTask[List[Traversable]]])] = Nil
 
   // Returns new agents to try to spawn, or boolean means reap this genertor
   def run(): Either[List[Agent], Boolean]
@@ -33,21 +34,43 @@ abstract class Generator(sim: Simulation, desired_starts: List[Edge], val end_ca
     val a = new Agent(sim.next_id, sim, start, sim.queues(start).safe_spawn_dist, route)
 
     // schedule whatever work the route needs done.
-    val delayed = new FutureTask[List[Traversable]](route.request_route(start, end))
-    Generator.worker_pool.execute(delayed)
-    pending :+= (a, delayed)
+    route.request_route(start, end) match {
+      case Some(task) => {
+        val delayed = new FutureTask[List[Traversable]](task)
+        Generator.worker_pool.execute(delayed)
+        pending :+= (a, Some(delayed))
+      }
+      case _ => {
+        // don't schedule anything
+        pending :+= (a, None)
+      }
+    }
   }
 
   // This is a nonblocking poll, of course
   def poll(): List[Agent] = {
-    val (done, still_pending) = pending.partition(a => a._2.isDone)
+    val (done, still_pending) = pending.partition(
+      a => a._2 match {
+        case Some(task) => {
+          if (task.isDone) {
+            a._1.route.got_route(task.get)
+            true
+          } else {
+            false
+          }
+        }
+        case None       => true
+      }
+    )
     pending = still_pending
-    done.foreach(a => a._1.route.got_route(a._2.get()))
     return done.map(a => a._1)
   }
 
   // And the blocking poll
-  def wait_for_all() = pending.foreach(a => a._2.get())
+  def wait_for_all() = pending.foreach(a => a._2 match {
+    case Some(task) => task.get
[…]

### 2012-02-21 utexas/sim/Intersection.scala

@@ -3,6 +3,7 @@ package utexas.sim
 import scala.collection.mutable.{HashMap => MutableMap}
 import scala.collection.mutable.{HashSet => MutableSet}
 import scala.collection.mutable.ListBuffer
+import scala.collection.immutable.{SortedSet, TreeSet}
 
 import utexas.map.{Vertex, Turn}
 
@@ -11,14 +12,14 @@ import utexas.{Util, cfg}
 // Reason about collisions from conflicting simultaneous turns.
 class Intersection(val v: Vertex) {
   //val policy: Policy = new NeverGoPolicy(this)
-  val policy: Policy = new StopSignPolicy(this)
+  //val policy: Policy = new StopSignPolicy(this)
+  val policy: Policy = new SignalCyclePolicy(this)
   //val policy: Policy = new ReservationPolicy(this)
-  //val policy: Policy = new SignalCyclePolicy(this)
 
   override def toString = "Intersection(" + v + ")"
 
   // Just delegate.
-  def should_stop(a: Agent, turn: Turn, far_away: Double) = policy.should_stop(a, turn, far_away)
+  def can_go(a: Agent, turn: Turn, far_away: Double) = policy.can_go(a, turn, far_away)
   def yield_lock(a: Agent) = policy.yield_lock(a)
   def unregister(a: Agent) = policy.unregister(a)
 
@@ -38,7 +39,7 @@ class Intersection(val v: Vertex) {
     // by observing that the conflicts-with relation is symmetric.
     for (t1 <- turns.keys; t2 <- turns.keys if t1.id < t2.id) {
       if (t1.conflicts(t2)) {
-        Util.log("!!! Collision1 in an intersection: " + t1 + " and " + t2)
+        Util.log("!!! Collision in an intersection: " + t1 + " and " + t2)
         assert(t2.conflicts(t1))
       }
     }
@@ -70,7 +71,7 @@ class Intersection(val v: Vertex) {
 }
 
 abstract class Policy(intersection: Intersection) {
-  def should_stop(a: Agent, turn: Turn, far_away: Double): Boolean
+  def can_go(a: Agent, turn: Turn, far_away: Double): Boolean
   def validate_entry(a: Agent, turn: Turn): Boolean
   def handle_exit(a: Agent, turn: Turn)
   def yield_lock(a: Agent)
@@ -85,7 +86,7 @@ abstract class Policy(intersection: Intersection) {
 
 // Great for testing to see if agents listen to this.
[…]

### 2012-02-21 utexas/sim/Route.scala

@@ -7,7 +7,7 @@ import utexas.{Util, cfg}
 abstract class Route() {
   // Define these
   def lookahead_step(n: Int): Option[Traversable]
-  def request_route(from: Edge, to: Edge): Callable[List[Traversable]]
+  def request_route(from: Edge, to: Edge): Option[Callable[List[Traversable]]]
 
   // Common to most, but can change. Should never become empty after got_route.
   var steps: List[Traversable] = Nil
@@ -34,11 +34,11 @@ class StaticRoute() extends Route() {
                                         else
                                           Some(steps(n))
   
-  override def request_route(from: Edge, to: Edge) = new Callable[List[Traversable]]() {
+  override def request_route(from: Edge, to: Edge) = Some(new Callable[List[Traversable]]() {
     def call(): List[Traversable] = {
       return Agent.sim.pathfind_astar(from, to)
     }
-  }
+  })
 }
 
 // DOOMED TO WALK FOREVER
@@ -65,9 +65,10 @@ class DrunkenRoute() extends Route() {
   }
 
   // Schedule empty work! :P
-  override def request_route(from: Edge, to: Edge) = new Callable[List[Traversable]]() {
-    // just start with where we are.
-    def call(): List[Traversable] = List(pick_turn(from))
+  override def request_route(from: Edge, to: Edge): Option[Callable[List[Traversable]]] = {
+    // no actual work to do
+    steps = List(pick_turn(from))
+    return None
   }
 
   // TODO choose a random point, or one leading in general direction of goal?
@@ -96,7 +97,7 @@ class DirectionalRoute extends DrunkenRoute() {
     }
   }
   
-  override def request_route(from: Edge, to: Edge): Callable[List[Traversable]] = {
+  override def request_route(from: Edge, to: Edge): Option[Callable[List[Traversable]]] = {
     end_point = to
     return super.request_route(from, to)
   }

### 2012-02-22 papers/itsc_2012/draft2.tex
worker threads dont do dumb work, deterministic cycled light policy with bad math
simulation section revised heavily

@@
 still permitting imperative style. Parts of AORTA, including extensions and clients, can
-just as easily be written in Java. The software has a documented API, is open-source, and
-contains all of the code for the experiments described in Section 6 -- it is
-designed for extensibility\footnote{http://code.google.com/p/road-rage}.
+just as easily be written in Java. The software has a documented API, is
+open-source\footnote{http://code.google.com/p/road-rage}, and contains all of
+the code for the experiments described in Section 6 -- it is designed for
+extensibility.
 
@@
 
-The map also has a geometry interpretation, used for visualization and physical
+The map also has a geometric interpretation, used for visualization and physical
 distances. A road has a sequence of points giving the line segments of the
@@
         information.
-  \item The strongly connected components of the digraph with the fewest
-        members are removed. Tarjan's algorithm locates these in linear time
-        (TODO cite). Disconnected lanes are spliced out of larger roads, leaving
-        unappealing geometric gaps, but correctly ensuring connectivity for
-        corner-case free pathfinding.
+  \item Because turns are heuristically placed, the graph is disconnected. The
+        strongly connected components of the digraph with the fewest members are
+        removed. Tarjan's algorithm locates these in linear time (TODO cite).
+        Disconnected lanes are spliced out of larger roads, leaving unappealing
+        geometric gaps, but correctly ensuring connectivity for corner-case free
+        pathfinding.
   \item The builder serializes the graph in a simple XML format.
@@
 for the duration of a ``tick'' (also referred to as ``step'') to achieve a new
-position and velocity. The tick duration $dt$ is configurable and fixed
-\commentpk{configurable \textit{and} fixed is confusing} for the simulation.
-Agents know this value and use it to reason about collision avoidance and
-satisfying other constraints, so setting this value to the average ``reaction
-time'' of a human driver may be appropriate. Each tick, the simulation performs
-the following steps:
+position and velocity. The tick duration $dt$ is fixed, enabling agents to
+reason about collision avoidance and satisfying other constraints. Setting this
+value to the average ``reaction time'' of a human driver may be appropriate for
+some experiments. Each tick, the simulation performs the following steps:
 
@@
   \item Introducing new agents into the map
-  \item Updating the position and velocity of agents
+  \item Updating the position and velocity of agents based on their choice the
+        previous tick
   \item Checking for collisions
@@
 traversable without appearing on the next. Intersection policies and avoiding
-collisions on the same traversable are the source of the other constraint in the
-simulation, but it is the job of a behavior to satisfy those. There are three
-main components of the simulation, each configurable by extending default
-classes: agent behaviors, routing strategies, and intersection policies. Each
-is described below.
+collisions on the same traversable provide the other constraints, but it is the
+job of an agent's behavior to satisfy those. There are three main components of
+the simulation, each configurable by extending default implementations of an
+interface: agent behaviors, routing strategies, and intersection policies.
 
-\commentd{justify arbitrary constants, like deaccel for cars;
-          systems ideas, or limitations.. trylocks, atomic sim update, deadlock
-          issues;
-          limit: no lane-changing}
+%\commentd{justify arbitrary constants, like deaccel for cars;
+%          systems ideas, or limitations.. trylocks, atomic sim update, deadlock
+%          issues;
+%          limit: no lane-changing}
 
@@
 
-\commentpk{I wonder if demand data generation in the ui should go here.}
-
 To introduce new agents into the simulation in real-time, users create
 \emph{generators} programatically or using the UI. A generator picks a random
-starting position and random goal from two sets of roads chosen by the user. A
-generator may cover the entire map, uniformly distributing traffic everywhere,
-or it can mimic rush-hour scenarios. Every tick, a generator creates some number
-of new agents.  If the agent's route policy needs work to be done (such as
-pathfinding), the generator either performs it immediately (incurring a delay in
-simulation if it is running) or sends it to a pool of worker threads to compute
-in the background. When a generator finishes the work for an agent, it promotes
-it to the ``ready'' state.
+starting position and random goal from two sets of roads chosen by the user,
+selectable by drawing a polygon with the mouse in the UI. A generator may cover
+the entire map, uniformly distributing traffic everywhere, or it can mimic
+rush-hour scenarios.
+
+Every tick, a generator creates some number of new agents. If the agent's route
+policy needs work to be done (such as pathfinding), the generator either
+performs it immediately (incurring a delay in simulation if it is running) or
+sends it to a pool of worker threads to compute in the background.
 
 Once an agent's route is ready, the agent waits alongside its starting edge (as
-if they are in a driveway or parking lot). At the beginning of each tick, they
-enter with zero initial speed, if it is safe to do so. The list of agents
-currently on the edge determines safety. Because behaviors must enforce speed
-limit along an edge by the time an agent enters the road, then if there is no
-agent within the worst-case distance (travel distance at the speed limit, plus
-stopping distance from this speed) before the spawn point, then the new agent
-may enter. To render unnecessary the prediction of which agents could enter the
-edge in the near future, there is a buffer space at the beginning of the edge
-where a new agent may never spawn. This disqualifies some short edges from being
-spawn points, but this is not a big limitation in practice.
+if it is in a driveway or parking lot). When the edge has no agents that could
+potentially crash with the new vehicle, the new agent enters the system with
+zero initial speed.  Because behaviors must enforce speed limit along an edge by
+the time an agent enters the road, then if there is no agent within the
+worst-case distance (traveling distance at the speed limit, plus stopping
+distance from this speed) before the spawn point, then the new agent may enter.
+To render unnecessary the prediction of which agents could enter the edge in the
+near future, there is a buffer space at the beginning of the edge where a new
+agent may never spawn. This disqualifies some short edges from being spawn
+points, but this is not a big limitation in practice.
 
@@
 
-There are two structures to detect collisions. All traversables (edges and the
-straight line-approximation of turns) maintain a \emph{queue} of occupying
-agents, ordering them so that the agent farthest along is at the head of the
-queue. A naive method of collision detection would detect an agent passing an
-agent on the same queue, or entering a new queue not at the end, immediately
-after an agent moves. However, agents take their steps sequentially and in an
-effectively random order. Ideally they could be reordered so that agents in the
-front of a queue always move first. Since the first agent may cross into the
-next queue, that introduces a dependency between traversables. A topological
-sort of a directed graph containing circuits\commentp{do you mean cycles?} is not possible though, and since
-we guarantee connectedness, there are necessarily circuits in maps.  Instead,
-collisions are checked after every agent has stepped forward in time, treating
-each time-step as atomic. Queues remember the original order of agents before a
-step, and afterwards can verify that the order is still legal, with new agents
-on the end and existing agents in the same order.\commentp{I don't
-  follow the point of the last sentence.}
+All traversables (edges and the straight line-approximation of turns) maintain
+an ordered \emph{queue} of occupying agents. Agents stepping forward in time
+happens atomically, so collision checking occurs after all agents have moved for
+some tick. This is done because otherwise false positives could occur when an
+agent steps before the agent in front of it, and ordering agents for updates is
+difficult near the crossings between edges. \commentd{Do I need to explain why
+it is difficult, or should I just stop after ``in front of it''?} A collision is
+defined as the reversal of order of agents before and after a step.
 
-Likewise, intersections check for collisions after everybody has moved by
-examining the turn each agent is performing and verifying that no other agent is
-simultaneously performing a conflicting turn.
+Likewise, intersections check for collisions by examining the turn each agent is
+performing and verifying that no other agent is simultaneously performing a
+conflicting turn.  AORTA has a low-granularity model of turn conflict. If two
+vehicles at any point along two different turns could ever come into contact,
+then the turns are always in conflict, no matter where the agent specifically is
+along that turn.  The alternative of tiling intersections \cite{JAIR08-dresner}
+would provide more granularity, but it adds computational complexity and would
+be impractical for many of the complex intersection geometries inferred from
+OSM.
 
-\commentpk{The following paragraph should not be here. It should be in intersection behaviors} AORTA has a low-granularity model of turn conflict. If two vehicles at any point
-along two different turns could ever come into contact, then the turns are
-always in conflict, no matter where the agent specifically is along that turn.
-The alternative of tiling intersections (like \cite{JAIR08-dresner} \commentpk{I made this same mistake a month ago, dont use citations as a noun} does) would
-provide more granularity, but it adds computational complexity and would be
-impractical for many of the complex intersection geometries inferred from OSM.
+When a collision occurs, the simulator throws warnings or halts, since this
+indicates a bug in some policy. A future alternative could be to mark the region
+as impassable and re-route other traffic away from the accident.
 
@@
     {scale=0.5}
-
-\commentp{What happens when a collision is detected? }
 
@@
 disappearing from the map (only when the agent is at rest and is done with its
-route) or setting an acceleration for the next step. In the future, another
-possible action will be initiating a lane-change or
-merge.\commentp{Say ``Were lane changing or merging modeled, that
-  would be another possible action.''}
+route) or setting an acceleration for the next step. Were lane changing or
+merging modeled, that would be another possible action.
 
@@
 collide with another agent or enter an intersection at the wrong time. Since it
-picks the fastest conservative (safe) choice of acceleration at each step, it
-could be easily extended to mimic human drivers by traveling at some random
-speed slower than this or to optimize fuel efficiency by tuning acceleration.
+picks the highest safe choice of acceleration at each step, it could be easily
+extended to mimic human drivers by traveling at some random speed slower than
+this or to optimize fuel efficiency by tuning acceleration.
 
@@
 
-Since agents know the fixed $dt$ duration, they may reason about what could
-happen during the next tick to avoid collision and illegal intersection entry.
-The conservative analysis is to determine worst case -- how far could an agent
-travel by the end of the next tick? If they accelerate as fast as possible for
-$dt$, this gives them the longest possible traveling distance and highest
-possible speed. Adding the stopping distance for this speed to the traveling
-distance gives the worst-case distance.
+\commentd{one diagram probably covers lots of this}
 
-\commentd{how much formula? pretty simple so far, but can certainly include them.
- should this part be worded like a proof? (because it can be)}
-\commentp{Use a formula if you can use it to mkae the description more
-  concise (and precise).}
+The default behavior maximizes accelerations while obeying speed limits,
+avoiding collisions, and obeying intersections. By consdering the highest speed
+achievable at the end of the next tick and the subsequent stopping distance at
+this speed, the behavior can scan sufficiently far ahead along several
+traversables to locate agents to avoid and intersections to query. The analysis
+can short-circuit at the first nearby agent and intersection requiring a stop,
+since it would be illegal to pass either.
 
-The worst-case distance may exceed the distance remaining on the agent's current
-traversable. That means they could start or finish a turn during the next tick.
-Because edges are often extremely short due to mis-marked ways in OSM (less than
-$???$ meters), they may travel through several traversables during the next
-move. Thus, any conservative behavior must look ahead until the worst-case
-distance is exhausted, asking the route strategy which turn an agent will choose
-in the future. At each traversable considered, there are 3 constraints to
-satisfy: obeying the intersection, avoiding collision with another agent, and
-obeying speed limits. Each constraint contributes to the max acceleration that
-will obey the limit, so ultimately the result of lookahead will be the minimum
-of these candidates.
+\subsubsection{Interaction with intersections}
 
-Obeying a road's speed limit is a simple constraint. Like many parameters of the
-map, speed limits are inferred from the road type claimed by the OSM
-source.
-
-\commentp{This section seems belabored.  Can't you just say that the
-  agent sets its acceleration to be the maximum possible that satifies
-  3 constraints: staying at or below the speed limit, avoiding
-  collisions, and not entering an intersection illegally.}
-
-\subsubsection{Obeying intersections}
+% TODO regroup? does this belong as a subsection, not subsub?
 
@@
 delay before entering the intersection). The intersection policy simply orders
-the agent to ``stop'' or ``continue.'' The lookahead considers no intersections
-beyond the first that issues ``stop'', since the behavior cannot legally proceed
-past that one anyway. The acceleration to avoid entering an intersection
+the agent to ``stop'' or ``continue.'' The acceleration to avoid entering an intersection
 attempts to end a configurable threshold back from the end of a traversable to
-account for crosswalks and as an epsilon value, since achieveing a distance
-along an edge equal to its length means spilling over into the next traversable,
-which the intersection has banned. Sometimes stopping behind this threshold is
-not possible -- namely, when the edge length is shorter -- and in that case,
-anywhere before the end is acceptable.
+account for crosswalks.
 
-\commentd{include some form of piyush's proof about accel to end?} \commentpk{I would say it is unnecessary}
+Intersection policies are one of the most intriguing aspects of the simulation
+to tweak, especially in light of autonomous vehicles. Current policies include
+traditional stop signs and traffic signals. Future policies could include stop
+lights that adaptively learn better timings and an AIM-based reservation system
+\cite{JAIR08-dresner}.
 
-\subsubsection{Optimal tailgating}
+\commentd{Actually, at least the stop sign policy is really mundane. Include its
+  description?}
 
-\commentd{diagram with overlapping distances}
+%The stop sign policy works by maintaining a current owner and a queue of waiting
+%agents. An agent can enter the queue only when it is within some threshold of
+%the start of its turn. After the agent at the head of the queue has waited for a
+%configurable pause, it obtains the lock and has exclusive access to the
+%intersection, releasing it only when it completes its turn.
 
-When the lookahead analysis is considering some future traversable, it only has
-to consider the agent closest to the start of it in the present. An edge may
-only be entered from a turn, so the intersection policy has the burden of
-ensuring no agent performs a turn that leads to the same edge simultaneously.
-Following an agent safely means maintaining a configurable following distance
-behind the agent \commentpk{not sure what you mean by configurable here}. Worst-case analysis is again used: the most dangerous move the
-agent in front could perform is decelerating as fast as possible
-during the next tick,
-and the worst-case traveling distance for the agent following is already known
-from lookahead (speeding up as much as possible).
+\commentd{describe traffic signal policy, once it's working fully}
+
+\subsection{Routing Strategies}
+
+Picking accelerations that optimize some objective functions is a separate task
+from picking a path to travel. Behaviors consult a route strategy to pick turns.
+Currently two simple implementations are available. One statically plans a route
+to some goal using standard A* search; the other performs a drunken walk, lazily
+planning random choices at each intersection as the lookahead engine needs to
+know. Future experiments in dynamic replanning or hierarchial planning can be
+performed without modifying any code other than a creating new route
+implementation. Route policies could also interact with a singleton overseer to
+have some sort of global insight.
+
+\subsection{Deadlock}
 
@@
 
-\commentd{diagram showing deadlock, or lookahead in gen eral}
-\commentd{i really hope to fix this issue soon, but it's not an immediately easy
-fix}
-\commentp{Cite (and ideally use ideas from) paper commented out in latex file
+%\commentp{Cite (and ideally use ideas from) paper commented out in latex file
     % @inproceedings{AAAI11-au,
@@
     % Month= {August},
-}
+%}
 
-\subsection{Routing Strategies}
+% TODO solve the above case, then figure out what to describe.
 
-Picking accelerations that optimize some objective functions is a separate task
-from picking a path to travel. Thus, the behavior described above can consult
-any route strategy. Currently two simple implementations are available. One
-statically plans a route to some goal using standard A* search; the other
-performs a drunken walk, lazily planning random choices at each intersection as
-the lookahead engine needs to know. Future experiments in dynamic replanning or
-hierarchial planning can be performed without modifying any code other than a
-creating new route implementation. Route policies could also interact with a
-singleton overseer to have some sort of global insight.
+%A form of deadlock is possible in a sequence of short edges and turns. Agent
+%$A_1$ acquires the lock to intersection $V_1$, but needs the lock to $V_2$ to
+%proceed. Meanwhile, $A_2$ has the lock for $V_2$ and needs the lock for $V_1$.
+%To deal with this circumstance, an agent's behavior can yield any held locks (a
+%notion generalized to any intersection policy) when its speed is $0$ and its
+%conservative acceleration is to remain at rest. The stop sign policy only
+%removes an agent's lock if it has not started its turn (because otherwise, the
+%agent is stalling in the intersection). The next time that agent polls, the
+%policy restores the agent to the front of the queue.
 
-% TODO and mike's variant that wanders in the right direction
-
-\subsection{Intersection Policies}
-
-Agents and behaviors (on behalf of an agent) poll an intersection as they
-approach it, asking if they should stop or not. A policy is free to prevent
-agents from performing conflicting turns concurrently by any means. Two main
-policies exist already, but implementing the interface for extensions is simple.
-
-The stop sign policy works by maintaining a current owner and a queue of waiting
-agents. An agent can enter the queue only when it is within some threshold of
-the start of its turn (a constraint satisfied by the default lookahead
-behavior), even if that is several (rather short) edges and turns away. After
-the agent at the head of the queue has waited for a configurable pause, it obtains the lock
-and has exclusive access to the intersection, releasing it only when it
-completes its turn.
-
-A form of deadlock is possible in a sequence of short edges and turns. Agent
-$A_1$ acquires the lock to intersection $V_1$, but needs the lock to $V_2$ to
-proceed. Meanwhile, $A_2$ has the lock for $V_2$ and needs the lock for $V_1$.
-To deal with this circumstance, an agent's behavior can yield any held locks (a
-notion generalized to any intersection policy) when its speed is $0$ and its
-conservative acceleration is to remain at rest. The stop sign policy only
-removes an agent's lock if it has not started its turn (because otherwise, the
-agent is stalling in the intersection). The next time that agent polls, the
-policy restores the agent to the front of the queue.
-
-\commentd{hopefully we'll have a neater solution than the above soon}
-
-\commentd{describe traffic signal policy, once it's working}
-
-Intersection policies are one of the most intriguing aspects of the simulation to
-tweak, especially in light of autonomous vehicles. Future policies could include stop lights that adaptively learn better
-timings and an AIM-based reservation system \cite{JAIR08-dresner}.
+%\commentd{hopefully we'll have a neater solution than the above soon}
 

### 2012-02-22 utexas/map/Traversable.scala

@@ -20,7 +20,7 @@ abstract class Traversable() {
       throw new Exception("Negative distance on a location?!")
     }
 
-    // TODO it's late, I am not going to write this functionally...
+    // TODO it's late at night, I am not going to write this functionally...
     var at = dist
     for (l <- lines) {
       if (at > l.length) {
@@ -35,7 +35,7 @@ abstract class Traversable() {
       }
     }
 
-    throw new Exception("Location is past the end of an edge! " + dist + " > " + length)
+    throw new Exception("Location is past the end of an edge! " + dist + " > " + length + "; at (dist left) = " + at + "; there are " + lines.size + "lines")
   }
 
   def start_pt = lines.head.start

### 2012-02-22 papers/itsc_2012/draft2.tex
small typos fixed

@@
 All traversables (edges and the straight line-approximation of turns) maintain
-an ordered \emph{queue} of occupying agents. Agents stepping forward in time
-happens atomically, so collision checking occurs after all agents have moved for
-some tick. This is done because otherwise false positives could occur when an
-agent steps before the agent in front of it, and ordering agents for updates is
+an ordered queue of occupying agents. Agents stepping forward in time happens
+atomically, so collision checking occurs after all agents have moved for some
+tick. This is done because otherwise false positives could occur when an agent
+steps before the agent in front of it, and ordering agents for updates is
 difficult near the crossings between edges. \commentd{Do I need to explain why
@@
 
-When a collision occurs, the simulator throws warnings or halts, since this
+When a collision occurs, the simulator throws warnings and halts, since this
 indicates a bug in some policy. A future alternative could be to mark the region
@@
 
-Each agent has a ``behavior'' governing it. When the agent travels past the end
+Each agent has a \emph{behavior} governing it. When the agent travels past the end
 of an edge during its step, the behavior picks the turn to pursue. Each tick,
@@
 
-The default behavior maximizes accelerations while obeying speed limits,
-avoiding collisions, and obeying intersections. By consdering the highest speed
+The default behavior maximizes acceleration while obeying speed limits,
+avoiding collisions, and obeying intersections. By considering the highest speed
 achievable at the end of the next tick and the subsequent stopping distance at
@@
 Currently two simple implementations are available. One statically plans a route
-to some goal using standard A* search; the other performs a drunken walk, lazily
-planning random choices at each intersection as the lookahead engine needs to
-know. Future experiments in dynamic replanning or hierarchial planning can be
-performed without modifying any code other than a creating new route
-implementation. Route policies could also interact with a singleton overseer to
-have some sort of global insight.
+to some goal using standard A* search (TODO cite); the other performs a drunken
+walk, lazily planning random choices at each intersection as the lookahead
+engine needs to know. Future experiments in dynamic replanning or hierarchial
+planning can be performed without modifying any code other than a creating new
+route implementation. Route policies could also interact with a singleton
+overseer to have some sort of global insight.
 

### 2012-02-24 papers/itsc_2012/draft2.tex
re-did arch diagram in a dumb uml program
UI draws agents between lane lines

@@
 
-\pix{architecture.pdf}{An illustration of AORTA's components.}{scale=0.3}
+\pix{architecture.png}
+    {An illustration of AORTA's components.}
+    {scale=0.2}
 
@@
 a curved road, but AORTA uses straight lines as an approximation. Lanes are
-physically placed a lane's width away from and parallel to the center line. (In
-the case of one-ways, there is no center line; instead, the lanes simply cluster
-around the invisible center lane, all pointing the same way.) Turns are modeled
-with a single straight line connecting the end of one edge to the start of
-another (so right turns often have $0$ length).
+parallel to the center line, some width apart from one another.  Turns are
+modeled with a single straight line connecting the end of one edge to the start
+of another (so right turns often have $0$ length).
 

### 2012-02-24 utexas/Util.scala

@@ -89,7 +89,7 @@ object cfg {
 
   // all distances should be in meters
   val doubles = List(
-    ("lane_width",     0.05, "Width of a lane", 0.01, 0.1),
+    ("lane_width",      0.5, "Width of a lane in meters", 0.01, 0.1),
     ("zoom_threshold",  5.0, "How close to zoom in before drawing details",    1.0, 15.0),
     ("epsilon",     0.00001, "What do we take as zero due to FP imprecision?", 0.0,  1.0),
     ("dt_s",            0.1, "The only dt in seconds an agent can experience", 0.1,  3.0),

### 2012-02-24 utexas/map/Edge.scala

@@ -173,6 +173,17 @@ class Line(var x1: Double, var y1: Double, var x2: Double, var y2: Double) {
       y1 + (height * percent)
     )
   }
+
+  def shift_line(off: Double): Line = {
+    // just move in the direction of the road (as given by the ordering of the
+    // points) plus 90 degrees clockwise
+    // TODO why does the angle() that respects inversion fail?
+    val theta = broken_angle + (math.Pi / 2)
+    val dx = off * cfg.lane_width * math.cos(theta)
+    val dy = off * cfg.lane_width * math.sin(theta)
+
+    return new Line(x1 + dx, y1 + dy, x2 + dx, y2 + dy)
+  }
 }
 
 object Direction extends Enumeration {

### 2012-02-24 utexas/map/make/Pass3.scala

@@ -11,7 +11,7 @@ import scala.collection.mutable.MutableList
 import utexas.map.{Road, Edge, Vertex, Turn, TurnType, Line, Coordinate, Ward,
                    Traversable}
 
-import utexas.Util
+import utexas.{Util, cfg}
 
 class Pass3(old_graph: PreGraph2) {
   Util.log("Multiplying and directing " + old_graph.edges.length + " edges")
@@ -33,32 +33,16 @@ class Pass3(old_graph: PreGraph2) {
       roads_per_vert.addBinding(r.v2, r)
 
       // pre-compute lines constituting the edges
-      if (r.is_oneway) {
-        val num = r.oneway_lanes.size
-        // These always make the oneway lanes center around the road's center
-        val offsets = if (num % 2 == 0)
-                        // even
-                        (-0.5 - ((num / 2) - 1)) until (0.5 + (num / 2)) by 1.0
-                      else
-                        // odd
-                        -(num / 2).toDouble until ((num / 2) + 1).toDouble by 1.0
-
-        // reverse because the most positive shift should be the rightmost lane
-        for ((e, offset) <- r.oneway_lanes.zip(offsets.reverse)) {
-          e.set_lines(for ((from, to) <- r.pairs_of_points)
-                      yield shift_line(offset, from, to))
-        }
-      } else {
-        for (e <- r.pos_lanes) {
-          e.set_lines(for ((from, to) <- r.pairs_of_points)
-                      yield shift_line(e.lane_offset, from, to))
-        }
-        for (e <- r.neg_lanes) {
-          val ls = for ((from, to) <- r.pairs_of_points)
-                   yield shift_line(e.lane_offset, to, from)
-          // TODO inefficient just because i wanted a two-liner?
-          e.set_lines(ls.reverse)
-        }
+      // the -0.5 lets there be nice lane lines between lanes
+      for (e <- r.pos_lanes) {
+        e.set_lines(for ((from, to) <- r.pairs_of_points)
+                    yield new Line(from, to).shift_line(e.lane_offset - 0.5))
+      }
+      for (e <- r.neg_lanes) {
+        val ls = for ((from, to) <- r.pairs_of_points)
+                 yield new Line(to, from).shift_line(e.lane_offset - 0.5)
+        // TODO inefficient just because i wanted a two-liner?
+        e.set_lines(ls.reverse)
       }
 
       // force line segments to meet up on the inside
[…]

### 2012-02-24 utexas/ui/MapCanvas.scala

@@ -117,7 +117,8 @@ class MapCanvas(sim: Simulation) extends ScrollingCanvas {
   private def build_fg_lines(): List[EdgeLine] = {
     val list = new ListBuffer[EdgeLine]()
     for (e <- sim.edges; l <- e.lines) {
-      val line = new EdgeLine(l, e)
+      // draw the lines on the borders of lanes, not in the middle
+      val line = new EdgeLine(l.shift_line(0.5), e)
       edge2lines.addBinding(e, line)
       list += line
     }
@@ -135,14 +136,14 @@ class MapCanvas(sim: Simulation) extends ScrollingCanvas {
   private val center_stroke     = new BasicStroke(0.1f)
   private val route_lane_stroke = new BasicStroke(0.45f)
   private val lane_stroke       = new BasicStroke(0.05f)
-  private val lane_width = 0.6f
   // TODO make this look cooler.
   private val drawing_stroke = new BasicStroke(
     5.0f, BasicStroke.CAP_BUTT, BasicStroke.JOIN_MITER, 1.0f, Array(1.0f), 0.0f
   )
 
   // pre-compute; we don't have more than max_lanes
-  private val strokes = (0 until cfg.max_lanes).map(n => new BasicStroke(lane_width * n.toFloat))
+  private val lane_line_width = 0.6f  // TODO cfg
+  private val strokes = (0 until cfg.max_lanes).map(n => new BasicStroke(lane_line_width * n.toFloat))
 
   def zoomed_in = zoom > cfg.zoom_threshold
 
@@ -183,7 +184,7 @@ class MapCanvas(sim: Simulation) extends ScrollingCanvas {
       // and the third layer (dashed center lines)
       g2d.setColor(Color.YELLOW)
       g2d.setStroke(center_stroke)
-      for (l <- roads_seen if !l.road.is_oneway) {
+      for (l <- roads_seen) {
         g2d.draw(l.line)
       }
 
@@ -247,7 +248,7 @@ class MapCanvas(sim: Simulation) extends ScrollingCanvas {
   def draw_road(g2d: Graphics2D, l: RoadLine) = {
     g2d.setColor(color_road(l.road))
     g2d.setStroke(strokes(l.road.num_lanes))
-    g2d.draw(l.line)
+    g2d.draw(l.bg_line)
   }
 
   def draw_edge(g2d: Graphics2D, l: EdgeLine) = {
@@ -763,11 +764,19 @@ sealed trait ScreenLine {
[…]

### 2012-02-25 papers/itsc_2012/draft2.tex
simple but vital clarification to paper

@@
 
-\commentd{is this clear now? it's not a multigraph and there are no loops, since
-  edges are lanes and turns. I have to think a bit about intersections as
-  vertices strictly, though...}
-
-A map is represented as a directed graph with \emph{lanes} and \emph{turns} as edges and
-\emph{intersections} as vertices. A \emph{road} groups one or two groups of
-lanes together, each group being a different direction of travel. Intersections
-map the possible turns from incoming to outgoing lanes. Since both lanes and
-turns support traffic, they are grouped together as the \emph{traversables} or
-edges, meaning they have some total length.
+A map is represented as a directed graph with \emph{lanes} as edges and
+\emph{turns} as vertices. A \emph{road} groups one or two groups of lanes
+together, each group being a different direction of travel.
+\emph{Intersections} map the possible turns from incoming to outgoing lanes.
+Since both lanes and turns support traffic, they are grouped together as the
+\emph{traversables}, meaning they have some geometric total length.
 

### 2012-02-26 utexas/sim/Agent.scala
gridlock 99% fixed for stop signs and traffic lights! (using a generalized lookahead engine and judicious cancelation of signal waiting)

@@ -166,7 +166,6 @@ class Agent(val id: Int, val graph: Graph, val start: Edge, val start_dist: Doub
     Util.log("At: " + at)
     Util.log("Speed: " + speed)
     Util.log("Next step's acceleration: " + target_accel)
-    Util.log("At end of edge? " + at_end_of_edge)
     Util.log("How long idle? " + how_long_idle)
     behavior.dump_info
     Util.log_pop
@@ -174,13 +173,20 @@ class Agent(val id: Int, val graph: Graph, val start: Edge, val start_dist: Doub
 
   def cur_queue = Agent.sim.queues(at.on)
 
+  // math queries for lookahead and such
+
   // stopping time comes from v_f = v_0 + a*t
   // negative accel because we're slowing down.
   def stopping_distance(s: Double = speed) = Util.dist_at_constant_accel(-max_accel, s / max_accel, s)
+  def max_next_speed = speed + (max_accel * cfg.dt_s)
+  def max_next_dist = Util.dist_at_constant_accel(max_accel, cfg.dt_s, speed)
+  // TODO clamp v_f at 0, since they cant deaccelerate into negative speed.
+  def min_next_dist = Util.dist_at_constant_accel(-max_accel, cfg.dt_s, speed)
+  def max_lookahead_dist = max_next_dist + stopping_distance(max_next_speed)
 
-  // If at.dist == at.on.length, then we'd actually end up on the next
-  // traversable. So accept something a few meters back.
-  def at_end_of_edge = at.dist_left <= cfg.end_threshold
+  def accel_to_achieve(target_speed: Double) = Util.accel_to_achieve(speed, target_speed)
+  // This directly follows from the distance traveled at constant accel
+  def accel_to_cover(dist: Double) = 2 * (dist - (speed * cfg.dt_s)) / (cfg.dt_s * cfg.dt_s)
 }
 
 // the singleton just lets us get at the simulation to look up queues

### 2012-02-26 utexas/sim/Behaviors.scala

@@ -58,27 +58,90 @@ class RouteFollowingBehavior(a: Agent, route: Route) extends Behavior(a) {
     Util.log("Route-following behavior")
   }
 
+  // Just a struct to encode all of this info
+  class LookaheadStep(
+    val lookahead_left: Double, val looked_ahead_so_far: Double,
+    val at: Traversable, val dist_left: Double, val steps_ahead: Int
+  ) {}
+
+  def lookahead(predict_dist: Double = a.max_lookahead_dist,
+                start_at: Traversable = a.at.on,
+                first_dist_left: Double = a.at.dist_left,
+                start_steps_ahead: Int = 0
+               ): List[LookaheadStep] =
+  {
+    var steps: List[LookaheadStep] = Nil
+
+    var lookahead_left = predict_dist
+    var looked_ahead_so_far = 0.0   // how far from a.at to beginning of step
+    var at = start_at
+    var dist_left = first_dist_left
+    var steps_ahead = start_steps_ahead
+
+    while (lookahead_left > 0.0) {
+      steps :+= new LookaheadStep(
+        lookahead_left, looked_ahead_so_far, at, dist_left, steps_ahead
+      )
+
+      lookahead_left -= dist_left
+      if (!route.lookahead_step(steps_ahead).isDefined) {
+        // done with route. stop considering possibilities immediately.
+        lookahead_left = -1
+      } else {
+        looked_ahead_so_far += dist_left
+        // order matters... steps_ahead will be what follows this new 'step'
+        at = route.lookahead_step(steps_ahead).get
+        steps_ahead += 1
+        // For all but the first step, we have all of its distance left.
+        dist_left = at.length
+      }
+    }
+
+    return steps
+  }
+
+  // true if it could happen
+  def check_for_gridlock(step: LookaheadStep): Boolean = {
+    // Look ahead from the start of the turn until follow_dist
+    // after the end of it too see if anybody's there.
+    val cautious_turn = route.lookahead_step(step.steps_ahead) match {
+      case Some(t: Turn) => t
+      case _       => throw new Exception("not a turn next?")
+    }
+    val cautious_edge = cautious_turn.to
+
+    val check_steps = lookahead(
+      // TODO + 0.5 as an epsilon...
+      predict_dist      = cautious_turn.length + cfg.follow_dist + 0.5,
+      start_at          = cautious_turn,
+      first_dist_left   = cautious_turn.length,
[…]

### 2012-02-26 utexas/sim/Generator.scala

@@ -28,10 +28,10 @@ abstract class Generator(sim: Simulation, desired_starts: List[Edge], val end_ca
 
   def add_specific_agent(start: Edge, end: Edge) = {
     // TODO how to decide?!
-    //val route = new StaticRoute()
+    val route = new StaticRoute()
     //val route = new DrunkenRoute()
     //val route = new DirectionalDrunkRoute()
-    var route = new DrunkenExplorerRoute()
+    //var route = new DrunkenExplorerRoute()
     val a = new Agent(sim.next_id, sim, start, sim.queues(start).safe_spawn_dist, route)
 
     // schedule whatever work the route needs done.

### 2012-02-26 utexas/sim/Intersection.scala

@@ -13,8 +13,8 @@ import utexas.{Util, cfg}
 class Intersection(val v: Vertex) {
   //val policy: Policy = new NeverGoPolicy(this)
   //val policy: Policy = new StopSignPolicy(this)
-  //val policy: Policy = new SignalCyclePolicy(this)
-  val policy: Policy = new ReservationPolicy(this)
+  val policy: Policy = new SignalCyclePolicy(this)
+  //val policy: Policy = new ReservationPolicy(this)
 
   override def toString = "Intersection(" + v + ")"
 
@@ -77,6 +77,7 @@ abstract class Policy(intersection: Intersection) {
   def yield_lock(a: Agent)
   def unregister(a: Agent)
   def current_greens(): Set[Turn] = Set()
+  def dump_info() = {}
 
   // Since we lookahead over small edges, we maybe won't/can't stop on the edge
   // right before the turn. As long as we validly stopped for us, then fine.
@@ -271,7 +272,11 @@ class ReservationPolicy(intersection: Intersection) extends Policy(intersection)
   override def current_greens = current_batch.turns.keys.toSet
 
   def yield_lock(a: Agent) = {}
-  def unregister(a: Agent) = {}
+  def unregister(a: Agent) = {
+    if (current_agents.contains(a)) {
+      // TODO what turn do they want to do? decrement that counter...
+    }
+  }
 }
 
 // Count how many agents are doing each type of turn, and add turns that don't
@@ -373,6 +378,22 @@ class SignalCyclePolicy(intersection: Intersection) extends Policy(intersection)
         start_waiting = Some(Agent.sim.tick)
       }
     }
+
+    if (start_waiting.isDefined) {
+      // Filter our agents who haven't started their turn and are not moving;
+      // although we told them they could go in this cycle, we can cancel --
+      // there's no danger of them entering because it's too late, since
+      // they've stopped.
+      for (pair <- current_agents) {
+        val a = pair._1
+        val t = pair._2
+        if (a.at.on != t && a.speed == 0.0) {
+          //Util.log("canceling " + pair)
+          current_agents -= pair
+        }
+        finish_waiting
+      }
+    }
     // TODO we want delay to be close to 0 mod duration, so maybe poke at
     // durations every now and then to get back on perfect schedule
 
@@ -387,8 +408,7 @@ class SignalCyclePolicy(intersection: Intersection) extends Policy(intersection)
     state_change
 
[…]

### 2012-02-26 utexas/ui/MapCanvas.scala

@@ -635,6 +635,9 @@ class MapCanvas(sim: Simulation) extends ScrollingCanvas {
             Util.log_push
             i.turns.foreach(pair => Util.log(pair._2 + " doing " + pair._1))
             Util.log_pop
+
+            // anything else
+            i.policy.dump_info
           }
           // This is what pressing 'd' when highlighting nothing means
           case _ => {

### 2012-02-26 utexas/ui/Viewer.scala

@@ -79,7 +79,7 @@ object Viewer extends SimpleSwingApplication {
       }
 
       contents += new Menu("Simulate") {
-        contents += new MenuItem("Spawn Agent") // TODO
+        //contents += new MenuItem("Spawn Agent") // TODO
         contents += new MenuItem(Action("Spawn Army") {
           canvas.handle_ev(EV_Action("spawn-army"))
         })

### 2012-02-26 utexas/sim/Behaviors.scala
agents creep forward properly to fix a close-intersection, unsynced-light issue

@@ -56,6 +56,12 @@ class RouteFollowingBehavior(a: Agent, route: Route) extends Behavior(a) {
 
   override def dump_info() = {
     Util.log("Route-following behavior")
+    // Just show a few steps ahead
+    Util.log_push
+    for (i <- 0 until 5) {
+      Util.log("Step: " + route.lookahead_step(i))
+    }
+    Util.log_pop
   }
 
   // Just a struct to encode all of this info
@@ -143,6 +149,7 @@ class RouteFollowingBehavior(a: Agent, route: Route) extends Behavior(a) {
 
     // the output.
     var stop_how_far_away: Option[Double] = None
+    var stop_at: Option[Traversable] = None
     var stopping_for_destination = false
     var follow_agent: Option[Agent] = None
     var follow_agent_how_far_away = 0.0   // gets set when follow_agent does.
@@ -210,12 +217,8 @@ class RouteFollowingBehavior(a: Agent, route: Route) extends Behavior(a) {
               // haven't started their turn yet, so wait for them to completely
               // finish first.
               case Some(a) => true
-              case _ => {
-                val gridlock_possible = check_for_gridlock(step)
-
-                // TODO so yield?
-                gridlock_possible
-              }
+              // TODO so yield?
+              case _ => check_for_gridlock(step)
             })
 
             stop_for_policy || stop_for_gridlock
@@ -224,6 +227,7 @@ class RouteFollowingBehavior(a: Agent, route: Route) extends Behavior(a) {
         if (stop_at_end) {
           keep_stopping = true
           stop_how_far_away = Some(how_far_away)
+          stop_at = Some(step.at)
         }
       }
 
@@ -241,17 +245,16 @@ class RouteFollowingBehavior(a: Agent, route: Route) extends Behavior(a) {
       // So, three possible constraints.
       val a1 = stop_how_far_away match {
         case Some(dist) => {
-          // in most normal circumstances, we take away end_threshold, since
-          // stopping right at the end of the edge makes us technically enter
-          // the intersection
-          // but when we're in the middle of a turn and that would make the
-          // distance be <= 0, we don't want to stop; that causes deadlock. it
-          // means the next edge is tiny. so just stop along it where we can.
-          // TODO lol uh, silly bug below? it's been there a while...
-          val go_this_dist = a.at.on match {
-            // just go halfway whatever's available.
-            case t: Turn if dist <= cfg.end_threshold => accel_to_end(dist / 2)
-            case _ => dist - cfg.end_threshold
[…]

### 2012-02-26 utexas/sim/Simulation.scala

@@ -76,11 +76,11 @@ class Simulation(roads: List[Road], edges: List[Edge], vertices: List[Vertex],
   var time_speed = 1.0
 
   // TODO cfg
-  def slow_down() = {
-    time_speed = math.max(0.5, time_speed - 0.5)
+  def slow_down(amount: Double = 0.5) = {
+    time_speed = math.max(0.5, time_speed - amount)
   }
-  def speed_up() = {
-    time_speed += 0.5
+  def speed_up(amount: Double = 0.5) = {
+    time_speed += amount
   }
 
   // Returns true if at least one generator is active

### 2012-02-26 utexas/ui/MapCanvas.scala

@@ -606,11 +606,19 @@ class MapCanvas(sim: Simulation) extends ScrollingCanvas {
         repaint
       }
       case EV_Key_Press(Key.OpenBracket) => {
-        sim.slow_down
+        sim.slow_down()
         status.time_speed.text = "%.1fx".format(sim.time_speed)
       }
       case EV_Key_Press(Key.CloseBracket) => {
-        sim.speed_up
+        sim.speed_up()
+        status.time_speed.text = "%.1fx".format(sim.time_speed)
+      }
+      case EV_Key_Press(Key.Minus) => {
+        sim.slow_down(10)
+        status.time_speed.text = "%.1fx".format(sim.time_speed)
+      }
+      case EV_Key_Press(Key.Equals) => {
+        sim.speed_up(10)
         status.time_speed.text = "%.1fx".format(sim.time_speed)
       }
       // general debug based on whatever you're hovering over

### 2012-02-26 papers/itsc_2012/draft2.tex
gridlock 99% fixed for stop signs and traffic lights! (using a generalized lookahead engine and judicious cancelation of signal waiting)
agents creep forward properly to fix a close-intersection, unsynced-light issue
no more ugly notion of 'yielding' lock to a stop sign, now that creep-forward behavior works.

@@
 
-\commentd{describe traffic signal policy, once it's working fully}
+\commentd{describe traffic signal policy, once it's working fully. the delay
+thing is maybe interesting.}
 
@@
 before starting it. This will prevent intersections from ever being blocked.
+
+\commentd{this is now fixed in most cases. describe, and describe a better
+  workaround for some cases: S turns}
+\commentd{and describe the full-capacity case and how that's hopeless anyway}
 

### 2012-02-26 utexas/sim/Behaviors.scala

@@ -217,7 +217,6 @@ class RouteFollowingBehavior(a: Agent, route: Route) extends Behavior(a) {
               // haven't started their turn yet, so wait for them to completely
               // finish first.
               case Some(a) => true
-              // TODO so yield?
               case _ => check_for_gridlock(step)
             })
 
@@ -272,24 +271,8 @@ class RouteFollowingBehavior(a: Agent, route: Route) extends Behavior(a) {
       // won't be going just yet,
       // IF we are not going to make a move this tick.
 
-      // This means we're holding a valuable resource...
-      // TODO Pretty sure this is no longer a problem! :D
-      if (a.speed == 0.0 && conservative_accel == 0.0) {
-        a.at.on match {
-          case t: Turn if a2 == 0.0 && a.how_long_idle >= 30.0 => {
-            // And the cause is an agent in front of us. If this situation persists,
-            // we could enter gridlock.
-            // Don't keep spamming the same message.
-            if (a.how_long_idle == 30.0) {
-              Util.log("Gridlock possible near %s, who is %.1f away from %s".format(a, follow_agent_how_far_away, follow_agent))
-            }
-          }
-          case _ =>
-        }
-
-        // let them ignore us if we're not owner
-        a.upcoming_intersections.foreach(p => p.yield_lock(a))
-      }
+      // No longer a problem, but detect a cause for gridlock here: when speed
+      // and conservative_accel are nil, especially when we're in a turn.
 
       // As the very last step, clamp based on our physical capabilities.
       return Act_Set_Accel(if (conservative_accel > 0)

### 2012-02-26 utexas/sim/Intersection.scala

@@ -12,15 +12,14 @@ import utexas.{Util, cfg}
 // Reason about collisions from conflicting simultaneous turns.
 class Intersection(val v: Vertex) {
   //val policy: Policy = new NeverGoPolicy(this)
-  //val policy: Policy = new StopSignPolicy(this)
-  val policy: Policy = new SignalCyclePolicy(this)
+  val policy: Policy = new StopSignPolicy(this)
+  //val policy: Policy = new SignalCyclePolicy(this)
   //val policy: Policy = new ReservationPolicy(this)
 
   override def toString = "Intersection(" + v + ")"
 
   // Just delegate.
   def can_go(a: Agent, turn: Turn, far_away: Double) = policy.can_go(a, turn, far_away)
-  def yield_lock(a: Agent) = policy.yield_lock(a)
   def unregister(a: Agent) = policy.unregister(a)
 
   // Multiple agents can be on the same turn; the corresponding queue will
@@ -74,7 +73,6 @@ abstract class Policy(intersection: Intersection) {
   def can_go(a: Agent, turn: Turn, far_away: Double): Boolean
   def validate_entry(a: Agent, turn: Turn): Boolean
   def handle_exit(a: Agent, turn: Turn)
-  def yield_lock(a: Agent)
   def unregister(a: Agent)
   def current_greens(): Set[Turn] = Set()
   def dump_info() = {}
@@ -89,7 +87,6 @@ class NeverGoPolicy(intersection: Intersection) extends Policy(intersection) {
   def can_go(a: Agent, turn: Turn, far_away: Double) = false
   def validate_entry(a: Agent, turn: Turn) = false
   def handle_exit(a: Agent, turn: Turn) = {}
-  def yield_lock(a: Agent) = {}
   def unregister(a: Agent) = {}
 }
 
@@ -99,40 +96,12 @@ class StopSignPolicy(intersection: Intersection) extends Policy(intersection) {
   // case, the first person has to pause a bit longer before continuing.
   var current_owner: Option[Agent] = None
   var queue = List[Agent]()
-  val yielders = new MutableSet[Agent]()
 
   def can_go(a: Agent, turn: Turn, far_away: Double): Boolean = {
-    // TODO remove once stagnation fixed.
-    Agent.sim.debug_agent match {
-      case Some(agent) if agent == a => {
-        Util.log(a + " wants to go, but locked from " + intersection.v + " by " + current_owner)
-        Util.log("  waiting: " + queue)
[…]

### 2012-02-27 utexas/sim/Queue.scala
lazy atomic queue-checking; in-place often suffices. considerable speedup.

@@ -11,12 +11,17 @@ import utexas.{Util, cfg}
 class Queue(t: Traversable) {
   // Descending by distance: head of list is front of traversable.
   var agents = List[Agent]()
+  var last_tick = -1.0              // last observed
   var prev_agents = List[Agent]()   // to verify no collisions occurred in a step
 
   def last = agents.lastOption
 
+  // Called lazily.
   def start_step() = {
-    prev_agents = agents
+    if (last_tick != Agent.sim.tick) {
+      prev_agents = agents
+      last_tick = Agent.sim.tick
+    }
   }
   
   // Check for collisions by detecting abnormal changes in ordering.
@@ -54,13 +59,27 @@ class Queue(t: Traversable) {
   def enter(a: Agent, dist: Double): Position = {
     // Just find our spot.
 
+    start_step  // lazily, if needed
+
     val (ahead, behind) = agents.partition(c => c.at.dist > dist)
     agents = ahead ++ List(a) ++ behind
+
+    // In-place check: we should enter at the end of the queue
+    if (!behind.isEmpty) {
+      Agent.sim.active_queues += this
+    }
 
     return Position(t, dist)
   }
 
   def exit(a: Agent) = {
+    start_step  // lazily, if needed
+
+    // In-place check: we should leave from the front of the queue
+    if (agents.head != a) {
+      Agent.sim.active_queues += this
+    }
+
     agents = agents.filter(c => c != a)
   }
 

### 2012-02-27 utexas/sim/Simulation.scala

@@ -43,10 +43,12 @@ class Simulation(roads: List[Road], edges: List[Edge], vertices: List[Vertex],
     agents.size, ready_to_spawn.size, generator_count, generators.size
   )
 
+  // Added by a queue that does an in-place check and thinks there could be an
+  // issue.
+  val active_queues = new MutableSet[Queue]   // TODO list?
   // It's expensive to go through every one of these without reason every
   // time. This banks on the fact that the number of agents is smaller, and
   // they're not evenly distributed through a huge map.
-  def active_queues(): Set[Queue] = agents.map(a => queues(a.at.on)).toSet
   def active_intersections(): Set[Intersection] = agents.flatMap(
     a => a.at.on match {
       case t: Turn => Some(intersections(t.vert))
@@ -126,9 +128,8 @@ class Simulation(roads: List[Road], edges: List[Edge], vertices: List[Vertex],
       // reacting. But that involves trusting that no simulation violations
       // could occur. ;)
 
-      //val t2 = Util.timer("queue start")
-      active_queues.foreach(q => q.start_step)
-      //t2.stop
+      // Queues will lazily start_step, remembering their current state, when
+      // they need to.
 
       //val t3 = Util.timer("agent step")
       agents.foreach(a => {
@@ -138,6 +139,7 @@ class Simulation(roads: List[Road], edges: List[Edge], vertices: List[Vertex],
       })
       //t3.stop
 
+      // Just check the ones we need to.
       //val t4 = Util.timer("queue stop")
       active_queues.foreach(q => q.end_step)
       //t4.stop
@@ -157,6 +159,9 @@ class Simulation(roads: List[Road], edges: List[Edge], vertices: List[Vertex],
       
       //Util.log_pop
       //t0.stop
+
+      // reset queues that need to be checked
+      active_queues.clear
     }
     return moved_count
   }

### 2012-02-27 utexas/ui/MapCanvas.scala

@@ -614,11 +614,11 @@ class MapCanvas(sim: Simulation) extends ScrollingCanvas {
         status.time_speed.text = "%.1fx".format(sim.time_speed)
       }
       case EV_Key_Press(Key.Minus) => {
-        sim.slow_down(10)
+        sim.slow_down(5)
         status.time_speed.text = "%.1fx".format(sim.time_speed)
       }
       case EV_Key_Press(Key.Equals) => {
-        sim.speed_up(10)
+        sim.speed_up(5)
         status.time_speed.text = "%.1fx".format(sim.time_speed)
       }
       // general debug based on whatever you're hovering over

### 2012-02-28 papers/itsc_2012/draft2.tex
lazy atomic queue-checking; in-place often suffices. considerable speedup.
measuring number of moves/second... very comforting number

@@
 
+\commentd{last sentence above... we're not not doing that yet, and we dont
+  discuss it further here.}
+
 Another popular open source traffic simulator that can use OSM data is SUMO
@@
 
+\commentd{space continuous? get terminology right}
+
 \begin{enumerate}
@@
 defined as the reversal of order of agents before and after a step.
+\commentd{lazy queue checking for speedup. mention briefly?}
 
@@
 
+\commentd{http://sumo.sourceforge.net/pdf/mieec1.pdf gives their measure as
+  agent updates / second. This is indep of dt and num of agents, assuming
+  nothing grows nonlinearly with more agents. our initial result on my box is
+  about 150,000 moves / second. so with 500 agents, runs 30x faster than
+  reality. with 5,000, runs about 3x faster. with 50,000 (yes, this works!),
+  runs at 1/3 the speed of reality (it takes 1 real second to simulate 0.3
+  seconds)}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2012-02-28 utexas/Util.scala

@@ -39,6 +39,17 @@ object Util {
   // return true 'p'% of the time. p is [0.0, 1.0]
   def percent(p: Double) = rand_double(0.0, 1.0) < p
 
+  // due to internationalization, printf doesn't have a way of adding commas
+  // after every 3 orders of mag
+  def comma_num(n: Int, pad: Boolean = true): String = {
+    return if (n < 1000 && pad)
+             "%03d".format(n)
+           else if (n < 1000)
+             "" + n
+           else
+             comma_num(n / 1000, pad = false) + "," + comma_num(n % 1000)
+  }
+
   // to meters/sec, that is. SI units.
   def mph_to_si(r: Double) = r * 0.44704
 

### 2012-02-28 utexas/sim/Headless.scala

@@ -45,21 +45,27 @@ object Headless {
     var last_real_time = 0.0
     var last_virtual_time = 0.0
     var max_movements = 0
+    var total_agent_steps = 0
     while (!sim.done) {
       // Print every 1 literal second
       val now = timer.so_far
       //val now = sim.tick    // or every 1 virtual second
       if (now - last_real_time >= 1.0) {
-        Util.log("At t=%.1f (%.1fs later): %s [%d agents moved]".format(
+        Util.log("At t=%.1f (%.1fs later): %s [%d agents moved] {%s moves/sec}".format(
           sim.tick, sim.tick - last_virtual_time, sim.describe_agents,
-          max_movements
+          max_movements, Util.comma_num(total_agent_steps)
         ))
         last_real_time = now
         last_virtual_time = sim.tick
         max_movements = 0
+        total_agent_steps = 0
       }
-      val moves = sim.step(cfg.dt_s)
-      max_movements = math.max(max_movements, moves)
+      sim.step(cfg.dt_s) match {
+        case (moved, total) => {
+          max_movements = math.max(max_movements, moved)
+          total_agent_steps += total
+        }
+      }
     }
     Util.log("Simulation took " + sim.tick + " virtual seconds")
     timer.stop

### 2012-02-28 utexas/sim/Intersection.scala

@@ -66,7 +66,8 @@ class Intersection(val v: Vertex) {
       Util.log("!!! " + a + " illegally entered intersection, going at " + a.speed)
       Util.log("  Incident was near " + t.from + " and " + t + " (vert " + t.from.to.id + ")")
       Util.log("  Origin lane length: " + t.from.length + "; time " + Agent.sim.tick)
-      assert(false)
+      Util.log("  Happened at " + Agent.sim.tick)
+      sys.exit
     }
   }
 

### 2012-02-28 utexas/sim/Simulation.scala

@@ -100,8 +100,8 @@ class Simulation(roads: List[Road], edges: List[Edge], vertices: List[Vertex],
     return changed
   }
 
-  // Returns the number of agents that moved
-  def step(dt_s: Double): Int = {
+  // Returns (the number of agents that moved, total number of agents processed)
+  def step(dt_s: Double): (Int, Int) = {
     pre_step
 
     // This value is dt in simulation time, not real time
@@ -110,6 +110,7 @@ class Simulation(roads: List[Road], edges: List[Edge], vertices: List[Vertex],
     tick += this_time
 
     var moved_count = 0
+    var total_count = 0
 
     // Agents can't react properly in the presence of huge time-steps. So chop
     // up this time-step into exactly consistent/equal pieces, if needed.
@@ -130,6 +131,7 @@ class Simulation(roads: List[Road], edges: List[Edge], vertices: List[Vertex],
         if (a.step(cfg.dt_s)) {
           moved_count += 1
         }
+        total_count += 1
       })
       //t3.stop
 
@@ -157,7 +159,7 @@ class Simulation(roads: List[Road], edges: List[Edge], vertices: List[Vertex],
       // reset queues that need to be checked
       active_queues.clear
     }
-    return moved_count
+    return (moved_count, total_count)
   }
 
   // True if we've correctly promoted into real agents. Does the work of

### 2012-03-07 papers/itsc_2012/draft2.tex
paper edits, filling out sections in simulation especially

@@
 \section{ARCHITECTURE}
+\label{sec:arch}
 
@@
 
-AORTA is divided into three modular components: the map model, a
-micro-simulation engine, and a UI. The map model transforms OSM maps into AORTA
-graphs, then answers pathfinding and geometry queries. The simulation engine
-adds a notion of agents, vehicle dynamics, and collisions. A headless mode
-enables running experiments without the overhead of visualization. Finally, the
-UI, which uses Swing, can interactively render the map and agents. For
-convenience, specific traffic patterns can be constructed between two regions of
-the map by simply drawing a polygon around each set of roads.
+AORTA is divided into three modular components: the map model, micro-simulation
+engine, and UI. The map model transforms OSM maps into AORTA graphs, then
+answers pathfinding and geometry queries. The simulation engine adds a notion of
+agents, vehicle dynamics, and collisions. Finally, the UI, which uses Swing, can
+interactively render the map and agents. A headless mode also enables running
+experiments without the overhead of visualization.
 
@@
 Machine. Scala provides the advantage of functional programming constructs while
-still permitting imperative style. Parts of AORTA, including extensions and clients, can
-just as easily be written in Java. The software has a documented API, is
+still permitting imperative style. Parts of AORTA, including extensions and
+clients, can easily be written in Java or Scala. The software, designed for
+extensibility, has a documented API, is
 open-source\footnote{http://code.google.com/p/road-rage}, and contains all of
-the code for the experiments described in Section 6 -- it is designed for
-extensibility.
+the code for the experiments described in Section \ref{sec:results}.
 
@@
 \section{MAP CONSTRUCTION FROM OSM}
+\label{sec:map}
 
 Our goal is to simulate traffic on existing infrastructure, not contrived,
-randomly generated maps. We do this by utilizing OpenStreetMaps, a community
-project offering user-editable, free maps. To run a simulation on a map from
-OSM, there is a one-time conversion process to transform the map into AORTA's
-format, which is also encoded with XML.
+randomly generated maps. Utilizing OpenStreetMaps, a community project offering
+user-editable, free XML-based maps, enables this. To run a simulation on a map
+from OSM, there is a one-time conversion process to transform the map into
+AORTA's format, which is also encoded with XML.
 
@@
 \emph{Intersections} map the possible turns from incoming to outgoing lanes.
-Since both lanes and turns support traffic, they are grouped together as the
-\emph{traversables}, meaning they have some geometric total length.
+Since both lanes and turns support traffic, they are grouped together as
+\emph{traversables}, meaning agents may exist on them.
 
 The map also has a geometric interpretation, used for visualization and physical
-distances. A road has a sequence of points giving the line segments of the
+distances. A road contains a sequence of points giving the line segments of the
 center yellow line. These points could also be interpreted as control points for
 a curved road, but AORTA uses straight lines as an approximation. Lanes are
-parallel to the center line, some width apart from one another.  Turns are
-modeled with a single straight line connecting the end of one edge to the start
+parallel to the center line, some fixed width apart from one another. Turns are
+modeled with a single straight line connecting the end of one lane to the start
 of another (so right turns often have $0$ length).
@@
 \subsection{Map Construction Passes}
-
-\commentd{gotta work on passive voice}
 
@@
         OSM's ``road type'' tag or an explicit number of lanes, when given.
-  \item Turns are next constructed between incoming and outgoing edges at each
-        vertex. Left, right, straight, and U-turns are heuristically chosen
-        based on the angles between edges. When several lanes all cross into
-        fewer lanes, merging is arbitrarily forced on the rightmost lanes.
-        Although inaccurate, these patterns are the best compromise. At the
-        time of writing, even sources like Google Maps (TODO cite) lack this
-        information.
-  \item Because turns are heuristically placed, the graph is disconnected. The
-        strongly connected components of the digraph with the fewest members are
-        removed. Tarjan's algorithm locates these in linear time (TODO cite).
+  \item Turns are next constructed between incoming and outgoing lanes at each
+        vertex. Left, right, straight, and U-turns are heuristically chosen based on
+        the angles between lanes. When several lanes all cross into fewer lanes,
+        merging is arbitrarily forced on the rightmost lanes.  Although inaccurate,
+        these patterns are the best compromise. At the time of writing, even sources
+        like Google Maps \commentd{TODO cite} lack this information.
+  \item Because turns are heuristically placed, the graph is usually
+        disconnected. Since the largest resulting map is desirable, the strongly
+        connected components of the digraph with the fewest members are removed.
         Disconnected lanes are spliced out of larger roads, leaving unappealing
@@
 \section{SIMULATION}
+\label{sec:simulation}
 
 AORTA uses microscopic simulation, modeling individual drivers as agents.
-Discrete-time simulation is used, meaning an agent accelerates at a fixed rate
-for the duration of a ``tick'' (also referred to as ``step'') to achieve a new
-position and velocity. The tick duration $dt$ is fixed, enabling agents to
-reason about collision avoidance and satisfying other constraints. Setting this
-value to the average ``reaction time'' of a human driver may be appropriate for
-some experiments. Each tick, the simulation performs the following steps:
-
-\commentd{space continuous? get terminology right}
+Simulation is discrete-time, meaning an agent accelerates at a fixed rate for
+the duration of a ``step'' to achieve a new position and velocity, and
+space-continuous, meaning roads are not tiled. The step duration $dt$ is fixed,
+enabling agents to reason about collision avoidance and satisfying other
+constraints. Setting this value to the average ``reaction time'' of a human
+driver may be appropriate for some experiments. Each step, the simulation
+performs the following:
 
@@
   \item Updating the position and velocity of agents based on their choice the
-        previous tick
+        previous step
   \item Checking for collisions
@@
 traversable without appearing on the next. Intersection policies and avoiding
-collisions on the same traversable provide the other constraints, but it is the
-job of an agent's behavior to satisfy those. There are three main components of
-the simulation, each configurable by extending default implementations of an
+collisions on the same traversable provide the other constraints, but the
+agent's behavior must satisfy these. There are three main components of the
+simulation, each configurable by extending built-in implementations of an
 interface: agent behaviors, routing strategies, and intersection policies.
 
-%\commentd{justify arbitrary constants, like deaccel for cars;
-%          systems ideas, or limitations.. trylocks, atomic sim update, deadlock
-%          issues;
-%          limit: no lane-changing}
+Currently lane-changing is not supported. Merging lanes is accomplished by
+treating the merge point as an intersection.
 
-\subsection{Spawning}
+\subsection{Spawning new agents}
 
-To introduce new agents into the simulation in real-time, users create
+To introduce new agents into the simulation in real-time, users can create
 \emph{generators} programatically or using the UI. A generator picks a random
-starting position and random goal from two sets of roads chosen by the user,
-selectable by drawing a polygon with the mouse in the UI. A generator may cover
-the entire map, uniformly distributing traffic everywhere, or it can mimic
-rush-hour scenarios.
+starting position and goal from two sets of roads. These roads can be selected
+by mouse with a bounding polygon in the UI, letting specific traffic patterns
+like rush-hour scenarios easily be constructed between two regions. A generator
+may also cover the entire map, uniformly distributing traffic everywhere.
 
-Every tick, a generator creates some number of new agents. If the agent's route
-policy needs work to be done (such as pathfinding), the generator either
+Every step, a generator creates some number of new agents. If the agent's route
+policy needs computation like pathfinding to operate, the generator either
 performs it immediately (incurring a delay in simulation if it is running) or
-sends it to a pool of worker threads to compute in the background.
+sends it to a pool of worker threads to process in the background.
 
-Once an agent's route is ready, the agent waits alongside its starting edge (as
-if it is in a driveway or parking lot). When the edge has no agents that could
+Once an agent's route is ready, the agent waits alongside its starting lane (as
+if it is in a driveway or parking lot). When the lane has no agents that could
 potentially crash with the new vehicle, the new agent enters the system with
-zero initial speed.  Because behaviors must enforce speed limit along an edge by
-the time an agent enters the road, then if there is no agent within the
+zero initial speed.  Because behaviors must enforce speed limit along a lane 
+from the moment an agent enters the road, then if there is no agent within the
 worst-case distance (traveling distance at the speed limit, plus stopping
 distance from this speed) before the spawn point, then the new agent may enter.
-To render unnecessary the prediction of which agents could enter the edge in the
-near future, there is a buffer space at the beginning of the edge where a new
-agent may never spawn. This disqualifies some short edges from being spawn
-points, but this is not a big limitation in practice.
+To avoid predicting which agents might enter the lane in the near future, there
+is a buffer space at the beginning of the lane where a new agent may never
+spawn. This disqualifies some short lanes from being spawn points, but this is
+not a big limitation in practice.
 
@@
 
-All traversables (edges and the straight line-approximation of turns) maintain
-an ordered queue of occupying agents. Agents stepping forward in time happens
-atomically, so collision checking occurs after all agents have moved for some
-tick. This is done because otherwise false positives could occur when an agent
-steps before the agent in front of it, and ordering agents for updates is
-difficult near the crossings between edges. \commentd{Do I need to explain why
-it is difficult, or should I just stop after ``in front of it''?} A collision is
-defined as the reversal of order of agents before and after a step.
-\commentd{lazy queue checking for speedup. mention briefly?}
+All traversables (lanes and the straight line-approximation of turns) maintain
+an ordered queue of occupying agents. One simulation step happens atomically,
+eaning collision checking occurs after all agents have moved. If this was not
+done, false positives could occur when an agent steps before the agent in front
+of it. A collision is defined as the reversal of order of agents before and
+after a step.
 
@@
 Each agent has a \emph{behavior} governing it. When the agent travels past the end
-of an edge during its step, the behavior picks the turn to pursue. Each tick,
+of a lane during its step, the behavior picks the turn to pursue. Each step,
 the behavior can make the agent react by performing one of two actions:
@@
 
-Currently, the primary behavior used by all agents is the ``route-following
-behavior.'' It is a generalized, baseline behavior that guarantees not to
-collide with another agent or enter an intersection at the wrong time. Since it
-picks the highest safe choice of acceleration at each step, it could be easily
-extended to mimic human drivers by traveling at some random speed slower than
-this or to optimize fuel efficiency by tuning acceleration.
-
-\subsubsection{Worst-case analysis and lookahead}
+Currently, the primary behavior used by all agents is a generalized, baseline
+behavior that guarantees not to collide with another agent or enter an
+intersection at the wrong time. Since it picks the highest safe choice of
+acceleration at each step, it could be easily extended to mimic human drivers by
+traveling at some random slower acceleration or to optimize fuel efficiency
+by tuning movement.
 
@@
 avoiding collisions, and obeying intersections. By considering the highest speed
-achievable at the end of the next tick and the subsequent stopping distance at
+achievable at the end of the next step and the subsequent stopping distance at
 this speed, the behavior can scan sufficiently far ahead along several
 traversables to locate agents to avoid and intersections to query. The analysis
-can short-circuit at the first nearby agent and intersection requiring a stop,
-since it would be illegal to pass either.
+short-circuits at the first nearby agent and intersection requiring a stop,
+since it would be illegal to pass another anyway.
 
-\subsubsection{Interaction with intersections}
-
-% TODO regroup? does this belong as a subsection, not subsub?
+\subsection{Intersection Policies}
 
@@
 delay before entering the intersection). The intersection policy simply orders
-the agent to ``stop'' or ``continue.'' The acceleration to avoid entering an intersection
-attempts to end a configurable threshold back from the end of a traversable to
-account for crosswalks.
+the agent to ``stop'' or ``proceed.'' The acceleration chosen to avoid entering
+an intersection attempts to account for crosswalks by ending a configurable
+threshold before the end of a traversable.
 
@@
 to tweak, especially in light of autonomous vehicles. Current policies include
-traditional stop signs and traffic signals. Future policies could include stop
-lights that adaptively learn better timings and an AIM-based reservation system
-\cite{JAIR08-dresner}.
-
-\commentd{Actually, at least the stop sign policy is really mundane. Include its
-  description?}
-
-%The stop sign policy works by maintaining a current owner and a queue of waiting
-%agents. An agent can enter the queue only when it is within some threshold of
-%the start of its turn. After the agent at the head of the queue has waited for a
-%configurable pause, it obtains the lock and has exclusive access to the
-%intersection, releasing it only when it completes its turn.
-
-\commentd{describe traffic signal policy, once it's working fully. the delay
-thing is maybe interesting.}
+traditional stop signs, traffic signals, and an AIM-inspired autonomous
+reservation system \cite{JAIR08-dresner}. The traffic signals are further
+extensible by assigning groups of non-conflicting turns at each intersection
+with some duration and offset. The experiment in section \ref{sec:results}
+features standard turn grouping (turns from parallel incoming directions are
+scheduled simultaneously, and left turns are grouped) and a breadth-first search
+heuristic for propagating timings. Future policies could include signals that
+adapt timings and refinements to the autonomous reservation policy.
 
@@
 
-Picking accelerations that optimize some objective functions is a separate task
-from picking a path to travel. Behaviors consult a route strategy to pick turns.
-Currently two simple implementations are available. One statically plans a route
-to some goal using standard A* search (TODO cite); the other performs a drunken
-walk, lazily planning random choices at each intersection as the lookahead
-engine needs to know. Future experiments in dynamic replanning or hierarchial
-planning can be performed without modifying any code other than a creating new
-route implementation. Route policies could also interact with a singleton
-overseer to have some sort of global insight.
+Picking accelerations that optimize objective functions is a separate task from
+picking a path to travel. Agent behaviors consult a route strategy to pick
+turns. Currently two simple implementations are available. One statically plans
+a route to some goal using standard A* search (\commentd{TODO cite}); the other
+performs a drunken walk, lazily planning random choices at each intersection as
+the lookahead engine needs to know. Future experiments in dynamic replanning or
+hierarchial planning can be performed without modifying any code other than a
+creating new route implementation. Route policies could also interact with a
+singleton overseer to gain some sort of global insight or with smart
+intersections to avoid congestion.
 
-\subsection{Deadlock}
+\subsection{Gridlock}
 
-Maintaining following distance from agents actually enables a form of deadlock
-in the simulation. If an agent is forced to stop in an intersection mid-turn due
-to its destination edge filling up with other agents, then nobody may cross the
-blocked intersection. With many agents in a small area, this situation may
-occur several times in interdependent ways. The solution currently being
-investigated will require behaviors to ensure a turn may be fully completed
-before starting it. This will prevent intersections from ever being blocked.
+Agents maintain a safe following distance from the next agent. If a lane is
+filled, then an agent may be forced to stop in an intersection mid-turn and
+block other traffic. The default behavior prevents this by refusing to start a
+turn before the lookahead engine guarantees no agent will trigger a premature
+stop.
 
-\commentd{this is now fixed in most cases. describe, and describe a better
-  workaround for some cases: S turns}
-\commentd{and describe the full-capacity case and how that's hopeless anyway}
+\commentd{wiki has a page naming gridlock. cite that for the name?}
 
-%\commentp{Cite (and ideally use ideas from) paper commented out in latex file
-    % @inproceedings{AAAI11-au,
-    % Author = {Tsz-Chiu Au and Neda Shahidi and Peter Stone},
-    % Booktitle= {Proceedings of the Twenty-Fifth Conference on Artificial Intelligence},
-    % Title = {Enforcing Liveness in Autonomous Traffic Management},
-    % Year = {2011},
-    % Month= {August},
-%}
+However, with very high volumes of traffic, it is possible for waiting agents to
+accumulate and fill lanes to their capacity. When this happens in a circular
+manner so that the agent at the front of each lane is waiting to perform a turn
+into another full lane with the same such head agent, gridlock occurs; no agent
+in the system will make progress unless an agent at the front decides to pick a
+different turn. At best, AORTA can detect this by maintaining a dependency graph
+that encodes the ``is following'' relation and finding cycles. It is not clear
+how to prevent or mend the deadlock without forcing agents to re-route.
 
-% TODO solve the above case, then figure out what to describe.
-
-%A form of deadlock is possible in a sequence of short edges and turns. Agent
-%$A_1$ acquires the lock to intersection $V_1$, but needs the lock to $V_2$ to
-%proceed. Meanwhile, $A_2$ has the lock for $V_2$ and needs the lock for $V_1$.
-%To deal with this circumstance, an agent's behavior can yield any held locks (a
-%notion generalized to any intersection policy) when its speed is $0$ and its
-%conservative acceleration is to remain at rest. The stop sign policy only
-%removes an agent's lock if it has not started its turn (because otherwise, the
-%agent is stalling in the intersection). The next time that agent polls, the
-%policy restores the agent to the front of the queue.
-
-%\commentd{hopefully we'll have a neater solution than the above soon}
+\commentd{is the above clear and an acceptable limitation? should i find other
+sources that have the same troubles?}
 
@@
 AORTA is designed for controlled experiments. Consequently, simulations are
-deterministically reproducible. A single seed for the pseudo-random number
-generator, the input graph, and the generators' configuration completely
-determine the outcome of any simulation. This determinism does break down when
-generators use worker threads, since routes may be computed in different orders,
-causing agents to be ready to enter the road at different times.
+deterministically reproducible. A seed for the pseudo-random number generator,
+the input graph, and the generators' configuration determine the outcome of any
+simulation. This determinism may break down when generators use worker threads,
+since routes may be computed in different orders, causing agents to enter the
+road at different times.
 
@@
 \section{EXPERIMENTAL RESULTS}
+\label{sec:results}
 
@@
 \section{CONCLUSIONS}
+\label{sec:conclusion}
 
@@
 
-\section*{ACKNOWLEDGMENT}
+\section*{ACKNOWLEDGMENTS}
 

### 2012-03-07 utexas/sim/Intersection.scala

@@ -12,9 +12,9 @@ import utexas.{Util, cfg}
 // Reason about collisions from conflicting simultaneous turns.
 class Intersection(val v: Vertex) {
   //val policy: Policy = new NeverGoPolicy(this)
-  val policy: Policy = new StopSignPolicy(this)
+  //val policy: Policy = new StopSignPolicy(this)
   //val policy: Policy = new SignalCyclePolicy(this)
-  //val policy: Policy = new ReservationPolicy(this)
+  val policy: Policy = new ReservationPolicy(this)
 
   override def toString = "Intersection(" + v + ")"
 

### 2012-03-07 papers/itsc_2012/draft2.tex
wrote a conclusion

@@
 
-\section{CONCLUSIONS}
+\section{CONCLUSION}
 \label{sec:conclusion}
 
-Write me. big future ideas:
+This paper has presented AORTA, a new city-scale traffic simulation framework
+that focuses on ease-of-use and policy configurability. In conjunction with Open
+Street Maps, AORTA allows researchers to repeat an experiment in any number of
+places with minimal effort. Future work will exploit the flexible infrastructure
+of interactions between agent behaviors and intersections by exploring dynamic
+replanning and adaptable signal timings. On top of these structures,
+applications could experiment with contra-flow \commentd{cite these all} and
+micro-tolling. There will also be an effort to improve AORTA as a framework by
+including a lane-changing model and preventing gridlock.
 
-\commentp{Proposed improvements to OSM data would be great for a
-  discussion or conclusion section at the end of the paper.}
-proposals to change OSM to fix these issues
-  projecting lines sucks... but all it really does is clutter UI
-  filtering out invalid ways in pass 1 is really not a clear process
-  number of lanes and speed limits... road types are wacky and wrong
-  turns are brittle.. detecting right-turn only lane is rare. shared center lane?
-    not part of our model, or OSM's.
-  removing sccs is not ideal
+AORTA's map construction process, although flexible, is fraught with heuristics,
+motivating proposed changes to OSM. Road type annotations imprecisely estimate
+the number of lanes and typical speed limits, resulting in incorrect graphs.
+Guessing the turns at each intersection makes common road features like right
+turn-only lanes and shared center left turn lanes hard to detect. There are many
+circumstances where AORTA simulates one intersection as several in close
+proximity because roads do not geometrically line up, but semantically they do.
+If OSM could encode this, the performance of intersection policies such as
+traffic signals would vastly improve in those regions.
 
-contraflow
-
-microtolling
-
-dynamic replanning
-
-autonomous intersections
+\commentd{Er, so how do I finally end this?}
 
@@
 
-% Notes.
-
-%possible background/links
-
-%http://www.caliper.com/transmodeler/default.htm
-%http://ieeexplore.ieee.org/stamp/stamp.jsp?tp=&arnumber=6082793
-%https://www.google.com/search?sourceid=chrome&ie=UTF-8&q=Agent-Based+Traffic+Simulation+and+Traffic+Signal+Timing+Optimization+with+GPU
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
 \bibliographystyle{IEEEtran}

### 2012-03-07 papers/itsc_2012/draft2.tex
fixed bibliography and some citations

@@
         these patterns are the best compromise. At the time of writing, even sources
-        like Google Maps \commentd{TODO cite} lack this information.
+        like Google Maps \cite{googleMaps} lack this information.
   \item Because turns are heuristically placed, the graph is usually
@@
 turns. Currently two simple implementations are available. One statically plans
-a route to some goal using standard A* search (\commentd{TODO cite}); the other
-performs a drunken walk, lazily planning random choices at each intersection as
-the lookahead engine needs to know. Future experiments in dynamic replanning or
+a route to some goal using standard A* search \cite{astar}; the other performs a
+drunken walk, lazily planning random choices at each intersection as the
+lookahead engine needs to know. Future experiments in dynamic replanning or
 hierarchial planning can be performed without modifying any code other than a
@@
 
-\commentd{wiki has a page naming gridlock. cite that for the name?}
-
 However, with very high volumes of traffic, it is possible for waiting agents to
@@
 manner so that the agent at the front of each lane is waiting to perform a turn
-into another full lane with the same such head agent, gridlock occurs; no agent
-in the system will make progress unless an agent at the front decides to pick a
-different turn. At best, AORTA can detect this by maintaining a dependency graph
-that encodes the ``is following'' relation and finding cycles. It is not clear
-how to prevent or mend the deadlock without forcing agents to re-route.
+into another full lane with the same such head agent, gridlock \cite{gridlock}
+occurs; no agent in the system will make progress unless an agent at the front
+decides to pick a different turn. At best, AORTA can detect this by maintaining
+a dependency graph that encodes the ``is following'' relation and finding
+cycles. It is not clear how to prevent or mend the deadlock without forcing
+agents to re-route.
 
@@
 replanning and adaptable signal timings. On top of these structures,
-applications could experiment with contra-flow \commentd{cite these all} and
-micro-tolling. There will also be an effort to improve AORTA as a framework by
-including a lane-changing model and preventing gridlock.
+applications could experiment with dynamic contra-flow \cite{ITSC11-hausknecht}
+and micro-tolling \commentd{need a cite}. There will also be an effort to
+improve AORTA as a framework by including a lane-changing model and preventing
+gridlock.
 
@@
 
+\commentd{Er, these don't look like they're alphabetically sorted.}
+
 \bibliographystyle{IEEEtran}

### 2012-03-07 papers/itsc_2012/generate.sh

@@ -13,7 +13,7 @@ for f in *.dot; do
 done
 
 pdflatex $INPUT
-bibtex ${INPUT%tex}
+bibtex ${INPUT%.tex}
 # pretty sure we have to re-run this twice to include the newly generated bib
 # and references
 pdflatex $INPUT

### 2012-03-07 papers/itsc_2012/draft2.tex
Removed a lot of passive voice, cleaned up wording, added some information.

@@
 (OSM) \cite{osm}, a Wikipedia-like interface for world maps. Researchers using
-AORTA can export the relevant map area from OSM that is then parsed by AORTA to
-create scale simulations of the real world. Using real world road data is
-advantageous for a number of reasons. First, researchers using AORTA for human
-traffic simulations would find it advantageous to also have the ability to use
-real-world roads. Second, even as infrastructural changes will be made to the
-road infrastructure to accommodate driver-less cars, it is reasonable to assume
-that the \textit{same} road network will still be in use.\commentp{I
-  don't understand this second point} Third, using
-real-world data from OSM allows setting up simulations for an area of interest
-in a matter of minutes, thereby allowing easy transfer of research results to
-any desired region.
-
-\commentd{The first reason for real world data almost feels too obvious to state
-like that}
-\commentp{I think all 3 reasons are similar.  I'd just say simplty
-  that AORTA's key features are being open source, easy to install and
-  get running, and flexibility to quickly and easily incorporate
-  models from real road networks around the world.}
+AORTA can export any relevant map area from OSM. This map is then parsed by AORTA to
+create scale simulations of the real world. A researcher employing AORTA can have a simulation
+of any part of the world for which OSM has data running in about five minutes
+(allowing for download time).  AORTA also has the advantage of being open source and
+easy to install so that researchers can make any tweaks they find necessary.
+\footnote{http://code.google.com/p/road-rage}
 
@@
 AORTA is divided into three modular components: the map model, micro-simulation
-engine, and UI. The map model transforms OSM maps into AORTA graphs, then
+engine, and user interface (UI). The map model transforms OSM maps into AORTA graphs, then
 answers pathfinding and geometry queries. The simulation engine adds a notion of
-agents, vehicle dynamics, and collisions. Finally, the UI, which uses Swing, can
+agents, vehicle dynamics, and collisions. Finally, the UI can
 interactively render the map and agents. A headless mode also enables running
@@
 
-Our goal is to simulate traffic on existing infrastructure, not contrived,
+One advantage of AORTA is that it simulates traffic on existing infrastructure rather than contrived,
 randomly generated maps. Utilizing OpenStreetMaps, a community project offering
@@
 A map is represented as a directed graph with \emph{lanes} as edges and
-\emph{turns} as vertices. A \emph{road} groups one or two groups of lanes
-together, each group being a different direction of travel.
+\emph{turns} as vertices. A \emph{road} groups one or two sets of lanes
+together, one set for each direction of travel.
 \emph{Intersections} map the possible turns from incoming to outgoing lanes.
@@
 \begin{enumerate}
-  \item OSM encodes many paths besides driveable roads, so these are first
-        filtered out. GPS locations common to multiple roads are marked, since
+  \item OSM encodes many paths besides driveable roads, so the builder first filters
+        these out. Next the builder marks GPS locations common to multiple roads, since
         they implicitly indicate intersections.
@@
         segments of roads between exactly two intersections.
-  \item Undirected roads are multiplied into directed lanes in each direction
-        (unless the road is marked one-way). The number of lanes is guessed from
-        OSM's ``road type'' tag or an explicit number of lanes, when given.
-  \item Turns are next constructed between incoming and outgoing lanes at each
-        vertex. Left, right, straight, and U-turns are heuristically chosen based on
-        the angles between lanes. When several lanes all cross into fewer lanes,
-        merging is arbitrarily forced on the rightmost lanes.  Although inaccurate,
+  \item The builder multiplies undirected roads into directed lanes in each direction
+        (unless the road is marked one-way). The builder guesses the number of lanes based on
+        OSM's ``road type'' tag or an explicit number of lanes, when that data is available.
+  \item The builder constructs turns between incoming and outgoing lanes at each
+        vertex. The builder uses an angle heuristic to choose between left, right, straight,
+        and U-turns. When several lanes all cross into fewer lanes,
+        the builder forces merging of the rightmost lanes.  Although inaccurate,
         these patterns are the best compromise. At the time of writing, even sources
         like Google Maps \cite{googleMaps} lack this information.
-  \item Because turns are heuristically placed, the graph is usually
-        disconnected. Since the largest resulting map is desirable, the strongly
-        connected components of the digraph with the fewest members are removed.
+  \item Because turns are heuristically placed, the graph is often
+        disconnected. In order to ensure that an agent will be able to reach any point in the map,
+        the builder removes all but the largest strongly connected component of the digraph.
         Disconnected lanes are spliced out of larger roads, leaving unappealing
@@
 \begin{enumerate}
-  \item Introducing new agents into the map
-  \item Updating the position and velocity of agents based on their choice the
-        previous step
-  \item Checking for collisions
-  \item Allowing each agent to react
+  \item Introduce new agents into the map
+  \item Update the position and velocity of agents based on their choice of
+  acceleration in the previous step
+  \item Check for collisions
+  \item Allow each agent to choose a reaction for the next step
 \end{enumerate}
@@
 collisions on the same traversable provide the other constraints, but the
-agent's behavior must satisfy these. There are three main components of the
+agent's behavior is responsible for obeying these. There are three main components of the
 simulation, each configurable by extending built-in implementations of an
@@
 
-Currently lane-changing is not supported. Merging lanes is accomplished by
+Currently agents cannot switch lanes. AORTA allows merging lanes by
 treating the merge point as an intersection.
@@
 \emph{generators} programatically or using the UI. A generator picks a random
-starting position and goal from two sets of roads. These roads can be selected
-by mouse with a bounding polygon in the UI, letting specific traffic patterns
-like rush-hour scenarios easily be constructed between two regions. A generator
+starting position and goal from two sets of roads. Users can select these roads 
+by mouse with a bounding polygon in the UI, allowing specific traffic patterns
+like rush-hour scenarios to be easily simulated. A generator
 may also cover the entire map, uniformly distributing traffic everywhere.
@@
 zero initial speed.  Because behaviors must enforce speed limit along a lane 
-from the moment an agent enters the road, then if there is no agent within the
-worst-case distance (traveling distance at the speed limit, plus stopping
-distance from this speed) before the spawn point, then the new agent may enter.
+from the moment an agent enters a road, it is safe for a new agent to enter a road
+if there is no other agent within the worst-case distance (traveling distance at the
+speed limit, plus stopping distance from this speed) preceeding the spawn point.
 To avoid predicting which agents might enter the lane in the near future, there
@@
 an ordered queue of occupying agents. One simulation step happens atomically,
-eaning collision checking occurs after all agents have moved. If this was not
-done, false positives could occur when an agent steps before the agent in front
-of it. A collision is defined as the reversal of order of agents before and
-after a step.
+meaning collision checking occurs after all agents have moved. If this were not
+the case, false positives could occur when an agent steps before the agent in front
+of it. A collision is defined as the reversal of order of agents during a step.
 
@@
 vehicles at any point along two different turns could ever come into contact,
-then the turns are always in conflict, no matter where the agent specifically is
+then the turns are always in conflict, no matter where the agents actually are
 along that turn. The alternative of tiling intersections \cite{JAIR08-dresner}
@@
 Each agent has a \emph{behavior} governing it. When the agent travels past the end
-of a lane during its step, the behavior picks the turn to pursue. Each step,
+of a lane during its step, the behavior picks the turn to take. At each step,
 the behavior can make the agent react by performing one of two actions:
-disappearing from the map (only when the agent is at rest and is done with its
+disappearing from the map (when the agent is at rest and is done with its
 route) or setting an acceleration for the next step. Were lane changing or
 merging modeled, that would be another possible action.
+\commentm{What's the difference between lane changing and merging?}
 
-Currently, the primary behavior used by all agents is a generalized, baseline
+Currently, all agents use a primary behavior that is a generalized, baseline
 behavior that guarantees not to collide with another agent or enter an
 intersection at the wrong time. Since it picks the highest safe choice of
-acceleration at each step, it could be easily extended to mimic human drivers by
-traveling at some random slower acceleration or to optimize fuel efficiency
+acceleration at each step, it could easily be extended to mimic human drivers by
+traveling at some random lesser acceleration or to optimize fuel efficiency
 by tuning movement.
@@
 short-circuits at the first nearby agent and intersection requiring a stop,
-since it would be illegal to pass another anyway.
+since it would be illegal to drive further anyway.
 
@@
 turns. Currently two simple implementations are available. One statically plans
-a route to some goal using standard A* search \cite{astar}; the other performs a
+a route to some goal using standard A* search \cite{astar}. The other performs a
 drunken walk, lazily planning random choices at each intersection as the
-lookahead engine needs to know. Future experiments in dynamic replanning or
-hierarchial planning can be performed without modifying any code other than a
-creating new route implementation. Route policies could also interact with a
+lookahead engine requires, choosing the direction that minimizes the Euclidean
+distance to the goal with 75\% probability. Researchers can experiment with
+dynamic replanning or hierarchial planning without modifying any code other than
+creating a new route implementation. Route policies could also interact with a
 singleton overseer to gain some sort of global insight or with smart
@@
 a dependency graph that encodes the ``is following'' relation and finding
-cycles. It is not clear how to prevent or mend the deadlock without forcing
+cycles. It is not clear how to prevent or mend deadlock without forcing
 agents to re-route.
@@
 deterministically reproducible. A seed for the pseudo-random number generator,
-the input graph, and the generators' configuration determine the outcome of any
+the input graph, and the generators' configuration fully determine the outcome of any
 simulation. This determinism may break down when generators use worker threads,
@@
   seconds)}
+\commentm{I typically have 160k-170k agent steps per second on my machine.  The
+	highest I have seen is upwards of 300k, but that was because most agents had
+	gridlocked...}
 
@@
 gridlock.
+\commentm{Perhaps mention concurency plans (with TACC) to simulate arbitrarily large maps?}
 
@@
 \commentd{Er, so how do I finally end this?}
+\commentm{We still have a bit of passive voice.  How much is acceptable?  I think
+	it sounds fine, but I imagine an English professor may disagree.}
 

### 2012-03-07 papers/itsc_2012/draft2.tex
architecture diagram finalized

@@
 \usepackage[pdftex]{graphicx}
-% TODO remember we need to take out all the color
+% TODO remember we need to meet color compliance:
+% http://its.papercept.net/conferences/support/general.php
 \newcommand{\pix}[3]{
@@
 
+% Diagrma made with gliffy.com. architecture.gxml should enable editing.
 \pix{architecture.png}
     {An illustration of AORTA's components.}
-    {scale=0.2}
+    {scale=0.18}
 
@@
   \item The builder constructs turns between incoming and outgoing lanes at each
-        vertex. The builder uses an angle heuristic to choose between left, right, straight,
+        junction. The builder uses an angle heuristic to choose between left, right, straight,
         and U-turns. When several lanes all cross into fewer lanes,
@@
 driver may be appropriate for some experiments. Each step, the simulation
-performs the following:
+does the following:
 
 \begin{enumerate}
-  \item Introduce new agents into the map
-  \item Update the position and velocity of agents based on their choice of
+  \item Introduces new agents into the map
+  \item Updates the position and velocity of agents based on their choice of
   acceleration in the previous step
-  \item Check for collisions
-  \item Allow each agent to choose a reaction for the next step
+  \item Checks for collisions
+  \item Allows each agent to choose a reaction for the next step
 \end{enumerate}
@@
 the behavior can make the agent react by performing one of two actions:
-disappearing from the map (when the agent is at rest and is done with its
-route) or setting an acceleration for the next step. Were lane changing or
-merging modeled, that would be another possible action.
-\commentm{What's the difference between lane changing and merging?}
+disappearing from the map (when the agent is at rest and is done with its route)
+or setting an acceleration for the next step. Were lane changing modeled, that
+would be another possible action.
 
@@
 applications could experiment with dynamic contra-flow \cite{ITSC11-hausknecht}
-and micro-tolling \commentd{need a cite}. There will also be an effort to
-improve AORTA as a framework by including a lane-changing model and preventing
-gridlock.
+and micro-tolling \commentd{Dr. Stone, should we just not mention this yet since
+it's not well-published?}. There will also be an effort to improve AORTA as a
+framework by including a lane-changing model and preventing gridlock.
 \commentm{Perhaps mention concurency plans (with TACC) to simulate arbitrarily large maps?}

### 2012-03-08 papers/itsc_2012/draft2.tex
gridlock diagram
2 more citations and scale

@@
 
-One advantage of AORTA is that it simulates traffic on existing infrastructure rather than contrived,
-randomly generated maps. Utilizing OpenStreetMaps, a community project offering
-user-editable, free XML-based maps, enables this. To run a simulation on a map
-from OSM, there is a one-time conversion process to transform the map into
-AORTA's format, which is also encoded with XML.
+One advantage of AORTA is that it simulates traffic on existing infrastructure
+rather than contrived, randomly generated maps. Utilizing OpenStreetMaps, a
+community project offering user-editable, free XML-based maps, enables this. To
+run a simulation on a map from OSM, there is a one-time conversion process to
+transform the map into AORTA's format, which is also encoded with XML.
 
@@
 achievable at the end of the next step and the subsequent stopping distance at
-this speed, the behavior can scan sufficiently far ahead along several
-traversables to locate agents to avoid and intersections to query. The analysis
-short-circuits at the first nearby agent and intersection requiring a stop,
-since it would be illegal to drive further anyway.
+this speed, the behavior's lookahead engine can scan sufficiently far ahead
+along several traversables to locate agents to avoid and intersections to query.
+The analysis short-circuits at the first nearby agent and intersection requiring
+a stop, since it would be illegal to drive further anyway.
 
@@
 
+\pix{gridlock.png}{Each of the red agents wants to turn left, so none of them
+can. The traffic jam will slowly build up around this intersection.}{scale=0.25}
+
 However, with very high volumes of traffic, it is possible for waiting agents to
@@
 
-\subsection{Determinism and other simulation properties}
+\subsection{Determinism and scale}
 
@@
 
-\commentd{give some informal or formal (with numbers) idea of scale / how many
-agents we can simulate in different map sizes} \commentpk{I would say that the the last point is fairly important}
-
-\commentd{http://sumo.sourceforge.net/pdf/mieec1.pdf gives their measure as
-  agent updates / second. This is indep of dt and num of agents, assuming
-  nothing grows nonlinearly with more agents. our initial result on my box is
-  about 150,000 moves / second. so with 500 agents, runs 30x faster than
-  reality. with 5,000, runs about 3x faster. with 50,000 (yes, this works!),
-  runs at 1/3 the speed of reality (it takes 1 real second to simulate 0.3
-  seconds)}
-\commentm{I typically have 160k-170k agent steps per second on my machine.  The
-	highest I have seen is upwards of 300k, but that was because most agents had
-	gridlocked...}
+The speed of simulation depends on the step duration $dt$, the map size, and the
+number of agents. A measure independent of these factors is the number of agent
+steps per second. On a $2.4$ GHz machine, this can be observed to average at
+about $150,000$, a number comparable to existing simulators \cite{SUMOthesis}.
+Further optimizations are planned.
 
@@
 If OSM could encode this, the performance of intersection policies such as
-traffic signals would vastly improve in those regions.
+traffic signals would vastly improve in those regions. Other simulators have
+identified similar issues before \cite{Krajzewicz_Hertkorn_Ringel_Wagner_2005}.
 

### 2012-03-09 papers/itsc_2012/final.tex
optimizing agent creation

--- 2012-03-08 papers/itsc_2012/draft2.tex
@@
 \label{sec:related_work}
+
+% just draw differences between us and them
 
@@
 
+mention all the fixed nuances, like pause time at stop sign
+
+  stop sign: FCFS, judged by stopping for the right time within the right dist of
+  the end of the intersection.
+
+  traffic signal: duration matters, turns grouped together, how timings are
+  propagated.
+
+  autonomous reservation: the greedy approach is a bit wretched. either time
+  out, or somehow count how many people are on lanes in the queued groups.
+  ideally we'd even recurse and factor in
+
 \commentd{upgrading stop signs to signals, optimizing timing?}
@@
 applications could experiment with dynamic contra-flow \cite{ITSC11-hausknecht}
-and micro-tolling \commentd{Dr. Stone, should we just not mention this yet since
-it's not well-published?}. There will also be an effort to improve AORTA as a
-framework by including a lane-changing model and preventing gridlock.
-\commentm{Perhaps mention concurency plans (with TACC) to simulate arbitrarily large maps?}
+or intelligent routing that avoids stopping. There will also be an effort to
+improve AORTA as a framework by including a lane-changing model and preventing
+gridlock. \commentm{Perhaps mention concurency plans (with TACC) to simulate
+arbitrarily large maps?}
 
@@
 
-\commentd{Er, these don't look like they're alphabetically sorted.}
+\commentd{Er, these don't look like they're alphabetically sorted. And why is
+the Google Maps entry so stretched?}
 

### 2012-03-09 papers/itsc_2012/generate.sh

@@ -3,7 +3,7 @@
 if [[ $# -eq 1 ]]; then
   INPUT=$1
 else
-  INPUT=draft2.tex
+  INPUT=final.tex
 fi
 
 # generate graphviz diagrams

### 2012-03-09 papers/itsc_2012/make-diff.sh

@@ -1,7 +1,7 @@
 #!/bin/sh
 
-V1=draft1.tex
-V2=peter.tex
+V1=peter.tex
+V2=draft2.tex
 
 latexdiff $V1 $V2 > diff.tex
 

### 2012-03-09 utexas/Util.scala

@@ -84,10 +84,11 @@ object Util {
   }
 }
 
+// use nanoseconds for both; it's nice and sensitive
 class Timer(msg: String) {
-  val start = System.currentTimeMillis
+  val start = System.nanoTime
 
-  def so_far = (System.currentTimeMillis - start) / 1000.0
+  def so_far = (System.nanoTime - start) / 1000000000.0
 
   def stop = {
     Util.log("\"" + msg + "\": " + so_far + "s")
@@ -95,7 +96,6 @@ class Timer(msg: String) {
 }
 
 class Stopwatch() {
-  // accumulation is more sensitive; use nanoseconds
   var from: Long = System.nanoTime
   var seconds: Double = 0.0
 

### 2012-03-09 utexas/sim/Generator.scala

@@ -13,10 +13,12 @@ object Generator {
   def shutdown = worker_pool.shutdown
 }
 
-abstract class Generator(sim: Simulation, desired_starts: List[Edge], val end_candidates: List[Edge])
+abstract class Generator(sim: Simulation, desired_starts: List[Edge], ends: List[Edge])
 {
-  // Prune the desired_starts for road type and length
-  val start_candidates = desired_starts.filter(e => sim.queues(e).ok_to_spawn)
+  // Prune the desired_starts for road type and length. Arrays have random
+  // access.
+  val start_candidates = desired_starts.filter(e => sim.queues(e).ok_to_spawn).toArray
+  val end_candidates = ends.toArray
 
   // there may be no task scheduled
   protected var pending: List[(Agent, Option[FutureTask[List[Traversable]]])] = Nil
@@ -39,11 +41,11 @@ abstract class Generator(sim: Simulation, desired_starts: List[Edge], val end_ca
       case Some(task) => {
         val delayed = new FutureTask[List[Traversable]](task)
         Generator.worker_pool.execute(delayed)
-        pending :+= (a, Some(delayed))
+        pending = (a, Some(delayed)) :: pending
       }
       case _ => {
         // don't schedule anything
-        pending :+= (a, None)
+        pending = (a, None) :: pending
       }
     }
   }

### 2012-03-09 utexas/Util.scala
diagnosing epsilon issue

@@ -128,9 +128,11 @@ object cfg {
     ("lane_width",      0.5, "Width of a lane in meters", 0.01, 0.1),
     ("zoom_threshold",  5.0, "How close to zoom in before drawing details",    1.0, 15.0),
     ("epsilon",     0.00001, "What do we take as zero due to FP imprecision?", 0.0,  1.0),
-    ("dt_s",            0.1, "The only dt in seconds an agent can experience", 0.1,  3.0),
+    ("dt_s",            1.0, "The only dt in seconds an agent can experience", 0.1,  3.0),
     // account for crosswalks, vehicle length...
     ("end_threshold",   5.0, "The end of a traversable is its length - this",  0.1,  3.0),
+    // when we don't have end_threshold left, go all the distance - this.
+    ("end_epsilon",   0.01, "The end of a traversable for short edges",  0.1,  3.0),
     // this kind of gives dimension to cars, actually
     ("follow_dist",     20.0, "Even if stopped, don't get closer than this", 0.1, 1.0),
     ("max_accel",       2.7, "A car's physical capability",                  0.5, 5.0),
@@ -140,7 +142,7 @@ object cfg {
 
   val ints = List(
     ("max_lanes",  20,  "The max total lanes a road could have", 1, 30),
-    ("army_size",  500, "How many agents to spawn by default", 1, 10000),
+    ("army_size",  5000, "How many agents to spawn by default", 1, 10000),
     ("signal_duration",  120, "How long for a traffic signal to go through all of its cycles", 4, 1200)
   ) map {c => c._1 -> new Int_Cfgable(c._2, c._3, c._4, c._5)} toMap
 
@@ -158,6 +160,7 @@ object cfg {
   def zoom_threshold  = doubles("zoom_threshold").value
   def epsilon         = doubles("epsilon").value
   def end_threshold   = doubles("end_threshold").value
+  def end_epsilon   = doubles("end_epsilon").value
   def follow_dist     = doubles("follow_dist").value
   def max_accel       = doubles("max_accel").value
   def pause_at_stop   = doubles("pause_at_stop").value

### 2012-03-09 utexas/map/Ward.scala

@@ -172,7 +172,7 @@ object Ward {
 
     def consider(ls: Set[Road]) = {
       // & is intersection, if unclear
-      if ((ls & major.toSet).size == 0) {
+      if ((ls & major.toSet).isEmpty) {
         for (r <- ls if !set(r)) {
           stack.push(r)
         }

### 2012-03-09 utexas/map/make/Pass3.scala

@@ -122,7 +122,7 @@ class Pass3(old_graph: PreGraph2) {
         val neg_lanes = r.neg_lanes.filter(l => !doomed_edges(l))
 
         // We can completely nix this road otherwise
-        if (pos_lanes.size != 0 || neg_lanes.size != 0) {
+        if ((!pos_lanes.isEmpty) || (!neg_lanes.isEmpty)) {
           // This will end up looking weird (gaps in roads), but just get rid of
           // the bad lanes. Gotta clean up lane numbers right now, IDs later.
           r.pos_lanes.clear
@@ -147,7 +147,7 @@ class Pass3(old_graph: PreGraph2) {
       for (v <- graph.vertices) {
         val turns = v.turns.filter(t => !doomed_turns(t))
         // If there are no turns left, nix this vertex completely
-        if (turns.size != 0) {
+        if (!turns.isEmpty) {
           v.turns.clear
           v.turns ++= turns
           keep_verts += v

### 2012-03-09 utexas/sim/Behaviors.scala

@@ -249,11 +249,15 @@ class RouteFollowingBehavior(a: Agent, route: Route) extends Behavior(a) {
           // we can. This handles a few stalemate cases with sequences of short
           // edges and possibly out-of-sync intersection policies.
           val last_stop = stop_at.get
+          assert(dist > cfg.end_epsilon)  // we should never get THIS close...
           val go_this_dist = last_stop match {
             // creep forward to halfway along the shorty edge.
-            case e: Edge if dist <= cfg.end_threshold => dist - (last_stop.length / 2.0)
+            case e: Edge if dist <= cfg.end_threshold => dist - cfg.end_epsilon
             // stop back appropriately.
             case _                                    => dist - cfg.end_threshold
+          }
+          if (a.id == 55) {
+            Util.log("... stopping " + go_this_dist)
           }
           accel_to_end(go_this_dist)
         }
@@ -264,6 +268,10 @@ class RouteFollowingBehavior(a: Agent, route: Route) extends Behavior(a) {
         case _       => Double.MaxValue
       }
       val a3 = a.accel_to_achieve(min_speed_limit)
+
+      if (a.id == 55) {
+        //Util.log("%.1f, %.1f, %.1f".format(a1, a2, a3))
+      }
 
       val conservative_accel = math.min(a1, math.min(a2, a3))
 
@@ -342,14 +350,25 @@ class RouteFollowingBehavior(a: Agent, route: Route) extends Behavior(a) {
                         else
                           math.max(0, try_speed)
 
-    val needed_accel = a.accel_to_achieve(desired_speed)
+    var needed_accel = a.accel_to_achieve(desired_speed)
 
     // TODO dumb epsilon bug again. fix this better.
-    val stop_speed = a.speed + (needed_accel * cfg.dt_s)
-    return if (stop_speed < 0)
-             needed_accel + 0.1
-           else
-             needed_accel
+    def stop_speed = a.speed + (needed_accel * cfg.dt_s)
+    def stop_dist = Util.dist_at_constant_accel(needed_accel, cfg.dt_s, a.speed) + a.stopping_distance(desired_speed)
+
+    while (stop_dist > how_far_away) {
+      Util.log(a.id + " slow to avoid too far... " + (how_far_away - stop_dist))
+      needed_accel -= cfg.epsilon
+      Util.log("now " + needed_accel + " compared to " + a.max_accel)
+    }
+    while (stop_speed < 0) {
[…]

### 2012-03-09 utexas/sim/Generator.scala

@@ -30,10 +30,10 @@ abstract class Generator(sim: Simulation, desired_starts: List[Edge], ends: List
 
   def add_specific_agent(start: Edge, end: Edge) = {
     // TODO how to decide?!
-    val route = new StaticRoute()
+    //val route = new StaticRoute()
     //val route = new DrunkenRoute()
     //val route = new DirectionalDrunkRoute()
-    //var route = new DrunkenExplorerRoute()
+    var route = new DrunkenExplorerRoute()
     val a = new Agent(sim.next_id, sim, start, sim.queues(start).safe_spawn_dist, route)
 
     // schedule whatever work the route needs done.

### 2012-03-09 utexas/sim/Intersection.scala

@@ -12,9 +12,9 @@ import utexas.{Util, cfg}
 // Reason about collisions from conflicting simultaneous turns.
 class Intersection(val v: Vertex) {
   //val policy: Policy = new NeverGoPolicy(this)
-  //val policy: Policy = new StopSignPolicy(this)
+  val policy: Policy = new StopSignPolicy(this)
   //val policy: Policy = new SignalCyclePolicy(this)
-  val policy: Policy = new ReservationPolicy(this)
+  //val policy: Policy = new ReservationPolicy(this)
 
   override def toString = "Intersection(" + v + ")"
 

### 2012-03-09 utexas/sim/Queue.scala

@@ -30,7 +30,7 @@ class Queue(t: Traversable) {
     // this breaks badly.
 
     // Everything's fine.
-    if (agents.size == 0) {
+    if (agents.isEmpty) {
       return
     }
 
@@ -45,7 +45,7 @@ class Queue(t: Traversable) {
 
     // If anybody from the old crowd is still here, that means all new agents
     // must be at the end. It suffices to check this:
-    if (old_crowd.size > 0 && old_crowd.head != agents.head) {
+    if ((!old_crowd.isEmpty) && old_crowd.head != agents.head) {
       Util.log("!!! New agents on " + t + " aren't at the tail!")
     }
 

### 2012-03-09 utexas/sim/policies/ReservationPolicy.scala

@@ -25,7 +25,7 @@ class ReservationPolicy(intersection: Intersection) extends Policy(intersection)
     if (current_batch.all_done) {
       // Time for the next reservation! If there is none, then keep
       // current_batch because it's empty anyway.
-      if (reservations.size != 0) {
+      if (!reservations.isEmpty) {
         current_batch = reservations.head
         reservations = reservations.tail
       }
@@ -107,7 +107,7 @@ class TurnBatch() {
       // existing turn
       tickets.addBinding(t, a)
       return true
-    } else if (tickets.keys.filter(c => t.conflicts(c)).size == 0) {
+    } else if (tickets.keys.filter(c => t.conflicts(c)).isEmpty) {
       // new turn that doesn't conflict
       tickets.addBinding(t, a)
       return true

### 2012-03-10 papers/itsc_2012/final.tex
diagnosing epsilon issue
seeing if things are even possible...

@@
 
-\commentd{upgrading stop signs to signals, optimizing timing?}
+controlled factors: several maps, agent spawning pattern
+independent: every intersection has 1 of 3 main policies, each can have tweaks
+output: average/total/whatever wasted time and intersection thruput
+
+key: all of this code is separated out, easy to edit, 200 lines. even if results
+suck, thats fine, we'll work on em. different results in diff places is maybe
+even good! one size doesnt fit all, suggests we need a mix and some adapting to
+maps. future experiment could be upgrading few at a time.
 

### 2012-03-10 utexas/Util.scala

@@ -128,7 +128,7 @@ object cfg {
     ("lane_width",      0.5, "Width of a lane in meters", 0.01, 0.1),
     ("zoom_threshold",  5.0, "How close to zoom in before drawing details",    1.0, 15.0),
     ("epsilon",     0.00001, "What do we take as zero due to FP imprecision?", 0.0,  1.0),
-    ("dt_s",            0.1, "The only dt in seconds an agent can experience", 0.1,  3.0),
+    ("dt_s",            1.0, "The only dt in seconds an agent can experience", 0.1,  3.0),
     // account for crosswalks, vehicle length...
     ("end_threshold",   5.0, "The end of a traversable is its length - this",  0.1,  3.0),
     // this kind of gives dimension to cars, actually
@@ -140,7 +140,7 @@ object cfg {
 
   val ints = List(
     ("max_lanes",  20,  "The max total lanes a road could have", 1, 30),
-    ("army_size",  500, "How many agents to spawn by default", 1, 10000),
+    ("army_size",  5000, "How many agents to spawn by default", 1, 10000),
     ("signal_duration",  120, "How long for a traffic signal to go through all of its cycles", 4, 1200)
   ) map {c => c._1 -> new Int_Cfgable(c._2, c._3, c._4, c._5)} toMap
 

### 2012-03-10 utexas/sim/Behaviors.scala

@@ -265,6 +265,10 @@ class RouteFollowingBehavior(a: Agent, route: Route) extends Behavior(a) {
       }
       val a3 = a.accel_to_achieve(min_speed_limit)
 
+      if (a.id == 1472) {
+        Util.log("  ~ %.1f, %.1f, %.1f".format(a1, a2, a3))
+      }
+
       val conservative_accel = math.min(a1, math.min(a2, a3))
 
       // let the other intersections that are letting us go know that we
@@ -322,6 +326,46 @@ class RouteFollowingBehavior(a: Agent, route: Route) extends Behavior(a) {
   // This is based on Piyush's proof.
   // how_far_away already includes end_threshold, if appropriate.
   private def accel_to_end(how_far_away: Double): Double = {
+    if (a.id == 1472) {
+      Util.log(" -- cover " + how_far_away + " from speed " + a.speed)
+    }
+
+    if (how_far_away > 0) {
+      // See if it's even possible, or if we're doomed.
+      // In other words, if we slam on our breaks now, how far do we travel?
+      val stop_dist_now = Util.dist_at_constant_accel(-a.max_accel, cfg.dt_s, a.speed)
+      if (stop_dist_now >= how_far_away) {
+        Util.log("!!!!!!! " + a.id + " is doomed: " + (how_far_away - stop_dist_now))
+      }
+    }
+
+    // What accel do we feed dist_at_constant_accel that makes it be equal to
+    // how_far_away?
+    // dist = (1/2)a(t^2) + (v_i)t
+    val direct_accel = 2.0 * (math.max(0.0, how_far_away) - (cfg.dt_s * a.speed)) / (cfg.dt_s * cfg.dt_s)
+    // make sure 1) this travels the correct amount
+      //assert(Util.dist_at_constant_accel(direct_accel, cfg.dt_s, a.speed) <= math.max(0.0, how_far_away))
+      val goal_dist = math.max(0.0, how_far_away)
+      // this does: (v_i)t + (1/2)a(t^2)
+      val would_dist = Util.dist_at_constant_accel(direct_accel, cfg.dt_s, a.speed)
+      if (would_dist > goal_dist) {
+        Util.log(would_dist + " exceeds " + goal_dist)
+      }
+    // make sure 2) this stops at speed 0
+      //assert(a.speed + (direct_accel * cfg.dt_s) >= 0.0)
+      val end_spd = a.speed + (direct_accel * cfg.dt_s)
+      if (end_spd < 0.0) {
+        Util.log("end speed " + end_spd)
[…]

### 2012-03-10 utexas/sim/Intersection.scala

@@ -12,9 +12,9 @@ import utexas.{Util, cfg}
 // Reason about collisions from conflicting simultaneous turns.
 class Intersection(val v: Vertex) {
   //val policy: Policy = new NeverGoPolicy(this)
-  //val policy: Policy = new StopSignPolicy(this)
+  val policy: Policy = new StopSignPolicy(this)
   //val policy: Policy = new SignalCyclePolicy(this)
-  val policy: Policy = new ReservationPolicy(this)
+  //val policy: Policy = new ReservationPolicy(this)
 
   override def toString = "Intersection(" + v + ")"
 
@@ -64,7 +64,7 @@ class Intersection(val v: Vertex) {
       Util.log("!!! " + a + " illegally entered intersection, going at " + a.speed)
       Util.log("  Incident was near " + t.from + " and " + t + " (vert " + t.from.to.id + ")")
       Util.log("  Origin lane length: " + t.from.length + "; time " + Agent.sim.tick)
-      Util.log("  Happened at " + Agent.sim.tick)
+      //policy.dump_info
       sys.exit
     }
   }

### 2012-03-10 utexas/Util.scala
reservation policy now preempts greedy cycles (haven't proven this helps yet)

@@ -141,7 +141,7 @@ object cfg {
   val ints = List(
     ("max_lanes",  20,  "The max total lanes a road could have", 1, 30),
     ("army_size",  500, "How many agents to spawn by default", 1, 10000),
-    ("signal_duration",  120, "How long for a traffic signal to go through all of its cycles", 4, 1200)
+    ("signal_duration",  60, "How long for a traffic signal to go through all of its cycles", 4, 1200)
   ) map {c => c._1 -> new Int_Cfgable(c._2, c._3, c._4, c._5)} toMap
 
   // TODO val colors = ... (I'm not kidding)
@@ -165,7 +165,7 @@ object cfg {
 
   def max_lanes       = ints("max_lanes").value
   def army_size       = ints("army_size").value
-  def sig_duration    = ints("signal_duration").value
+  def signal_duration    = ints("signal_duration").value
   
   def dt_s            = doubles("dt_s").value
 }

### 2012-03-10 utexas/sim/policies/GreenFlood.scala

@@ -39,7 +39,7 @@ class GreenFlood(sim: Simulation) {
     val turn_groups = Cycle.grouped_left_turn_sets(edge.to)
 
     // Schedule all cycles equally
-    val duration: Double = cfg.sig_duration / turn_groups.size.toDouble
+    val duration: Double = cfg.signal_duration / turn_groups.size.toDouble
 
     // Find the group including the turn our starting edge leads to
     val first_group = turn_groups.find(g => g.contains(edge.leads_to.head)).get

### 2012-03-10 utexas/sim/policies/ReservationPolicy.scala

@@ -8,7 +8,7 @@ import scala.collection.mutable.{HashSet => MutableHashSet}
 import utexas.sim.{Intersection, Policy, Agent}
 import utexas.map.Turn
 
-import utexas.Util
+import utexas.{Util, cfg}
 
 // FIFO based on request, batched by non-conflicting turns.  Possible liveness
 // violation, since new agents can pour into the current_turns, and the ones
@@ -17,17 +17,26 @@ import utexas.Util
 // coloring.
 class ReservationPolicy(intersection: Intersection) extends Policy(intersection) {
   var current_batch = new TurnBatch()
+  var lock_cur_batch = false    // because we're trying to preempt
   var reservations = List[TurnBatch]()
   // used to determine if it's an agent's first requent or not
   val current_agents = new MutableHashSet[Agent]()
+  // When did the first group of reservations start waiting?
+  private var others_started_waiting = -1.0
 
   def shift_batches() = {
     if (current_batch.all_done) {
+      lock_cur_batch = false
       // Time for the next reservation! If there is none, then keep
       // current_batch because it's empty anyway.
       if (reservations.size != 0) {
         current_batch = reservations.head
         reservations = reservations.tail
+        if (!reservations.isEmpty) {
+          // If we had a previous preempt notification on the way, just ignore
+          // it when it arrives
+          dont_be_greedy
+        }
       }
     }
   }
@@ -46,7 +55,7 @@ class ReservationPolicy(intersection: Intersection) extends Policy(intersection)
 
     if (first_req) {
       current_agents += a
-      if (current_batch.add_ticket(a, turn)) {
+      if ((!lock_cur_batch) && current_batch.add_ticket(a, turn)) {
         return true
       } else {
         // A conflicting turn. Add it to the reservations.
@@ -56,6 +65,11 @@ class ReservationPolicy(intersection: Intersection) extends Policy(intersection)
           // new batch!
           val batch = new TurnBatch()
           batch.add_ticket(a, turn)
+          // Make sure these guys don't wait too long. Schedule the event system
+          // to poke us after a certain delay.
+          if (reservations.isEmpty) {
+            dont_be_greedy
+          }
           reservations :+= batch
         }
 
@@ -92,6 +106,32 @@ class ReservationPolicy(intersection: Intersection) extends Policy(intersection)
       Util.log(t + ": " + current_batch.tickets(t))
     }
     Util.log_pop
+  }
+
+  def dont_be_greedy() = {
[…]

### 2012-03-11 utexas/map/make/Pass1.scala
scenario saving/loading so everything can be easelessly resimulated

@@ -47,7 +47,7 @@ class Pass1(fn: String) {
 
   def run(): PreGraph1 = {
     // fill out graph with roads and collect all info
-    match_events( new XMLEventReader( Source.fromFile(fn) ) )
+    match_events( new XMLEventReader(Source.fromFile(fn)))
 
     Util.log("Normalizing graph coordinates")
     graph.normalize()
@@ -209,7 +209,7 @@ class Pass1(fn: String) {
           }
         }
 
-        case _ => {}
+        case _ =>
       }
     })
     Util.log("")

### 2012-03-11 utexas/sim/Simulation.scala

@@ -2,8 +2,12 @@ package utexas.sim
 
 import scala.annotation.tailrec
 import scala.collection.immutable.{SortedSet, TreeSet}
-import scala.collection.mutable.{MutableList, PriorityQueue}
+import scala.collection.mutable.{MutableList, PriorityQueue, ListBuffer}
 import scala.collection.mutable.{HashSet => MutableSet}
+import java.io.FileWriter
+import scala.io.Source
+import scala.xml.MetaData
+import scala.xml.pull._
 
 import utexas.map.{Graph, Road, Edge, Vertex, Ward, Turn}
 import utexas.map.make.Reader
@@ -33,7 +37,7 @@ class Simulation(roads: List[Road], edges: List[Edge], vertices: List[Vertex],
   val intersections = vertices.map(v => v -> new Intersection(v)).toMap
   var agents: SortedSet[Agent] = new TreeSet[Agent]
   var ready_to_spawn: List[Agent] = Nil
-  val generators = new MutableSet[Generator]
+  private var generators: SortedSet[Generator] = new TreeSet[Generator]
   private var generator_count = 0   // just for informational/UI purposes
   private var id_cnt = -1
 
@@ -43,6 +47,11 @@ class Simulation(roads: List[Road], edges: List[Edge], vertices: List[Vertex],
   def next_id(): Int = {
     id_cnt += 1
     return id_cnt
+  }
+
+  def add_gen(g: Generator) = {
+    generators += g
+    Simulation.generators += g
   }
 
   // Be sure to call pre_step at least once to poke all the generators
@@ -62,7 +71,7 @@ class Simulation(roads: List[Road], edges: List[Edge], vertices: List[Vertex],
 
   // Just a convenience method.
   def spawn_army(total: Int) = {
-    generators += new FixedSizeGenerator(this, edges, edges, total)
+    add_gen(new FixedSizeGenerator(this, edges, edges, total))
   }
 
   def wait_for_all_generators() = {
@@ -92,7 +101,16 @@ class Simulation(roads: List[Road], edges: List[Edge], vertices: List[Vertex],
 
   // Returns true if at least one generator is active
   def pre_step(): Boolean = {
-    // First, give generators a chance to introduce more agents into the system
+    // First, fire any scheduled callbacks (usually intersections or
+    // overseers)
+    // (It's good to do this first because they might create generators as part
+    // of resimulation)
+    while (!events.isEmpty && tick >= events.head.at) {
[…]

### 2012-03-11 utexas/ui/MapCanvas.scala

@@ -727,12 +727,12 @@ class MapCanvas(sim: Simulation) extends ScrollingCanvas {
                 // Fixed
                 prompt_int("How many agents?") match {
                   case Some(num) => {
-                    sim.generators += new FixedSizeGenerator(
+                    sim.add_gen(new FixedSizeGenerator(
                       sim,
                       polygon_roads1.toList.flatMap(r => r.all_lanes),
                       polygon_roads2.toList.flatMap(r => r.all_lanes),
                       num
-                    )
+                    ))
                   }
                   case _ =>
                 }
@@ -741,12 +741,12 @@ class MapCanvas(sim: Simulation) extends ScrollingCanvas {
                 // Continuous
                 prompt_double("How often (in simulation-time seconds) do you want one new agent?") match {
                   case Some(time) => {
-                    sim.generators += new ContinuousGenerator(
+                    sim.add_gen(new ContinuousGenerator(
                       sim,
                       polygon_roads1.toList.flatMap(r => r.all_lanes),
                       polygon_roads2.toList.flatMap(r => r.all_lanes),
                       time
-                    )
+                    ))
                   }
                   case _ =>
                 }

### 2012-03-11 utexas/ui/Viewer.scala

@@ -2,6 +2,7 @@ package utexas.ui
 
 import swing._  // TODO figure out exactly what
 import java.awt.Color
+import swing.Dialog
 
 import utexas.sim.{Simulation, Headless}
 import utexas.Util
@@ -40,6 +41,11 @@ object Viewer extends SimpleSwingApplication {
       contents += new Menu("File") {
         contents += new MenuItem(Action("Configuration") {
           popup_config
+        })
+        contents += new MenuItem(Action("Save scenario for later resimulation") {
+          val fn = "resim_log"
+          Simulation.save_log(fn)
+          Dialog.showMessage(message = "Scenario saved in '" + fn + "'")
         })
         contents += new Separator
         contents += new MenuItem(Action("Quit") {

### 2012-03-13 papers/itsc_2012/final.tex
reservation policy now preempts greedy cycles (haven't proven this helps yet)
scenario saving/loading so everything can be easelessly resimulated
amping up results script to make comparisons

@@
 still permitting imperative style. Parts of AORTA, including extensions and
-clients, can easily be written in Java or Scala. The software, designed for
-extensibility, has a documented API, is
-open-source\footnote{http://code.google.com/p/road-rage}, and contains all of
-the code for the experiments described in Section \ref{sec:results}.
+clients, can easily be written in Java or Scala. The software is
+open-source\footnote{http://code.google.com/p/road-rage} and easily extensible.
+The stop sign policy used in Section \ref{sec:results}'s experiments is
+implemented in just $70$ lines of code.
 
@@
 
+% TODO in fact, this is caused by lack of lane changing!!!
+% but that's scary... bad routes really do cause it.
+
 However, with very high volumes of traffic, it is possible for waiting agents to
@@
 about $150,000$, a number comparable to existing simulators \cite{SUMOthesis}.
-Further optimizations are planned.
+When $dt = 0.1$ seconds, this means $15,000$ agents can be comfortably simulated
+in real-time.  Further optimizations are planned.
+
+\commentd{if i have a day after the paper's done, I can make this number
+better.}
 
@@
 
-  autonomous reservation: the greedy approach is a bit wretched. either time
-  out, or somehow count how many people are on lanes in the queued groups.
-  ideally we'd even recurse and factor in
+  autonomous reservation: the greedy approach has a timeout, but ideally we'd
+  ount how many people are on lanes in the queued groups.  and recurse and
+  factor in all agents "dependent" on something
 
@@
 output: average/total/whatever wasted time and intersection thruput
+
+map:
+
+contunuous gen, 1 new /sec. so stabe agent num is testament to policy.
+
+stop sign: stable around 800
+signal: noticed one side can be starved for a long time. even when it's their
+        cycle, theres no room in the first guy's dest lane, so they all waste the cycle.
+        stable around 1200.
+reservation: stable around 400. nobody's ever stopping.
 

### 2012-03-13 process_stats.py

@@ -1,69 +1,97 @@
 #!/usr/bin/python2
 
+import sys
+import fileinput
 from collections import defaultdict
 import numpy as np
 import pylab
 
+if len(sys.argv) == 1:
+  print "Give me stats logs!"
+  sys.exit()
+
+experiments = list()
+# The first key for all of these is experiment.
 # s1
-lag_per_a = defaultdict(list)
-lag_per_i = defaultdict(list)
-all_lags = list()
+lag_per_a = defaultdict(lambda: defaultdict(list))
+lag_per_i = defaultdict(lambda: defaultdict(list))
+all_lags = defaultdict(list)
 # s2
-avg_speed = list()
+avg_speed = defaultdict(list)
 # s3
-thruput_per_i = defaultdict(list)
+thruput_per_i = defaultdict(lambda: defaultdict(list))
 # s4
 # TODO find a clever way to pass in list of pairs, unpack, and wrap in a
 # subroutine
-active_per_time_x = list()
-active_per_time_y = list()
+active_per_time_x = defaultdict(list)
+active_per_time_y = defaultdict(list)
 
-input = open("stats_log")
-for line in input:
-  if line.startswith("s1"):
+# Grab input from everywhere first
+exp = ""    # what experiment?
+for line in fileinput.input():
+  if fileinput.isfirstline():
+    exp = line.strip()
+    experiments.append(exp)
+  elif line.startswith("s1"):
     nil, a, i, lag, time = line.strip().split(" ")
     lag = float(lag)
-    lag_per_a[a].append(lag)
-    lag_per_i[i].append(lag)
-    all_lags.append(lag)
+    lag_per_a[exp][a].append(lag)
+    lag_per_i[exp][i].append(lag)
+    all_lags[exp].append(lag)
   elif line.startswith("s2"):
     nil, a, time, dist = line.strip().split(" ")
-    avg_speed.append(float(dist) / float(time))
+    avg_speed[exp].append(float(dist) / float(time))
   elif line.startswith("s3"):
     nil, i, requests, entered, time = line.strip().split(" ")
-    thruput_per_i[i].append(float(entered) / float(requests))
+    thruput_per_i[exp][i].append(float(entered) / float(requests))
   elif line.startswith("s4"):
     nil, time, cnt = line.strip().split(" ")
-    active_per_time_x.append(int(time))
-    active_per_time_y.append(int(cnt))
-input.close()
[…]

### 2012-03-13 utexas/Util.scala

@@ -213,11 +213,19 @@ object Stats {
   var use_log = false
   var use_print = false
 
+  def experiment_name(name: String) = {
+    if (use_log) {
+      assert(log == null)
+      log = new FileWriter("stats_log")   // TODO use right buffering
+      log.write(name + "\n")
+    }
+    if (use_print) {
+      Util.log("Experiment name: " + name)
+    }
+  }
+
   def record(item: Measurement) = {
     if (use_log) {
-      if (log == null) {
-        log = new FileWriter("stats_log")   // TODO use right buffering
-      }
       log.write(item.toString + "\n")
     }
     if (use_print) {

### 2012-03-13 utexas/sim/Headless.scala

@@ -15,6 +15,7 @@ object Headless {
     var rng = System.currentTimeMillis
     var diff_rng = false
     var load_scenario = ""
+    var exp_name = "Experiment " + rng
 
     if (args.size % 2 != 0) {
       Util.log("Command-line parameters must be pairs of key => value")
@@ -28,9 +29,12 @@ object Headless {
         case "--log_stats"   => { Stats.use_log = value == "1" }
         case "--run_for"     => { run_for = value.toDouble }
         case "--scenario"    => { load_scenario = value }
+        case "--name"        => { exp_name = value }
         case _               => { Util.log("Unknown argument: " + key); sys.exit }
       }
     }
+
+    Stats.experiment_name(exp_name)
 
     if (load_scenario.isEmpty) {
       Util.init_rng(rng)

### 2012-03-13 utexas/ui/MapCanvas.scala

@@ -314,10 +314,14 @@ class MapCanvas(sim: Simulation) extends ScrollingCanvas {
   }
 
   def draw_agent(g2d: Graphics2D, a: Agent) = {
-    if (!agent_colors.contains(a)) {
-      agent_colors(a) = GeomFactory.rand_color
+    if (a.speed == 0.0) {
+      g2d.setColor(Color.RED)
+    } else {
+      if (!agent_colors.contains(a)) {
+        agent_colors(a) = GeomFactory.rand_color
+      }
+      g2d.setColor(agent_colors(a))
     }
-    g2d.setColor(agent_colors(a))
     if (zoomed_in) {
       // TODO cfg. just tweak these by sight.
       val vehicle_length = 0.5  // along the edge

### 2012-03-13 papers/itsc_2012/final.tex
we have a results section finally!

--- 2012-03-09 papers/itsc_2012/final.tex
@@
 
-% \long\def\commentp#1{}
-% \long\def\commentpk#1{}
-% \long\def\commentm#1{}
-% \long\def\commentd#1{}
+%\long\def\commentp#1{}
+%\long\def\commentpk#1{}
+%\long\def\commentm#1{}
+%\long\def\commentd#1{}
 
@@
 
-\pix{map_model.png}
-    {Roads contain several lanes, and intersections contain several turns
-     between lanes.}
-    {scale=0.5}
+%\pix{map_model.png}
+%    {Roads contain several lanes, and intersections contain several turns
+%     between lanes.}
+%    {scale=0.5}
 
@@
 occurs; no agent in the system will make progress unless an agent at the front
-decides to pick a different turn. At best, AORTA can detect this by maintaining
-a dependency graph that encodes the ``is following'' relation and finding
-cycles. It is not clear how to prevent or mend deadlock without forcing
-agents to re-route.
+decides to pick a different turn. AORTA can detect this by finding cycles in the
+graph of the ``is following'' relation.
 
-\commentd{is the above clear and an acceptable limitation? should i find other
-sources that have the same troubles?}
+In many observed cases, the gridlock is caused by AORTA's present lack of
+lane-changing support: agents loop around these intersections in order to enter
+a lane adjacent to their original lane. Since these routes are still legal, it
+is desirable to prevent or mend this without forcing different routes. The
+conditions under which every vehicle eventually may proceed are known.
+\cite{AAAI11-au}.
 
@@
 
-mention all the fixed nuances, like pause time at stop sign
+A particular focus of AORTA is intersection policies. To demonstrate how much
+policy affects agent performance, three policies are evaluated: stop sign,
+traffic signal, and autonomous reservation. These are compared with all other
+parameters fixed:
 
-  stop sign: FCFS, judged by stopping for the right time within the right dist of
-  the end of the intersection.
+\begin{enumerate}
+  \item the map (a $6$ by $7$ mile slice of downtown Austin, Texas
+        \footnote{bounded by the Mopac and Springdale Road horizontally,
+        and Oltorf Street and Koenig Lane vertically})
+  \item $dt = 0.1$ seconds
+  \item the agents spawned and their routes -- a continuous generator creates
+        one new agent every $1$ simulation second, with a uniformly random
+        starting position and goal
+  \item the duration of the experiment: one hour
+\end{enumerate}
 
-  traffic signal: duration matters, turns grouped together, how timings are
-  propagated.
+\subsection{Experiment parameters}
 
-  autonomous reservation: the greedy approach is a bit wretched. either time
-  out, or somehow count how many people are on lanes in the queued groups.
-  ideally we'd even recurse and factor in
+Each policy has parameters, so the configuration used is relevant. Stop signs
+accept agents FCFS, refusing agents until they have idled at $0$ speed for $1$
+second. This delay approximates human reaction time. The traffic signal policy
+requires timings and groups of simultaneous turns for each intersection. The
+configuration tested here uses a breadth-first search from one major road's
+lanes to try to synchronize green signals for the agents near that lane to not
+stop for several intersections. This heuristic needs improvement in how it
+handles the remaining groups of turns that are scheduled after the first
+sequence. Finally, the autonomous reservation policy allows new agents to join
+existing groups of agents if their turn is compatible with the others. To
+prevent one heavy direction of traffic from hogging the intersection
+indefinitely, a timeout preempts and cycles through reservations. Ideally this
+should be based on the relative numbers of agents depending on their
+reservation.
 
-\commentd{upgrading stop signs to signals, optimizing timing?}
+\subsection{Results}
+
+\pix{avg_lag_agent_atx.png}
+    {The distribution of average delay reveals that traffic signals can cause the most lag.}
+    {scale=0.35}
+
+If an intersection immediately permits an agent to proceed and grants other
+agents in the same lane similar access (because this new agent must still follow
+these others), then the new agent may simply proceed at full speed limit without
+pausing. This gives the least possible delay spent along some lane, so any extra
+time is defined as lag and is used to measure intersection performance. On
+average, reservation policies cause less than $1$ second of delay at each
+intersection, stop signs cause about $7$ seconds lag, and traffic signals cause
+$16$ seconds.
+
+\pix{agent_cnt_atx.png}
+    {Autonomous reservations outperform stop signs, and traffic signal gridlocked.}
+    {scale=0.35}
+
+The number of agents changes because the continuous generator introduces a new
+agent every second. This count reaches a steady state when the generator
+introduces an agent at the same rate as older agents finish their route. As one
+measure of intersection policy efficiency, it reveals that autonomous
+reservations are twice as efficient as the competition, with a steady-state of
+about $400$ agents. The count for traffic signals continues to increase because
+gridlock occurred in one segment of the map, causing agents to not finish their
+route. Unexpected starvation was also observed: although the signals cycle
+through turns regularly, agents cannot proceed when the previous cycle caused a
+destination lane to fill and become blocked.
+
+Each policy has its trade-offs in terms of performance and physical hardware
+cost, suggesting a mixed approach, perhaps with intersection policies adapting
+as traffic conditions shift. Researchers could conduct this experiment and many
+others without needing to understand most of AORTA's code-base. The largest
+intersection policy, traffic signals with the flooding heuristic for timing
+assignment, is only $400$ lines of Scala.
 

### 2012-03-13 utexas/ui/MapCanvas.scala

@@ -314,6 +314,7 @@ class MapCanvas(sim: Simulation) extends ScrollingCanvas {
   }
 
   def draw_agent(g2d: Graphics2D, a: Agent) = {
+    // TODO and idle time is at least a second
     if (a.speed == 0.0) {
       g2d.setColor(Color.RED)
     } else {

### 2012-03-13 papers/itsc_2012/final.tex
added a bunch of comments to section III and IV

@@
 
-\commentd{consider re-grouping sections... maybe this is all 'methodology'?}
+\commentd{consider re-grouping sections... maybe this is all 'methodology'?} \commentpk{I think it is fine as it is}
 
@@
 AORTA's implementation uses Scala, a language implemented on the Java Virtual
-Machine. Scala provides the advantage of functional programming constructs while
-still permitting imperative style. Parts of AORTA, including extensions and
-clients, can easily be written in Java or Scala. The software is
-open-source\footnote{http://code.google.com/p/road-rage} and easily extensible.
-The stop sign policy used in Section \ref{sec:results}'s experiments is
-implemented in just $70$ lines of code.
+Machine \commentpk{cite a link to Scala here}. Scala provides the advantage of
+functional programming constructs while still permitting imperative style.
+Parts of AORTA, including extensions and clients, can easily be written in Java
+or Scala. The software is
+open-source\footnote{http://code.google.com/p/road-rage} and easily extensible
+\commentpk{maybe "Extensions and clients built on top of AORTA can be written
+in either Java or Scala, as the software is
+open-source\footnote{\commentpk{needs more description}
+http://code.google.com/p/road-rage} and easily extensible}. \commentpk {For
+instance, the} The stop sign policy used in Section \ref{sec:results}'s
+experiments \commentpk{Sec VI's experiments is awk, maybe "used in experiments
+described in Sec VI"} is implemented in just $70$ \commentpk{remove \$\$ in
+latex code, here and elsewhere} lines of code.
+
+\commentpk{need to reference the architecture figure here. the text in the
+figure also needs to be bigger}
 
@@
 One advantage of AORTA is that it simulates traffic on existing infrastructure
-rather than contrived, randomly generated maps. Utilizing OpenStreetMaps, a
-community project offering user-editable, free XML-based maps, enables this. To
-run a simulation on a map from OSM, there is a one-time conversion process to
-transform the map into AORTA's format, which is also encoded with XML.
+rather than contrived, randomly \commentpk{"contrived, randomly" is a bit
+weird, maybe "contrived or randomly"?} generated maps. Utilizing OpenStreetMaps
+\commentpk{remove trailing s in OpenStreetMaps}, a community project offering
+user-editable, free XML-based maps, enables this. To run a simulation on a map
+from OSM, there is a one-time conversion process to transform the map into
+AORTA's format, which is also encoded with XML. \commentpk{why convert it to
+another XML format? you might want to put an additional line outlining the
+purpose of this step}
 
@@
 
-The map also has a geometric interpretation, used for visualization and physical
-distances. A road contains a sequence of points giving the line segments of the
-center yellow line. These points could also be interpreted as control points for
-a curved road, but AORTA uses straight lines as an approximation. Lanes are
-parallel to the center line, some fixed width apart from one another. Turns are
-modeled with a single straight line connecting the end of one lane to the start
-of another (so right turns often have $0$ length).
+The map also has a geometric interpretation, used for visualization and
+\commentpk{measuring?} physical distances. A road contains a sequence of points
+giving the line segments of the center yellow line \commentpk{if you want to
+use "yellow", you need to reference a figure. I would recommend putting it this
+way "center line of the road (yellow lines in Fig. XX)"}. These points could
+also be interpreted as control points for a curved road, but AORTA uses
+straight lines as an approximation. Lanes are parallel to the center line, some
+fixed width apart from one another \commentpk{no need to mention that the roads
+could be interpreted as a spline, instead say "AORTA interprets roads as
+straight lines between these points, and projects parallel lanes out using a
+fixed width"}. Turns are modeled with a single straight line connecting the end
+of one lane to the start of another (so right turns often have $0$ length).
 
@@
   \item OSM encodes many paths besides driveable roads, so the builder first filters
-        these out. Next the builder marks GPS locations common to multiple roads, since
-        they implicitly indicate intersections.
+        these out.
+Next the builder \commentpk{you have not mentioned a builder before, maybe mention it in the last line of the intro to this section} marks GPS locations common to multiple roads, since
+        they implicitly indicate intersections \commentpk{do you guys use GPS locations or OSM unique identifiers, what about overpasses etc.?}.
   \item OSM \emph{ways} (sequences of GPS locations) include many
@@
         (unless the road is marked one-way). The builder guesses the number of lanes based on
-        OSM's ``road type'' tag or an explicit number of lanes, when that data is available.
+        OSM's ``road type'' \commentpk{If you have space, elaborate that you use heuristics on the road type attribute, and the 4 different road types are residential etc.} tag or an explicit number of lanes, when that data is available.
   \item The builder constructs turns between incoming and outgoing lanes at each
-        junction. The builder uses an angle heuristic to choose between left, right, straight,
-        and U-turns. When several lanes all cross into fewer lanes,
+        junction \commentpk{intersection, keep the terminology constant}. The builder uses an angle heuristic to choose between left, right, straight,
+        and U-turns \commentpk{preceding sentence unclear to me}. When several lanes all cross into fewer lanes,
         the builder forces merging of the rightmost lanes.  Although inaccurate,
         these patterns are the best compromise. At the time of writing, even sources
-        like Google Maps \cite{googleMaps} lack this information.
+        like Google Maps \cite{googleMaps} lack this information \commentpk{the last sentence is tricky.
+unless you have a source that states this, I would avoid this claim.
+Linking to google maps directly is not correct}.
   \item Because turns are heuristically placed, the graph is often

### 2012-03-13 papers/itsc_2012/final.tex
added some comments to Section V

@@
 the duration of a ``step'' to achieve a new position and velocity, and
-space-continuous, meaning roads are not tiled. The step duration $dt$ is fixed,
-enabling agents to reason about collision avoidance and satisfying other
-constraints. Setting this value to the average ``reaction time'' of a human
-driver may be appropriate for some experiments. Each step, the simulation
-does the following:
+space-continuous, meaning roads are not tiled \commentpk{recommend breaking up
+preceding sentence into 2. I lost track haflway}. The step duration $dt$ is
+fixed, enabling agents to reason about collision avoidance and satisfying other
+constraints \commentpk{preceding sentence is unclear}. Setting this value to
+the average ``reaction time'' of a human driver may be appropriate for some
+\commentpk{"human driver" instead of some?} experiments. Each step, the
+simulation does the following:
 
@@
   \item Checks for collisions
-  \item Allows each agent to choose a reaction for the next step
+  \item Allows each agent to choose a reaction \commentpk {should be action} for the next step
 \end{enumerate}
@@
 collisions on the same traversable provide the other constraints, but the
-agent's behavior is responsible for obeying these. There are three main components of the
-simulation, each configurable by extending built-in implementations of an
-interface: agent behaviors, routing strategies, and intersection policies.
+agent's behavior is responsible for obeying these \commentpk{if these "other"
+constraints are not enforced, clearly specify. Something like "Agent behaviors
+are responsible for following intersection policies as directed by an
+intersection, and are responsible for not colliding with other agents" probably
+sounds a bit better}. There are three main components of the simulation, each
+configurable by extending built-in implementations of an interface: agent
+behaviors, routing strategies, and intersection policies.
 
-Currently agents cannot switch lanes. AORTA allows merging lanes by
-treating the merge point as an intersection.
+Currently agents cannot switch lanes. AORTA allows merging lanes by treating
+the merge point as an intersection. \commentpk{i take it this is an orphan
+paragraph?}
+
+\commentpk{this section is a bit incoherent, and might need to be reorganized.
+should the constraints be mentioned here explicitly, or maybe just added to
+agent behaviors? What are you introducing in the next few subsections? I can
+see A-E working together as an explanation of the workings of the simulator,
+and would make more sense right after the enumeration above. Section G can be
+moved into the experiments section. Once you are done with this primary
+explanation, Section F should be talked about separately under something along
+the lines of current limitations of the simulator.}
 
@@
 \emph{generators} programatically or using the UI. A generator picks a random
-starting position and goal from two sets of roads. Users can select these roads 
-by mouse with a bounding polygon in the UI, allowing specific traffic patterns
-like rush-hour scenarios to be easily simulated. A generator
-may also cover the entire map, uniformly distributing traffic everywhere.
+starting position and goal from two sets of roads \commentpk{preceding sentence
+is ambiguous, maybe go with "The generator produces a desired amount of traffic
+by initializing agents from random locations inside a \textit{start} polygon to
+random destinations inside a \textit{goal} polygon, as demarcated by the
+user"}. Users can select these roads by mouse with a bounding polygon in the
+UI, allowing specific traffic patterns like rush-hour scenarios to be easily
+simulated.  A generator \commentpk{polygon} may also cover the entire map,
+uniformly distributing traffic everywhere.
 
@@
 performs it immediately (incurring a delay in simulation if it is running) or
-sends it to a pool of worker threads to process in the background.
+sends it to a pool of worker threads to process in the background \commentpk{if multiple threads are available?}.
 
@@
 zero initial speed.  Because behaviors must enforce speed limit along a lane
-from the moment an agent enters a road, it is safe for a new agent to enter a road
-if there is no other agent within the worst-case distance (traveling distance at the
-speed limit, plus stopping distance from this speed) preceeding the spawn point.
-To avoid predicting which agents might enter the lane in the near future, there
-is a buffer space at the beginning of the lane where a new agent may never
-spawn. This disqualifies some short lanes from being spawn points, but this is
-not a big limitation in practice.
+from the moment an agent enters a road, it is safe for a new agent to enter a
+road if there is no other agent within the worst-case distance (traveling
+distance at the speed limit, plus stopping distance from this speed) preceeding
+the spawn point.  To avoid predicting which agents might enter the lane in the
+near future, there is a buffer space at the beginning of the lane where a new
+agent may never spawn. This disqualifies some short lanes from being spawn
+points, but this is not a big limitation in practice. \commentpk{I understand
+your concern about explaining the entire methodology, but the last 2 lines are
+unnecessary and somewhat distracting}
 

### 2012-03-13 papers/itsc_2012/final.tex
referencing figures and reducing wasted vspace

@@
 % http://its.papercept.net/conferences/support/general.php
-\newcommand{\pix}[3]{
-  \begin{figure}[h]
-    \centering \includegraphics[#3]{#1}
-    \caption{#2}
-  \end{figure}
-}
 
@@
 \begin{abstract} 
-Autonomous vehicles have seen great advancements in recent years and
+Autonomous vehicles have seen great advancements in recent years, and
 such vehicles are now ever closer to being commercially available.
@@
 
-\pix{downtown_atx.png}{A simulation of downtown Austin, Texas.}{scale=0.25}
+\begin{figure}[h]
+  \centering \includegraphics[width=\linewidth]{downtown_atx.png}
+  \caption{A simulation of downtown Austin, Texas.}
+  \label{fig:ui_screenshot}
+  \vspace{-10pt}
+\end{figure}
+
+% TODO reference the screenshot
 
@@
 
-\commentd{consider re-grouping sections... maybe this is all 'methodology'?} \commentpk{I think it is fine as it is}
-
-% Diagrma made with gliffy.com. architecture.gxml should enable editing.
-\pix{architecture.png}
-    {An illustration of AORTA's components.}
-    {scale=0.18}
+% Diagram made with gliffy.com. architecture.gxml should enable editing.
+\begin{figure*}
+  \centering \includegraphics[scale=0.3]{architecture.png}
+  \caption{A summary of AORTA's components.}
+  \label{fig:arch}
+  \vspace{-20pt}
+\end{figure*}
 
@@
 interactively render the map and agents. A headless mode also enables running
-experiments without the overhead of visualization.
+experiments without the overhead of visualization. The interaction between
+modules is visualized in figure \ref{fig:arch}.
 
@@
 latex code, here and elsewhere} lines of code.
-
-\commentpk{need to reference the architecture figure here. the text in the
-figure also needs to be bigger}
 
@@
 then the turns are always in conflict, no matter where the agents actually are
-along that turn. The alternative of tiling intersections \cite{JAIR08-dresner}
-would provide more granularity, but it adds computational complexity and would
-be impractical for many of the complex intersection geometries inferred from
-OSM.
+along that turn. This is visualized in figure \ref{fig:conflicts}. The
+alternative of tiling intersections \cite{JAIR08-dresner} would provide more
+granularity, but it adds computational complexity and would be impractical for
+many of the complex intersection geometries inferred from OSM.
 
@@
 
-\pix{turn_conflicts.png}
-    {An agent cannot turn left while agents on the perpendicular road cross}
-    {scale=0.5}
+\begin{figure}[h]
+  \centering \includegraphics[scale=0.5]{turn_conflicts.png}
+  \caption{An agent cannot turn left while agents on the perpendicular road cross}
+  \label{fig:conflicts}
+  \vspace{-15pt}
+\end{figure}
 
@@
 
-\pix{gridlock.png}{Each of the red agents wants to turn left, so none of them
-can. The traffic jam will slowly build up around this intersection.}{scale=0.25}
-
-% TODO in fact, this is caused by lack of lane changing!!!
-% but that's scary... bad routes really do cause it.
-
 However, with very high volumes of traffic, it is possible for waiting agents to
@@
 occurs; no agent in the system will make progress unless an agent at the front
-decides to pick a different turn. AORTA can detect this by finding cycles in the
-graph of the ``is following'' relation.
+decides to pick a different turn. Figure \ref{fig:gridlock} demonstrates this.
+AORTA can detect this by finding cycles in the graph of the ``is following''
+relation.
+
+\begin{figure}[h]
+  \centering \includegraphics[scale=0.25]{gridlock.png}
+  \caption{Each of the red agents wants to turn left, so none of them
+           can. The traffic jam will slowly build up around this intersection.}
+  \label{fig:gridlock}
+  \vspace{-10pt}
+\end{figure}
 
@@
 prevent one heavy direction of traffic from hogging the intersection
-indefinitely, a timeout preempts and cycles through reservations. Ideally this
+indefinitely, a timeout preemptively cycles through reservations. Ideally this
 should be based on the relative numbers of agents depending on their
-reservation.
+group of reservations.
 
@@
 
-\pix{avg_lag_agent_atx.png}
-    {The distribution of average delay reveals that traffic signals can cause the most lag.}
-    {scale=0.35}
+\begin{figure}[h]
+  \centering \includegraphics[width=\linewidth]{avg_lag_agent_atx.png}
+  \caption{The distribution of average delay reveals that traffic signals can
+           cause the most lag.}
+  \label{fig:avg_lag}
+  \vspace{-10pt}
+\end{figure}
 
@@
 pausing. This gives the least possible delay spent along some lane, so any extra
-time is defined as lag and is used to measure intersection performance. On
-average, reservation policies cause less than $1$ second of delay at each
-intersection, stop signs cause about $7$ seconds lag, and traffic signals cause
-$16$ seconds.
+time is defined as lag and is used to measure intersection performance. Figure
+\ref{fig:avg_lag} graphs the average of this delay per agent. On average,
+reservation policies cause less than $1$ second of delay at each intersection,
+stop signs cause about $7$ seconds lag, and traffic signals cause $16$ seconds.
 
-\pix{agent_cnt_atx.png}
-    {Autonomous reservations outperform stop signs, and traffic signal gridlocked.}
-    {scale=0.35}
+\begin{figure}[h]
+  \centering \includegraphics[width=\linewidth]{agent_cnt_atx.png}
+  \caption{Autonomous reservations outperform stop signs, and traffic signal
+           gridlocked.}
+  \label{fig:agent_cnt}
+  \vspace{-10pt}
+\end{figure}
 
@@
 agent every second. This count reaches a steady state when the generator
-introduces an agent at the same rate as older agents finish their route. As one
-measure of intersection policy efficiency, it reveals that autonomous
-reservations are twice as efficient as the competition, with a steady-state of
-about $400$ agents. The count for traffic signals continues to increase because
+introduces an agent at the same rate as older agents finish their route. Figure
+\ref{agent_cnt} reveals that autonomous reservations let agents reach their
+destination twice as quickly as the competition, with a steady-state of about
+$400$ agents. The count for traffic signals continues to increase because
 gridlock occurred in one segment of the map, causing agents to not finish their

### 2012-03-13 process_stats.py

@@ -79,9 +79,9 @@ def average(dat):
   return [np.mean(ls) for ls in dat.values()]
 
 # Results from s1
-hgram(average, lag_per_a, "Average lag per agent", "agents")
-hgram(average, lag_per_i, "Average lag per intersection", "intersections")
-hgram(identity, all_lags, "Average lag", "events")
+hgram(average, lag_per_a, "Average lag per agent (seconds)", "agents")
+hgram(average, lag_per_i, "Average lag per intersection (seconds)", "intersections")
+hgram(identity, all_lags, "Average lag (seconds)", "events")
 # Results from s2
 hgram(identity, avg_speed, "Average speed (m/s) of each agent", "agents")
 # Results from s3

### 2012-03-14 papers/itsc_2012/final.tex
section 3 and 4 revised based on comments

@@
 AORTA's implementation uses Scala, a language implemented on the Java Virtual
-Machine \commentpk{cite a link to Scala here}. Scala provides the advantage of
-functional programming constructs while still permitting imperative style.
-Parts of AORTA, including extensions and clients, can easily be written in Java
-or Scala. The software is
-open-source\footnote{http://code.google.com/p/road-rage} and easily extensible
-\commentpk{maybe "Extensions and clients built on top of AORTA can be written
-in either Java or Scala, as the software is
-open-source\footnote{\commentpk{needs more description}
-http://code.google.com/p/road-rage} and easily extensible}. \commentpk {For
-instance, the} The stop sign policy used in Section \ref{sec:results}'s
-experiments \commentpk{Sec VI's experiments is awk, maybe "used in experiments
-described in Sec VI"} is implemented in just $70$ \commentpk{remove \$\$ in
-latex code, here and elsewhere} lines of code.
+Machine \cite{scala}. Scala provides the advantage of functional programming
+constructs while still permitting imperative style.  Extensions and clients
+built on top of AORTA can be written in either Java or Scala. The software is
+open-source \footnote{source code at http://code.google.com/p/road-rage} and
+easily extensible. For instance, the stop sign sign policy used in experiments
+described in section \ref{sec:result} is implemented in just 70 lines of code.
 
@@
 One advantage of AORTA is that it simulates traffic on existing infrastructure
-rather than contrived, randomly \commentpk{"contrived, randomly" is a bit
-weird, maybe "contrived or randomly"?} generated maps. Utilizing OpenStreetMaps
-\commentpk{remove trailing s in OpenStreetMaps}, a community project offering
-user-editable, free XML-based maps, enables this. To run a simulation on a map
-from OSM, there is a one-time conversion process to transform the map into
-AORTA's format, which is also encoded with XML. \commentpk{why convert it to
-another XML format? you might want to put an additional line outlining the
-purpose of this step}
+rather than randomly generated maps. Utilizing OpenStreetMap \cite{osm}, a
+community project offering user-editable, free XML-based maps, enables this. To
+run a simulation on a map from OSM, there is a one-time conversion process to
+transform the map into AORTA's format, which encodes information about lanes and
+turns not contained in the OSM source.
 
@@
 The map also has a geometric interpretation, used for visualization and
-\commentpk{measuring?} physical distances. A road contains a sequence of points
-giving the line segments of the center yellow line \commentpk{if you want to
-use "yellow", you need to reference a figure. I would recommend putting it this
-way "center line of the road (yellow lines in Fig. XX)"}. These points could
-also be interpreted as control points for a curved road, but AORTA uses
-straight lines as an approximation. Lanes are parallel to the center line, some
-fixed width apart from one another \commentpk{no need to mention that the roads
-could be interpreted as a spline, instead say "AORTA interprets roads as
-straight lines between these points, and projects parallel lanes out using a
-fixed width"}. Turns are modeled with a single straight line connecting the end
-of one lane to the start of another (so right turns often have $0$ length).
+measuring physical distances. A road contains a sequence of points, defining the
+center line that divides the road into two directions of travel. AORTA
+interprets these points as straight lines and projects parallel lanes out using
+a fixed width. Turns are modeled with a single straight line connecting the end
+of one lane to the start of another (so right turns often have 0 length).
 
@@
 
-Map construction is split into several passes, each feeding the next.
+The builder works by incrementally transforming OSM graphs into AORTA graphs,
+with each pass feeding the next.
 \begin{enumerate}
   \item OSM encodes many paths besides driveable roads, so the builder first filters
-        these out.
-Next the builder \commentpk{you have not mentioned a builder before, maybe mention it in the last line of the intro to this section} marks GPS locations common to multiple roads, since
-        they implicitly indicate intersections \commentpk{do you guys use GPS locations or OSM unique identifiers, what about overpasses etc.?}.
+        these out. Next the builder marks GPS nodes common to multiple roads, since
+        they implicitly indicate intersections. Overpasses do not share nodes
+        with the roads they cross \cite{osmOverpass}.
   \item OSM \emph{ways} (sequences of GPS locations) include many
@@
         segments of roads between exactly two intersections.
-  \item The builder multiplies undirected roads into directed lanes in each direction
-        (unless the road is marked one-way). The builder guesses the number of lanes based on
-        OSM's ``road type'' \commentpk{If you have space, elaborate that you use heuristics on the road type attribute, and the 4 different road types are residential etc.} tag or an explicit number of lanes, when that data is available.
+  \item The builder multiplies undirected roads into directed lanes in each
+        direction (unless the road is marked one-way). The builder guesses the
+        number of lanes based on OSM's ``road type'' tag or an explicit number of
+        lanes, when that data is available.
   \item The builder constructs turns between incoming and outgoing lanes at each
-        junction \commentpk{intersection, keep the terminology constant}. The builder uses an angle heuristic to choose between left, right, straight,
-        and U-turns \commentpk{preceding sentence unclear to me}. When several lanes all cross into fewer lanes,
-        the builder forces merging of the rightmost lanes.  Although inaccurate,
-        these patterns are the best compromise. At the time of writing, even sources
-        like Google Maps \cite{googleMaps} lack this information \commentpk{the last sentence is tricky.
-unless you have a source that states this, I would avoid this claim.
-Linking to google maps directly is not correct}.
-  \item Because turns are heuristically placed, the graph is often
-        disconnected. In order to ensure that an agent will be able to reach any point in the map,
-        the builder removes all but the largest strongly connected component of the digraph.
-        Disconnected lanes are spliced out of larger roads, leaving unappealing
-        geometric gaps, but correctly ensuring connectivity for corner-case free
-        pathfinding.
+        intersection. The angle between the two lanes determines whether the turn is
+        a left, right, straight, or U-turn. When several lanes all cross into fewer
+        lanes, the builder forces merging of the rightmost lanes. Although
+        inaccurate, these patterns are the best compromise with the available data.
+  \item Because turns are heuristically placed, the graph is often disconnected.
+        In order to ensure that an agent will be able to reach any point in the map,
+        the builder removes all but the largest strongly connected component of the
+        digraph.  Disconnected lanes are spliced out of larger roads, leaving
+        unappealing geometric gaps, but correctly ensuring connectivity for
+        corner-case free pathfinding.
   \item The builder serializes the graph in a simple XML format.
@@
 space-continuous, meaning roads are not tiled \commentpk{recommend breaking up
-preceding sentence into 2. I lost track haflway}. The step duration $dt$ is
+preceding sentence into 2. I lost track haflway}. The step duration dt is
 fixed, enabling agents to reason about collision avoidance and satisfying other
@@
 
-The speed of simulation depends on the step duration $dt$, the map size, and the
+The speed of simulation depends on the step duration dt, the map size, and the
 number of agents. A measure independent of these factors is the number of agent
-steps per second. On a $2.4$ GHz machine, this can be observed to average at
-about $150,000$, a number comparable to existing simulators \cite{SUMOthesis}.
-When $dt = 0.1$ seconds, this means $15,000$ agents can be comfortably simulated
+steps per second. On a 2.4 GHz machine, this can be observed to average at
+about 150,000, a number comparable to existing simulators \cite{SUMOthesis}.
+When dt = 0.1 seconds, this means 15,000 agents can be comfortably simulated
 in real-time.  Further optimizations are planned.
@@
 \begin{enumerate}
-  \item the map (a $6$ by $7$ mile slice of downtown Austin, Texas
+  \item the map (a 6 by 7 mile slice of downtown Austin, Texas
         \footnote{bounded by the Mopac and Springdale Road horizontally,
         and Oltorf Street and Koenig Lane vertically})
-  \item $dt = 0.1$ seconds
+  \item dt = 0.1 seconds
   \item the agents spawned and their routes -- a continuous generator creates
-        one new agent every $1$ simulation second, with a uniformly random
+        one new agent every 1 simulation second, with a uniformly random
         starting position and goal
@@
 Each policy has parameters, so the configuration used is relevant. Stop signs
-accept agents FCFS, refusing agents until they have idled at $0$ speed for $1$
+accept agents FCFS, refusing agents until they have idled at 0 speed for 1
 second. This delay approximates human reaction time. The traffic signal policy
@@
 \ref{fig:avg_lag} graphs the average of this delay per agent. On average,
-reservation policies cause less than $1$ second of delay at each intersection,
-stop signs cause about $7$ seconds lag, and traffic signals cause $16$ seconds.
+reservation policies cause less than 1 second of delay at each intersection,
+stop signs cause about 7 seconds lag, and traffic signals cause 16 seconds.
 
@@
 destination twice as quickly as the competition, with a steady-state of about
-$400$ agents. The count for traffic signals continues to increase because
+400 agents. The count for traffic signals continues to increase because
 gridlock occurred in one segment of the map, causing agents to not finish their
@@
 intersection policy, traffic signals with the flooding heuristic for timing
-assignment, is only $400$ lines of Scala.
+assignment, is only 400 lines of Scala.
 

### 2012-03-14 papers/itsc_2012/final.tex
splitting and reorganizing simulation section

@@
 easily extensible. For instance, the stop sign sign policy used in experiments
-described in section \ref{sec:result} is implemented in just 70 lines of code.
+described in section \ref{sec:results} is implemented in just 70 lines of code.
 
@@
 AORTA uses microscopic simulation, modeling individual drivers as agents.
-Simulation is discrete-time, meaning an agent accelerates at a fixed rate for
-the duration of a ``step'' to achieve a new position and velocity, and
-space-continuous, meaning roads are not tiled \commentpk{recommend breaking up
-preceding sentence into 2. I lost track haflway}. The step duration dt is
-fixed, enabling agents to reason about collision avoidance and satisfying other
-constraints \commentpk{preceding sentence is unclear}. Setting this value to
-the average ``reaction time'' of a human driver may be appropriate for some
-\commentpk{"human driver" instead of some?} experiments. Each step, the
-simulation does the following:
+Time is modeled discretely, meaning an agent accelerates at a fixed rate for
+the duration of a ``step'' to achieve a new position and velocity. Space is
+continuous, meaning agents occupy a moving interval of a lane, rather than one
+fixed tile of the road. Each step lasts the same fixed duration \textit{dt}.
+This lets agents reason about how far they will travel for a given choice of
+acceleration before being able to choose a different acceleration. Setting
+\textit{dt} to the ``reaction time'' of a human driver may be appropriate for
+experiments modeling humans.
+
+Currently agents cannot switch lanes. AORTA allows merging lanes by treating
+the merge point as an intersection. \commentpk{i take it this is an orphan
+paragraph?} \commentd{yeah, where should it belong?}
+
+Each step, the simulation does the following:
 
@@
   \item Checks for collisions
-  \item Allows each agent to choose a reaction \commentpk {should be action} for the next step
+  \item Allows each agent to choose an action for the next step
 \end{enumerate}
 
-Agents have a few enforced constraints -- they cannot exceed their physical
-acceleration capabilities, travel in reverse, or travel past the end of a
-traversable without appearing on the next. Intersection policies and avoiding
-collisions on the same traversable provide the other constraints, but the
-agent's behavior is responsible for obeying these \commentpk{if these "other"
-constraints are not enforced, clearly specify. Something like "Agent behaviors
-are responsible for following intersection policies as directed by an
-intersection, and are responsible for not colliding with other agents" probably
-sounds a bit better}. There are three main components of the simulation, each
-configurable by extending built-in implementations of an interface: agent
-behaviors, routing strategies, and intersection policies.
-
-Currently agents cannot switch lanes. AORTA allows merging lanes by treating
-the merge point as an intersection. \commentpk{i take it this is an orphan
-paragraph?}
-
-\commentpk{this section is a bit incoherent, and might need to be reorganized.
-should the constraints be mentioned here explicitly, or maybe just added to
-agent behaviors? What are you introducing in the next few subsections? I can
-see A-E working together as an explanation of the workings of the simulator,
-and would make more sense right after the enumeration above. Section G can be
-moved into the experiments section. Once you are done with this primary
-explanation, Section F should be talked about separately under something along
-the lines of current limitations of the simulator.}
+Each step (except 2) will be described below, followed by limitations of the
+simulator. Section \ref{sec:config} will describe how researchers may easily
+modify relevant portions of AORTA.
 
@@
 To introduce new agents into the simulation in real-time, users can create
-\emph{generators} programatically or using the UI. A generator picks a random
-starting position and goal from two sets of roads \commentpk{preceding sentence
-is ambiguous, maybe go with "The generator produces a desired amount of traffic
-by initializing agents from random locations inside a \textit{start} polygon to
-random destinations inside a \textit{goal} polygon, as demarcated by the
-user"}. Users can select these roads by mouse with a bounding polygon in the
-UI, allowing specific traffic patterns like rush-hour scenarios to be easily
-simulated.  A generator \commentpk{polygon} may also cover the entire map,
-uniformly distributing traffic everywhere.
+\emph{generators} programatically or using the UI. The generator produces the
+desired amount of traffic by initializing agents from random locations inside a
+\textit{start} polygon to random destinations within a \textit{goal} polygon.
+Users can draw these polyons by mouse in the UI, allowing specific traffic
+patterns like rush-hour scenarios to be easily simulated. A generator polygon
+may also cover the entire map, uniformly distributing traffic everywhere.
 
@@
 policy needs computation like pathfinding to operate, the generator either
-performs it immediately (incurring a delay in simulation if it is running) or
-sends it to a pool of worker threads to process in the background \commentpk{if multiple threads are available?}.
+performs it immediately (incurring a delay in simulation if it is running) or,
+if possible, sends it to a pool of worker threads to process in the background.
 
@@
 distance at the speed limit, plus stopping distance from this speed) preceeding
-the spawn point.  To avoid predicting which agents might enter the lane in the
-near future, there is a buffer space at the beginning of the lane where a new
-agent may never spawn. This disqualifies some short lanes from being spawn
-points, but this is not a big limitation in practice. \commentpk{I understand
-your concern about explaining the entire methodology, but the last 2 lines are
-unnecessary and somewhat distracting}
+the spawn point.% To avoid predicting which agents might enter the lane in the
+%near future, there is a buffer space at the beginning of the lane where a new
+%agent may never spawn. This disqualifies some short lanes from being spawn
+%points, but this is not a big limitation in practice. \commentpk{I understand
+%your concern about explaining the entire methodology, but the last 2 lines are
+%unnecessary and somewhat distracting}
 
@@
 \subsection{Agent Reactions}
+\label{sec:behaviors}
 
-Each agent has a \emph{behavior} governing it. When the agent travels past the end
-of a lane during its step, the behavior picks the turn to take. At each step,
-the behavior can make the agent react by performing one of two actions:
+Each agent has a \emph{behavior} governing it, responsible for obeying
+intersection policies and avoiding collisions. When the agent travels past the
+end of a lane during its step, the behavior picks the turn to take. At each
+step, the behavior can make the agent react by performing one of two actions:
 disappearing from the map (when the agent is at rest and is done with its route)
@@
 
-\commentd{one diagram probably covers lots of this}
-
 The default behavior maximizes acceleration while obeying speed limits,
@@
 a stop, since it would be illegal to drive further anyway.
+
+\subsection{Simulation Limitations: Gridlock}
+
+Agents maintain a safe following distance from the next agent. If a lane is
+filled, then an agent may be forced to stop in an intersection mid-turn and
+block other traffic. The default behavior prevents this by refusing to start a
+turn before the lookahead engine guarantees no agent will trigger a premature
+stop.
+
+However, with very high volumes of traffic, it is possible for waiting agents to
+accumulate and fill lanes to their capacity. When this happens in a circular
+manner so that the agent at the front of each lane is waiting to perform a turn
+into another full lane with the same such head agent, gridlock \cite{gridlock}
+occurs; no agent in the system will make progress unless an agent at the front
+decides to pick a different turn. Figure \ref{fig:gridlock} demonstrates this.
+AORTA can detect this by finding cycles in the graph of the ``is following''
+relation.
+
+\begin{figure}[h]
+  \centering \includegraphics[scale=0.25]{gridlock.png}
+  \caption{Each of the red agents wants to turn left, so none of them
+           can. The traffic jam will slowly build up around this intersection.}
+  \label{fig:gridlock}
+  \vspace{-10pt}
+\end{figure}
+
+In many observed cases, the gridlock is caused by AORTA's present lack of
+lane-changing support: agents loop around these intersections in order to enter
+a lane adjacent to their original lane. Since these routes are still legal, it
+is desirable to prevent or mend this without forcing different routes. The
+conditions under which every vehicle eventually may proceed are known.
+\cite{AAAI11-au}.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{CONFIGURABILITY}
+\label{sec:config}
+
+There are three main components of the simulation, each configurable by
+extending built-in implementations of an interface: agent behaviors, routing
+strategies, and intersection policies. Behaviors have been described already in
+section \ref{sec:behaviors}.
 
@@
 
-\subsection{Gridlock}
-
-Agents maintain a safe following distance from the next agent. If a lane is
-filled, then an agent may be forced to stop in an intersection mid-turn and
-block other traffic. The default behavior prevents this by refusing to start a
-turn before the lookahead engine guarantees no agent will trigger a premature
-stop.
-
-However, with very high volumes of traffic, it is possible for waiting agents to
-accumulate and fill lanes to their capacity. When this happens in a circular
-manner so that the agent at the front of each lane is waiting to perform a turn
-into another full lane with the same such head agent, gridlock \cite{gridlock}
-occurs; no agent in the system will make progress unless an agent at the front
-decides to pick a different turn. Figure \ref{fig:gridlock} demonstrates this.
-AORTA can detect this by finding cycles in the graph of the ``is following''
-relation.
-
-\begin{figure}[h]
-  \centering \includegraphics[scale=0.25]{gridlock.png}
-  \caption{Each of the red agents wants to turn left, so none of them
-           can. The traffic jam will slowly build up around this intersection.}
-  \label{fig:gridlock}
-  \vspace{-10pt}
-\end{figure}
-
-In many observed cases, the gridlock is caused by AORTA's present lack of
-lane-changing support: agents loop around these intersections in order to enter
-a lane adjacent to their original lane. Since these routes are still legal, it
-is desirable to prevent or mend this without forcing different routes. The
-conditions under which every vehicle eventually may proceed are known.
-\cite{AAAI11-au}.
-
-\subsection{Determinism and scale}
-
-AORTA is designed for controlled experiments. Consequently, simulations are
-deterministically reproducible. A seed for the pseudo-random number generator,
-the input graph, and the generators' configuration fully determine the outcome of any
-simulation. This determinism may break down when generators use worker threads,
-since routes may be computed in different orders, causing agents to enter the
-road at different times.
-
-\commentd{saving a ``scenario'' can soon be a feature}
-
-\commentd{and how will we solve this issue with threads and determinism?}
-
-The speed of simulation depends on the step duration dt, the map size, and the
-number of agents. A measure independent of these factors is the number of agent
-steps per second. On a 2.4 GHz machine, this can be observed to average at
-about 150,000, a number comparable to existing simulators \cite{SUMOthesis}.
-When dt = 0.1 seconds, this means 15,000 agents can be comfortably simulated
-in real-time.  Further optimizations are planned.
-
-\commentd{if i have a day after the paper's done, I can make this number
-better.}
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \end{enumerate}
+
+Each experiment is deterministically reproducible. A seed for the pseudo-random
+number generator, the input graph, and the generators' configuration fully
+determine the outcome of any simulation \footnote{This determinism may break
+  down when generators use worker threads, since routes may be computed in
+different orders, causing agents to enter the road at different times.}. After
+setting up a scenario in the UI, users can save this configuration for
+resimulation.
+
+The speed of simulation depends on the step duration dt, the map size, and the
+number of agents. A measure independent of these factors is the number of agent
+steps per second. On a 2.4 GHz machine, this can be observed to average at about
+150,000, a number comparable to existing simulators \cite{SUMOthesis}. When dt
+= 0.1 seconds, this means 15,000 agents can be comfortably simulated in
+real-time. Further optimizations are planned.
 
@@
 introduces an agent at the same rate as older agents finish their route. Figure
-\ref{agent_cnt} reveals that autonomous reservations let agents reach their
+\ref{fig:agent_cnt} reveals that autonomous reservations let agents reach their
 destination twice as quickly as the competition, with a steady-state of about
-400 agents. The count for traffic signals continues to increase because
-gridlock occurred in one segment of the map, causing agents to not finish their
-route. Unexpected starvation was also observed: although the signals cycle
-through turns regularly, agents cannot proceed when the previous cycle caused a
+400 agents. The count for traffic signals continues to increase because gridlock
+occurred in one segment of the map, causing agents to not finish their route.
+Unexpected starvation was also observed: although the signals cycle through
+turns regularly, agents cannot proceed when the previous cycle caused a
 destination lane to fill and become blocked.

### 2012-03-14 papers/itsc_2012/final.tex
tweaked some wording
some comments in section V

@@
 
-%\long\def\commentp#1{}
-%\long\def\commentpk#1{}
-%\long\def\commentm#1{}
-%\long\def\commentd#1{}
+% \long\def\commentp#1{}
+% \long\def\commentpk#1{}
+% \long\def\commentm#1{}
+% \long\def\commentd#1{}
 
@@
 Like any other microscopic traffic simulator, AORTA needs maps to run
-simulations. One of AORTA's key features is that it generates maps using road data available from OpenStreetMap
-(OSM) \cite{osm}, a Wikipedia-like interface for world maps. Researchers using
-AORTA can export any relevant map area from OSM. This map is then parsed by AORTA to
-create scale simulations of the real world. A researcher employing AORTA can have a simulation
-of any part of the world for which OSM has data running in about five minutes
-(allowing for download time).  AORTA also has the advantage of being open source and
-easy to install so that researchers can make any tweaks they find necessary.
+simulations. One of AORTA's key features is that it generates maps using road
+data available from OpenStreetMap (OSM) \cite{osm}, a Wikipedia-like interface
+for world maps. Researchers using AORTA can export any relevant map area from
+OSM. This map is then parsed by AORTA to create scale simulations of the real
+world. A researcher employing AORTA can have a simulation of any part of the
+world for which OSM has data running in about five minutes (allowing for
+download time).  AORTA also has the advantage of being open source and easy to
+install so that researchers can make any tweaks they find necessary.
 \footnote{http://code.google.com/p/road-rage}
@@
 AORTA is divided into three modular components: the map model, micro-simulation
-engine, and user interface (UI). The map model transforms OSM maps into AORTA graphs, then
-answers pathfinding and geometry queries. The simulation engine adds a notion of
-agents, vehicle dynamics, and collisions. Finally, the UI can
+engine, and user interface (UI). The map model transforms OSM maps into AORTA
+graphs, then answers pathfinding and geometry queries. The simulation engine
+adds a notion of agents, vehicle dynamics, and collisions. Finally, the UI can
 interactively render the map and agents. A headless mode also enables running
@@
 Currently agents cannot switch lanes. AORTA allows merging lanes by treating
-the merge point as an intersection. \commentpk{i take it this is an orphan
-paragraph?} \commentd{yeah, where should it belong?}
+the merge point as an intersection. \commentpk{move this to the limitations
+section, and rename the section to "Current limitations". Explain gridlock as a
+main limitation, and put this in somewhere}
 
@@
   \item Checks for collisions
-  \item Allows each agent to choose an action for the next step
+  \item Allows each agent's behavior to choose an action for the next step
 \end{enumerate}
@@
 potentially crash with the new vehicle, the new agent enters the system with
-zero initial speed. Because behaviors must enforce speed limit along a lane
+zero initial speed. \commentpk {For space, remove the following sentence as well, and merge this paragraph into thr previous one} Because behaviors must enforce speed limit along a lane
 from the moment an agent enters a road, it is safe for a new agent to enter a
@@
 distance at the speed limit, plus stopping distance from this speed) preceeding
-the spawn point.% To avoid predicting which agents might enter the lane in the
+the spawn point. % To avoid predicting which agents might enter the lane in the
 %near future, there is a buffer space at the beginning of the lane where a new
@@
 All traversables (lanes and the straight line-approximation of turns) maintain
-an ordered queue of occupying agents. One simulation step happens atomically,
+an ordered queue of occupying agents. \commentpk{remove the following 2 sentences} One simulation step happens atomically,
 meaning collision checking occurs after all agents have moved. If this were not
 the case, false positives could occur when an agent steps before the agent in front
-of it. A collision is defined as the reversal of order of agents during a step.
+of it.
+A collision is defined as the reversal of order of agents during a step.
+\commentpk{maybe change this to "A collision is detected as the reversal of order of agents after a time step has completed.
+Also do we still have point vehicles only?}
 
@@
 
-\subsection{Agent Reactions}
+\subsection{Agent Behavior}
 \label{sec:behaviors}
@@
 or setting an acceleration for the next step. Were lane changing modeled, that
-would be another possible action.
+would be another possible action. \commentpk {remove last line - we'll bundle it up in limitations}
 
@@
 a stop, since it would be illegal to drive further anyway.
+
+\commentpk{The last 2 paragraphs need to move into the configurability section, as they are not part of the simulator directly. Similarly there should be a subsection on intersection behaviors here}
 

### 2012-03-14 papers/itsc_2012/final.tex
fig 2 downsized

--- 2012-03-14 papers/itsc_2012/final.tex
@@
 % Diagram made with gliffy.com. architecture.gxml should enable editing.
-\begin{figure*}
+\begin{figure}
   \centering \includegraphics[scale=0.3]{architecture.png}
@@
   \vspace{-20pt}
-\end{figure*}
+\end{figure}
 
@@
 
-Dr. Quinlan has certainly been helpful.
-
-What credits do we need to give to OSM?
-\commentp{Just citing it is good enough.}
+The authors would like to thank Dr. Michael Quinlan for his support during the
+seminal stages of AORTA and all OpenStreetMap contributers for providing free
+maps.
 
@@
 
-\commentd{Er, these don't look like they're alphabetically sorted. And why is
-the Google Maps entry so stretched?}
-
+% They're sorted by order of citation, not alphabetically. IEEEtran.bst confirms
+% this is the correct "numerical citation style."
 \bibliographystyle{IEEEtran}

### 2012-03-14 papers/itsc_2012/final.tex
suggested changes to sect 5

--- 2012-03-14 papers/itsc_2012/final.tex
@@
 
-Currently agents cannot switch lanes. AORTA allows merging lanes by treating
-the merge point as an intersection. \commentpk{i take it this is an orphan
-paragraph?} \commentd{yeah, where should it belong?}
-
 Each step, the simulation does the following:
@@
 if possible, sends it to a pool of worker threads to process in the background.
-
 Once an agent's route is ready, the agent waits alongside its starting lane (as
@@
 potentially crash with the new vehicle, the new agent enters the system with
-zero initial speed. Because behaviors must enforce speed limit along a lane
-from the moment an agent enters a road, it is safe for a new agent to enter a
-road if there is no other agent within the worst-case distance (traveling
-distance at the speed limit, plus stopping distance from this speed) preceeding
-the spawn point.% To avoid predicting which agents might enter the lane in the
+zero initial speed. % Because behaviors must enforce speed limit along a lane
+%from the moment an agent enters a road, it is safe for a new agent to enter a
+%road if there is no other agent within the worst-case distance (traveling
+%distance at the speed limit, plus stopping distance from this speed) preceeding
+%the spawn point. To avoid predicting which agents might enter the lane in the
 %near future, there is a buffer space at the beginning of the lane where a new
@@
 All traversables (lanes and the straight line-approximation of turns) maintain
-an ordered queue of occupying agents. One simulation step happens atomically,
-meaning collision checking occurs after all agents have moved. If this were not
-the case, false positives could occur when an agent steps before the agent in front
-of it. A collision is defined as the reversal of order of agents during a step.
+an ordered queue of occupying agents. A collision is detected when two agents
+reverse order after a time step has completed.
+\commentpk{Also do we still have point vehicles only?}
+\commentd{Technically yes, but we could add in a check to make sure the fronts
+  of every vehicle aren't too close to any other vehicle. It would just be
+  needlessly expensive, and we trust the following-distance behavior.}
 
@@
 \subsection{Agent Behavior}
-\label{sec:behaviors}
 
@@
 disappearing from the map (when the agent is at rest and is done with its route)
-or setting an acceleration for the next step. Were lane changing modeled, that
-would be another possible action.
+or setting an acceleration for the next step.
 
-Currently, all agents use a primary behavior that is a generalized, baseline
-behavior that guarantees not to collide with another agent or enter an
-intersection at the wrong time. Since it picks the highest safe choice of
-acceleration at each step, it could easily be extended to mimic human drivers by
-traveling at some random lesser acceleration or to optimize fuel efficiency
-by tuning movement.
+\subsection{Intersection Policies}
 
-The default behavior maximizes acceleration while obeying speed limits,
-avoiding collisions, and obeying intersections. By considering the highest speed
-achievable at the end of the next step and the subsequent stopping distance at
-this speed, the behavior's lookahead engine can scan sufficiently far ahead
-along several traversables to locate agents to avoid and intersections to query.
-The analysis short-circuits at the first nearby agent and intersection requiring
-a stop, since it would be illegal to drive further anyway.
+A behavior's interaction with an intersection policy amounts to polling it with
+the turn the agent wants to perform, the distance away the agent currently is,
+and how long the agent has been waiting (so policies can enforce a required
+delay before entering the intersection). The intersection policy simply orders
+the agent to ``stop'' or ``proceed.'' The acceleration chosen to avoid entering
+an intersection attempts to account for crosswalks by ending a configurable
+threshold before the end of a traversable.
 
-\subsection{Simulation Limitations: Gridlock}
+\subsection{Simulation Limitations}
+
+Currently agents cannot change lanes. However, AORTA supports roads forcibly
+merging by treating the merge point as an intersection. The lack of
+lane-changing causes a second issue: gridlock.
 
@@
 extending built-in implementations of an interface: agent behaviors, routing
-strategies, and intersection policies. Behaviors have been described already in
-section \ref{sec:behaviors}.
+strategies, and intersection policies.
+
+\section{Agent Behaviors}
+
+Currently, all agents use a primary behavior that is a generalized, baseline
+behavior that guarantees not to collide with another agent or enter an
+intersection at the wrong time. Since it picks the highest safe choice of
+acceleration at each step, it could easily be extended to mimic human drivers by
+traveling at some random lesser acceleration or to optimize fuel efficiency
+by tuning movement.
+
+\commentd{So is this paragraph below still relevant?}
+The default behavior maximizes acceleration while obeying speed limits,
+avoiding collisions, and obeying intersections. By considering the highest speed
+achievable at the end of the next step and the subsequent stopping distance at
+this speed, the behavior's lookahead engine can scan sufficiently far ahead
+along several traversables to locate agents to avoid and intersections to query.
+The analysis short-circuits at the first nearby agent and intersection requiring
+a stop, since it would be illegal to drive further anyway.
 
@@
 extensible by assigning groups of non-conflicting turns at each intersection
-with some duration and offset. The experiment in section \ref{sec:results}
-features standard turn grouping (turns from parallel incoming directions are
-scheduled simultaneously, and left turns are grouped) and a breadth-first search
-heuristic for propagating timings. Future policies could include signals that
-adapt timings and refinements to the autonomous reservation policy.
+with some duration and offset. Future policies could include signals that adapt
+timings and refinements to the autonomous reservation policy.
 
@@
 
-\subsection{Experiment parameters}
+\subsection{Intersection Policies Evaluated}
 
-Each policy has parameters, so the configuration used is relevant. Stop signs
-accept agents FCFS, refusing agents until they have idled at 0 speed for 1
-second. This delay approximates human reaction time. The traffic signal policy
-requires timings and groups of simultaneous turns for each intersection. The
-configuration tested here uses a breadth-first search from one major road's
-lanes to try to synchronize green signals for the agents near that lane to not
-stop for several intersections. This heuristic needs improvement in how it
-handles the remaining groups of turns that are scheduled after the first
-sequence. Finally, the autonomous reservation policy allows new agents to join
-existing groups of agents if their turn is compatible with the others. To
-prevent one heavy direction of traffic from hogging the intersection
-indefinitely, a timeout preemptively cycles through reservations. Ideally this
-should be based on the relative numbers of agents depending on their
-group of reservations.
+Stop signs accept agents FCFS, refusing agents until they have idled at 0 speed
+for 1 second. This delay approximates human reaction time.
+
+The traffic signal policy requires timings and groups of simultaneous turns for
+each intersection. The heuristic evaluated uses standard turn groups (turns from
+parallel incoming directions are scheduled simultaneously, and left turns are
+grouped) and employs propagates timings with a breadth-first search. The goal is
+to synchronize green signals so agents in some lanes do not stop for several
+intersections.
+
+\commentd{advice on wording?}
+Finally, in anticipation of the advent of autonomous vehicles, an autonomous
+reservation policy inspired by AIM \cite{JAIR08-dresner} is tested. This policy
+allows new agents to join existing groups of agents if their turn is compatible
+with the others. To prevent one heavy direction of traffic from hogging the
+intersection indefinitely, a timeout preemptively cycles through reservations.
+Ideally this should be based on the relative numbers of agents depending on
+their group of reservations. \commentd{take out the last sentence? all of the
+policies could use some work.}
 

### 2012-03-14 papers/itsc_2012/final.tex
Edits to abstract and intro

--- 2012-03-14 papers/itsc_2012/final.tex
@@
 
-\commentd{sounds fine to me, only thing is that we haven't actually focused on
-  dynamic re-planning or any kind of overseer yet, although theres nothing
-  stopping us from trying that idea again}
-\commentp{If it's not implemented, we shouldn't talk about it.}
-
 \begin{abstract} 
@@
 intersections. These behaviors allow agents to interact with each
-other as well as with an external \textit{overseer} that is
-responsible for optimizing traffic flow. Overseers can interact with
-these agents to allow for dynamic route re-planning to reduce traffic
-load, optimized intersection navigation, and improved car following.
+other as well as local intersections in any manner required for a
+simulation of interest.
 \commentpk{rest devoted to application}
@@
 The arrival of driver-less cars presents several challenges that need to be
-addressed in the near term. Road infrastructure needs to be updated to better
-support autonomous vehicles.\commentp{Is this necessarily the case?
-  It may be useful, but is it a need?} A number of legal and safety concerns need to be
+addressed in the near term. Road infrastructure changes could provide important
+improvements for autonomous vehicles. A number of legal and safety concerns need to be
 addressed in the context of driver-less cars. Human drivers need new
 methodologies to interact with autonomous drivers on public roads. At the same
-time, driver-less cars provide opportunities of solving traffic problems in
+time, driver-less cars provide opportunities for solving traffic problems in
 ways that were never before possible. For instance, autonomous cars can be
@@
 % Diagram made with gliffy.com. architecture.gxml should enable editing.
-\begin{figure}
+\begin{figure*}
   \centering \includegraphics[scale=0.3]{architecture.png}
@@
   \vspace{-20pt}
-\end{figure}
+\end{figure*}
 
@@
 
-The authors would like to thank Dr. Michael Quinlan for his support during the
-seminal stages of AORTA and all OpenStreetMap contributers for providing free
-maps.
+Dr. Quinlan has certainly been helpful.
+
+What credits do we need to give to OSM?
+\commentp{Just citing it is good enough.}
 
@@
 
-% They're sorted by order of citation, not alphabetically. IEEEtran.bst confirms
-% this is the correct "numerical citation style."
+\commentd{Er, these don't look like they're alphabetically sorted. And why is
+the Google Maps entry so stretched?}
+
 \bibliographystyle{IEEEtran}

### 2012-03-14 papers/itsc_2012/final.tex
a few mistakes fixed

@@
 
-Likewise, intersections check for collisions by examining the turn each agent is
+Intersections check for collisions by examining the turn each agent is
 performing and verifying that no other agent is simultaneously performing a
@@
 delay before entering the intersection). The intersection policy simply orders
-the agent to ``stop'' or ``proceed.'' The acceleration chosen to avoid entering
-an intersection attempts to account for crosswalks by ending a configurable
-threshold before the end of a traversable.
+the agent to ``stop'' or ``proceed.'' \commentd{I'm not sure this fits in the
+default behavior later anymore; just delete it?} The acceleration chosen to
+avoid entering an intersection attempts to account for crosswalks by ending a
+configurable threshold before the end of a traversable.
 
@@
 
-\section{Agent Behaviors}
+\subsection{Agent Behaviors}
 
@@
 \subsection{Intersection Policies}
-
-A behavior's interaction with an intersection policy amounts to polling it with
-the turn the agent wants to perform, the distance away the agent currently is,
-and how long the agent has been waiting (so policies can enforce a required
-delay before entering the intersection). The intersection policy simply orders
-the agent to ``stop'' or ``proceed.'' The acceleration chosen to avoid entering
-an intersection attempts to account for crosswalks by ending a configurable
-threshold before the end of a traversable.
 

### 2012-03-14 papers/itsc_2012/final.tex
Edits to abstract and intro

@@
 other as well as local intersections in any manner required for a
-simulation of interest.
-\commentpk{rest devoted to application}
+simulation of interest.  This is demonstrated in a simple experiment
+for autonomous vehicles demonstrated at the end of this paper.
 \end{abstract}
@@
 addressed in the near term. Road infrastructure changes could provide important
-improvements for autonomous vehicles. A number of legal and safety concerns need to be
-addressed in the context of driver-less cars. Human drivers need new
-methodologies to interact with autonomous drivers on public roads. At the same
-time, driver-less cars provide opportunities for solving traffic problems in
-ways that were never before possible. For instance, autonomous cars can be
-quickly diverted to make way for emergency vehicles responding to an accident.
-Even day-to-day traffic issues such as congestion could potentially be greatly
-reduced by effective management of traffic flow. This work addresses
-some of these problems.\commentp{Be more specific in this paragraph.
-  Which of these challenges are eing addressed elsewhere (cite), and
-  which are addressed in this paper?}
-\commentp{Overall, this paragraph is a bit scattered.  I suggest
-  something more like the following (after the first sentence).
-  Several of the challenges are already being addressed.  For example
-  .....
-  This paper focusses on the specific challenge(s) of....}
-
-\commentd{nevada even made autonomous cars legal:
-http://www.pcmag.com/article2/0,2817,2400400,00.asp}
-
-\commentd{another example long-term usage: fuel. I went to a talk by
-  somebody on the engr side about powermiling and optimizing battery usage}
-
-\commentd{we don't support lane-changing yet, and overseers arent re-implemented
-yet either}
-
-\commentp{legal issues are addressed here:  http://www.path.berkeley.edu/PATH/Publications/PDF/PRR/2009/PRR-2009-28.pdf}
+improvements to traffic involving autonomous vehicles. A number of legal and
+safety concerns need to be addressed in the context of driver-less cars. Human
+drivers need new methodologies to interact with autonomous drivers on public roads.
+AORTA aims to solve some of these problems by providing a sandbox in which such
+methodologies can be found and tested for safety and efficiency before being
+deployed in the real world.  In this manner, AORTA aids in the effort to find
+intermediate stages between entirely human and entirely autonomous traffic.
 

### 2012-03-14 papers/itsc_2012/final.tex
revising sect 6

--- 2012-03-14 papers/itsc_2012/final.tex
@@
 reverse order after a time step has completed.
-\commentpk{Also do we still have point vehicles only?}
-\commentd{Technically yes, but we could add in a check to make sure the fronts
-  of every vehicle aren't too close to any other vehicle. It would just be
-  needlessly expensive, and we trust the following-distance behavior.}
 
@@
 
-\commentd{So is this paragraph below still relevant?}
-The default behavior maximizes acceleration while obeying speed limits,
-avoiding collisions, and obeying intersections. By considering the highest speed
-achievable at the end of the next step and the subsequent stopping distance at
-this speed, the behavior's lookahead engine can scan sufficiently far ahead
-along several traversables to locate agents to avoid and intersections to query.
-The analysis short-circuits at the first nearby agent and intersection requiring
-a stop, since it would be illegal to drive further anyway.
-
 \subsection{Intersection Policies}
@@
 
-Picking accelerations that optimize objective functions is a separate task from
-picking a path to travel. Agent behaviors consult a route strategy to pick
-turns. Currently two simple implementations are available. One statically plans
-a route to some goal using standard A* search \cite{astar}. The other performs a
-drunken walk, lazily planning random choices at each intersection as the
-lookahead engine requires, choosing the direction that minimizes the Euclidean
-distance to the goal with 75\% probability. Researchers can experiment with
-dynamic replanning or hierarchial planning without modifying any code other than
-creating a new route implementation. Route policies could also interact with a
-singleton overseer to gain some sort of global insight or with smart
-intersections to avoid congestion.
+To pick turns, agent behaviors consult a route strategy. Two simple
+implementations exist already: a static route using A* \cite{astar} and a
+drunken walk that probabilistically moves towards the destination.  Researchers
+can experiment with dynamic replanning or hierarchial planning without modifying
+any code other than creating a new route implementation. Route policies could
+also interact with a singleton overseer to gain some sort of global insight or
+with smart intersections to avoid congestion.
 
@@
 A particular focus of AORTA is intersection policies. To demonstrate how much
-policy affects agent performance, three policies are evaluated: stop sign,
+policy affects agent performance \commentd{performance is a vague term, but I
+don't want to talk about dependent vars just yet}, three policies are evaluated: stop sign,
 traffic signal, and autonomous reservation. These are compared with all other
@@
 parallel incoming directions are scheduled simultaneously, and left turns are
-grouped) and employs propagates timings with a breadth-first search. The goal is
-to synchronize green signals so agents in some lanes do not stop for several
+grouped) and employs breadth-first search to propagate timings. The goal is to
+synchronize green signals so agents in some lanes do not stop for several
 intersections.
@@
 Each policy has its trade-offs in terms of performance and physical hardware
-cost, suggesting a mixed approach, perhaps with intersection policies adapting
-as traffic conditions shift. Researchers could conduct this experiment and many
-others without needing to understand most of AORTA's code-base. The largest
-intersection policy, traffic signals with the flooding heuristic for timing
-assignment, is only 400 lines of Scala.
+cost \commentd{do we want to raise this issue?}, suggesting a mixed approach,
+perhaps with intersection policies adapting as traffic conditions shift.
+Researchers could conduct this experiment and many others without needing to
+understand most of AORTA's code-base. The largest intersection policy, traffic
+signals with the flooding heuristic for timing assignment, is only 400 lines of
+Scala.
 

### 2012-03-14 papers/itsc_2012/final.tex
Edits to abstract and intro

@@
 simulation of interest.  This is demonstrated in a simple experiment
-for autonomous vehicles demonstrated at the end of this paper.
+for autonomous vehicles presented at the end of this paper.
 \end{abstract}
@@
 drivers need new methodologies to interact with autonomous drivers on public roads.
-AORTA aims to solve some of these problems by providing a sandbox in which such
+This work aims to solve some of these problems by providing a sandbox in which such
 methodologies can be found and tested for safety and efficiency before being
-deployed in the real world.  In this manner, AORTA aids in the effort to find
+deployed in the real world.  In this manner, the present work aids in the effort to find
 intermediate stages between entirely human and entirely autonomous traffic.
@@
 \textit{behaviors} that these agents follow to interact with one
-another. To optimize traffic flow, external \textit{overseers} can
-coordinate with these agents to provide recommendations such as route
-suggestions or intersection signaling policies.  An advantage of such
-an approach is that by assigning human-like behaviors such as the
-``car following model'' \cite{brackstone1999car} to vehicles, and
-traffic-signal-like policies to intersection, AORTA can just as easily
-simulate human traffic.
+another. An advantage of such an approach is that by assigning
+human-like behaviors such as the ``car following model''
+\cite{brackstone1999car} to vehicles, and traffic-signal-like
+policies to intersections, AORTA can just as easily simulate human
+traffic.
 
@@
 
-\commentpk{Needs 1 more para based on final application}
-
-% Section 2 will review other traffic simulators. Section 3 will introduce the
-% overall architecture of AORTA. Section 4 describes how OSM maps are used.
-% Section 5 explains the simulation engine and the properties AORTA posses that
-% are useful for general traffic experiments. Section 6 gives some initial
-% results from such an experiment, and Section 7 concludes.
+After exploring the state of the art \ref{sec:related_work} and explaining
+the architecture of AORTA \ref{sec:arch} as well as its use of OSM \ref{sec:map},
+we will describe simulation in AORTA \ref{sec:simulation} and how to extend it
+\ref{sec:config}, concluding with an example experiment \ref{sec:results}.
+In this final experiment, AORTA shows the benefits of an AIM-inspired
+\cite{JAIR08-dresner} reservation intersection policy over traditional
+traffic signals and stop signs.
 

### 2012-03-14 papers/itsc_2012/final.tex
Fixed section refs in intro

@@
 
-After exploring the state of the art \ref{sec:related_work} and explaining
-the architecture of AORTA \ref{sec:arch} as well as its use of OSM \ref{sec:map},
-we will describe simulation in AORTA \ref{sec:simulation} and how to extend it
-\ref{sec:config}, concluding with an example experiment \ref{sec:results}.
+After exploring the state of the art in section \ref{sec:related_work} and explaining
+the architecture of AORTA (section \ref{sec:arch}) as well as its use of OSM (section \ref{sec:map}),
+we will describe simulation in AORTA and how to extend it in sections \ref{sec:simulation} and
+\ref{sec:config} respectively. We conclude with an example experiment in section \ref{sec:results}.
 In this final experiment, AORTA shows the benefits of an AIM-inspired

### 2012-03-14 papers/itsc_2012/final.tex
heading changes
fig 5's axis cropped and intro to sec 7 reorganized

@@
 
-\subsection{Agent Behavior}
+\subsection{Agent Behaviors}
 
@@
 
-\section{CONFIGURABILITY}
+\section{EXTENDING AORTA}
 \label{sec:config}
@@
 don't want to talk about dependent vars just yet}, three policies are evaluated: stop sign,
-traffic signal, and autonomous reservation. These are compared with all other
-parameters fixed:
+traffic signal, and autonomous reservation.
 
-\begin{enumerate}
-  \item the map (a 6 by 7 mile slice of downtown Austin, Texas
-        \footnote{bounded by the Mopac and Springdale Road horizontally,
-        and Oltorf Street and Koenig Lane vertically})
-  \item dt = 0.1 seconds
-  \item the agents spawned and their routes -- a continuous generator creates
-        one new agent every 1 simulation second, with a uniformly random
-        starting position and goal
-  \item the duration of the experiment: one hour
-\end{enumerate}
+\subsection{Experimental Setup}
 
-Each experiment is deterministically reproducible. A seed for the pseudo-random
-number generator, the input graph, and the generators' configuration fully
-determine the outcome of any simulation \footnote{This determinism may break
-  down when generators use worker threads, since routes may be computed in
-different orders, causing agents to enter the road at different times.}. After
-setting up a scenario in the UI, users can save this configuration for
-resimulation.
+The experiment simulates a 6 by 7 mile slice of downtown Austin,
+Texas\footnote{bounded by the Mopac and Springdale Road horizontally, and Oltorf
+Street and Koenig Lane vertically}, for one hour. The step duration is fixed at
+0.1 seconds. While the intersection policies vary, the agents spawned and their
+routes remain the same. A generator creates one new agent every 1 simulation
+second, with a uniformly random starting position and goal.
+
+These experiments are deterministically reproducible. A seed for the
+pseudo-random number generator, the input graph, and the generators'
+configuration fully determine the outcome of any simulation \footnote{This
+  determinism may break down when generators use worker threads, since routes
+  may be computed in different orders, causing agents to enter the road at
+different times.}. After setting up a scenario in the UI, users can save this
+configuration for resimulation.
 

### 2012-03-14 papers/itsc_2012/final.tex
Addressed comments in related works

--- 2012-03-14 papers/itsc_2012/final.tex
@@
 time spent in transit. This goal is somewhat different from that of AORTA,
-where the demand supplied by individuals is taken as is. Instead, AORTA aims to
-reduce traffic congestion through interactions with individuals in real-time
-based on the current demand.
-
-\commentd{last sentence above... we're not not doing that yet, and we dont
-  discuss it further here.}
+where the demand supplied by individuals is taken as is.  Rather than simply improving
+routes, AORTA allows entirely different behaviors for both vehicles and intersections.
 
@@
 based intersection precedence, or through more efficient methods designed
-specifically for autonomous vehicles \cite{JAIR08-dresner}.
-%\commentpk{does the above difference seem a bit pedantic?}
+specifically for autonomous vehicles \cite{JAIR08-dresner}. Based on experience
+with SUMO, AORTA also displays more robust handling of data exported directly from OSM.
 
@@
 target autonomous traffic systems. One such system is AIM
-\cite{JAIR08-dresner}, which aims to optimize traffic flow between
+\cite{JAIR08-dresner}, which aims to optimize traffic flow of
 autonomous vehicles at a given intersection. AIM studies problems at
@@
 while AORTA has the potential to apply some of the same research at a
-city-wide scale to real-world road networks. A few simulators focus on
-autonomous vehicles with sensors and actuators
+city-wide scale to real-world road networks.
+
+A few simulators focus on autonomous vehicles with sensors and actuators
 \cite{figueiredo2009approach}. These simulators are used to ensure
@@
 other vehicles and infrastructure. On the other hand, AORTA assumes
-that autonomous vehicles are capable of reasonable driving, and
-focuses on interactions between agents to optimize traffic flow.
+that autonomous vehicles have reasonable sensory input and mechanical
+output, and instead focuses on interactions between an agent and its
+surrounding agents and intersections.
 
@@
 minimizing overall wait time \cite{manikonda2001autonomous}.
-
-\commentd{and we can do the above too. we talk about how to make the current
-  behavior more flawed and humanlike in the agent reactions section.}
 

### 2012-03-14 papers/itsc_2012/final.tex
intermediate version of edited draft

@@
 \title{\LARGE \bf
-Approximately Orchestrated Road Traffic Automata\\
-a General Traffic Simulation Framework
+Approximately Orchestrated Road Traffic Automata:\\
+Large-scale Traffic Simulation for Autonomous Vehicles
 }
@@
 \begin{abstract}
-Autonomous vehicles have seen great advancements in recent years, and
-such vehicles are now ever closer to being commercially available.
-The advent of driver-less cars provides opportunities for optimizing
-traffic in ways not possible before. This paper introduces an open
-source multi-agent traffic simulator called AORTA, which stands for
-\textit{Approximately Orchestrated Road Traffic Automata}, and
-demonstrates its usage for optimizing autonomous traffic at a
-city-wide scale. AORTA creates scale simulations by generating maps
-using publicly available road data from \textit{OpenStreetMap}
-(OSM). Simulations of specific regions throughout the world can be set
-up in the few minutes it takes to export data for the region of
-interest from OSM. AORTA focuses on interactions between agents by
-defining \textit{behaviors} for individual vehicles and
-intersections. These behaviors allow agents to interact with each
-other as well as local intersections in any manner required for a
-simulation of interest.  This is demonstrated in a simple experiment
-for autonomous vehicles presented at the end of this paper.
+
+Autonomous vehicles have seen great advancements in recent years, and such
+vehicles are now ever closer to being commercially available.  The advent of
+driverless cars provides opportunities for optimizing traffic in ways not
+possible before. This paper introduces an open source multi-agent microscopic
+traffic simulator called AORTA, which stands for \textit{Approximately
+Orchestrated Road Traffic Automata}, designed for optimizing autonomous traffic
+at a city-wide scale. AORTA creates scale simulations of the real world by
+generating maps using publicly available road data from \textit{OpenStreetMap}
+(OSM). This allows simulations to be set up through AORTA for a desired region
+anywhere in the world in a matter of minutes. AORTA allows for optimizing
+traffic by creating more intelligent behaviors for individual drivers (agents),
+as well as by defining new intersection policies that can be followed by these
+autonomous agents. These agent behaviors define how agents interact with one
+another, follow intersection policies, and select the next action to take. This
+paper demonstrates a simple applications using AORTA through an experiment
+testing an intersection policy designed for autonomous vehicles at a city-wide
+scale.
+
 \end{abstract}
@@
 In 2004, the farthest distance traveled by a vehicle autonomously in the DARPA
-Grand Challenge was $11.9km$ \cite{cnnGrandChallenge2004}. By 2007, six of the
-competing teams completed the $96km$ course set for the DARPA Urban Challenge
+Grand Challenge was 11.9km \cite{cnnGrandChallenge2004}. By 2007, six of the
+competing teams completed the 96km course set for the DARPA Urban Challenge
 \cite{spectrumUrbanChallenge2007}. They did so while following the same traffic
@@
 and following correct intersection precedence order. Since then, Google's
-driver-less cars have clocked more than $250,000km$ on public roads in urban
+driverless cars have clocked more than 250,000km on public roads in urban
 California, USA \cite{tedThrun2011}. In 2010, researchers from the University
@@
 completion of all these milestones suggests that autonomous cars are here to
-stay, and are ever closer to becoming commercially available.
+stay, and are ever closer to becoming commercially available. With the arrival
+of autonomous cars, it also becomes possible to optimize traffic in ways not
+possible for human drivers.
 
-The arrival of driver-less cars presents several challenges that need to be
-addressed in the near term. Road infrastructure changes could provide important
-improvements to traffic involving autonomous vehicles. A number of legal and
-safety concerns need to be addressed in the context of driver-less cars. Human
-drivers need new methodologies to interact with autonomous drivers on public roads.
-This work aims to solve some of these problems by providing a sandbox in which such
-methodologies can be found and tested for safety and efficiency before being
-deployed in the real world.  In this manner, the present work aids in the effort to find
-intermediate stages between entirely human and entirely autonomous traffic.
+% The arrival of driverless cars presents several challenges that need to be
+% addressed in the near term. Road infrastructure changes could provide important
+% improvements to traffic involving autonomous vehicles. A number of legal and
+% safety concerns need to be addressed in the context of driverless cars. Human
+% drivers need new methodologies to interact with autonomous drivers on public roads.
+% This work aims to solve some of these problems by providing a sandbox in which such
+% methodologies can be found and tested for safety and efficiency before being
+% deployed in the real world.  In this manner, the present work aids in the effort to find
+% intermediate stages between entirely human and entirely autonomous traffic.
 
-\begin{figure}[h]
+\begin{figure}[t]
   \centering \includegraphics[width=\linewidth]{downtown_atx.png}
@@
   \label{fig:ui_screenshot}
-  \vspace{-10pt}
+  \vspace{-15pt}
 \end{figure}
@@
 
-This paper introduces an open source multi-agent microscopic traffic
-simulator called AORTA, which stands for \textit{Approximately
-Orchestrated Road Traffic Automata}. The AORTA simulator focuses on
-the interactions between individual vehicles and intersections, both
-of which act as \textit{agents} in AORTA's framework. Vehicles
-interact with other vehicles to pass and change lanes, and interact
-with intersections to decide when to cross them. Researchers
-developing applications on top of the AORTA simulator define
-\textit{behaviors} that these agents follow to interact with one
-another. An advantage of such an approach is that by assigning
-human-like behaviors such as the ``car following model''
-\cite{brackstone1999car} to vehicles, and traffic-signal-like
-policies to intersections, AORTA can just as easily simulate human
-traffic.
+This paper introduces an open source multi-agent microscopic traffic simulator
+called AORTA, which stands for \textit{Approximately Orchestrated Road Traffic
+Automata}, a platform designed for testing autonomous vehicle behaviors and
+intersection policies. Autonomous vehicles, termed \textit{agents}, use
+behaviors to interact with one another, follow intersection policies, and
+decide on both long term and short term actions. AORTA allows for the creation
+of intersection policies on top of a reservation based intersection manager
+\cite{JAIR08-dresner}, and these policies designate when agents can cross the
+intersection. AORTA's goal is to allow for the definition of new agent
+behaviors and intersection policies to optimize autonomous traffic.
+Additionally, by assigning human-like behaviors such as the ``car following
+model'' \cite{brackstone1999car} to agents, and traffic-signal-like policies to
+intersections, AORTA could potentially simulate human traffic.
 
 Like any other microscopic traffic simulator, AORTA needs maps to run
-simulations. One of AORTA's key features is that it generates maps using road
-data available from OpenStreetMap (OSM) \cite{osm}, a Wikipedia-like interface
-for world maps. Researchers using AORTA can export any relevant map area from
-OSM. This map is then parsed by AORTA to create scale simulations of the real
-world. A researcher employing AORTA can have a simulation of any part of the
-world for which OSM has data running in about five minutes (allowing for
-download time).  AORTA also has the advantage of being open source and easy to
-install so that researchers can make any tweaks they find necessary.
-\footnote{http://code.google.com/p/road-rage}
+simulations. One of AORTA's key features is that it generates maps using real
+road data available from OpenStreetMap (OSM) \cite{osm}, a Wikipedia-like
+interface for world maps. A map for any desired city in the world can be
+downloaded, which is then parsed by AORTA to set up a scale simulation of the
+real world in a few minutes. AORTA is available open-source and is easily
+extensible \footnote{Code available at
+\texttt{http://code.google.com/p/road-rage}} \commentpk{could also change this
+to a citation}, making it easier for users to test out a number of agent
+behaviors and intersection policies.
 
-After exploring the state of the art in section \ref{sec:related_work} and explaining
-the architecture of AORTA (section \ref{sec:arch}) as well as its use of OSM (section \ref{sec:map}),
-we will describe simulation in AORTA and how to extend it in sections \ref{sec:simulation} and
-\ref{sec:config} respectively. We conclude with an example experiment in section \ref{sec:results}.
-In this final experiment, AORTA shows the benefits of an AIM-inspired
-\cite{JAIR08-dresner} reservation intersection policy over traditional
-traffic signals and stop signs.
+This paper explores the state-of-the-art in traffic simulators in section
+\ref{sec:related_work}, followed by a description of AORTA's architecture in
+section \ref{sec:arch}.  Use of OSM data in AORTA is explained in section
+\ref{sec:map}, along with a description of the simulator and the default
+behaviors available in AORTA in section \ref{sec:simulation}. An experiment
+investigating the benefit of an alternate intersection policy designed for
+autonomous vehicles is presented in \ref{sec:results}. The paper then concludes
+with a discussion on future work.
 
@@
 
-Computational processing power has made excellent advancements in the
-last two decades. Parallel computing has proved significant for
-traffic simulation through the use of Geographic Information System
-(GIS) software \cite{pursula1999simulation}. Advanced computing
-techniques have enabled microscopic models of traffic simulation to
-generate results at a meaningful scale (city-wide or greater). As a
-result, a number of good micro-simulators have been introduced in the
-past decade. We review a number of such simulators in this section
-along with other relevant work.
+Computational processing power has made excellent advancements in the last two
+decades. Parallel computing has proved significant for traffic simulation
+through the use of Geographic Information System (GIS) software
+\cite{pursula1999simulation}. \commentpk{preceding sentence doesnt paraphrase
+the ref correctly. double check}. Advanced computing techniques have enabled
+microscopic models of traffic simulation to generate results at a meaningful
+scale (city-wide or greater). As a result, a number of good micro-simulators
+have been introduced in the past decade. We review a number of such simulators
+in this section along with other relevant work.
 
@@
 time spent in transit. This goal is somewhat different from that of AORTA,
-where the demand supplied by individuals is taken as is.  Rather than simply improving
-routes, AORTA allows entirely different behaviors for both vehicles and intersections.
+where the demand supplied by individuals is taken as is. Rather than simply
+improving routes, AORTA allows creating entirely different behaviors for both
+vehicles and intersections.
 
@@
 specifically for autonomous vehicles \cite{JAIR08-dresner}. Based on experience
-with SUMO, AORTA also displays more robust handling of data exported directly from OSM.
+with SUMO, AORTA also displays more robust handling of data exported directly
+from OSM. \commentpk{I am worried about the last line. It might be true, but it
+is a fairly strong claim - we should put a citation of us handling something
+that has been documented as not being handled via SUMO}
 
-On the other hand, there are a number of simulators that specifically
-target autonomous traffic systems. One such system is AIM
-\cite{JAIR08-dresner}, which aims to optimize traffic flow of
-autonomous vehicles at a given intersection. AIM studies problems at
-the scale of roughly a dozen intersections arranged in a regular grid,
-while AORTA has the potential to apply some of the same research at a
-city-wide scale to real-world road networks.
+On the other hand, there are a number of simulators that specifically target
+autonomous traffic systems. One such system is AIM \cite{JAIR08-dresner}, which
+aims to optimize traffic flow of autonomous vehicles at a given intersection.
+AIM studies problems at the scale of roughly a dozen intersections arranged in
+a regular grid, while AORTA has the potential to apply some of the same
+research at a city-wide scale to real-world road networks.
 
-A few simulators focus on autonomous vehicles with sensors and actuators
-\cite{figueiredo2009approach}. These simulators are used to ensure
-that autonomous vehicles are behaving correctly while interacting with
-other vehicles and infrastructure. On the other hand, AORTA assumes
-that autonomous vehicles have reasonable sensory input and mechanical
-output, and instead focuses on interactions between an agent and its
-surrounding agents and intersections.
+Simulators have also focused on autonomous vehicles with sensors and actuators
+\cite{figueiredo2009approach}. These simulators are used to ensure that
+autonomous vehicles are behaving correctly while interacting with other
+vehicles and infrastructure, given their configuration of sensors and
+actuators. On the other hand, AORTA assumes that autonomous vehicles have
+reasonable sensory input and mechanical output, and instead focuses on
+interactions between an agent and its surrounding agents and intersections.
 
-Autonomous agents have also been used in the past to model human
-traffic. For instance, one approach models human merging and lane
-changing behavior through the use of autonomous agents interacting
-with one another \cite{hidas2002modelling}. Another approach attempted
-to improve traffic congestion for human drivers by using a network of
-autonomous intersections.  These intersections have the ability to
-interact with one another and execute a dynamic traffic signal policy
-minimizing overall wait time \cite{manikonda2001autonomous}.
+Autonomous agents have also been used in the past to model human traffic. For
+instance, one approach models human merging and lane changing behavior through
+the use of autonomous agents interacting with one another
+\cite{hidas2002modelling}. Another approach attempted to improve traffic
+congestion for human drivers by using a network of autonomous intersections.
+These intersections have the ability to interact with one another and execute a
+dynamic traffic signal policy minimizing overall wait time
+\cite{manikonda2001autonomous}.
 
@@
 built on top of AORTA can be written in either Java or Scala. The software is
-open-source \footnote{source code at http://code.google.com/p/road-rage} and
-easily extensible. For instance, the stop sign sign policy used in experiments
-described in section \ref{sec:results} is implemented in just 70 lines of code.
+open-source and easily extensible. For instance, the stop sign sign policy used
+in experiments described in section \ref{sec:results} is implemented in just 70
+lines of code.
+
+Section \ref{sec:map} and \ref{sec:simulation} describe map construction from
+OSM data and the simulation engine used in the AORTA framework respectively.
 
@@
 A map is represented as a directed graph with \emph{lanes} as edges and
-\emph{turns} as vertices. A \emph{road} groups one or two sets of lanes
-together, one set for each direction of travel.
-\emph{Intersections} map the possible turns from incoming to outgoing lanes.
-Since both lanes and turns support traffic, they are grouped together as
-\emph{traversables}, meaning agents may exist on them.
+\emph{turns} as vertices. A \emph{road} groups a set of lanes together, one set
+for each direction of travel. \emph{Intersections} map the possible turns from
+incoming to outgoing lanes.  Since both lanes and turns support traffic, they
+are grouped together as \emph{traversables}, meaning agents may exist on them.
 
@@
   \item OSM encodes many paths besides driveable roads, so the builder first filters
-        these out. Next the builder marks GPS nodes common to multiple roads, since
+        these out. Next, the builder marks GPS nodes common to multiple roads, since
         they implicitly indicate intersections. Overpasses do not share nodes

### 2012-03-15 papers/itsc_2012/final.tex
checking in edits to sections I through III based on Chiu's and Piyush's comments - also added acknowledgements blurb

@@
 Computational processing power has made excellent advancements in the last two
-decades. Parallel computing has proved significant for traffic simulation
-through the use of Geographic Information System (GIS) software
-\cite{pursula1999simulation}. \commentpk{preceding sentence doesnt paraphrase
-the ref correctly. double check}. Advanced computing techniques have enabled
-microscopic models of traffic simulation to generate results at a meaningful
-scale (city-wide or greater). As a result, a number of good micro-simulators
-have been introduced in the past decade. We review a number of such simulators
-in this section along with other relevant work.
+decades. Parallel computing and the use of GPUs have enabled microscopic models
+of traffic simulation to generate results at a meaningful scale (city-wide or
+greater) \cite{nagel1994microscopic,shen2011agent}. As a result, a number of
+multi-agent micro-simulators have been introduced in the past decade. We review
+relevant simulators and other related work in this section.
 
-A number of traffic simulators already make use of OSM data. MATSim is one such
+OSM data has been used by traffic simulators in the past. MATSim is one such
 multiagent simulator that focuses on performing large-scale simulations in a
@@
 Another popular open source traffic simulator that can use OSM data is SUMO
-\cite{SUMO2011}, which shares the same goals as AORTA and is fairly similar at
-a high level. SUMO has been used to study the effect of automated
-transportation systems, route planning of individual vehicles, as well as
-dynamically adapting traffic signal policies to increase traffic efficiency.
-AORTA and SUMO differ in the manner in which intersections are handled. SUMO
-uses traditional traffic signals, while AORTA adopts a much more general
-\textit{reservation} system. A reservation tells a car when it is safe to take
-its desired path through the intersection. Reservations can be allotted to
-vehicles through more traditional means such as traffic signals or stop-sign
-based intersection precedence, or through more efficient methods designed
-specifically for autonomous vehicles \cite{JAIR08-dresner}. Based on experience
-with SUMO, AORTA also displays more robust handling of data exported directly
-from OSM. \commentpk{I am worried about the last line. It might be true, but it
-is a fairly strong claim - we should put a citation of us handling something
-that has been documented as not being handled via SUMO}
+\cite{SUMO2011}, which shares the similar goals as AORTA. SUMO has been used to
+study the effect of automated transportation systems, route planning of
+individual vehicles, as well as dynamically adapting traffic signal policies to
+increase traffic efficiency.  AORTA and SUMO differ in the manner in which
+intersections are handled. SUMO uses traditional traffic signals, while AORTA
+adopts a much more general \textit{reservation} system. A reservation tells a
+car when it is safe to take its desired path through the intersection.
+Reservations can be allotted to vehicles through more traditional means such as
+traffic signals or stop-sign based intersection precedence, or through more
+efficient methods designed specifically for autonomous vehicles
+\cite{JAIR08-dresner}.
 
-On the other hand, there are a number of simulators that specifically target
-autonomous traffic systems. One such system is AIM \cite{JAIR08-dresner}, which
-aims to optimize traffic flow of autonomous vehicles at a given intersection.
-AIM studies problems at the scale of roughly a dozen intersections arranged in
-a regular grid, while AORTA has the potential to apply some of the same
-research at a city-wide scale to real-world road networks.
+Efforts have been made to optimize traffic flow in the context of autonomous
+vehicles. AIM \cite{JAIR08-dresner} is one such approach that aims to optimize
+traffic flow of autonomous vehicles at a given intersection. The AIM approach
+has been applied to a regular grid of four intersections
+\cite{IROS11-hausknecht}, while AORTA has the potential to apply some of the
+same research at a city-wide scale to real-world road networks.
 
@@
 
+This work has taken place in the Learning Agents Research Group (LARG) at the
+Artificial Intelligence Laboratory, The University of Texas at Austin. LARG
+research is supported in part by grants from the National Science Foundation
+(IIS-0917122), ONR (N00014-09-1-0658), and the Federal Highway Administration
+(DTFH61-07-H-00030).
+
 The authors would like to thank Dr. Michael Quinlan for his support during the
-seminal stages of AORTA and all OpenStreetMap contributers for providing free
-maps.
+seminal stages of AORTA and all OpenStreetMap contributers for collecting the
+map data used in this work. Efforts by the Freshman Research Initiative,
+College of Natural Sciences, UT Austin have allowed for the development of this
+research.
 

### 2012-03-15 papers/itsc_2012/generate.sh

@@ -1,6 +1,6 @@
 #!/bin/sh
 
-if [[ $# -eq 1 ]]; then
+if [ $# -eq 1 ]; then
   INPUT=$1
 else
   INPUT=final.tex

### 2012-03-15 papers/itsc_2012/final.tex
edits to section IV

@@
 \long\def\commentd#1{{\bf **Dustin: #1**}}
+\long\def\commentc#1{{\bf **Chiu: #1**}}
 
@@
 % \long\def\commentd#1{}
+% \long\def\commentc#1{}
 
@@
 
-One advantage of AORTA is that it simulates traffic on existing infrastructure
-rather than randomly generated maps. Utilizing OpenStreetMap \cite{osm}, a
-community project offering user-editable, free XML-based maps, enables this. To
-run a simulation on a map from OSM, there is a one-time conversion process to
-transform the map into AORTA's format, which encodes information about lanes and
-turns not contained in the OSM source.
+One of the key features of AORTA is the simulation of traffic on existing road
+network data from OpenStreetMap \cite{osm}. The data from OSM is not directly
+suitable for running a simulation and is pre-processed through a \emph{map
+builder} utility. The map model generated by the builder is described in sec.
+\ref{sec:mapmodel} and the process of refining OSM data is enumerated in sec.
+\ref{sec:mapconstruction}.
 
 \subsection{Map Model}
+\label{sec:mapmodel}
 
-%\pix{map_model.png}
+\commentpk{I remember you guys mentioning there was a paper for SUMO describing
+a similar process, we should ref it ass to why OSM data is not directly suitable}
+
+% \pix{map_model.png}
 %    {Roads contain several lanes, and intersections contain several turns
@@
 are grouped together as \emph{traversables}, meaning agents may exist on them.
+\commentc{What about parking lots?}
 
 The map also has a geometric interpretation, used for visualization and
-measuring physical distances. A road contains a sequence of points, defining the
-center line that divides the road into two directions of travel. AORTA
+measuring physical distances. A road contains a sequence of points, defining
+the center line that divides the road into two directions of travel. AORTA
 interprets these points as straight lines and projects parallel lanes out using
 a fixed width. Turns are modeled with a single straight line connecting the end
-of one lane to the start of another (so right turns often have 0 length).
+of one lane to the start of another. \commentc{Illustrate through a figure?}
 
 \subsection{Map Construction Passes}
+\label{sec:mapconstruction}
 
-The builder works by incrementally transforming OSM graphs into AORTA graphs,
-with each pass feeding the next.
+The builder works incrementally while parsing OSM graphs into the map model,
+with each pass feeding the next. The map model is then stored in an XML format.
 \begin{enumerate}
   \item OSM encodes many paths besides driveable roads, so the builder first filters
-        these out. Next, the builder marks GPS nodes common to multiple roads, since
+        these out. Next, the builder marks OSM nodes common to multiple roads, since
         they implicitly indicate intersections. Overpasses do not share nodes
         with the roads they cross \cite{osmOverpass}.
-  \item OSM \emph{ways} (sequences of GPS locations) include many
+  \item OSM \emph{ways} (sequences of nodes) include many
         intersections, so the builder next splits ways into undirected
@@
         a left, right, straight, or U-turn. When several lanes all cross into fewer
-        lanes, the builder forces merging of the rightmost lanes. Although
-        inaccurate, these patterns are the best compromise with the available data.
+        lanes, the builder forces merging of the rightmost lanes. 
   \item Because turns are heuristically placed, the graph is often disconnected.
@@
         corner-case free pathfinding.
-  \item The builder serializes the graph in a simple XML format.
 \end{enumerate}

### 2012-03-15 papers/itsc_2012/final.tex
intermediate commit for sec V

@@
 
-Section \ref{sec:map} and \ref{sec:simulation} describe map construction from
+Sections \ref{sec:map} and \ref{sec:simulation} describe map construction from
 OSM data and the simulation engine used in the AORTA framework respectively.
@@
 \commentpk{I remember you guys mentioning there was a paper for SUMO describing
-a similar process, we should ref it ass to why OSM data is not directly suitable}
+a similar process, we should ref it as to why OSM data is not directly suitable}
 
@@
 
-AORTA uses microscopic simulation, modeling individual drivers as agents.
-Time is modeled discretely, meaning an agent accelerates at a fixed rate for
-the duration of a ``step'' to achieve a new position and velocity. Space is
-continuous, meaning agents occupy a moving interval of a lane, rather than one
-fixed tile of the road. Each step lasts the same fixed duration \textit{dt}.
-This lets agents reason about how far they will travel for a given choice of
-acceleration before being able to choose a different acceleration. Setting
-\textit{dt} to the ``reaction time'' of a human driver may be appropriate for
-experiments modeling humans.
+Different elements of the simulator are described as follows:
 
-Each step, the simulation does the following:
+\subsection{Simulation Dynamics}
+
+AORTA uses microscopic simulation, modeling individual drivers as point agents
+following a simple acceleration-based dynamics model \commentpk{need ideas on
+framing the last sentence better}.  Time is modeled discretely, and an
+agent accelerates at a fixed rate for the duration of a ``step'' to achieve a
+new position and velocity. Space is continuous, meaning agents occupy a moving
+interval of a lane, rather than one fixed tile of the road. Each time step
+lasts for the same fixed duration, and the simulation does the following after
+each time step:
+
+% \textit{dt}.
+% This lets agents reason about how far they will travel for a given choice of
+% acceleration before being able to choose a different acceleration. Setting
+% \textit{dt} to the ``reaction time'' of a human driver may be appropriate for
+% experiments modeling humans.
+% 
+% Each step, the simulation does the following:
 
@@
 
-Each step (except 2) will be described below, followed by limitations of the
-simulator. Section \ref{sec:config} will describe how researchers may easily
-modify relevant portions of AORTA.
+\commentpk{need to check the ordering of above enumeration}
 
@@
 if it is in a driveway or parking lot). When the lane has no agents that could
-potentially crash with the new vehicle, the new agent enters the system with
-zero initial speed. % Because behaviors must enforce speed limit along a lane
+potentially crash with the new vehicle, the new agent enters the system. % with
+%zero initial speed. Because behaviors must enforce speed limit along a lane
 %from the moment an agent enters a road, it is safe for a new agent to enter a
@@
 then the turns are always in conflict, no matter where the agents actually are
-along that turn. This is visualized in figure \ref{fig:conflicts}. The
-alternative of tiling intersections \cite{JAIR08-dresner} would provide more
-granularity, but it adds computational complexity and would be impractical for
-many of the complex intersection geometries inferred from OSM.
+along that turn. This is visualized in figure \ref{fig:conflicts}. Further work
+is required before higher granularity can be introduced through space-time
+tiling intersections \cite{JAIR08-dresner}, due to some of the complex
+intersection geometries inferred from OSM \commentpk{If there is an image for
+one such complex intersection geometry, we can place it next to the current fig
+3 using a subfloat}.
 
-When a collision occurs, the simulator throws warnings and halts, since this
-indicates a bug in some policy. A future alternative could be to mark the region
-as impassable and re-route other traffic away from the accident.
+% When a collision occurs, the simulator throws warnings and halts, since this
+% indicates a bug in some policy. A future alternative could be to mark the region
+% as impassable and re-route other traffic away from the accident.
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{EXTENDING AORTA}
+\subsection{Extending AORTA}
 \label{sec:config}
@@
 strategies, and intersection policies.
-
-\subsection{Agent Behaviors}
 
@@
 
-\subsection{Intersection Policies}
-
 Intersection policies are one of the most intriguing aspects of the simulation
@@
 timings and refinements to the autonomous reservation policy.
-
-\subsection{Routing Strategies}
 

### 2012-03-15 papers/itsc_2012/final.tex
added some of dr chius comments and responded to a few

@@
 driverless cars provides opportunities for optimizing traffic in ways not
-possible before. This paper introduces an open source multi-agent microscopic
+possible before. This paper introduces an open source multiagent microscopic
 traffic simulator called AORTA, which stands for \textit{Approximately
@@
 possible for human drivers. 
+\commentd{If we need space, I agree with Dr Chiu, some of this could be trimmed}
 
@@
 
-This paper introduces an open source multi-agent microscopic traffic simulator
+This paper introduces an open source multigent microscopic traffic simulator
 called AORTA, which stands for \textit{Approximately Orchestrated Road Traffic
 Automata}, a platform designed for testing autonomous vehicle behaviors and
-intersection policies. Autonomous vehicles, termed \textit{agents}, use
+intersection policies. Autonomous vehicles, termed agents, use
 behaviors to interact with one another, follow intersection policies, and
 decide on both long term and short term actions. AORTA allows for the creation
-of intersection policies on top of a reservation based intersection manager
-\cite{JAIR08-dresner}, and these policies designate when agents can cross the
+of intersection policies that designate when agents can cross the
 intersection. AORTA's goal is to allow for the definition of new agent
@@
 intersections, AORTA could potentially simulate human traffic.
+\commentd{Previously, ``AORTA allows for the creation
+of intersection policies on top of a reservation based intersection manager
+\cite{JAIR08-dresner}, and these policies designate when agents can cross the
+intersection.'' Actually, reservations are just one policy. Do we want to focus
+on the general case here, or autonomous specifically?}
 
@@
 simulations. One of AORTA's key features is that it generates maps using real
-road data available from OpenStreetMap (OSM) \cite{osm}, a Wikipedia-like
-interface for world maps. A map for any desired city in the world can be
-downloaded, which is then parsed by AORTA to set up a scale simulation of the
-real world in a few minutes. AORTA is available open-source and is easily
+road data available from OpenStreetMap (OSM) \cite{osm}, a moderated,
+user-editable interface for world maps. A map for any desired city in the world
+can be downloaded, which is then parsed by AORTA to set up a scale simulation of
+the real world in a few minutes. AORTA is available open-source and is easily
 extensible \footnote{Code available at
@@
 greater) \cite{nagel1994microscopic,shen2011agent}. As a result, a number of
-multi-agent micro-simulators have been introduced in the past decade. We review
+multiagent micro-simulators have been introduced in the past decade. We review
 relevant simulators and other related work in this section.
@@
 \cite{JAIR08-dresner}.
+\commentd{Again, ``reservation'' feels like an overloaded term here, meaning the
+general notion of being able to cross, as opposed to the specific policy.}
 
@@
 a similar process, we should ref it as to why OSM data is not directly suitable}
+\commentd{We cite it in the conclusion because they had similar critiques about
+  OSM. I don't think we have a source for why OSM isn't directly suitable,
+  although that source also describes a conversion process.}
 
@@
 \commentc{What about parking lots?}
+\commentd{They're not something relevant for us to model. I don't think we need
+to mention this.}
 
@@
 of one lane to the start of another. \commentc{Illustrate through a figure?}
+\commentd{maybe a subfloat by the turn conflicts, sure}
 
@@
 
+\commentd{Is this incomplete?}
 Different elements of the simulator are described as follows:
@@
 AORTA uses microscopic simulation, modeling individual drivers as point agents
-following a simple acceleration-based dynamics model \commentpk{need ideas on
-framing the last sentence better}.  Time is modeled discretely, and an
-agent accelerates at a fixed rate for the duration of a ``step'' to achieve a
-new position and velocity. Space is continuous, meaning agents occupy a moving
-interval of a lane, rather than one fixed tile of the road. Each time step
-lasts for the same fixed duration, and the simulation does the following after
-each time step:
+with a simple acceleration-based dynamics model. Time is modeled discretely,
+and an agent accelerates at a fixed rate for the duration of a ``step'' to
+achieve a new position and velocity. Space is continuous, meaning agents occupy
+a moving interval of a lane, rather than one fixed tile of the road. Each time
+step lasts for the same fixed duration, and the simulation does the following
+for each time step:
 
@@
 
+\commentd{Mike, I left off here.}
+
 Each agent has a \emph{behavior} governing it, responsible for obeying
@@
 manner so that the agent at the front of each lane is waiting to perform a turn
-into another full lane with the same such head agent, gridlock \cite{gridlock}
+into another full lane with the same such head agent, gridlock \cite{AAAI11-au}
 occurs; no agent in the system will make progress unless an agent at the front
@@
 a lane adjacent to their original lane. Since these routes are still legal, it
-is desirable to prevent or mend this without forcing different routes. The
-conditions under which every vehicle eventually may proceed are known.
-\cite{AAAI11-au}.
+is desirable to prevent or mend this without forcing different routes. However,
+choosing alternative moves (by changing lanes or rerouting) is one sufficient,
+but not necessarily required, condition for liveness \cite{AAAI11-au}.
+\commentd{I misinterpreted the results of the paper originally. Is this a safer
+and more relevant claim to make?}
 
@@
 
-There are three main components of the simulation, each configurable by
-extending built-in implementations of an interface: agent behaviors, routing
-strategies, and intersection policies.
+There are three main configurable components of the simulation, each with at
+least one existing implementation: agent behaviors, routing strategies, and
+intersection policies.
 
@@
 
-Intersection policies are one of the most intriguing aspects of the simulation
+Intersection policies are one of the most interesting aspects of the simulation
 to tweak, especially in light of autonomous vehicles. Current policies include
@@
 timings and refinements to the autonomous reservation policy.
+\commentc{You can mention about using optimized traffic signal timings generated
+  by commercial traffic signal optimization packages such as SYNCHRO (cite
+  Husch, D. and Albeck, J. (2004) Trafficware Synchro 6 User Guide, TrafficWare,
+  Albany, California.)
+}
 
@@
 any code other than creating a new route implementation. Route policies could
-also interact with a singleton overseer to gain some sort of global insight or
-with smart intersections to avoid congestion.
+also interact with a single central manager to gain some sort of global insight or
+with autonomous intersections to avoid congestion.
+\commentd{I can cite a paper for hierarchial planning. As for why, the 2
+  mentioned are just examples of future work we intend to do, and I want to make
+  the point that implementing both would be easy from a software engineering
+  perspective.}
 
@@
 A particular focus of AORTA is intersection policies. To demonstrate how much
-policy affects agent performance \commentd{performance is a vague term, but I
-don't want to talk about dependent vars just yet}, three policies are evaluated: stop sign,
-traffic signal, and autonomous reservation.
+policy affects the throughput of intersections and total performance of the
+transportation system, three policies are evaluated: stop sign, traffic signal,
+and autonomous reservation. Every intersection uses the same policy.
 
@@
 second, with a uniformly random starting position and goal.
+\commentd{Chiu asked about the distribution of incoming agents. Is ``uniformly
+random start and end'' not clear?}
 
@@
 
-Stop signs accept agents FCFS, refusing agents until they have idled at 0 speed
-for 1 second. This delay approximates human reaction time.
+Stop signs accept agents first-come, first-served, refusing agents until they
+have idled at 0 speed for 1 second. This delay approximates human reaction time.
 
@@
 intersections.
+\commentd{chiu recommends more details with how we propagate timings. Thoughts,
+Mike?}
 
@@
 reservation policy inspired by AIM \cite{JAIR08-dresner} is tested. This policy
-allows new agents to join existing groups of agents if their turn is compatible
-with the others. To prevent one heavy direction of traffic from hogging the
+groups agents with compatible turns together, allowing new agents to join
+existing groups. To prevent one heavy direction of traffic from hogging the
 intersection indefinitely, a timeout preemptively cycles through reservations.
-Ideally this should be based on the relative numbers of agents depending on
-their group of reservations. \commentd{take out the last sentence? all of the
-policies could use some work.}
 
@@
 \end{figure}
+\commentc{This figure only shows the distributions for a particular
+  traffic level only. People would wonder what is the traffic level.
+Actually people are more interested to see how it varies with
+different traffic levels so a few more graphs like this at
+different traffic level is helpful.}
+\commentc{Remove ``In this range'' in the y-label.}
+\commentc{Replace ``lag'' with ``delay'' in the x-label}}
 
@@
 stop signs cause about 7 seconds lag, and traffic signals cause 16 seconds.
+\commentd{We need to be clear that our traffic lights suck, so this number is
+pretty biased.}
+
+\commentc{Reviewers will certainly complain the lack of the details of the
+  parameters you used, which implies your results can be somewhat ``biased''.
+  Honestly, I don't think you should show this results in
+  this paper because it is somewhere irrelevant to the topic.  Perhaps
+  you should emphasize that these experiments intend to show what you can do
+with your simulator.
+}
 
@@
   \caption{Autonomous reservations outperform stop signs, and traffic signal
-           gridlocked.}
+           gridlocked. \commentc{This label is not helpful. Clarify ``active'' in y label.}}
   \label{fig:agent_cnt}
@@
 
-The number of agents changes because the continuous generator introduces a new
-agent every second. This count reaches a steady state when the generator
-introduces an agent at the same rate as older agents finish their route. Figure
-\ref{fig:agent_cnt} reveals that autonomous reservations let agents reach their
-destination twice as quickly as the competition, with a steady-state of about
-400 agents. The count for traffic signals continues to increase because gridlock
-occurred in one segment of the map, causing agents to not finish their route.
-Unexpected starvation was also observed: although the signals cycle through
-turns regularly, agents cannot proceed when the previous cycle caused a
-destination lane to fill and become blocked.
+The number of agents driving at some time changes because the continuous
+generator introduces a new agent every second. This count reaches a steady
+state when the generator introduces an agent at the same rate as older agents
+finish their route. Figure \ref{fig:agent_cnt} reveals that autonomous
+reservations let agents reach their destination twice as quickly as the
+competition, with a steady-state of about 400 agents. The count for traffic
+signals continues to increase because gridlock occurred in one segment of the
+map, causing agents to not finish their route.  Unexpected starvation was also
+observed: although the signals cycle through turns regularly, agents cannot
+proceed when the previous cycle caused a destination lane to fill and become
+blocked.
 
 Each policy has its trade-offs in terms of performance and physical hardware
-cost \commentd{do we want to raise this issue?}, suggesting a mixed approach,
-perhaps with intersection policies adapting as traffic conditions shift.
-Researchers could conduct this experiment and many others without needing to
-understand most of AORTA's code-base. The largest intersection policy, traffic
-signals with the flooding heuristic for timing assignment, is only 400 lines of
-Scala.
+cost \commentd{do we want to raise this issue?}, suggesting an approach that
+controls different intersections with different polices, perhaps adapting
+policies adapting as traffic conditions shift.  Researchers could conduct this
+experiment and many others without needing to understand most of AORTA's
+code-base. The largest intersection policy, traffic signals with the flooding
+heuristic for timing assignment, is only 400 lines of code in Scala.
 
@@
 
-AORTA's map construction process, although flexible, is fraught with heuristics,
-motivating proposed changes to OSM. Road type annotations imprecisely estimate
+AORTA's map construction process, although flexible, cannot completely account
+for incompletedness in OSM data.  Road type annotations imprecisely estimate
 the number of lanes and typical speed limits, resulting in incorrect graphs.
 Guessing the turns at each intersection makes common road features like right
-turn-only lanes and shared center left turn lanes hard to detect. There are many
-circumstances where AORTA simulates one intersection as several in close
+turn-only lanes and shared center left turn lanes hard to detect. There are
+many circumstances where AORTA simulates one intersection as several in close
 proximity because roads do not geometrically line up, but semantically they do.

### 2012-03-15 papers/itsc_2012/final.tex
Confirmed comparisons to matSIM and SUMO

--- 2012-03-15 papers/itsc_2012/final.tex
@@
 supplied to MATSim in the form of \textit{plans} that an individual may intend
-to follow in a given day. MATSim aims to iteratively improve these plans though
-offline computation to maximize traffic throughput. In other words, it helps
-individuals understand the best way to go about their daily routine to minimize
-time spent in transit. This goal is somewhat different from that of AORTA,
-where the demand supplied by individuals is taken as is. Rather than simply
-improving routes, AORTA allows creating entirely different behaviors for both
-vehicles and intersections.
+to follow in a given day. MATSim aims to iteratively improve these plans to
+minimize traffic delays while minimizing the total amount of time by which agents
+are late to their destinations. This goal is somewhat different from that of AORTA,
+where the routes (demand) supplied by individuals is taken as is and does not have
+a pre-determined ideal time of arrival. Rather than improving agent routes, AORTA
+allows creating entirely different behaviors for both vehicles and intersections
+in order to simulate a different environment instead of optimizing plans for a
+society in a static environment.
 
 Another popular open source traffic simulator that can use OSM data is SUMO
-\cite{SUMO2011}, which shares the similar goals as AORTA. SUMO has been used to
+\cite{SUMO2011}, which shares many similar goals with AORTA. SUMO has been used to
 study the effect of automated transportation systems, route planning of
@@
 \cite{JAIR08-dresner}.
+\commentm{Given that SUMO is also open-source and seems to be used fairly commonly
+in the traffic community, our only real advantage over it is actually our choice
+of language.  AORTA is much easier to use and extend primarily because we use
+a higher level language.  (As we mention, our intersections are also more general,
+but I think that's actually less important.)  I'm not sure of how to handle this.}
 

### 2012-03-15 papers/itsc_2012/final.tex

--- 2012-03-15 papers/itsc_2012/final.tex
@@
 \cite{JAIR08-dresner}.
+\commentm{Given that SUMO is also open-source and seems to be used fairly commonly
+in the traffic community, our only real advantage over it is actually our choice
+of language.  AORTA is much easier to use and extend primarily because we use
+a higher level language.  (As we mention, our intersections are also more general,
+but I think that's actually less important.)  I'm not sure of how to handle this.}
 \commentd{Again, ``reservation'' feels like an overloaded term here, meaning the
@@
 \commentc{Remove ``In this range'' in the y-label.}
-\commentc{Replace ``lag'' with ``delay'' in the x-label}}
+\commentc{Replace ``lag'' with ``delay'' in the x-label}
 

### 2012-03-15 papers/itsc_2012/final.tex
crazy intersection example added

@@
 \usepackage[pdftex]{graphicx}
+\usepackage{subfig}
 % TODO remember we need to meet color compliance:
@@
 a fixed width. Turns are modeled with a single straight line connecting the end
-of one lane to the start of another. \commentc{Illustrate through a figure?}
-\commentd{maybe a subfloat by the turn conflicts, sure}
+of one lane to the start of another.
 
@@
 tiling intersections \cite{JAIR08-dresner}, due to some of the complex
-intersection geometries inferred from OSM \commentpk{If there is an image for
-one such complex intersection geometry, we can place it next to the current fig
-3 using a subfloat}.
+intersection geometries inferred from OSM. An example of such a difficult
+intersection is shown in figure \ref{fig:crazyIntersection}.
 
@@
 \begin{figure}[h]
-  \centering \includegraphics[scale=0.5]{turn_conflicts.png}
-  \caption{An agent cannot turn left while agents on the perpendicular road cross}
+  \centering
+  \subfloat[An agent cannot turn left while agents on the perpendicular road
+  cross]{
+    \includegraphics[scale=0.4]{turn_conflicts.png}
     \label{fig:conflicts}
+  }
+  ~ % horizontal spacing, mostly for the sake of the captions
+  \subfloat[A difficult intersection to tile in New Orleans, Louisiana. Note
+  there are no overpasses pictured.]{
+    \includegraphics[scale=0.4]{crazy_intersection.png}
+    \label{fig:crazyIntersection}
+  }
+  \caption{Examples of AORTA's modeling of intersections and turns}
   \vspace{-15pt}

### 2012-03-15 process_stats.py
fixed 2 graphs per suggestion and thickened lines

@@ -53,14 +53,14 @@ colors = ['b', 'r', 'g']
 
 def hgram(transform, data, xlabel, yobject):
   pylab.xlabel(xlabel, fontsize='x-large')
-  pylab.ylabel("Number of " + yobject + " in this range", fontsize='x-large')
+  pylab.ylabel("Number of " + yobject, fontsize='x-large')
   alpha = 1
   alpha_step = 0.9 / len(data)
   color = iter(colors)
   for exp, dat in data.items():
     ls = transform(dat)
     pylab.hist(ls, bins=50, histtype='stepfilled', label=exp, alpha=alpha,
-               color=next(color), range=(0, 25))
+               color=next(color))
     alpha -= alpha_step
     print exp + ": " + xlabel
     print "-" * len(xlabel)
@@ -79,9 +79,9 @@ def average(dat):
   return [np.mean(ls) for ls in dat.values()]
 
 # Results from s1
-hgram(average, lag_per_a, "Average lag per agent (seconds)", "agents")
-hgram(average, lag_per_i, "Average lag per intersection (seconds)", "intersections")
-hgram(identity, all_lags, "Average lag (seconds)", "events")
+hgram(average, lag_per_a, "Average delay per agent (seconds)", "agents")
+hgram(average, lag_per_i, "Average delay per intersection (seconds)", "intersections")
+hgram(identity, all_lags, "Average delay (seconds)", "events")
 # Results from s2
 hgram(identity, avg_speed, "Average speed (m/s) of each agent", "agents")
 # Results from s3
@@ -90,8 +90,8 @@ hgram(average, thruput_per_i,
       "intersections")
 # Results from s4
 pylab.xlabel("Time (seconds)", fontsize='x-large')
-pylab.ylabel("Number of active agents", fontsize='x-large')
+pylab.ylabel("Number of agents in the map", fontsize='x-large')
 for exp in experiments:
-  pylab.plot(active_per_time_x[exp], active_per_time_y[exp], label=exp)
+  pylab.plot(active_per_time_x[exp], active_per_time_y[exp], label=exp, linewidth=6)
 pylab.legend()
 pylab.show()

### 2012-03-15 papers/itsc_2012/final.tex
fixed 2 graphs per suggestion and thickened lines
tweaks to abstract, intro, and sect 5 up till agent behaviors

@@
 \usepackage[pdftex]{graphicx}
-\usepackage{subfig}
+\usepackage[font={footnotesize}]{caption,subfig}
 % TODO remember we need to meet color compliance:
@@
 \title{\LARGE \bf
-Approximately Orchestrated Road Traffic Automata:\\
+Approximately Orchestrated Routing and Transportation Analyzer:\\
 Large-scale Traffic Simulation for Autonomous Vehicles
@@
 traffic simulator called AORTA, which stands for \textit{Approximately
-Orchestrated Road Traffic Automata}, designed for optimizing autonomous traffic
-at a city-wide scale. AORTA creates scale simulations of the real world by
-generating maps using publicly available road data from \textit{OpenStreetMap}
-(OSM). This allows simulations to be set up through AORTA for a desired region
-anywhere in the world in a matter of minutes. AORTA allows for optimizing
-traffic by creating more intelligent behaviors for individual drivers (agents),
-as well as by defining new intersection policies that can be followed by these
-autonomous agents. These agent behaviors define how agents interact with one
-another, follow intersection policies, and select the next action to take. This
-paper demonstrates a simple applications using AORTA through an experiment
-testing an intersection policy designed for autonomous vehicles at a city-wide
-scale.
+Orchestrated Routing and Transportation Analyzer}, designed for optimizing
+autonomous traffic at a city-wide scale. AORTA creates scale simulations of the
+real world by generating maps using publicly available road data from
+\textit{OpenStreetMap} (OSM). This allows simulations to be set up through AORTA
+for a desired region anywhere in the world in a matter of minutes. AORTA allows
+for traffic optimization by creating intelligent behaviors for individual driver
+agents and intersection policies to be followed by these agents. These behaviors
+and policies define how agents interact with one another, control when they
+cross intersections, and route agents to their destination. This paper
+demonstrates a simple application using AORTA through an experiment testing an
+intersection policy designed for autonomous vehicles at a city-wide scale.
 
@@
 
-% theme: easy simulation in your city, in 2 minutes!
-% be sure to emphasize open source!
-% contribution: autonomous car / human car sim on EXISTING INFRASTRUCTURE
+% TODO theme: easy simulation in your city, in 2 minutes!
 
@@
 competing teams completed the 96km course set for the DARPA Urban Challenge
-\cite{spectrumUrbanChallenge2007}. They did so while following the same traffic
-laws followed by human drivers, navigating along with other moving vehicles,
-and following correct intersection precedence order. Since then, Google's
-driverless cars have clocked more than 250,000km on public roads in urban
-California, USA \cite{tedThrun2011}. In 2010, researchers from the University
-of Parma successfully completed an autonomous \textit{intercontinental} run
-from Parma, Italy to Shanghai, China \cite{cnnVislab2010}. The successful
-completion of all these milestones suggests that autonomous cars are here to
-stay, and are ever closer to becoming commercially available. With the arrival
-of autonomous cars, it also becomes possible to optimize traffic in ways not
-possible for human drivers. 
-\commentd{If we need space, I agree with Dr Chiu, some of this could be trimmed}
+\cite{spectrumUrbanChallenge2007}. They did so while obeying the traffic laws
+followed by human drivers, navigating along with other moving vehicles, and
+following correct intersection precedence order. Since then, Google's driverless
+cars have clocked more than 250,000km on public roads in urban California, USA
+\cite{tedThrun2011}. In 2010, researchers from the University of Parma
+successfully completed an autonomous \textit{intercontinental} run from Parma,
+Italy to Shanghai, China \cite{cnnVislab2010}. The successful completion of all
+these milestones suggests that autonomous cars are here to stay, and are ever
+closer to becoming commercially available. With the arrival of autonomous cars,
+it also becomes possible to optimize traffic in ways not possible for human
+drivers. 
+%\commentd{If we need space, I agree with Dr Chiu, some of this could be trimmed}
 
@@
 This paper introduces an open source multigent microscopic traffic simulator
-called AORTA, which stands for \textit{Approximately Orchestrated Road Traffic
-Automata}, a platform designed for testing autonomous vehicle behaviors and
-intersection policies. Autonomous vehicles, termed agents, use
-behaviors to interact with one another, follow intersection policies, and
-decide on both long term and short term actions. AORTA allows for the creation
-of intersection policies that designate when agents can cross the
-intersection. AORTA's goal is to allow for the definition of new agent
-behaviors and intersection policies to optimize autonomous traffic.
-Additionally, by assigning human-like behaviors such as the ``car following
-model'' \cite{brackstone1999car} to agents, and traffic-signal-like policies to
+called AORTA, which stands for \textit{Approximately Orchestrated Routing and
+Transportation Analyzer}, a platform designed for testing autonomous vehicle
+behaviors and intersection policies. Autonomous vehicles, termed agents, use
+behaviors to interact with one another, follow intersection policies, and decide
+on both long term and short term actions. AORTA allows for the creation of
+intersection policies that designate when agents can cross the intersection.
+AORTA's goal is to allow for the definition of new agent behaviors and
+intersection policies to optimize autonomous traffic.  Additionally, by
+assigning human-like behaviors such as the ``car following model''
+\cite{brackstone1999car} to agents, and traffic-signal-like policies to
 intersections, AORTA could potentially simulate human traffic.
-\commentd{Previously, ``AORTA allows for the creation
-of intersection policies on top of a reservation based intersection manager
-\cite{JAIR08-dresner}, and these policies designate when agents can cross the
-intersection.'' Actually, reservations are just one policy. Do we want to focus
-on the general case here, or autonomous specifically?}
+\commentd{Previously, ``AORTA allows for the creation of intersection policies
+  on top of a reservation based intersection manager \cite{JAIR08-dresner}, and
+  these policies designate when agents can cross the intersection.'' Actually,
+  reservations are just one policy. Do we want to focus on the general case
+here, or autonomous specifically?}
 
@@
 
-\commentd{Is this incomplete?}
-Different elements of the simulator are described as follows:
+This section describes the simulator's agent model and its mechanisms for
+creating agents, verifying safety, and allowing agents and intersections to
+control themselves. Comments on current limitations and how to extend the
+simulator follow.
 
@@
 
-% \textit{dt}.
-% This lets agents reason about how far they will travel for a given choice of
-% acceleration before being able to choose a different acceleration. Setting
-% \textit{dt} to the ``reaction time'' of a human driver may be appropriate for
-% experiments modeling humans.
-% 
-% Each step, the simulation does the following:
-
 \begin{enumerate}
@@
 \end{enumerate}
-
-\commentpk{need to check the ordering of above enumeration}
 
@@
 
-Every step, a generator creates some number of new agents. If the agent's route
-policy needs computation like pathfinding to operate, the generator either
-performs it immediately (incurring a delay in simulation if it is running) or,
-if possible, sends it to a pool of worker threads to process in the background.
-Once an agent's route is ready, the agent waits alongside its starting lane (as
-if it is in a driveway or parking lot). When the lane has no agents that could
-potentially crash with the new vehicle, the new agent enters the system. % with
-%zero initial speed. Because behaviors must enforce speed limit along a lane
+Every step, a generator creates some number of new agents. The generator either
+immediately performs any timely computation required by the agent's route
+policy, such as pathfinding, or delegates it to a pool of worker threads to
+process in the background.  Once an agent's route is ready, the agent waits
+alongside its starting lane (as if it is in a driveway or parking lot). When the
+lane has no agents that could potentially crash with the new vehicle, the new
+agent enters the system.
+%with zero initial speed. Because behaviors must enforce speed limit along a lane
 %from the moment an agent enters a road, it is safe for a new agent to enter a
@@
 
-All traversables (lanes and the straight line-approximation of turns) maintain
-an ordered queue of occupying agents. A collision is detected when two agents
-reverse order after a time step has completed.
+All traversables (lanes and turns) maintain an ordered queue of occupying
+agents. A collision is detected when two agents reverse order after a time step
+has completed.
 
-Intersections check for collisions by examining the turn each agent is
-performing and verifying that no other agent is simultaneously performing a
-conflicting turn. AORTA has a low-granularity model of turn conflict. If two
-vehicles at any point along two different turns could ever come into contact,
-then the turns are always in conflict, no matter where the agents actually are
-along that turn. This is visualized in figure \ref{fig:conflicts}. Further work
-is required before higher granularity can be introduced through space-time
-tiling intersections \cite{JAIR08-dresner}, due to some of the complex
-intersection geometries inferred from OSM. An example of such a difficult
+Intersections check for collisions by verifying no two agents are simultaenously
+performing a conflicting turn. AORTA has a low-granularity model of turn
+conflict. If two vehicles at any point along two different turns could ever come
+into contact, then the turns are always in conflict, no matter where the agents
+actually are along that turn. This is visualized in figure \ref{fig:conflicts}.
+Further work is required before higher granularity can be introduced through
+space-time tiling intersections \cite{JAIR08-dresner}, due to some of the
+complex intersection geometries inferred from OSM. An example of such a complex
 intersection is shown in figure \ref{fig:crazyIntersection}.
@@
 different traffic level is helpful.}
-\commentc{Remove ``In this range'' in the y-label.}
-\commentc{Replace ``lag'' with ``delay'' in the x-label}
 
@@
   \caption{Autonomous reservations outperform stop signs, and traffic signal
-           gridlocked. \commentc{This label is not helpful. Clarify ``active'' in y label.}}
+           gridlocked.}
   \label{fig:agent_cnt}

### 2012-03-15 papers/itsc_2012/final.tex
revised usage of 'reservation' and middle of section 5

@@
 intersections, AORTA could potentially simulate human traffic.
-\commentd{Previously, ``AORTA allows for the creation of intersection policies
-  on top of a reservation based intersection manager \cite{JAIR08-dresner}, and
-  these policies designate when agents can cross the intersection.'' Actually,
-  reservations are just one policy. Do we want to focus on the general case
-here, or autonomous specifically?}
 
@@
 intersections are handled. SUMO uses traditional traffic signals, while AORTA
-adopts a much more general \textit{reservation} system. A reservation tells a
-car when it is safe to take its desired path through the intersection.
-Reservations can be allotted to vehicles through more traditional means such as
-traffic signals or stop-sign based intersection precedence, or through more
-efficient methods designed specifically for autonomous vehicles
-\cite{JAIR08-dresner}.
-\commentm{Given that SUMO is also open-source and seems to be used fairly commonly
-in the traffic community, our only real advantage over it is actually our choice
-of language.  AORTA is much easier to use and extend primarily because we use
-a higher level language.  (As we mention, our intersections are also more general,
-but I think that's actually less important.)  I'm not sure of how to handle this.}
-\commentd{Again, ``reservation'' feels like an overloaded term here, meaning the
-general notion of being able to cross, as opposed to the specific policy.}
+allows intersections to grant agents entry by any general policy. These policies
+can use traditional schemes such as traffic signals or stop-sign based
+precedence, or more efficient methods designed specifically for autonomous
+vehicles \cite{JAIR08-dresner}.
+\commentd{Are we sure SUMO is not easily configurable in this way?}
 
@@
 
-\commentd{Mike, I left off here.}
-
 Each agent has a \emph{behavior} governing it, responsible for obeying
-intersection policies and avoiding collisions. When the agent travels past the
-end of a lane during its step, the behavior picks the turn to take. At each
-step, the behavior can make the agent react by performing one of two actions:
-disappearing from the map (when the agent is at rest and is done with its route)
-or setting an acceleration for the next step.
+intersection policies and avoiding collisions. Before each step, the behavior
+chooses one of two actions for the agent to perform: disappearing from the map
+(when the agent is at rest and is done with its route) or accelerating at some
+rate. Behaviors query current and upcoming lanes and intersections to find
+other nearby agents and determine all constraints that must be satisfied for a
+safe choice. The behavior also picks turns once an agent reaches the end of a
+lane by consulting a route strategy that finds a path to the destination.
 
@@
 
-A behavior's interaction with an intersection policy amounts to polling it with
-the turn the agent wants to perform, the distance away the agent currently is,
-and how long the agent has been waiting (so policies can enforce a required
-delay before entering the intersection). The intersection policy simply orders
-the agent to ``stop'' or ``proceed.'' \commentd{I'm not sure this fits in the
-default behavior later anymore; just delete it?} The acceleration chosen to
-avoid entering an intersection attempts to account for crosswalks by ending a
-configurable threshold before the end of a traversable.
+A behavior interacts with an intersection policy by sending it the agent's
+desired turn, the agent's speed, the current distance from the start of the
+intersection, and the length of time the agent has been idling at zero speed.
+The policy replies by ordering the agent to ``stop'' or ``proceed''. The
+behavior will continue to statelessly poll the policy every step until the agent
+begins the turn. The policy is responsible for never changing a reply of
+approval to denial when the agent could not safely honor the change. The policy
+is free to allow concurrent turns in any arrangement by different agents as long
+as no two simultaneous turns conflict.
 

### 2012-03-15 papers/itsc_2012/final.tex
added citation to hpa

@@
 behavior will continue to statelessly poll the policy every step until the agent
-begins the turn. The policy is responsible for never changing a reply of
-approval to denial when the agent could not safely honor the change. The policy
-is free to allow concurrent turns in any arrangement by different agents as long
-as no two simultaneous turns conflict.
+begins the turn. The policy may approve an agent for entry one step and deny it
+the next, so long as the agent could safely brake without danger of entering the
+intersection. The policy is free to allow concurrent turns in any arrangement
+by different agents as long as no two simultaneous turns conflict.
 
-\subsection{Simulation Limitations}
+\subsection{Current Limitations}
 
@@
 timings and refinements to the autonomous reservation policy.
-\commentc{You can mention about using optimized traffic signal timings generated
-  by commercial traffic signal optimization packages such as SYNCHRO (cite
-  Husch, D. and Albeck, J. (2004) Trafficware Synchro 6 User Guide, TrafficWare,
-  Albany, California.)
-}
 
@@
 drunken walk that probabilistically moves towards the destination. Researchers
-can experiment with dynamic replanning or hierarchial planning without modifying
-any code other than creating a new route implementation. Route policies could
-also interact with a single central manager to gain some sort of global insight or
-with autonomous intersections to avoid congestion.
-\commentd{I can cite a paper for hierarchial planning. As for why, the 2
-  mentioned are just examples of future work we intend to do, and I want to make
-  the point that implementing both would be easy from a software engineering
-  perspective.}
+can experiment with dynamic replanning or hierarchial planning
+\cite{Botea04nearoptima} without modifying any code other than creating a new
+route implementation. Route policies could also interact with a single central
+manager to gain some sort of global insight or with autonomous intersections to
+avoid congestion.
 
@@
 Mike?}
+\commentc{You can mention about using optimized traffic signal timings generated
+  by commercial traffic signal optimization packages such as SYNCHRO (cite
+  Husch, D. and Albeck, J. (2004) Trafficware Synchro 6 User Guide, TrafficWare,
+  Albany, California.)
+}
 

### 2012-03-15 papers/itsc_2012/final.tex
misc tweaks

@@
 agent enters the system.
-%with zero initial speed. Because behaviors must enforce speed limit along a lane
-%from the moment an agent enters a road, it is safe for a new agent to enter a
-%road if there is no other agent within the worst-case distance (traveling
-%distance at the speed limit, plus stopping distance from this speed) preceeding
-%the spawn point. To avoid predicting which agents might enter the lane in the
-%near future, there is a buffer space at the beginning of the lane where a new
-%agent may never spawn. This disqualifies some short lanes from being spawn
-%points, but this is not a big limitation in practice. \commentpk{I understand
-%your concern about explaining the entire methodology, but the last 2 lines are
-%unnecessary and somewhat distracting}
 
@@
 desired turn, the agent's speed, the current distance from the start of the
-intersection, and the length of time the agent has been idling at zero speed.
-The policy replies by ordering the agent to ``stop'' or ``proceed''. The
-behavior will continue to statelessly poll the policy every step until the agent
-begins the turn. The policy may approve an agent for entry one step and deny it
-the next, so long as the agent could safely brake without danger of entering the
-intersection. The policy is free to allow concurrent turns in any arrangement
-by different agents as long as no two simultaneous turns conflict.
+intersection, and the length of time the agent has been idling at rest. The
+policy replies by ordering the agent to ``stop'' or ``proceed''. The behavior
+will continue to statelessly poll the policy every step until the agent begins
+the turn. The policy may approve an agent for entry one step and deny it the
+next, so long as the agent could safely brake without danger of entering the
+intersection. The policy is free to allow concurrent turns in any arrangement by
+different agents as long as no two simultaneous turns conflict.
 
@@
 a lane adjacent to their original lane. Since these routes are still legal, it
-is desirable to prevent or mend this without forcing different routes. However,
-choosing alternative moves (by changing lanes or rerouting) is one sufficient,
-but not necessarily required, condition for liveness \cite{AAAI11-au}.
-\commentd{I misinterpreted the results of the paper originally. Is this a safer
-and more relevant claim to make?}
+is desirable to prevent or mend this without forcing different routes.
+However, choosing alternative moves (by changing lanes or rerouting) is an
+important constituent of a sufficient condition for the liveness of a road
+network cite{AAAI11-au}.
 
@@
 
-A particular focus of AORTA is intersection policies. To demonstrate how much
+A particular focus of AORTA is intersection policies. To demonstrate how
 policy affects the throughput of intersections and total performance of the
@@
 
-The experiment simulates a 6 by 7 mile slice of downtown Austin,
+The experiment simulates a 9$\times$11km$^{2}$ slice of downtown Austin,
 Texas\footnote{bounded by the Mopac and Springdale Road horizontally, and Oltorf
@@
 second, with a uniformly random starting position and goal.
-\commentd{Chiu asked about the distribution of incoming agents. Is ``uniformly
-random start and end'' not clear?}
 
@@
 Stop signs accept agents first-come, first-served, refusing agents until they
-have idled at 0 speed for 1 second. This delay approximates human reaction time.
+have idled at rest for 1 second. This delay approximates human reaction time.
 
 The traffic signal policy requires timings and groups of simultaneous turns for
-each intersection. The heuristic evaluated uses standard turn groups (turns from
-parallel incoming directions are scheduled simultaneously, and left turns are
-grouped) and employs breadth-first search to propagate timings. The goal is to
-synchronize green signals so agents in some lanes do not stop for several
-intersections.
-\commentd{chiu recommends more details with how we propagate timings. Thoughts,
-Mike?}
-\commentc{You can mention about using optimized traffic signal timings generated
-  by commercial traffic signal optimization packages such as SYNCHRO (cite
-  Husch, D. and Albeck, J. (2004) Trafficware Synchro 6 User Guide, TrafficWare,
-  Albany, California.)
-}
+each intersection. The heuristic currently assigning these arbitrarily picks one
+major road, schedules turns from that road to others, then repeats the process
+from all reachable roads using a breadth-first search. By estimating how long
+performing one turn and traveling along the next lane takes, the timing of the
+next intersection can be scheduled so that agents from the first road continue
+traveling without stopping for several consecutive intersections. This heuristic
+is far from optimal, meaning the performance of traffic signals described below
+could be improved, perhaps by incorporating packages such as SYNCHRO
+\cite{synchro}.
 
-\commentd{advice on wording?}
-Finally, in anticipation of the advent of autonomous vehicles, an autonomous
-reservation policy inspired by AIM \cite{JAIR08-dresner} is tested. This policy
-groups agents with compatible turns together, allowing new agents to join
-existing groups. To prevent one heavy direction of traffic from hogging the
-intersection indefinitely, a timeout preemptively cycles through reservations.
+The last policy tested is an autonomous reservation policy inspired by AIM
+\cite{JAIR08-dresner}. This policy groups agents with compatible turns together,
+allowing new agents to join existing groups. To prevent one heavy direction of
+traffic from hogging the intersection indefinitely, a timeout preemptively
+cycles through reservations.
 
@@
 reservation policies cause less than 1 second of delay at each intersection,
-stop signs cause about 7 seconds lag, and traffic signals cause 16 seconds.
-\commentd{We need to be clear that our traffic lights suck, so this number is
-pretty biased.}
+stop signs cause about 7 seconds lag, and traffic signals using the heuristic
+described above cause 16 seconds. These results will vary with different
+traffic levels and policy configurations, so the important demonstration is that
+AORTA can be used as a framework for evaluating these policies.
 

### 2012-03-15 papers/itsc_2012/final.tex
cite typo

@@
 important constituent of a sufficient condition for the liveness of a road
-network cite{AAAI11-au}.
+network \cite{AAAI11-au}.
 

### 2012-03-15 papers/itsc_2012/final.tex
rephrasing conclusion and clarifying purpose of results

@@
 graphs, then answers pathfinding and geometry queries. The simulation engine
-adds a notion of agents, vehicle dynamics, and collisions. Finally, the UI can
-interactively render the map and agents. A headless mode also enables running
+adds a notion of agents, vehicle dynamics, and collisions. Finally, the UI
+interactively renders the map and agents. A headless mode also exists to run
 experiments without the overhead of visualization. The interaction between
@@
 suitable for running a simulation and is pre-processed through a \emph{map
-builder} utility. The map model generated by the builder is described in sec.
-\ref{sec:mapmodel} and the process of refining OSM data is enumerated in sec.
-\ref{sec:mapconstruction}.
+builder} utility. Section \ref{sec:mapmodel} describes the map model generated
+by the builder, and section \ref{sec:mapconstruction} describes the process of
+refining OSM data.
 
@@
 \label{sec:mapmodel}
-
-\commentpk{I remember you guys mentioning there was a paper for SUMO describing
-a similar process, we should ref it as to why OSM data is not directly suitable}
-\commentd{We cite it in the conclusion because they had similar critiques about
-  OSM. I don't think we have a source for why OSM isn't directly suitable,
-  although that source also describes a conversion process.}
-
-% \pix{map_model.png}
-%    {Roads contain several lanes, and intersections contain several turns
-%     between lanes.}
-%    {scale=0.5}
 
@@
 are grouped together as \emph{traversables}, meaning agents may exist on them.
-\commentc{What about parking lots?}
-\commentd{They're not something relevant for us to model. I don't think we need
-to mention this.}
 
@@
         a left, right, straight, or U-turn. When several lanes all cross into fewer
-        lanes, the builder forces merging of the rightmost lanes. 
+        lanes, the builder forces merging at the rightmost lanes. 
   \item Because turns are heuristically placed, the graph is often disconnected.
@@
         digraph. Disconnected lanes are spliced out of larger roads, leaving
-        unappealing geometric gaps, but correctly ensuring connectivity for
-        corner-case free pathfinding.
+        geometric gaps, but correctly ensuring connectivity.
 \end{enumerate}
@@
 
-This section describes the simulator's agent model and its mechanisms for
+This section describes AORTA's agent model and the simulator's mechanisms for
 creating agents, verifying safety, and allowing agents and intersections to
@@
 the turn. The policy may approve an agent for entry one step and deny it the
-next, so long as the agent could safely brake without danger of entering the
+next, as long as the agent could safely brake without danger of entering the
 intersection. The policy is free to allow concurrent turns in any arrangement by
@@
 
-Agents maintain a safe following distance from the next agent. If a lane is
+Each agent maintains a safe following distance from the next. If a lane is
 filled, then an agent may be forced to stop in an intersection mid-turn and
@@
 into another full lane with the same such head agent, gridlock \cite{AAAI11-au}
-occurs; no agent in the system will make progress unless an agent at the front
-decides to pick a different turn. Figure \ref{fig:gridlock} demonstrates this.
-AORTA can detect this by finding cycles in the graph of the ``is following''
-relation.
+occurs: no agent in the system will make progress unless an agent at the front
+decides to pick a different turn.  AORTA can detect this by finding cycles in
+the graph of the ``is following'' relation.
 
@@
 In many observed cases, the gridlock is caused by AORTA's present lack of
-lane-changing support: agents loop around these intersections in order to enter
-a lane adjacent to their original lane. Since these routes are still legal, it
-is desirable to prevent or mend this without forcing different routes.
-However, choosing alternative moves (by changing lanes or rerouting) is an
-important constituent of a sufficient condition for the liveness of a road
-network \cite{AAAI11-au}.
+lane-changing support: agents loop around intersections such as those pictured
+in figure \ref{fig:gridlock} in order to enter a lane adjacent to their original
+lane.  Since these routes are still legal, it is desirable to prevent or mend
+this problem without forcing different routes. However, choosing alternative
+moves (by changing lanes or rerouting) is an important constituent of a
+sufficient condition for the liveness of a road network \cite{AAAI11-au}.
 
@@
 implementations exist already: a static route using A* \cite{astar} and a
-drunken walk that probabilistically moves towards the destination. Researchers
-can experiment with dynamic replanning or hierarchial planning
-\cite{Botea04nearoptima} without modifying any code other than creating a new
+drunken walk that probabilistically moves towards the destination. Users can
+experiment with dynamic replanning or hierarchial planning
+\cite{Botea04nearoptimal} without modifying any code other than creating a new
 route implementation. Route policies could also interact with a single central
@@
 
-A particular focus of AORTA is intersection policies. To demonstrate how
-policy affects the throughput of intersections and total performance of the
-transportation system, three policies are evaluated: stop sign, traffic signal,
-and autonomous reservation. Every intersection uses the same policy.
+To demonstrate how easily AORTA can be utilized to evaluate different
+intersection policies, this section presents a simple experiment showing how
+different policies affect the throughput of intersections. In this experiment,
+every intersection in the map uses either a stop sign, traffic signal, or
+autonomous reservation policy.
 
@@
 \end{figure}
-\commentc{This figure only shows the distributions for a particular
-  traffic level only. People would wonder what is the traffic level.
-Actually people are more interested to see how it varies with
-different traffic levels so a few more graphs like this at
-different traffic level is helpful.}
 
@@
 AORTA can be used as a framework for evaluating these policies.
-
-\commentc{Reviewers will certainly complain the lack of the details of the
-  parameters you used, which implies your results can be somewhat ``biased''.
-  Honestly, I don't think you should show this results in
-  this paper because it is somewhere irrelevant to the topic.  Perhaps
-  you should emphasize that these experiments intend to show what you can do
-with your simulator.
-}
 
@@
 
-Each policy has its trade-offs in terms of performance and physical hardware
-cost \commentd{do we want to raise this issue?}, suggesting an approach that
-controls different intersections with different polices, perhaps adapting
-policies adapting as traffic conditions shift.  Researchers could conduct this
-experiment and many others without needing to understand most of AORTA's
-code-base. The largest intersection policy, traffic signals with the flooding
-heuristic for timing assignment, is only 400 lines of code in Scala.
+Users could conduct this experiment and many others without needing to
+understand most of AORTA's code-base. The largest intersection policy, traffic
+signals with the flooding heuristic for timing assignment, is only 400 lines of
+code in Scala.
 
@@
 This paper has presented AORTA, a new city-scale traffic simulation framework
-that focuses on ease-of-use and policy configurability. In conjunction with Open
-Street Maps, AORTA allows researchers to repeat an experiment in any number of
-places with minimal effort. Future work will exploit the flexible infrastructure
-of interactions between agent behaviors and intersections by exploring dynamic
-replanning and adaptable signal timings. On top of these structures,
-applications could experiment with dynamic contra-flow \cite{ITSC11-hausknecht}
-or intelligent routing that avoids stopping. There will also be an effort to
-improve AORTA as a framework by including a lane-changing model and preventing
-gridlock. \commentm{Perhaps mention concurency plans (with TACC) to simulate
-arbitrarily large maps?}
+that focuses on configurability. In conjunction with Open Street Maps, AORTA
+allows users to repeat an experiment in different places with minimal effort.
+Future work will exploit the flexible infrastructure of interactions between
+agent behaviors and intersections by exploring dynamic replanning and adaptable
+signal timings. On top of these structures, applications could experiment with
+ideas such dynamic contra-flow \cite{ITSC11-hausknecht}.  There will also be an
+effort to improve AORTA as a framework by including a lane-changing model,
+preventing gridlock, and supporting the resources of supercomputing clusters.
 
 AORTA's map construction process, although flexible, cannot completely account
-for incompletedness in OSM data.  Road type annotations imprecisely estimate
-the number of lanes and typical speed limits, resulting in incorrect graphs.
-Guessing the turns at each intersection makes common road features like right
-turn-only lanes and shared center left turn lanes hard to detect. There are
-many circumstances where AORTA simulates one intersection as several in close
-proximity because roads do not geometrically line up, but semantically they do.
-If OSM could encode this, the performance of intersection policies such as
-traffic signals would vastly improve in those regions. Other simulators have
-identified similar issues before \cite{Krajzewicz_Hertkorn_Ringel_Wagner_2005}.
+for incompletedness in OSM data. Road type annotations imprecisely imply the
+number of lanes and typical speed limits. Heuristically enumerating the turns at
+each intersection often misses common features like right turn-only lanes and
+shared center left turn lanes. There are many circumstances where AORTA
+simulates one intersection as several in close proximity because roads do not
+geometrically line up, but functionally they do.  If OSM encoded this data, the
+performance of intersection policies and the realism of the simulation would
+improve. Other simulators have identified similar issues before
+\cite{Krajzewicz_Hertkorn_Ringel_Wagner_2005}.
 
 \commentd{Er, so how do I finally end this?}
-\commentm{We still have a bit of passive voice.  How much is acceptable?  I think
-	it sounds fine, but I imagine an English professor may disagree.}
 

### 2012-03-15 papers/itsc_2012/final.tex
adding houston results, WIP

@@
 The last policy tested is an autonomous reservation policy inspired by AIM
-\cite{JAIR08-dresner}. This policy groups agents with compatible turns together,
-allowing new agents to join existing groups. To prevent one heavy direction of
-traffic from hogging the intersection indefinitely, a timeout preemptively
-cycles through reservations.
+\cite{JAIR08-dresner}. Because agents explicitly communicate with intersections
+to schedule their turn, this policy is only suitable for autonomous drivers.
+This policy groups agents with compatible turns together, allowing new agents to
+join existing groups. To prevent one heavy direction of traffic from hogging the
+intersection indefinitely, a timeout preemptively cycles through reservations.
 
@@
 
-\begin{figure}[h]
-  \centering \includegraphics[width=\linewidth]{avg_lag_agent_atx.png}
-  \caption{The distribution of average delay reveals that traffic signals can
-           cause the most lag.}
-  \label{fig:avg_lag}
-  \vspace{-10pt}
-\end{figure}
+\begin{figure*}[h]
+  \centering
+  \subfloat[Average delay in Austin]{
+    \includegraphics[width=0.5\linewidth]{avg_lag_agent_atx.png}
+    \label{fig:avg_lag_atx}
+  }
+  \subfloat[Average delay in Houston]{
+    \includegraphics[width=0.5\linewidth]{avg_lag_agent_houston.png}
+    \label{fig:avg_lag_houston}
+  }
+\end{figure*}
+\begin{figure*}[h]
+  \subfloat[Number of agents driving in Austin]{
+    \includegraphics[width=0.5\linewidth]{agent_cnt_atx.png}
+    \label{fig:agent_cnt_atx}
+  }
+  \subfloat[Number of agents driving in Houston]{
+    \includegraphics[width=0.5\linewidth]{agent_cnt_houston.png}
+    \label{fig:agent_cnt_houston}
+  }
+  \caption{Autonomous reservations outperform stop signs and traffic signals in
+           these experiments.}
+  %\vspace{-10pt}
+\end{figure*}
 
@@
 AORTA can be used as a framework for evaluating these policies.
-
-\begin{figure}[h]
-  \centering \includegraphics[width=\linewidth]{agent_cnt_atx.png}
-  \caption{Autonomous reservations outperform stop signs, and traffic signal
-           gridlocked.}
-  \label{fig:agent_cnt}
-  \vspace{-10pt}
-\end{figure}
 

### 2012-03-15 papers/itsc_2012/final.tex
rewording stuff

@@
 
-To demonstrate how easily AORTA can be utilized to evaluate different
-intersection policies, this section presents a simple experiment showing how
-different policies affect the throughput of intersections. In this experiment,
-every intersection in the map uses either a stop sign, traffic signal, or
-autonomous reservation policy.
+This section presents a simple experiment showing how different policies affect
+the throughput of intersections. In this experiment, every intersection in the
+map uses either a stop sign, traffic signal, or autonomous reservation policy.
+The results demonstrate that policies affect total system performance and that
+AORTA is a useful tool for evaluating this relation.
 
@@
 Texas\footnote{bounded by the Mopac and Springdale Road horizontally, and Oltorf
-Street and Koenig Lane vertically}, for one hour. The step duration is fixed at
-0.1 seconds. While the intersection policies vary, the agents spawned and their
-routes remain the same. A generator creates one new agent every 1 simulation
-second, with a uniformly random starting position and goal.
+Street and Koenig Lane vertically}, and a 15$\times$13km$^{2}$ slice of downtown
+Houston, Texas \footnote{bounded by Ella Boulevard and Eastex Freeway
+horizontally, and West Alabama Street and Crosstimbers Street vertically}, for
+one hour. The step duration is fixed at 0.1 seconds. While the intersection
+policies vary, the agents spawned and their routes remain the same.  A generator
+creates one new agent every 1 simulation second, with a uniformly random
+starting position and goal.
 
@@
 
-If an intersection immediately permits an agent to proceed and grants other
-agents in the same lane similar access (because this new agent must still follow
-these others), then the new agent may simply proceed at full speed limit without
-pausing. This gives the least possible delay spent along some lane, so any extra
-time is defined as lag and is used to measure intersection performance. Figure
-\ref{fig:avg_lag} graphs the average of this delay per agent. On average,
-reservation policies cause less than 1 second of delay at each intersection,
-stop signs cause about 7 seconds lag, and traffic signals using the heuristic
-described above cause 16 seconds. These results will vary with different
-traffic levels and policy configurations, so the important demonstration is that
-AORTA can be used as a framework for evaluating these policies.
+If an agent travels along a lane at the speed limit and does not slow down to
+wait at an intersection, then it traverses the lane in an optimal amount of
+time. If the intersection orders it to stop or if other agents in the same lane
+have slowed down by the intersection's orders, then the intersection has caused
+some delay. This delay can be used as one measure of intersection performance.
+Figures \ref{fig:avg_lag_atx} and \ref{fig:avg_lag_houston} graph the average of
+this delay per agent. On average, reservation policies cause less than 1 second
+of delay at each intersection, stop signs cause about 7 seconds lag, and traffic
+signals using the heuristic described above cause 16 seconds. These results will
+vary with different traffic levels and policy configurations, so the important
+demonstration is that AORTA can be used as a framework for evaluating these
+policies.
 

### 2012-03-15 papers/itsc_2012/final.tex
fixed issued with multi-clumn figure in paper
minor edits to sec I through IV

@@
 real world by generating maps using publicly available road data from
-\textit{OpenStreetMap} (OSM). This allows simulations to be set up through AORTA
-for a desired region anywhere in the world in a matter of minutes. AORTA allows
-for traffic optimization by creating intelligent behaviors for individual driver
-agents and intersection policies to be followed by these agents. These behaviors
-and policies define how agents interact with one another, control when they
-cross intersections, and route agents to their destination. This paper
-demonstrates a simple application using AORTA through an experiment testing an
-intersection policy designed for autonomous vehicles at a city-wide scale.
+\textit{OpenStreetMap} (OSM). This allows simulations to be set up through
+AORTA for a desired region anywhere in the world in a matter of minutes. AORTA
+allows for traffic optimization by creating intelligent behaviors for
+individual driver agents and intersection policies to be followed by these
+agents. These behaviors and policies define how agents interact with one
+another, control when they cross intersections, and route agents to their
+destination. This paper demonstrates a simple application using AORTA through
+an experiment testing intersection policies at a city-wide scale.
 
@@
 
+% In 2004, the farthest distance traveled by a vehicle autonomously in the DARPA
+% Grand Challenge was 11.9km \cite{cnnGrandChallenge2004}. By 2007, six of the
+
 Autonomous vehicle technology has made tremendous progress in the last decade.
-In 2004, the farthest distance traveled by a vehicle autonomously in the DARPA
-Grand Challenge was 11.9km \cite{cnnGrandChallenge2004}. By 2007, six of the
-competing teams completed the 96km course set for the DARPA Urban Challenge
-\cite{spectrumUrbanChallenge2007}. They did so while obeying the traffic laws
-followed by human drivers, navigating along with other moving vehicles, and
-following correct intersection precedence order. Since then, Google's driverless
-cars have clocked more than 250,000km on public roads in urban California, USA
-\cite{tedThrun2011}. In 2010, researchers from the University of Parma
-successfully completed an autonomous \textit{intercontinental} run from Parma,
-Italy to Shanghai, China \cite{cnnVislab2010}. The successful completion of all
-these milestones suggests that autonomous cars are here to stay, and are ever
-closer to becoming commercially available. With the arrival of autonomous cars,
-it also becomes possible to optimize traffic in ways not possible for human
-drivers. 
-%\commentd{If we need space, I agree with Dr Chiu, some of this could be trimmed}
+In 2007, six of the competing teams completed the 96km course set for the DARPA
+Urban Challenge In 2004, the farthest distance traveled by a vehicle
+autonomously in the DARPA \cite{spectrumUrbanChallenge2007}. They did so while
+obeying the traffic laws followed by human drivers, navigating along with other
+moving vehicles, and following correct intersection precedence order. Since
+then, Google's driverless cars have clocked more than 250,000km on public roads
+in urban California, USA \cite{tedThrun2011}. In 2010, researchers from the
+University of Parma successfully completed an autonomous
+\textit{intercontinental} run from Parma, Italy to Shanghai, China
+\cite{cnnVislab2010}. The successful completion of all these milestones
+suggests that autonomous cars are here to stay, and are ever closer to becoming
+commercially available. With the arrival of autonomous cars, it also becomes
+possible to optimize traffic in ways not possible for human drivers. 
+%\commentd{If we need space, I agree with Dr Chiu, some of this could be
+%trimmed}
 
@@
   \centering \includegraphics[width=\linewidth]{downtown_atx.png}
-  \caption{A simulation of downtown Austin, Texas.}
+  \caption{A simulation of autonomous agents in downtown Austin, Texas, as visualized through the UI available within the AORTA simulator}
   \label{fig:ui_screenshot}
@@
 behaviors and intersection policies. Autonomous vehicles, termed agents, use
-behaviors to interact with one another, follow intersection policies, and decide
-on both long term and short term actions. AORTA allows for the creation of
-intersection policies that designate when agents can cross the intersection.
-AORTA's goal is to allow for the definition of new agent behaviors and
-intersection policies to optimize autonomous traffic. Additionally, by
-assigning human-like behaviors such as the ``car following model''
-\cite{brackstone1999car} to agents, and traffic-signal-like policies to
-intersections, AORTA could potentially simulate human traffic.
+behaviors to interact with one another, follow intersection policies, and
+decide on both long term and short term actions. Intersection policies
+deisgnate when it is safe for an agent to cross an intersection.  AORTA's goal
+is to allow for the definition of new agent behaviors and intersection policies
+to optimize autonomous traffic. Additionally, by assigning human-like behaviors
+such as the ``car following model'' \cite{brackstone1999car} to agents, and
+traffic-signal-like policies to intersections, AORTA could potentially simulate
+human traffic.
+
+% AORTA allows for the creation of
+% intersection policies that designate when agents can cross the intersection.
 
@@
 user-editable interface for world maps. A map for any desired city in the world
-can be downloaded, which is then parsed by AORTA to set up a scale simulation of
-the real world in a few minutes. AORTA is available open-source and is easily
-extensible \footnote{Code available at
-\texttt{http://code.google.com/p/road-rage}} \commentpk{could also change this
-to a citation}, making it easier for users to test out a number of agent
-behaviors and intersection policies.
+can be downloaded, which is then parsed by AORTA to set up a scale simulation
+of the real world in a few minutes. AORTA is available open-source and is
+easily extensible \footnote{Code available at
+\texttt{http://code.google.com/p/road-rage}}, making it easier for users to
+test out a number of agent behaviors and intersection policies in a short
+timespan.
 
@@
 section \ref{sec:arch}.  Use of OSM data in AORTA is explained in section
-\ref{sec:map}, along with a description of the simulator and the default
-behaviors available in AORTA in section \ref{sec:simulation}. An experiment
-investigating the benefit of an alternate intersection policy designed for
-autonomous vehicles is presented in \ref{sec:results}. The paper then concludes
-with a discussion on future work.
+\ref{sec:map}, along with a description of the simulator in section
+\ref{sec:simulation}. A simple experiment is presented in section
+\ref{sec:results} demonstrating an application built on top of AORTA, that
+evaluates different intersection policies in a large scale scenario. The paper
+then concludes with a discussion on future work.
 
@@
 to follow in a given day. MATSim aims to iteratively improve these plans to
-minimize traffic delays while minimizing the total amount of time by which agents
-are late to their destinations. This goal is somewhat different from that of AORTA,
-where the routes (demand) supplied by individuals is taken as is and does not have
-a pre-determined ideal time of arrival. Rather than improving agent routes, AORTA
-allows creating entirely different behaviors for both vehicles and intersections
-in order to simulate a different environment instead of optimizing plans for a
-society in a static environment.
+minimize traffic delays while minimizing the total amount of time by which
+agents are late to their destinations. This goal is somewhat different from
+that of AORTA, where the routes (demand) supplied by individuals is taken as is
+and does not have a pre-determined ideal time of arrival. Rather than improving
+agent routes, AORTA allows creating entirely different behaviors for both
+vehicles and intersections in order to simulate a different environment instead
+of optimizing plans for a society in a static environment \commentpk{I think
+MIke made these changes last night, but I am not sure if the last few sentences
+are clear to me}.
 
@@
 
-Simulators have also focused on autonomous vehicles with sensors and actuators
-\cite{figueiredo2009approach}. These simulators are used to ensure that
-autonomous vehicles are behaving correctly while interacting with other
-vehicles and infrastructure, given their configuration of sensors and
-actuators. On the other hand, AORTA assumes that autonomous vehicles have
-reasonable sensory input and mechanical output, and instead focuses on
-interactions between an agent and its surrounding agents and intersections.
+% Simulators have also focused on autonomous vehicles with sensors and actuators
+% \cite{figueiredo2009approach}. These simulators are used to ensure that
+% autonomous vehicles are behaving correctly while interacting with other
+% vehicles and infrastructure, given their configuration of sensors and
+% actuators. On the other hand, AORTA assumes that autonomous vehicles have
+% reasonable sensory input and mechanical output, and instead focuses on
+% interactions between an agent and its surrounding agents and intersections.
 
@@
   \label{fig:arch}
-  \vspace{-20pt}
+  \vspace{-15pt}
 \end{figure}
@@
   cross]{
-    \includegraphics[scale=0.4]{turn_conflicts.png}
+    \includegraphics[width=0.42\columnwidth]{turn_conflicts.png}
     \label{fig:conflicts}
@@
   there are no overpasses pictured.]{
-    \includegraphics[scale=0.4]{crazy_intersection.png}
+    \includegraphics[width=0.4\columnwidth]{crazy_intersection.png}
     \label{fig:crazyIntersection}
@@
 
+\begin{figure*}[t]
+  \centering
+  \subfloat[Average delay in Austin]{
+    \label{fig:avg_lag_atx}
+    \includegraphics[width=0.4\textwidth]{avg_lag_agent_atx.png}
+  }\,
+  \subfloat[Average delay in Houston]{
+    \label{fig:avg_lag_houston}
+    \includegraphics[width=0.4\textwidth]{avg_lag_agent_houston.png}
+  }\\
+  \subfloat[Number of agents driving in Austin]{
+    \label{fig:agent_cnt_atx}
+    \includegraphics[width=0.4\textwidth]{agent_cnt_atx.png}
+  }\,
+  \subfloat[Number of agents driving in Houston]{
+    \label{fig:agent_cnt_houston}
+    \includegraphics[width=0.4\textwidth]{agent_cnt_houston.png}
+  }
+  \caption{Autonomous reservations outperform stop signs and traffic signals in
+           these experiments.}
+  \label{fig:results}
+  \vspace{-15pt}
+\end{figure*}
+
 This section presents a simple experiment showing how different policies affect
@@
 \subsection{Results}
-
-\begin{figure*}[h]
-  \centering
-  \subfloat[Average delay in Austin]{
-    \includegraphics[width=0.5\linewidth]{avg_lag_agent_atx.png}
-    \label{fig:avg_lag_atx}
-  }
-  \subfloat[Average delay in Houston]{
-    \includegraphics[width=0.5\linewidth]{avg_lag_agent_houston.png}
-    \label{fig:avg_lag_houston}
-  }
-\end{figure*}
-\begin{figure*}[h]
-  \subfloat[Number of agents driving in Austin]{
-    \includegraphics[width=0.5\linewidth]{agent_cnt_atx.png}
-    \label{fig:agent_cnt_atx}
-  }
-  \subfloat[Number of agents driving in Houston]{
-    \includegraphics[width=0.5\linewidth]{agent_cnt_houston.png}
-    \label{fig:agent_cnt_houston}
-  }
-  \caption{Autonomous reservations outperform stop signs and traffic signals in
-           these experiments.}
-  %\vspace{-10pt}
-\end{figure*}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
+\small
 \section*{ACKNOWLEDGMENTS}

### 2012-03-15 papers/itsc_2012/final.tex
edits to results

@@
 \subsection{Intersection Policies}
+\label{sec:policies}
 
@@
 
-The experiment simulates a 9$\times$11km$^{2}$ slice of downtown Austin,
-Texas\footnote{bounded by the Mopac and Springdale Road horizontally, and Oltorf
-Street and Koenig Lane vertically}, and a 15$\times$13km$^{2}$ slice of downtown
-Houston, Texas \footnote{bounded by Ella Boulevard and Eastex Freeway
-horizontally, and West Alabama Street and Crosstimbers Street vertically}, for
-one hour. The step duration is fixed at 0.1 seconds. While the intersection
-policies vary, the agents spawned and their routes remain the same.  A generator
-creates one new agent every 1 simulation second, with a uniformly random
-starting position and goal.
+The experiment simulates one hour of traffic in two cities: a
+9$\times$11km$^{2}$ slice of downtown Austin, Texas\footnote{bounded by the
+Mopac and Springdale Road horizontally, and Oltorf Street and Koenig Lane
+vertically}, and a 15$\times$13km$^{2}$ slice of downtown Houston, Texas
+\footnote{bounded by Ella Boulevard and Eastex Freeway horizontally, and West
+Alabama Street and Crosstimbers Street vertically}.  The step duration is fixed
+at 0.1 seconds. While the intersection policies vary, the agents spawned and
+their routes remain the same.  A generator creates one new agent every 1
+simulation second, with a uniformly random starting position and goal.
 
@@
 Figures \ref{fig:avg_lag_atx} and \ref{fig:avg_lag_houston} graph the average of
-this delay per agent. On average, reservation policies cause less than 1 second
-of delay at each intersection, stop signs cause about 7 seconds lag, and traffic
-signals using the heuristic described above cause 16 seconds. These results will
-vary with different traffic levels and policy configurations, so the important
+this delay per agent. Similar trends are observed in both cities: an average of
+less than 1 second of delay at autonomous reservation intersections, 7 (Austin)
+or 8 (Houston) seconds at stop signs, and 16 (Austin) or 17 (Houston) seconds at
+traffic signals using the heuristic described above.  These results will vary
+with different traffic levels and policy configurations, so the important
 demonstration is that AORTA can be used as a framework for evaluating these
@@
 The number of agents driving at some time changes because the continuous
-generator introduces a new agent every second. This count reaches a steady
-state when the generator introduces an agent at the same rate as older agents
-finish their route. Figure \ref{fig:agent_cnt} reveals that autonomous
-reservations let agents reach their destination twice as quickly as the
-competition, with a steady-state of about 400 agents. The count for traffic
+generator introduces a new agent every second. This count reaches a steady state
+when the generator introduces an agent at the same rate as older agents finish
+their route. Figure \ref{fig:agent_cnt_atx} reveals that autonomous reservations
+let agents reach their destination twice as quickly as the competition in
+Austin, with a steady-state of about 400 agents. There, the count for traffic
 signals continues to increase because gridlock occurred in one segment of the
-map, causing agents to not finish their route.  Unexpected starvation was also
-observed: although the signals cycle through turns regularly, agents cannot
-proceed when the previous cycle caused a destination lane to fill and become
-blocked.
+map, causing agents to not finish their route. Figure
+\ref{fig:agent_cnt_houston} revals a similar trend in Houston, except that
+gridlock occurred both during the trial with traffic signals and stop signs.
+Unexpected starvation was also observed: although the signals cycle through
+turns regularly, agents cannot proceed when the previous cycle caused a
+destination lane to fill and become blocked.
 
-Users could conduct this experiment and many others without needing to
-understand most of AORTA's code-base. The largest intersection policy, traffic
-signals with the flooding heuristic for timing assignment, is only 400 lines of
-code in Scala.
+Repeating this or other experiments in more locations only requires those
+locations to be exported from the OSM website. Editing the intersection policies
+is similarly convenient, as users can ignore most of AORTA's code-base and just
+change the logic described in section \ref{sec:policies}.  For reference, the
+most complex intersection policy, traffic signals with the flooding heuristic
+for timing assignment, is only 400 lines of code in Scala.
 

### 2012-03-15 papers/itsc_2012/final.tex
checked edits and comments all the way to sec V

--- 2012-03-15 papers/itsc_2012/final.tex
@@
 Currently agents cannot change lanes. However, AORTA supports roads forcibly
-merging by treating the merge point as an intersection. The lack of
-lane-changing causes a second issue: gridlock.
+merging by treating the merge point as an intersection \commentpk{preceding
+statement is somewhat unclear}. The lack of lane-changing causes a second
+issue: gridlock.
 
@@
 traditional stop signs, traffic signals, and an AIM-inspired autonomous
-reservation system \cite{JAIR08-dresner}. The traffic signals are further
+reservation policy \cite{JAIR08-dresner}. The traffic signals are further
 extensible by assigning groups of non-conflicting turns at each intersection

### 2012-03-15 papers/itsc_2012/final.tex
checked edits and comments all the way to sec V

@@
 In 2007, six of the competing teams completed the 96km course set for the DARPA
-Urban Challenge In 2004, the farthest distance traveled by a vehicle
-autonomously in the DARPA \cite{spectrumUrbanChallenge2007}. They did so while
+Urban Challenge \cite{spectrumUrbanChallenge2007}. They did so while
 obeying the traffic laws followed by human drivers, navigating along with other
@@
 \begin{figure}[t]
-  \centering \includegraphics[width=\linewidth]{downtown_atx.png}
+  \centering \includegraphics[width=0.9\linewidth]{downtown_atx.png}
   \caption{A simulation of autonomous agents in downtown Austin, Texas, as visualized through the UI available within the AORTA simulator}
   \label{fig:ui_screenshot}
-  \vspace{-15pt}
+  \vspace{-20pt}
 \end{figure}

### 2012-03-15 papers/itsc_2012/final.tex
changed matsim comparison in related works

@@
 supplied to MATSim in the form of \textit{plans} that an individual may intend
-to follow in a given day. MATSim aims to iteratively improve these plans to
-minimize traffic delays while minimizing the total amount of time by which
-agents are late to their destinations. This goal is somewhat different from
-that of AORTA, where the routes (demand) supplied by individuals is taken as is
-and does not have a pre-determined ideal time of arrival. Rather than improving
-agent routes, AORTA allows creating entirely different behaviors for both
-vehicles and intersections in order to simulate a different environment instead
-of optimizing plans for a society in a static environment \commentpk{I think
-MIke made these changes last night, but I am not sure if the last few sentences
-are clear to me}.
+to follow in a given day. MATSim aims to improve these plans by attempting to
+minimize the total amount of time by which individuals are late to their
+destinations. This goal is somewhat different from that of AORTA, where the
+routes (demand) supplied by individuals is taken as is, and each agent has a
+specified destination at a given point in time. Instead, AORTA focuses on
+vehicle behaviors and intersection policies that can help this agent reach its
+destination faster.
+
+% This goal is somewhat different from that of AORTA, where the
+% routes (demand) supplied by individuals is taken as is and does not have a
+% pre-determined ideal time of arrival. Rather than improving agent routes, AORTA
+% allows creating entirely different behaviors for both vehicles and
+% intersections in order to simulate a different environment instead of
+% optimizing plans for a society in a static environment \commentpk{I think MIke
+% made these changes last night, but I am not sure if the last few sentences are
+% clear to me}.
 

### 2012-03-15 papers/itsc_2012/final.tex
edits through the end of the paper

@@
 9$\times$11km$^{2}$ slice of downtown Austin, Texas\footnote{bounded by the
-Mopac and Springdale Road horizontally, and Oltorf Street and Koenig Lane
-vertically}, and a 15$\times$13km$^{2}$ slice of downtown Houston, Texas
-\footnote{bounded by Ella Boulevard and Eastex Freeway horizontally, and West
-Alabama Street and Crosstimbers Street vertically}.  The step duration is fixed
-at 0.1 seconds. While the intersection policies vary, the agents spawned and
-their routes remain the same.  A generator creates one new agent every 1
-simulation second, with a uniformly random starting position and goal.
+Mopac, Springdale Road, Oltorf Street and Koenig Lane}, and a
+15$\times$13km$^{2}$ slice of downtown Houston, Texas\footnote{bounded by Ella
+Boulevard, Eastex Freeway, West Alabama Street and Crosstimbers Street}.  The
+step duration is fixed at 0.1 seconds. While the intersection policies vary,
+the agents spawned and their routes remain the same.  A generator creates one
+new agent every 1 simulation second, with a uniformly random starting position
+and goal.
 
@@
 different times.}. After setting up a scenario in the UI, users can save this
-configuration for resimulation.
+configuration for re-simulation.
 
@@
 number of agents. A measure independent of these factors is the number of agent
-steps per second. On a 2.4 GHz machine, this can be observed to average at about
-150,000, a number comparable to existing simulators \cite{SUMOthesis}. When dt
-= 0.1 seconds, this means 15,000 agents can be comfortably simulated in
-real-time. Further optimizations are planned.
+steps per second. On a 2.4 GHz machine, this can be observed to average at
+about 150,000, a number comparable to existing simulators \cite{SUMOthesis}.
+When the time step is 0.1 seconds, this means 15,000 agents can be comfortably
+simulated in real-time. Further optimizations are planned.
 
@@
 effort to improve AORTA as a framework by including a lane-changing model,
-preventing gridlock, and supporting the resources of supercomputing clusters.
+preventing gridlock, and supporting more agents through parallel computing.
 
 AORTA's map construction process, although flexible, cannot completely account
-for incompletedness in OSM data. Road type annotations imprecisely imply the
+for incompleteness in OSM data. Road type annotations imprecisely imply the
 number of lanes and typical speed limits. Heuristically enumerating the turns at
@@
 
-\commentd{Er, so how do I finally end this?}
-
-\addtolength{\textheight}{-12cm}  % This command serves to balance the column lengths
-                                  % on the last page of the document manually. It shortens
-                                  % the textheight of the last page by a suitable amount.
-                                  % This command does not take effect until the next page
-                                  % so it should come on the page before the last. Make
-                                  % sure that you do not shorten the textheight too much.
+% \addtolength{\textheight}{-12cm}  % This command serves to balance the column lengths
+%                                   % on the last page of the document manually. It shortens
+%                                   % the textheight of the last page by a suitable amount.
+%                                   % This command does not take effect until the next page
+%                                   % so it should come on the page before the last. Make
+%                                   % sure that you do not shorten the textheight too much.
 

### 2012-03-15 papers/itsc_2012/final.tex
moved osm critique and tweaking sumo prgrph

@@
 
-% TODO better name!
 \title{\LARGE \bf
@@
 
-% TODO theme: easy simulation in your city, in 2 minutes!
-
-% In 2004, the farthest distance traveled by a vehicle autonomously in the DARPA
-% Grand Challenge was 11.9km \cite{cnnGrandChallenge2004}. By 2007, six of the
-
 Autonomous vehicle technology has made tremendous progress in the last decade.
@@
 possible to optimize traffic in ways not possible for human drivers.
-%\commentd{If we need space, I agree with Dr Chiu, some of this could be
-%trimmed}
 
@@
   \centering \includegraphics[width=0.9\linewidth]{downtown_atx.png}
-  \caption{A simulation of autonomous agents in downtown Austin, Texas, as visualized through the UI available within the AORTA simulator}
+  \caption{Visualizing autonomous agents in downtown Austin, Texas, with AORTA's UI}
   \label{fig:ui_screenshot}
@@
 called AORTA, which stands for \textit{Approximately Orchestrated Routing and
-Transportation Analyzer}, a platform designed for testing autonomous vehicle
-behaviors and intersection policies. Autonomous vehicles, termed agents, use
-behaviors to interact with one another, follow intersection policies, and
-decide on both long term and short term actions. Intersection policies
-deisgnate when it is safe for an agent to cross an intersection.  AORTA's goal
-is to allow for the definition of new agent behaviors and intersection policies
-to optimize autonomous traffic. Additionally, by assigning human-like behaviors
-such as the ``car following model'' \cite{brackstone1999car} to agents, and
+Transportation Analyzer}. AORTA is a platform designed for testing autonomous
+vehicle behaviors and intersection policies. Autonomous vehicles, termed agents,
+use behaviors to interact with one another, follow intersection policies, and
+decide on both long term and short term actions. Intersection policies deisgnate
+when it is safe for an agent to cross an intersection.  AORTA's goal is to allow
+for the definition of new agent behaviors and intersection policies to optimize
+autonomous traffic. Additionally, by assigning human-like behaviors such as the
+``car following model'' \cite{brackstone1999car} to agents, and
 traffic-signal-like policies to intersections, AORTA could potentially simulate
 human traffic.
-
-% AORTA allows for the creation of
-% intersection policies that designate when agents can cross the intersection.
 
@@
 \ref{sec:map}, along with a description of the simulator in section
-\ref{sec:simulation}. A simple experiment is presented in section
-\ref{sec:results} demonstrating an application built on top of AORTA, that
-evaluates different intersection policies in a large scale scenario. The paper
-then concludes with a discussion on future work.
+\ref{sec:simulation}. Section \ref{sec:results} demonstrates an application
+built on top of AORTA and presents a simple experiment evaluating different
+intersection policies in a large scale scenario. The paper then concludes with a
+discussion on future work.
 
@@
 \label{sec:related_work}
-
-% just draw differences between us and them
 
@@
 destinations. This goal is somewhat different from that of AORTA, where the
-routes (demand) supplied by individuals is taken as is, and each agent has a
-specified destination at a given point in time. Instead, AORTA focuses on
-vehicle behaviors and intersection policies that can help this agent reach its
-destination faster.
-
-% This goal is somewhat different from that of AORTA, where the
-% routes (demand) supplied by individuals is taken as is and does not have a
-% pre-determined ideal time of arrival. Rather than improving agent routes, AORTA
-% allows creating entirely different behaviors for both vehicles and
-% intersections in order to simulate a different environment instead of
-% optimizing plans for a society in a static environment \commentpk{I think MIke
-% made these changes last night, but I am not sure if the last few sentences are
-% clear to me}.
+routes (demand) supplied by individuals is taken as is.  Instead, AORTA focuses
+on vehicle behaviors and intersection policies that can help this agent reach
+its destination faster.
 
@@
 individual vehicles, as well as dynamically adapting traffic signal policies to
-increase traffic efficiency.  AORTA and SUMO differ in the manner in which
-intersections are handled. SUMO uses traditional traffic signals, while AORTA
-allows intersections to grant agents entry by any general policy. These policies
-can use traditional schemes such as traffic signals or stop-sign based
-precedence, or more efficient methods designed specifically for autonomous
-vehicles \cite{JAIR08-dresner}.
-\commentd{Are we sure SUMO is not easily configurable in this way?}
+increase traffic efficiency. While SUMO has a flexible system for managing
+traditional traffic signals, AORTA's general intersection policies are designed
+with autonomous vehicles in mind. These policies can use traditional schemes
+such as traffic signals or stop-sign based precedence, or more efficient methods
+designed specifically for autonomous vehicles \cite{JAIR08-dresner}. In
+addition, AORTA's core implementation, written in Scala \cite{scala}, differs
+from that of SUMO, written in C++, in that it uses a higher-level language.
 
@@
 \end{enumerate}
+
+\subsection{Limitations with OSM}
+
+AORTA's map construction process, although flexible, cannot completely account
+for incompleteness in OSM data. Road type annotations imprecisely imply the
+number of lanes and typical speed limits. Heuristically enumerating the turns at
+each intersection often misses common features like right turn-only lanes and
+shared center left turn lanes. There are many circumstances where AORTA
+simulates one intersection as several in close proximity because roads do not
+geometrically line up, but functionally they do. If OSM encoded this data, the
+the realism of the simulation would improve. Other simulators have identified
+similar issues before \cite{Krajzewicz_Hertkorn_Ringel_Wagner_2005}.
 
@@
 
-AORTA's map construction process, although flexible, cannot completely account
-for incompleteness in OSM data. Road type annotations imprecisely imply the
-number of lanes and typical speed limits. Heuristically enumerating the turns at
-each intersection often misses common features like right turn-only lanes and
-shared center left turn lanes. There are many circumstances where AORTA
-simulates one intersection as several in close proximity because roads do not
-geometrically line up, but functionally they do.  If OSM encoded this data, the
-performance of intersection policies and the realism of the simulation would
-improve. Other simulators have identified similar issues before
-\cite{Krajzewicz_Hertkorn_Ringel_Wagner_2005}.
-
 % \addtolength{\textheight}{-12cm}  % This command serves to balance the column lengths

### 2012-03-15 papers/itsc_2012/final.tex
spell check

@@
 use behaviors to interact with one another, follow intersection policies, and
-decide on both long term and short term actions. Intersection policies deisgnate
+decide on both long term and short term actions. Intersection policies designate
 when it is safe for an agent to cross an intersection.  AORTA's goal is to allow
@@
 test out a number of agent behaviors and intersection policies in a short
-timespan.
+time span.
 
@@
 \begin{enumerate}
-  \item OSM encodes many paths besides driveable roads, so the builder first filters
+  \item OSM encodes many paths besides drivable roads, so the builder first filters
         these out. Next, the builder marks OSM nodes common to multiple roads, since
@@
 \textit{start} polygon to random destinations within a \textit{goal} polygon.
-Users can draw these polyons by mouse in the UI, allowing specific traffic
+Users can draw these polygons by mouse in the UI, allowing specific traffic
 patterns like rush-hour scenarios to be easily simulated. A generator polygon
@@
 
-Intersections check for collisions by verifying no two agents are simultaenously
+Intersections check for collisions by verifying no two agents are simultaneously
 performing a conflicting turn. AORTA has a low-granularity model of turn
@@
 drunken walk that probabilistically moves towards the destination. Users can
-experiment with dynamic replanning or hierarchial planning
+experiment with dynamic replanning or hierarchical planning
 \cite{Botea04nearoptimal} without modifying any code other than creating a new
@@
 
-The speed of simulation depends on the step duration dt, the map size, and the
+The speed of simulation depends on the step duration $dt$, the map size, and the
 number of agents. A measure independent of these factors is the number of agent
@@
 map, causing agents to not finish their route. Figure
-\ref{fig:agent_cnt_houston} revals a similar trend in Houston, except that
+\ref{fig:agent_cnt_houston} reveals a similar trend in Houston, except that
 gridlock occurred both during the trial with traffic signals and stop signs.
@@
 The authors would like to thank Dr. Michael Quinlan for his support during the
-seminal stages of AORTA and all OpenStreetMap contributers for collecting the
+seminal stages of AORTA and all OpenStreetMap contributors for collecting the
 map data used in this work. Efforts by the Freshman Research Initiative,

### 2012-03-15 papers/itsc_2012/final.tex
my last wording tweaks

@@
 Autonomous vehicles have seen great advancements in recent years, and such
-vehicles are now ever closer to being commercially available. The advent of
+vehicles are now closer than ever to being commercially available. The advent of
 driverless cars provides opportunities for optimizing traffic in ways not
@@
 autonomous traffic. Additionally, by assigning human-like behaviors such as the
-``car following model'' \cite{brackstone1999car} to agents, and
+``car following model'' \cite{brackstone1999car} to agents and
 traffic-signal-like policies to intersections, AORTA could potentially simulate
@@
 builder} utility. Section \ref{sec:mapmodel} describes the map model generated
-by the builder, and section \ref{sec:mapconstruction} describes the process of
-refining OSM data.
+by the builder, section \ref{sec:mapconstruction} describes the process of
+refining OSM data, and section \ref{sec:osmlimits} describes limitations with
+this process.
 
@@
 \subsection{Limitations with OSM}
+\label{sec:osmlimits}
 
@@
 alongside its starting lane (as if it is in a driveway or parking lot). When the
-lane has no agents that could potentially crash with the new vehicle, the new
+lane has no agents that could potentially crash into the new vehicle, the new
 agent enters the system.
@@
 intersection, and the length of time the agent has been idling at rest. The
-policy replies by ordering the agent to ``stop'' or ``proceed''. The behavior
+policy replies by ordering the agent to \textit{stop} or \textit{proceed}. The behavior
 will continue to statelessly poll the policy every step until the agent begins
@@
 
-Currently agents cannot change lanes. However, AORTA supports roads forcibly
-merging by treating the merge point as an intersection \commentpk{preceding
-statement is somewhat unclear}. The lack of lane-changing causes a second
-issue: gridlock.
-
+Currently agents cannot change lanes. This causes a second issue: gridlock.
 Each agent maintains a safe following distance from the next. If a lane is
@@
 determinism may break down when generators use worker threads, since routes
-may be computed in different orders, causing agents to enter the road at
+could be computed in different orders, causing agents to enter the road at
 different times.}. After setting up a scenario in the UI, users can save this

### 2012-03-15 papers/itsc_2012/final.tex
we were poking out of the bottom margin.
piyush caught 2 bugs

@@
         the builder removes all but the largest strongly connected component of the
-        digraph. Disconnected lanes are spliced out of larger roads, leaving
+        directed graph. Disconnected lanes are spliced out of larger roads, leaving
         geometric gaps, but correctly ensuring connectivity.
@@
 manner so that the agent at the front of each lane is waiting to perform a turn
-into another full lane with the same such head agent, gridlock \cite{AAAI11-au}
-occurs: no agent in the system will make progress unless an agent at the front
-decides to pick a different turn.  AORTA can detect this by finding cycles in
-the graph of the ``is following'' relation.
+into another full lane with a similar agent at the front, gridlock
+\cite{AAAI11-au} occurs: no agent in the system will make progress unless an
+agent at the front decides to pick a different turn. AORTA has the ability to
+find such cycles.
 
@@
   \centering \includegraphics[scale=0.25]{gridlock.png}
-  \caption{Each of the red agents wants to turn left, so none of them
-           can. The traffic jam will slowly build up around this intersection.}
+  \caption{Each of the red agents wants to turn left, so none of them can.}
   \label{fig:gridlock}

### 2012-03-16 papers/itsc_2012/post_submit.tex
made mikes changes for the future

--- 2012-03-15 papers/itsc_2012/final.tex
@@
 of the real world in a few minutes. AORTA is available open-source and is
-easily extensible \footnote{Code available at
+easily extensible\footnote{Code available at
 \texttt{http://code.google.com/p/road-rage}}, making it easier for users to
@@
 routes (demand) supplied by individuals is taken as is. Instead, AORTA focuses
-on vehicle behaviors and intersection policies that can help this agent reach
-its destination faster.
+on vehicle behaviors and intersection policies to improve the execution of
+routes.
 
@@
 incoming to outgoing lanes. Since both lanes and turns support traffic, they
-are grouped together as \emph{traversables}, meaning agents may exist on them.
+are both called \emph{traversables}, meaning agents may exist on them.
 
@@
 Intersections check for collisions by verifying no two agents are simultaneously
-performing a conflicting turn. AORTA has a low-granularity model of turn
+performing conflicting turns. AORTA has a low-granularity model of turn
 conflict. If two vehicles at any point along two different turns could ever come
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\small
 \section*{ACKNOWLEDGMENTS}

### 2012-03-16 papers/itsc_2012/generate.sh

@@ -3,7 +3,7 @@
 if [ $# -eq 1 ]; then
   INPUT=$1
 else
-  INPUT=final.tex
+  INPUT=post_submit.tex
 fi
 
 # generate graphviz diagrams

### 2012-03-18 papers/itsc_2012/submitted2.tex
results from worker threads deterministically passed into sim

--- 2012-03-16 papers/itsc_2012/post_submit.tex
@@
 configuration fully determine the outcome of any simulation \footnote{This
-determinism may break down when generators use worker threads, since routes
-could be computed in different orders, causing agents to enter the road at
-different times.}. After setting up a scenario in the UI, users can save this
+determinism may break down when generators use worker threads while the
+simulation is running, since tasks may finish and agents may enter the road at
+different times}. After setting up a scenario in the UI, users can save this
 configuration for re-simulation.

### 2012-03-18 utexas/aorta/sim/Agent.scala

@@ -40,6 +40,11 @@ class Agent(val id: Int, val graph: Graph, val start: Edge, val start_dist: Doub
   def step(dt_s: Double): Boolean = {
     assert(dt_s == cfg.dt_s)
 
+    // To confirm determinism, enable one of these (more precision in doubles is
+    // more likely to differ) and diff the logs.
+    //Util.log(this + " at " + Agent.sim.tick + " is at " + at)
+    //Util.log(this + " at " + "%.1f".format(Agent.sim.tick) + " is at " + at.on + " len " + "%.1f".format(at.dist))
+
     if (speed == 0.0 && target_accel == 0.0) {
       if (idle_since == -1.0) {
         idle_since = Agent.sim.tick
@@ -47,11 +52,6 @@ class Agent(val id: Int, val graph: Graph, val start: Edge, val start_dist: Doub
       // short-circuit... we're not going anywhere.
       return false
     }
-
-    // To confirm determinism, enable one of these (more precision in doubles is
-    // more likely to differ) and diff the logs.
-    //Util.log(this + " at " + Agent.sim.tick + " is at " + at)
-    //Util.log(this + " at " + "%.1f".format(Agent.sim.tick) + " is at " + at.on + " len " + "%.1f".format(at.dist))
 
     val start_on = at.on
     val old_dist = at.dist

### 2012-03-18 utexas/aorta/sim/Generator.scala

@@ -5,6 +5,7 @@
 package utexas.aorta.sim
 
 import java.util.concurrent.{Executors, FutureTask, Callable}
+import scala.collection.mutable.ListBuffer
 
 import utexas.aorta.map.{Edge, Traversable}
 
@@ -52,7 +53,7 @@ extends Ordered[Generator]
   def serialize_ls(ls: Array[Edge]) = ls.map(e => e.id).mkString(",")
 
   // there may be no task scheduled
-  protected var pending: List[(Agent, Option[FutureTask[List[Traversable]]])] = Nil
+  protected var pending = new ListBuffer[(Agent, Option[FutureTask[List[Traversable]]])]
 
   // Returns new agents to try to spawn, or boolean means reap this genertor
   def run(): Either[List[Agent], Boolean]
@@ -74,19 +75,24 @@ extends Ordered[Generator]
       case Some(task) => {
         val delayed = new FutureTask[List[Traversable]](task)
         Generator.worker_pool.execute(delayed)
-        pending = (a, Some(delayed)) :: pending
+        pending += ((a, Some(delayed)))
       }
       case _ => {
         // don't schedule anything
-        pending = (a, None) :: pending
+        pending += ((a, None))
       }
     }
   }
 
   // This is a nonblocking poll, of course
   def poll(): List[Agent] = {
-    val (done, still_pending) = pending.partition(
-      a => a._2 match {
+    // Even though worker threads will complete tasks out of order, can force
+    // determinism by scanning through in order and stopping at the first that
+    // isn't done
+    val done = new ListBuffer[Agent]()
+    while (!pending.isEmpty) {
+      val a = pending.head
+      val ready = a._2 match {
         case Some(task) => {
           if (task.isDone) {
             a._1.route.got_route(task.get)
@@ -97,17 +103,21 @@ extends Ordered[Generator]
         }
         case None       => true
       }
-    )
-    pending = still_pending
-    return done.map(a => a._1)
+      if (ready) {
+        done += a._1
+        pending = pending.tail
+      } else {
+        // stop immediately!
+        return done.toList
+      }
+    }
+    return done.toList
   }
 
   // And the blocking poll
   def wait_for_all() = {
     var cnt = 0
-    // go in reverse because we're prepending to pending (haha) now, and the
-    // first scheduled is probably the first to be done
-    pending.reverse.foreach(a => {
+    pending.foreach(a => {
       print("\r" + Util.indent + "Done with " + cnt + " routes")
       a._2 match {
         case Some(task) => task.get

### 2012-03-18 utexas/aorta/sim/analysis/Gridlock.scala

@@ -56,7 +56,14 @@ object Gridlock {
     // First and foremost...
     Agent.sim.schedule(Agent.sim.tick + check_time, { Gridlock.detect_cycles })
 
+    /*for (e <- deps.keys) {
+      Util.log(e.id + " depends on " + deps(e).id + " with cnt " + dep_counter((e, deps(e))))
+    }
+    Util.log("")
+    */
+
     val visited = new MutableSet[Edge]()
+    val in_gridlock = new MutableSet[Edge]()
     for (e <- deps.keys) {
       if (!visited(e)) {
         // flood from a
@@ -73,8 +80,15 @@ object Gridlock {
           }
           if (this_group(cur)) {
             // Cycle! Gridlock!
-            Util.log("Gridlock detected at " + Agent.sim.tick + " among: " + this_group)
+            // Don't report it if we're discovering more things that lead to
+            // already-reported gridlock. & is intersect.
+            if ((in_gridlock & this_group).isEmpty) {
+              Util.log("Gridlock detected at " + Agent.sim.tick + " among: " + this_group)
+            } else {
+              Util.log("not reporting new case with " + cur)
+            }
             found_gridlock = true   // break the loop
+            in_gridlock ++= this_group
             // TODO pause if we're in UI?
           }
           this_group += cur

### 2012-03-26 papers/itsc_2012/peter-submitted.tex
added peter's comments on the current submitted version

--- 2012-03-18 papers/itsc_2012/submitted2.tex
@@
 easily extensible\footnote{Code available at
-\texttt{http://code.google.com/p/road-rage}}, making it easier for users to
-test out a number of agent behaviors and intersection policies in a short
-time span.
+\texttt{http://code.google.com/p/road-rage}}, making it easier\commentp{easier
+is a comparitive word, so this should be either "easy" or "easier than...."}
+for users to test out a number of agent behaviors and intersection policies in
+a short time span.
 
@@
 experiments without the overhead of visualization. The interaction between
-modules is visualized in figure \ref{fig:arch}.
+modules is visualized in \commentp{all figure and section references should
+have capital F and S respectively}\commentpk{I tend to use Fig.}figure
+\ref{fig:arch}.
 
@@
 the center line that divides the road into two directions of travel. AORTA
-interprets these points as straight lines and projects parallel lanes out using
-a fixed width. Turns are modeled with a single straight line connecting the end
-of one lane to the start of another.
+interprets these points as straight lines \commentp{line segments instead of
+lines} and projects parallel lanes out using a fixed width. Turns are modeled
+\commentp{approximated} with a single straight line \commentp{segment}
+connecting the end of one lane to the start of another.
 
@@
         directed graph. Disconnected lanes are spliced out of larger roads, leaving
-        geometric gaps, but correctly ensuring connectivity.
+        geometric gaps, but correctly ensuring connectivity. \commentp{This paragraph was generally unclear, and makes it sound like you are removing huge chunks of the graph}
 \end{enumerate}
@@
 for incompleteness in OSM data. Road type annotations imprecisely imply the
-number of lanes and typical speed limits. Heuristically enumerating the turns at
-each intersection often misses common features like right turn-only lanes and
-shared center left turn lanes. There are many circumstances where AORTA
-simulates one intersection as several in close proximity because roads do not
-geometrically line up, but functionally they do. If OSM encoded this data, the
-the realism of the simulation would improve. Other simulators have identified
-similar issues before \cite{Krajzewicz_Hertkorn_Ringel_Wagner_2005}.
+number of lanes and typical speed limits. Heuristically enumerating the turns
+at each intersection often misses common features like right turn-only lanes
+and shared center left turn lanes. There are many circumstances where AORTA
+simulates one intersection \commentp{makes it sound like aorta's fault} as
+several in close proximity because roads do not geometrically line up, but
+functionally they do. If OSM encoded this data, the the\commentp{extra the}
+realism of the simulation would improve. Other simulators have identified
+\commentp{"dealt with" instead of "identified"} similar issues before
+\cite{Krajzewicz_Hertkorn_Ringel_Wagner_2005}.
 
@@
 \label{sec:simulation}
+
+\commentp{The images in this section need better labels, plus maybe print
+better in black and white. For instance, the gridlock issue - since b/w does
+not show red explicitly, maybe you should mark the loop in the image. In Fig
+3a) the green arrowhead is hidden etc.}
 
@@
 
-Currently agents cannot change lanes. This causes a second issue: gridlock.
-Each agent maintains a safe following distance from the next. If a lane is
-filled, then an agent may be forced to stop in an intersection mid-turn and
-block other traffic. The default behavior prevents this by refusing to start a
-turn before the lookahead engine guarantees no agent will trigger a premature
-stop.
+Currently agents cannot change lanes. This causes a second \commentp{potential}
+issue: gridlock.  Each agent maintains a safe following distance from the next.
+If a lane is filled, then an agent may be forced to stop in an intersection
+mid-turn and block other traffic. The default behavior prevents this by
+refusing to start a turn before the lookahead engine guarantees no agent will
+trigger a premature stop.
 
@@
 lane-changing support: agents loop around intersections such as those pictured
-in figure \ref{fig:gridlock} in order to enter a lane adjacent to their original
-lane.  Since these routes are still legal, it is desirable to prevent or mend
-this problem without forcing different routes. However, choosing alternative
+in figure \ref{fig:gridlock} in order to enter a lane adjacent to their
+original lane.  Since these routes are still legal, it is desirable to prevent
+or mend this problem without forcing different routes \commentp{the need for
+forccing alternate routes could be clearer}. However, choosing alternative
 moves (by changing lanes or rerouting) is an important constituent of a
@@
 signal timings. On top of these structures, applications could experiment with
-ideas such dynamic contra-flow \cite{ITSC11-hausknecht}.  There will also be an
-effort to improve AORTA as a framework by including a lane-changing model,
-preventing gridlock, and supporting more agents through parallel computing.
+ideas such \commentp{as} dynamic contra-flow \cite{ITSC11-hausknecht}.  There
+will also be an effort to improve AORTA as a framework by including a
+lane-changing model, preventing gridlock, and supporting more agents through
+parallel computing.
 

### 2012-03-29 papers/itsc_2012/submitted2.tex
made changes recommended by dr stone

@@
 user-editable interface for world maps. A map for any desired city in the world
-can be downloaded, which is then parsed by AORTA to set up a scale simulation
-of the real world in a few minutes. AORTA is available open-source and is
-easily extensible\footnote{Code available at
-\texttt{http://code.google.com/p/road-rage}}, making it easier for users to
-test out a number of agent behaviors and intersection policies in a short
-time span.
+can be downloaded, which is then parsed by AORTA to set up a scale simulation of
+the real world in a few minutes. AORTA is available open-source and is easily
+extensible\footnote{Code available at
+\texttt{http://code.google.com/p/road-rage}}, making it easy for users to test
+out a number of agent behaviors and intersection policies in a short time span.
 
-This paper explores the state-of-the-art in traffic simulators in section
+This paper explores the state-of-the-art in traffic simulators in Section
 \ref{sec:related_work}, followed by a description of AORTA's architecture in
-section \ref{sec:arch}.  Use of OSM data in AORTA is explained in section
-\ref{sec:map}, along with a description of the simulator in section
+section \ref{sec:arch}.  Use of OSM data in AORTA is explained in Section
+\ref{sec:map}, along with a description of the simulator in Section
 \ref{sec:simulation}. Section \ref{sec:results} demonstrates an application
@@
 experiments without the overhead of visualization. The interaction between
-modules is visualized in figure \ref{fig:arch}.
+modules is visualized in Fig. \ref{fig:arch}.
 
@@
 open-source and easily extensible. For instance, the stop sign sign policy used
-in experiments described in section \ref{sec:results} is implemented in just 70
+in experiments described in Section \ref{sec:results} is implemented in just 70
 lines of code.
@@
 builder} utility. Section \ref{sec:mapmodel} describes the map model generated
-by the builder, section \ref{sec:mapconstruction} describes the process of
-refining OSM data, and section \ref{sec:osmlimits} describes limitations with
+by the builder, Section \ref{sec:mapconstruction} describes the process of
+refining OSM data, and Section \ref{sec:osmlimits} describes limitations with
 this process.
@@
 The map also has a geometric interpretation, used for visualization and
-measuring physical distances. A road contains a sequence of points, defining
-the center line that divides the road into two directions of travel. AORTA
-interprets these points as straight lines and projects parallel lanes out using
-a fixed width. Turns are modeled with a single straight line connecting the end
-of one lane to the start of another.
+measuring physical distances. A road contains a sequence of points, defining the
+center line that divides the road into two directions of travel. AORTA
+interprets these points as straight line segmentss and projects parallel lanes
+out using a fixed width. Turns are approximated with a single straight line
+segment connecting the end of one lane to the start of another.
 
@@
         lanes, the builder forces merging at the rightmost lanes. 
-  \item Because turns are heuristically placed, the graph is often disconnected.
-        In order to ensure that an agent will be able to reach any point in the map,
-        the builder removes all but the largest strongly connected component of the
-        directed graph. Disconnected lanes are spliced out of larger roads, leaving
-        geometric gaps, but correctly ensuring connectivity.
+  \item Due to OSM's lack of turn data, the directed graph may not be connected.
+        The builder removes small disconnected portions, leaving the largest
+        strongly connected component.
 \end{enumerate}
@@
 each intersection often misses common features like right turn-only lanes and
-shared center left turn lanes. There are many circumstances where AORTA
-simulates one intersection as several in close proximity because roads do not
-geometrically line up, but functionally they do. If OSM encoded this data, the
-the realism of the simulation would improve. Other simulators have identified
-similar issues before \cite{Krajzewicz_Hertkorn_Ringel_Wagner_2005}.
+shared center left turn lanes. In some circumstances, one intersection appears
+to be several in close proximity, since OSM only has geometric, not functional,
+data on how roads meet. If OSM encoded this information, the realism of the
+simulation would improve. Other simulators have dealt with similar issues before
+\cite{Krajzewicz_Hertkorn_Ringel_Wagner_2005}.
 
@@
 into contact, then the turns are always in conflict, no matter where the agents
-actually are along that turn. This is visualized in figure \ref{fig:conflicts}.
+actually are along that turn. This is visualized in Fig. \ref{fig:conflicts}.
 Further work is required before higher granularity can be introduced through
@@
 complex intersection geometries inferred from OSM. An example of such a complex
-intersection is shown in figure \ref{fig:crazyIntersection}.
+intersection is shown in Fig. \ref{fig:crazyIntersection}.
 
@@
 
-Currently agents cannot change lanes. This causes a second issue: gridlock.
-Each agent maintains a safe following distance from the next. If a lane is
-filled, then an agent may be forced to stop in an intersection mid-turn and
-block other traffic. The default behavior prevents this by refusing to start a
-turn before the lookahead engine guarantees no agent will trigger a premature
+Currently agents cannot change lanes. This causes a second potential issue:
+gridlock.  Each agent maintains a safe following distance from the next. If a
+lane is filled, then an agent may be forced to stop in an intersection mid-turn
+and block other traffic. The default behavior prevents this by refusing to start
+a turn before the lookahead engine guarantees no agent will trigger a premature
 stop.
@@
 lane-changing support: agents loop around intersections such as those pictured
-in figure \ref{fig:gridlock} in order to enter a lane adjacent to their original
-lane.  Since these routes are still legal, it is desirable to prevent or mend
-this problem without forcing different routes. However, choosing alternative
-moves (by changing lanes or rerouting) is an important constituent of a
-sufficient condition for the liveness of a road network \cite{AAAI11-au}.
+in Fig. \ref{fig:gridlock} in order to enter a lane adjacent to their original
+lane. Since such routes are still legal, it is desirable to prevent or mend this
+problem without requesting different routes. However, choosing alternative moves
+(by changing lanes or rerouting) is an important constituent of a sufficient
+condition for the liveness of a road network \cite{AAAI11-au}.
 
@@
 when the generator introduces an agent at the same rate as older agents finish
-their route. Figure \ref{fig:agent_cnt_atx} reveals that autonomous reservations
+their route. Fig. \ref{fig:agent_cnt_atx} reveals that autonomous reservations
 let agents reach their destination twice as quickly as the competition in
@@
 signals continues to increase because gridlock occurred in one segment of the
-map, causing agents to not finish their route. Figure
+map, causing agents to not finish their route. Fig.
 \ref{fig:agent_cnt_houston} reveals a similar trend in Houston, except that
@@
 is similarly convenient, as users can ignore most of AORTA's code-base and just
-change the logic described in section \ref{sec:policies}.  For reference, the
+change the logic described in Section \ref{sec:policies}.  For reference, the
 most complex intersection policy, traffic signals with the flooding heuristic
@@
 signal timings. On top of these structures, applications could experiment with
-ideas such dynamic contra-flow \cite{ITSC11-hausknecht}.  There will also be an
-effort to improve AORTA as a framework by including a lane-changing model,
+ideas such as dynamic contra-flow \cite{ITSC11-hausknecht}. There will also be
+an effort to improve AORTA as a framework by including a lane-changing model,
 preventing gridlock, and supporting more agents through parallel computing.

### 2012-03-29 papers/itsc_2012/submitted2.tex
Couple of minor grammatical changes

@@
 the real world in a few minutes. AORTA is available open-source and is easily
-extensible\footnote{Code available at
-\texttt{http://code.google.com/p/road-rage}}, making it easy for users to test
+extensible,\footnote{Code available at
+\texttt{http://code.google.com/p/road-rage}} making it easy for users to test
 out a number of agent behaviors and intersection policies in a short time span.
@@
 built on top of AORTA can be written in either Java or Scala. The software is
-open-source and easily extensible. For instance, the stop sign sign policy used
+open-source and easily extensible. For instance, the stop sign policy used
 in experiments described in Section \ref{sec:results} is implemented in just 70
@@
 center line that divides the road into two directions of travel. AORTA
-interprets these points as straight line segmentss and projects parallel lanes
+interprets these points as straight line segments and projects parallel lanes
 out using a fixed width. Turns are approximated with a single straight line
@@
 shared center left turn lanes. In some circumstances, one intersection appears
-to be several in close proximity, since OSM only has geometric, not functional,
-data on how roads meet. If OSM encoded this information, the realism of the
-simulation would improve. Other simulators have dealt with similar issues before
-\cite{Krajzewicz_Hertkorn_Ringel_Wagner_2005}.
+to be several in close proximity, since OSM has only geometric data, as opposed
+to functional data, on how roads meet. If OSM encoded functional information,
+the realism of the simulation would improve. Other simulators have dealt with
+similar issues \cite{Krajzewicz_Hertkorn_Ringel_Wagner_2005}.
 
@@
 conflict. If two vehicles at any point along two different turns could ever come
-into contact, then the turns are always in conflict, no matter where the agents
-actually are along that turn. This is visualized in Fig. \ref{fig:conflicts}.
+into contact, then the turns are always in conflict, regardless of where the agents
+actually are along the turns. This is visualized in Fig. \ref{fig:conflicts}.
 Further work is required before higher granularity can be introduced through
@@
 simulation is running, since tasks may finish and agents may enter the road at
-different times}. After setting up a scenario in the UI, users can save this
+different times.}. After setting up a scenario in the UI, users can save this
 configuration for re-simulation.

### 2012-08-29 papers/itsc_2012/submitted2.tex

@@
 Autonomous vehicle technology has made tremendous progress in the last decade.
-In 2007, six of the competing teams completed the 96km course set for the DARPA
+In 2007, six of the competing teams completed the 96 km course set for the DARPA
 Urban Challenge \cite{spectrumUrbanChallenge2007}. They did so while
@@
 moving vehicles, and following correct intersection precedence order. Since
-then, Google's driverless cars have clocked more than 250,000km on public roads
+then, Google's driverless cars have clocked more than 250,000 km on public roads
 in urban California, USA \cite{tedThrun2011}. In 2010, researchers from the
@@
 study the effect of automated transportation systems, route planning of
-individual vehicles, as well as dynamically adapting traffic signal policies to
+individual vehicles, and dynamically adapting traffic signal policies to
 increase traffic efficiency. While SUMO has a flexible system for managing
@@
 
-Currently agents cannot change lanes. This causes a second potential issue:
-gridlock.  Each agent maintains a safe following distance from the next. If a
-lane is filled, then an agent may be forced to stop in an intersection mid-turn
-and block other traffic. The default behavior prevents this by refusing to start
-a turn before the lookahead engine guarantees no agent will trigger a premature
+In the currently released version of the simulator, the ability for agents to
+change lanes has not yet been implemented. As a result, gridlock potentially
+occurs. Each agent maintains a safe following distance from the next. If a lane
+is filled, then an agent may be forced to stop in an intersection mid-turn and
+block other traffic. The default behavior prevents this by refusing to start a
+turn before the lookahead engine guarantees no agent will trigger a premature
 stop.
@@
 agent at the front decides to pick a different turn. AORTA has the ability to
-find such cycles.
+detect such cyclic dependencies.
 
@@
 The experiment simulates one hour of traffic in two cities: a
-9$\times$11km$^{2}$ slice of downtown Austin, Texas\footnote{bounded by the
+9$\times$11 km$^{2}$ slice of downtown Austin, Texas\footnote{bounded by the
 Mopac, Springdale Road, Oltorf Street and Koenig Lane}, and a
-15$\times$13km$^{2}$ slice of downtown Houston, Texas\footnote{bounded by Ella
+15$\times$13 km$^{2}$ slice of downtown Houston, Texas\footnote{bounded by Ella
 Boulevard, Eastex Freeway, West Alabama Street and Crosstimbers Street}.  The

### 2012-11-27 papers/thesis/microtolling.tex
some notes and a script on braess paradox

The underlying aim of microtolling is to optimize total travel time among all
drivers. The costs assigned to roads in explanations of Braess' paradox account
for the time to cross the road. The paradox occurs because all rational agents
try to take a new shortcut road with a certain configuration, but if only a few
drivers deviate from the original routes and take the shortcut, shouldn't total
travel time decrease? (I wrote a quick brute-force solver that evaluates total
time for every combination of drivers on each route, and so far, the answer is
no, but there must be a mistake somewhere.) Let the "capacity" of the shortcut
be this critical maximum of drivers that can take the new route and not decrease
the total travel time.

Assuming the critical number of drivers who take the shortcut can be calculated,
the question becomes how to choose the lucky drivers who get the faster route.
A lottery isn't "fair." Microtolling is an alternative. Rational agents minimize
the cost of their route, so to limit the number who take the faster route, costs
should be based on both time and a toll. A new assumption is required: drivers
have different time/dollar ratios; only some drivers are willing to pay a higher
toll for a faster route.

Some questions arise:
1) How does the tolling work?
  a) The shortcut is free/cheap until the capacity is reached, then a high toll
     is imposed. (Unfair because it's first-come, first-serve.)
  b) Everybody pays a toll proportional to the number of drivers on the
     shortcut. If it's empty at first for some driver but more people join while
     that driver is still using it, that driver pays the new high cost. This
     encourages usage at non-peak times.
  c) Once the shortcut is at capacity, do we allow more drivers to join? They
     would get an individual lower travel time, but decrease the total travel
     time of the system. (Is this even possible? Intuitively, yes, because a
     light could remain green for one car to cross quickly, even though the
     other phase has more demand waiting to cross.)
  d) If c) is true, then does an auction work better than a fixed toll?

2) How to generalize to large networks with more than just time costs for roads?
   Intersection costs depend on the flow in connected roads; does this raise any
   fundamental new complication, or can crossing an intersection be modeled as
   just another edge with a cost?

3) Time costs that are just constant or just linear are both unrealistic. The
   time to cross a road is constant (distance / speed limit) until there's
   congestion due to an intersection. This can maybe be modeled by a simple
   piecewise function (constant up to the road's capacity, then some function of
   the number of cars.)

Related work

http://theory.stanford.edu/~tim/papers/nd_hard.pdf
detecting braess paradox and finding the optimal roads to remove is NP-hard

http://economics.mit.edu/files/302
may have done lots of this general analysis already

### 2012-11-27 papers/thesis/braess.cpp

@@ -1,0 +1,62 @@
+/*
+ * "Two roads diverged in a city, and I--
+ * I took the one less traveled by because it had a lower toll,
+ * And that has made all the difference to the overall travel time."
+ */
+
+#include <iostream>
+#include <limits>
+
+using namespace std;
+
+// Arguments are the number of drivers taking each possible route.
+double compute_cost(int n1, int n2, int n3) {
+  // How many are taking the upper and lower path?
+  int u = n1 + n2;
+  int d = n3;
+
+  // The cost of taking each route
+  double c1 = (u / 10.0) + 4.50;
+  double c2 = (u / 10.0) + 0 + (d / 10.0);
+  double c3 = 4.50 + (d / 10.0);
+
+  return (n1 * c1) + (n2 * c2) + (n3 * c3);
+}
+
+void describe(int n1, int n2, int n3) {
+  cout << "  cost(n1 = " << n1 << ", n2 = " << n2 << ", n3 = " << n3 << ") = "
+       << compute_cost(n1, n2, n3) << endl;
+}
+
+// Try every combination of drivers along the 3 routes.
+void brute_force(int N) {
+  double min = numeric_limits<double>::max();
+  int min_n1, min_n2, min_n3;
+
+  for (int n1 = 0; n1 <= N; ++n1) {
+    for (int n2 = 0; n2 <= N - n1; ++n2) {
+      int n3 = N - n1 - n2;
+      double cost = compute_cost(n1, n2, n3);
+      if (cost < min) {
+        min = cost;
+        min_n1 = n1;
+        min_n2 = n2;
+        min_n3 = n3;
+      }
+    }
+  }
+
+  cout << "Optimal:\n";
+  describe(min_n1, min_n2, min_n3);
+
+  cout << "Without the shortcut:\n";
+  describe(N / 2, 0, N / 2);
+
+  cout << "If everyone is greedy:\n";
+  describe(0, N, 0);
+}
+
+int main () {
+  brute_force(10000);
+  return 0;
+}
